# snRNAseq: microglia reprocessing and QC

We re-run SCTransform on the microglia subset using the `glmGamPoi` backend,
followed by Harmony integration across batches, UMAP and Leiden-style
clustering. Outlier clusters with non-microglia transcriptional signatures
or anomalous QC are subsequently dropped.

```{r snrnaseq-recluster, fig.width = 10, fig.height = 4}
processed_path <- file.path(params$cache_dir, "microglia_seurat_processed.rds")

microglia_seurat <- cache_or_run(processed_path, overwrite = params$recompute_snrnaseq, {
  obj <- microglia_seurat
  obj <- Seurat::SCTransform(
    obj, assay = "RNA",
    vst.flavor = "v2", method = "glmGamPoi",
    vars.to.regress = c("percent_mt", "percent_contam"),
    verbose = FALSE
  )
  obj <- Seurat::RunPCA(obj, assay = "SCT", npcs = 30, verbose = FALSE)
  obj <- harmony::RunHarmony(
    obj, group.by.vars = c("batch", "sex"),
    assay.use = "SCT", reduction.use = "pca",
    reduction.save = "harmony", verbose = FALSE
  )
  obj <- Seurat::RunUMAP(obj, reduction = "harmony", dims = 1:20, verbose = FALSE)
  obj <- Seurat::FindNeighbors(obj, reduction = "harmony", dims = 1:20, verbose = FALSE)
  for (res in c(0.2, 0.4, 0.6)) {
    obj <- Seurat::FindClusters(obj, resolution = res, verbose = FALSE)
  }
  obj
})

Seurat::Idents(microglia_seurat) <- "SCT_snn_res.0.4"
p_umap <- Seurat::DimPlot(microglia_seurat, group.by = "SCT_snn_res.0.4",
                          label = TRUE, repel = TRUE) +
  ggtitle("Microglia clusters (Harmony + SCT v2, res 0.4)")
p_gt <- Seurat::DimPlot(microglia_seurat, group.by = "genotype",
                        cols = genotype_colours) +
  ggtitle("Microglia by genotype")
p_umap | p_gt
```

## Marker-based QC and cluster pruning

```{r snrnaseq-markers, fig.width = 9, fig.height = 6}
canonical <- list(
  Microglia    = c("Cx3cr1", "P2ry12", "Tmem119", "Trem2", "Csf1r", "Hexb"),
  DAM          = c("Cst7", "Apoe", "Lpl", "Itgax", "Spp1", "Tyrobp"),
  IFN          = c("Ifit3", "Isg15", "Stat1", "Oasl2"),
  Proliferative= c("Mki67", "Top2a", "Cenpf"),
  Oligo_contam = c("Mbp", "Plp1", "Mog"),
  Neuron_contam= c("Snap25", "Stmn2", "Rbfox3"),
  Astro_contam = c("Aqp4", "Gfap", "Slc1a2")
)
# Map symbols to row IDs (ENSEMBL).
symbol_to_id <- function(syms) {
  hit <- symbol_map$ensembl[match(syms, symbol_map$symbol)]
  setNames(hit[!is.na(hit)], syms[!is.na(hit)])
}
canonical_ids <- lapply(canonical, symbol_to_id)
canonical_ids <- canonical_ids[lengths(canonical_ids) > 0]

# Use SCT data slot for visual inspection.
DefaultAssay(microglia_seurat) <- "SCT"
present <- unique(unlist(canonical_ids))
present <- present[present %in% rownames(microglia_seurat)]

dot_data <- Seurat::DotPlot(microglia_seurat, features = present,
                            group.by = "SCT_snn_res.0.4")$data
dot_data$symbol <- symbol_map$symbol[match(as.character(dot_data$features.plot),
                                           symbol_map$ensembl)]
dot_data$module <- NA_character_
for (m in names(canonical_ids)) {
  dot_data$module[dot_data$features.plot %in% canonical_ids[[m]]] <- m
}
ggplot(dot_data,
       aes(x = symbol, y = id, size = pct.exp, colour = avg.exp.scaled)) +
  geom_point() +
  scale_colour_gradient2(low = "#3a4cc0", mid = "white", high = "#b40426") +
  facet_grid(. ~ module, scales = "free_x", space = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = NULL, y = "cluster (res 0.4)", title = "Canonical markers by cluster")
```

```{r snrnaseq-prune}
# Score each cluster on contamination markers vs microglia markers.
score_module <- function(obj, ids, name) {
  ids <- ids[ids %in% rownames(obj)]
  if (length(ids) == 0) return(rep(NA_real_, ncol(obj)))
  Matrix::colMeans(GetAssayData(obj, assay = "SCT", layer = "data")[ids, , drop = FALSE])
}
micro_score    <- score_module(microglia_seurat, canonical_ids$Microglia)
oligo_score    <- score_module(microglia_seurat, canonical_ids$Oligo_contam)
neuron_score   <- score_module(microglia_seurat, canonical_ids$Neuron_contam)
astro_score    <- score_module(microglia_seurat, canonical_ids$Astro_contam)

cluster_scores <- microglia_seurat@meta.data |>
  tibble::as_tibble(rownames = "cell") |>
  dplyr::mutate(
    micro = micro_score, oligo = oligo_score,
    neuron = neuron_score, astro = astro_score
  ) |>
  dplyr::group_by(SCT_snn_res.0.4) |>
  dplyr::summarise(
    n = dplyr::n(),
    micro_mean    = mean(micro,    na.rm = TRUE),
    oligo_mean    = mean(oligo,    na.rm = TRUE),
    neuron_mean   = mean(neuron,   na.rm = TRUE),
    astro_mean    = mean(astro,    na.rm = TRUE),
    pct_mt        = mean(percent_mt,     na.rm = TRUE),
    pct_contam    = mean(percent_contam, na.rm = TRUE),
    nFeature_mean = mean(nFeature_RNA,   na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(n))
cluster_scores

# Define outlier rule: clusters with contamination signal > microglia, or with
# extreme pct_contam / pct_mt, or fewer than 50 nuclei.
outlier_clusters <- cluster_scores |>
  dplyr::filter(
    n < 50 |
    pct_contam > quantile(cluster_scores$pct_contam, 0.95) + 5 |
    (oligo_mean  > micro_mean) |
    (neuron_mean > micro_mean) |
    (astro_mean  > micro_mean)
  ) |>
  dplyr::pull(SCT_snn_res.0.4) |>
  as.character()
cat("Outlier clusters dropped:\n"); print(outlier_clusters)

keep <- !(as.character(microglia_seurat$SCT_snn_res.0.4) %in% outlier_clusters)
microglia_seurat <- microglia_seurat[, keep]
cat(sprintf("After pruning: %d nuclei across %d clusters.\n",
            ncol(microglia_seurat),
            length(unique(microglia_seurat$SCT_snn_res.0.4))))
```

## Microglia substate annotation

```{r snrnaseq-substate, fig.width = 9, fig.height = 5}
# Annotate four substates (homeostatic, DAM, IFN, proliferative) via
# AddModuleScore -> argmax. Logic is centralised in label_microglia_states()
# so the per-substate NEBULA build script gets identical labels.
microglia_seurat <- label_microglia_states(microglia_seurat,
                                           symbol_map = symbol_map)

state_tab <- table(microglia_seurat$state, microglia_seurat$genotype)
state_prop <- prop.table(state_tab, margin = 2)
print(round(state_prop * 100, 2))

state_df <- as.data.frame(state_prop)
names(state_df) <- c("state", "genotype", "prop")
ggplot(state_df, aes(genotype, prop, fill = state)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  labs(x = NULL, y = "% of microglia", fill = "state",
       title = "Microglia substate composition by genotype")
```
