# snRNAseq DE: pseudobulk per genotype_batch

We aggregate raw RNA counts per `genotype_batch` (16 pseudo-samples = 4 per
genotype) and fit `limma-voom` with the contrast matrix defined in
`R/helpers.R`. The microglia population is used as a whole; substate-specific
DE is run secondarily.

```{r snrnaseq-de}
de_sn_path <- file.path(params$cache_dir, "de_snrnaseq.rds")
de_snrnaseq <- cache_or_run(de_sn_path, overwrite = params$recompute_snrnaseq, {
  pb <- build_pseudobulk(microglia_seurat, sample_col = "genotype_batch",
                         covariate_cols = c("genotype", "batch", "sex"))
  meta <- pb$meta
  meta$genotype <- factor(as.character(meta$genotype), levels = genotype_levels)
  meta$batch    <- factor(meta$batch)
  # 0 + genotype gives clean genotype coefficients; batch is added as covariate
  # to absorb per-batch shifts. Sex is captured by batch (each batch is mixed).
  design <- model.matrix(~ 0 + genotype + batch, data = meta)
  colnames(design) <- sub("^genotype", "", colnames(design))
  colnames(design) <- gsub("[^A-Za-z0-9_]", "_", colnames(design))
  cm <- make_contrast_matrix(design)
  fit <- fit_limma_voom(pb$counts, group = meta$genotype, design = design,
                        contrasts = cm, min_count = 10)
  # Attach gene symbols.
  for (cn in names(fit$top)) {
    fit$top[[cn]] <- dplyr::left_join(fit$top[[cn]],
      symbol_map, by = c("gene" = "ensembl"))
  }
  list(meta = meta, design = design, contrasts = cm, fit = fit)
})

cat(sprintf("snRNAseq pseudobulk: %d samples, %d genes retained after filter.\n",
            nrow(de_snrnaseq$meta), de_snrnaseq$fit$kept))

# Summary of significant genes per contrast (FDR < 0.05).
sn_summary <- purrr::imap_dfr(de_snrnaseq$fit$top, function(tbl, cn) {
  data.frame(
    contrast = cn,
    n_total  = nrow(tbl),
    n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
    n_up     = sum(tbl$adj.P.Val < 0.05 & tbl$logFC > 0, na.rm = TRUE),
    n_down   = sum(tbl$adj.P.Val < 0.05 & tbl$logFC < 0, na.rm = TRUE)
  )
})
knitr::kable(sn_summary, caption = "snRNAseq pseudobulk DE: significant gene counts")
```

```{r snrnaseq-volcano, fig.width = 10, fig.height = 5}
volcano <- function(tbl, title) {
  tbl <- tbl |> dplyr::filter(is.finite(logFC), is.finite(adj.P.Val))
  tbl$sig <- tbl$adj.P.Val < 0.05 & abs(tbl$logFC) > 0.5
  top <- tbl |> dplyr::arrange(adj.P.Val) |> dplyr::filter(sig) |> head(15)
  ggplot(tbl, aes(logFC, -log10(adj.P.Val))) +
    geom_point(aes(colour = sig), size = 0.6, alpha = 0.7) +
    scale_colour_manual(values = c(`FALSE` = "grey70", `TRUE` = "#b40426"),
                        guide = "none") +
    ggrepel::geom_text_repel(data = top, aes(label = symbol),
                             size = 3, max.overlaps = 50) +
    theme_bw() +
    labs(title = title, x = "log2 FC", y = "-log10 adj.P")
}
(volcano(de_snrnaseq$fit$top$nlgf_in_maptki, "snRNAseq: NLGF in MAPTKI") |
 volcano(de_snrnaseq$fit$top$nlgf_in_p301s,  "snRNAseq: NLGF in P301S"))
(volcano(de_snrnaseq$fit$top$interaction,    "snRNAseq: interaction (tau-dependent NLGF)") |
 volcano(de_snrnaseq$fit$top$tau_alone,      "snRNAseq: tau alone"))
```

## snRNAseq DE: single-cell NB mixed model (NEBULA)

Pseudobulk reduces 26k cells to 16 pseudo-samples, which under-powers the
interaction contrast. We refit at single-cell resolution with `nebula`
(negative-binomial gamma mixed model, log-normal random-effect method)
using a 2x2 factorial parameterisation: `tau` and `nlgf` main effects
plus the `tau:nlgf` interaction. The random intercept on
`genotype_batch` absorbs batch and sex (both fully nested within
`genotype_batch`). Three contrasts (`tau_alone`, `nlgf_in_maptki`,
`interaction`) are direct model coefficients; the other two are computed
via the per-gene covariance matrix returned by `nebula(..., covariance =
TRUE)`. All logFC are converted to log2 for downstream comparability.

```{r snrnaseq-nebula}
nebula_path <- file.path(params$cache_dir, "de_snrnaseq_nebula.rds")
de_snrnaseq_nebula <- cache_or_run(nebula_path, overwrite = params$recompute_snrnaseq, {
  fit_nebula_microglia(
    microglia_seurat,
    id_col       = "genotype_batch",
    genotype_col = "genotype",
    assay        = "RNA",
    layer        = "counts",
    min_cell_frac = 0.01,
    ncore        = max(1L, parallel::detectCores() - 1L),
    symbol_map   = symbol_map
  )
})

cat(sprintf("NEBULA: %d cells x %d genes; %d / %d converged.\n",
            de_snrnaseq_nebula$n_cells, de_snrnaseq_nebula$n_genes,
            sum(de_snrnaseq_nebula$fit$convergence == 1, na.rm = TRUE),
            length(de_snrnaseq_nebula$fit$convergence)))

neb_summary <- purrr::imap_dfr(de_snrnaseq_nebula$top, function(tbl, cn) {
  data.frame(
    contrast = cn,
    n_total  = nrow(tbl),
    n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
    n_up     = sum(tbl$adj.P.Val < 0.05 & tbl$logFC > 0, na.rm = TRUE),
    n_down   = sum(tbl$adj.P.Val < 0.05 & tbl$logFC < 0, na.rm = TRUE)
  )
})
knitr::kable(neb_summary,
             caption = "snRNAseq NEBULA DE: significant gene counts (FDR < 0.05)")
```

```{r snrnaseq-nebula-volcano, fig.width = 10, fig.height = 5}
(volcano(de_snrnaseq_nebula$top$nlgf_in_maptki, "NEBULA: NLGF in MAPTKI") |
 volcano(de_snrnaseq_nebula$top$nlgf_in_p301s,  "NEBULA: NLGF in P301S"))
(volcano(de_snrnaseq_nebula$top$interaction,    "NEBULA: interaction (tau-dependent NLGF)") |
 volcano(de_snrnaseq_nebula$top$tau_in_nlgf,    "NEBULA: tau in NLGF"))
```

### Pseudobulk vs NEBULA: power gain on the interaction

To quantify the power gain on the interaction contrast we align the two
fits on shared genes and plot logFC and `-log10(P)`. A higher density off
the y = x line in the right-hand panel signals that NEBULA assigns more
significance to a comparable effect size than pseudobulk does.

```{r snrnaseq-pseudobulk-vs-nebula, fig.width = 10, fig.height = 4}
compare_contrast <- function(cn, title) {
  pb <- de_snrnaseq$fit$top[[cn]] |>
    dplyr::transmute(gene, logFC_pb = logFC, p_pb = P.Value)
  nb <- de_snrnaseq_nebula$top[[cn]] |>
    dplyr::transmute(gene, logFC_nb = logFC, p_nb = P.Value)
  cmp <- dplyr::inner_join(pb, nb, by = "gene") |>
    dplyr::filter(is.finite(logFC_pb), is.finite(logFC_nb),
                  is.finite(p_pb), is.finite(p_nb), p_pb > 0, p_nb > 0)
  rho_lfc <- suppressWarnings(cor(cmp$logFC_pb, cmp$logFC_nb,
                                  method = "spearman"))
  rho_p   <- suppressWarnings(cor(-log10(cmp$p_pb), -log10(cmp$p_nb),
                                  method = "spearman"))
  p1 <- ggplot(cmp, aes(logFC_pb, logFC_nb)) +
    geom_hline(yintercept = 0, colour = "grey70") +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_abline(slope = 1, intercept = 0, colour = "#b40426",
                linetype = "dashed") +
    geom_point(alpha = 0.25, size = 0.5) +
    labs(x = "pseudobulk log2FC", y = "NEBULA log2FC",
         title = paste(title, "| logFC"),
         subtitle = sprintf("Spearman rho = %.3f, n = %d",
                            rho_lfc, nrow(cmp))) +
    theme_bw()
  p2 <- ggplot(cmp, aes(-log10(p_pb), -log10(p_nb))) +
    geom_abline(slope = 1, intercept = 0, colour = "#b40426",
                linetype = "dashed") +
    geom_point(alpha = 0.25, size = 0.5) +
    labs(x = "-log10 pseudobulk P", y = "-log10 NEBULA P",
         title = paste(title, "| -log10 P"),
         subtitle = sprintf("Spearman rho = %.3f", rho_p)) +
    theme_bw()
  p1 | p2
}
compare_contrast("interaction",    "interaction")
compare_contrast("nlgf_in_p301s",  "nlgf_in_p301s")
compare_contrast("nlgf_in_maptki", "nlgf_in_maptki")
```
