## snRNAseq DE: cross-method triangulation of the interaction

NEBULA places the interaction at FDR ~ 0.06, borderline negative. To
check that this is not an artefact of a single method's machinery, we
re-fit the same interaction contrast with four further methods under a
shared 2x2 factorial parameterisation (`tau`, `nlgf`, `tau:nlgf`):

* **edgeR-QLF pseudobulk** (Quasi-likelihood F-test), the engine
  behind `muscat::pbDS(method = "edgeR")`.
* **DESeq2 pseudobulk** (NB-Wald with empirical Bayes dispersion
  shrinkage), the engine behind `muscat::pbDS(method = "DESeq2")`.
* **variancePartition::dream pseudobulk** with `(1 | batch)` random
  intercept, the engine behind `dreamlet`; differs from limma-voom by
  treating batch as a random effect.
* **glmmTMB single-cell** NB GLMM with `(1 | genotype_batch)` random
  intercept, same statistical family as NEBULA but implemented via
  Laplace approximation (TMB) instead of NEBULA's LN method.

Together with the original limma-voom pseudobulk and NEBULA, this gives
six methods spanning two scales (single-cell, pseudobulk) and three
shrinkage families (limma-style, NB-Wald, NB GLMM). If all six agree on
"no FDR-significant interaction", the negative call is robust.

```{r snrnaseq-triangulation-pb}
sn_tri_pb_path <- file.path(params$cache_dir,
                            "de_snrnaseq_triangulation_pb.rds")
sn_tri_pb <- cache_or_run(sn_tri_pb_path,
                          overwrite = params$recompute_snrnaseq, {
  pb_obj <- build_pseudobulk(microglia_seurat,
                             sample_col = "genotype_batch",
                             covariate_cols = c("genotype", "batch", "sex"))
  meta <- pb_obj$meta
  meta$genotype <- factor(as.character(meta$genotype), levels = genotype_levels)
  meta$batch    <- factor(meta$batch)
  meta$tau      <- as.integer(meta$genotype %in% c("P301S", "NLGF_P301S"))
  meta$nlgf     <- as.integer(meta$genotype %in% c("NLGF_MAPTKI", "NLGF_P301S"))
  meta$tau_nlgf <- meta$tau * meta$nlgf
  fd <- factorial_design(meta, "genotype", "batch")
  list(
    edger  = fit_edger_qlf(pb_obj$counts, fd$design, fd$contrasts,
                            min_count = 10, symbol_map = symbol_map),
    deseq2 = fit_deseq2_pb(pb_obj$counts, meta, fd$design, fd$contrasts,
                            min_count = 10, symbol_map = symbol_map),
    dream  = fit_dream_pb(pb_obj$counts, meta, fd$contrasts,
                           min_count = 10, symbol_map = symbol_map)
  )
})

tri_pb_summary <- purrr::imap_dfr(sn_tri_pb, function(m, mname) {
  purrr::imap_dfr(m$top, function(tbl, cn) {
    data.frame(
      method   = mname,
      contrast = cn,
      n_genes  = nrow(tbl),
      n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
      min_padj = signif(suppressWarnings(min(tbl$adj.P.Val, na.rm = TRUE)), 3)
    )
  })
})
knitr::kable(tri_pb_summary,
             caption = "Pseudobulk triangulation: significant gene counts (FDR<0.05)")
```

```{r snrnaseq-triangulation-glmmtmb}
sn_tri_tmb_path <- file.path(params$cache_dir,
                             "de_snrnaseq_glmmtmb.rds")
sn_tri_tmb <- cache_or_run(sn_tri_tmb_path,
                           overwrite = params$recompute_snrnaseq, {
  fit_glmmtmb_microglia(
    microglia_seurat,
    id_col        = "genotype_batch",
    genotype_col  = "genotype",
    assay         = "RNA",
    layer         = "counts",
    min_cell_frac = 0.01,
    ncore         = max(1L, parallel::detectCores() - 1L),
    symbol_map    = symbol_map
  )
})

cat(sprintf(
  "glmmTMB: %d cells x %d genes; %d converged (%.1f%%) for the interaction.\n",
  sn_tri_tmb$n_cells, sn_tri_tmb$n_genes,
  sum(sn_tri_tmb$top$interaction$convergence == 1, na.rm = TRUE),
  100 * mean(sn_tri_tmb$top$interaction$convergence == 1, na.rm = TRUE)))

tmb_summary <- purrr::imap_dfr(sn_tri_tmb$top, function(tbl, cn) {
  data.frame(
    contrast = cn,
    n_genes  = nrow(tbl),
    n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
    min_padj = signif(suppressWarnings(min(tbl$adj.P.Val, na.rm = TRUE)), 3)
  )
})
knitr::kable(tmb_summary,
             caption = "glmmTMB single-cell: significant gene counts (FDR<0.05)")
```

```{r snrnaseq-triangulation-summary}
# Consolidate all six methods.
method_tops <- list(
  limma_voom_pb = de_snrnaseq$fit$top,
  nebula_sc     = de_snrnaseq_nebula$top,
  edger_qlf_pb  = sn_tri_pb$edger$top,
  deseq2_pb     = sn_tri_pb$deseq2$top,
  dream_pb      = sn_tri_pb$dream$top,
  glmmtmb_sc    = sn_tri_tmb$top
)

collect_method <- function(top_list, mname) {
  purrr::imap_dfr(top_list, function(tbl, cn) {
    tibble::tibble(
      method   = mname,
      contrast = cn,
      n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
      min_padj = signif(suppressWarnings(min(tbl$adj.P.Val, na.rm = TRUE)), 3)
    )
  })
}
full_summary_long <- purrr::imap_dfr(method_tops, collect_method)

n_sig_wide <- full_summary_long |>
  dplyr::select(method, contrast, n_sig) |>
  tidyr::pivot_wider(names_from = method, values_from = n_sig) |>
  dplyr::arrange(contrast)
min_padj_wide <- full_summary_long |>
  dplyr::select(method, contrast, min_padj) |>
  tidyr::pivot_wider(names_from = method, values_from = min_padj) |>
  dplyr::arrange(contrast)

knitr::kable(n_sig_wide,
             caption = "Triangulation: number of genes FDR<0.05 per method")
knitr::kable(min_padj_wide,
             caption = "Triangulation: minimum adjusted P per method")

write_tsv_safe(full_summary_long,
               file.path(params$results_dir, "triangulation_summary.tsv"))
```

```{r snrnaseq-triangulation-interaction, fig.width = 11, fig.height = 5}
# Interaction-specific concordance. Inner-join all six on the gene-id set
# they share, then plot pairwise Spearman correlations of logFC and -log10P.
int_tbls <- purrr::imap(method_tops, function(top_list, mname) {
  top_list$interaction |>
    dplyr::filter(is.finite(logFC), is.finite(P.Value), P.Value > 0) |>
    dplyr::transmute(gene,
                     "{mname}_logFC"   := logFC,
                     "{mname}_negLogP" := -log10(P.Value))
})
int_joined <- Reduce(function(a, b) dplyr::inner_join(a, b, by = "gene"),
                     int_tbls)
cat(sprintf("Interaction contrast: %d genes common to all six methods.\n",
            nrow(int_joined)))

lfc_mat <- as.matrix(int_joined[, grep("_logFC$",   names(int_joined))])
plp_mat <- as.matrix(int_joined[, grep("_negLogP$", names(int_joined))])
colnames(lfc_mat) <- sub("_logFC$",   "", colnames(lfc_mat))
colnames(plp_mat) <- sub("_negLogP$", "", colnames(plp_mat))
rho_l <- cor(lfc_mat, method = "spearman", use = "pairwise.complete.obs")
rho_p <- cor(plp_mat, method = "spearman", use = "pairwise.complete.obs")

corp <- function(mat, title) {
  df <- as.data.frame.table(mat, responseName = "rho")
  ggplot(df, aes(Var1, Var2, fill = rho)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", rho)), size = 3) +
    scale_fill_gradient2(low = "#3a4cc0", mid = "white", high = "#b40426",
                         midpoint = 0.5, limits = c(0, 1)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(x = NULL, y = NULL, title = title, fill = "Spearman rho")
}
(corp(rho_l, "Interaction logFC concordance") |
 corp(rho_p, "Interaction -log10(P) concordance"))
```

```{r snrnaseq-triangulation-tophits}
# Union of top-20 by P.Value per method; report -log10(P) for each method
# so the reader can see which hits replicate.
top_per_method <- purrr::imap(method_tops, function(top_list, mname) {
  top_list$interaction |>
    dplyr::filter(is.finite(P.Value), P.Value > 0) |>
    dplyr::arrange(P.Value) |>
    head(20) |>
    dplyr::pull(gene)
})
union_genes <- unique(unlist(top_per_method))
cat(sprintf("Union of top-20 interaction hits across six methods: %d genes.\n",
            length(union_genes)))

plogp_top <- purrr::imap_dfc(method_tops, function(top_list, mname) {
  tbl <- top_list$interaction
  v   <- -log10(pmax(tbl$P.Value[match(union_genes, tbl$gene)], 1e-300))
  tibble::tibble("{mname}" := v)
})
plogp_top$gene   <- union_genes
plogp_top$symbol <- symbol_map$symbol[match(union_genes, symbol_map$ensembl)]
plogp_top <- plogp_top |>
  dplyr::relocate(gene, symbol) |>
  dplyr::mutate(mean_negLogP = rowMeans(
    as.matrix(plogp_top[, names(method_tops)]), na.rm = TRUE)) |>
  dplyr::arrange(dplyr::desc(mean_negLogP))
knitr::kable(head(plogp_top, 25), digits = 2,
             caption = "Top interaction hits: -log10(P) per method (sorted by mean)")
write_tsv_safe(plogp_top,
               file.path(params$results_dir, "triangulation_top_interaction.tsv"))
```
