## snRNAseq DE: per-substate NEBULA (localisation)

The whole-microglia NEBULA fit reported a borderline interaction
(min FDR 0.063) with `Colec12`, `Plac9` and a small OXPHOS-leaning
pathway signal driving the divergence narrative. The natural follow-up
question is *which substate* (homeostatic, DAM, IFN, proliferative)
carries each candidate signal. We therefore re-fit the 2x2 factorial
NEBULA model independently within each microglia state and align the
five canonical contrasts side-by-side. The per-substate gene universe is
filtered to genes expressed in >= 5% of cells in that state (cf. 1% for
whole-microglia), reflecting the smaller cell counts per state.

```{r snrnaseq-substate-nebula}
per_state_path <- file.path(params$cache_dir, "de_snrnaseq_nebula_per_state.rds")
de_snrnaseq_nebula_per_state <- cache_or_run(per_state_path,
                                             overwrite = params$recompute_snrnaseq, {
  fit_nebula_per_state(
    microglia_seurat,
    state_col     = "state",
    id_col        = "genotype_batch",
    genotype_col  = "genotype",
    assay         = "RNA",
    layer         = "counts",
    min_cell_frac = 0.05,
    ncore         = max(1L, parallel::detectCores() - 1L),
    symbol_map    = symbol_map
  )
})

per_state_states <- names(de_snrnaseq_nebula_per_state)
cat(sprintf("Per-substate NEBULA fits: %s.\n",
            paste(per_state_states, collapse = ", ")))
for (st in per_state_states) {
  fit <- de_snrnaseq_nebula_per_state[[st]]
  cat(sprintf("  %-15s n_cells=%5d  n_genes=%5d  converged=%d/%d.\n",
              st, fit$n_cells, fit$n_genes,
              sum(fit$fit$convergence == 1, na.rm = TRUE),
              length(fit$fit$convergence)))
}
```

```{r snrnaseq-substate-nebula-summary}
contrast_levels <- c("nlgf_in_maptki", "nlgf_in_p301s",
                     "interaction", "tau_alone", "tau_in_nlgf")
per_state_summary <- purrr::imap_dfr(de_snrnaseq_nebula_per_state, function(fit, st) {
  purrr::imap_dfr(fit$top, function(tbl, cn) {
    data.frame(
      state    = st,
      contrast = cn,
      n_genes  = nrow(tbl),
      n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
      n_up     = sum(tbl$adj.P.Val < 0.05 & tbl$logFC > 0, na.rm = TRUE),
      n_down   = sum(tbl$adj.P.Val < 0.05 & tbl$logFC < 0, na.rm = TRUE),
      min_padj = suppressWarnings(min(tbl$adj.P.Val, na.rm = TRUE))
    )
  })
})
per_state_summary$contrast <- factor(per_state_summary$contrast,
                                     levels = contrast_levels)
per_state_summary <- per_state_summary[order(per_state_summary$state,
                                             per_state_summary$contrast), ]
per_state_summary$min_padj[!is.finite(per_state_summary$min_padj)] <- NA_real_
knitr::kable(per_state_summary, digits = 3,
             caption = "Per-substate NEBULA DE summary (FDR < 0.05)")
write_tsv_safe(per_state_summary,
               file.path(params$results_dir, "nebula_per_state_summary.tsv"))
```

```{r snrnaseq-substate-interaction-volcano, fig.width = 11, fig.height = 7}
# Volcano of the interaction contrast per substate. Same scale across
# panels so a stronger signal in one state shows up as taller points.
int_panels <- purrr::imap(de_snrnaseq_nebula_per_state, function(fit, st) {
  volcano(fit$top$interaction, sprintf("Interaction (%s)", st))
})
patchwork::wrap_plots(int_panels, ncol = 2)
```

```{r snrnaseq-substate-localisation}
# Localise the whole-microglia interaction signal: for the union of
#   (a) top 20 NEBULA whole-microglia interaction hits by raw P, and
#   (b) the union of top-20 per-method hits from the 6-method
#       triangulation (already saved as `union_genes`),
# look up per-state interaction logFC and -log10(P). Highlights which
# substate carries each candidate divergence gene.
nebula_int_whole <- de_snrnaseq_nebula$top$interaction |>
  dplyr::filter(is.finite(P.Value)) |>
  dplyr::arrange(P.Value) |>
  head(20) |>
  dplyr::pull(gene)
target_genes <- unique(c(nebula_int_whole, union_genes))

# Symbols for display.
target_symbols <- symbol_map$symbol[match(target_genes, symbol_map$ensembl)]

per_state_int <- purrr::imap_dfr(de_snrnaseq_nebula_per_state, function(fit, st) {
  tbl <- fit$top$interaction
  idx <- match(target_genes, tbl$gene)
  data.frame(
    gene        = target_genes,
    symbol      = target_symbols,
    state       = st,
    logFC       = tbl$logFC[idx],
    P.Value     = tbl$P.Value[idx],
    adj.P.Val   = tbl$adj.P.Val[idx],
    convergence = tbl$convergence[idx]
  )
})

# Whole-microglia reference for each gene, joined alongside.
whole_ref <- de_snrnaseq_nebula$top$interaction |>
  dplyr::filter(gene %in% target_genes) |>
  dplyr::select(gene, whole_logFC = logFC,
                whole_P = P.Value, whole_padj = adj.P.Val)

per_state_int <- dplyr::left_join(per_state_int, whole_ref, by = "gene")

# Wide table for display: one row per gene, columns per state for
# -log10(P) and signed-logFC (sign captures direction of divergence).
loc_wide <- per_state_int |>
  dplyr::mutate(neg_log10p = -log10(pmax(P.Value, 1e-300))) |>
  dplyr::select(gene, symbol, state, logFC, neg_log10p,
                whole_P, whole_padj) |>
  tidyr::pivot_wider(
    id_cols = c(gene, symbol, whole_P, whole_padj),
    names_from = state,
    values_from = c(logFC, neg_log10p),
    names_glue = "{state}_{.value}"
  ) |>
  dplyr::mutate(
    max_negLogP_state = apply(
      dplyr::pick(dplyr::ends_with("_neg_log10p")), 1,
      function(x) {
        nm <- names(x)
        if (all(!is.finite(x))) return(NA_character_)
        sub("_neg_log10p$", "", nm[which.max(x)])
      }
    ),
    max_negLogP = apply(
      dplyr::pick(dplyr::ends_with("_neg_log10p")), 1,
      function(x) suppressWarnings(max(x, na.rm = TRUE))
    )
  ) |>
  dplyr::arrange(dplyr::desc(max_negLogP))

knitr::kable(head(loc_wide, 25), digits = 3,
             caption = paste0(
               "Per-substate localisation of the interaction contrast: ",
               "for each candidate gene, -log10(P) and logFC per state ",
               "plus the whole-microglia reference. ",
               "Sorted by maximum per-state -log10(P)."))
write_tsv_safe(loc_wide,
               file.path(params$results_dir, "nebula_per_state_localisation.tsv"))
```

```{r snrnaseq-substate-spotlight, fig.width = 10, fig.height = 5}
# Spotlight visualisation: pick the six genes with the most concentrated
# per-state interaction signal (largest max_negLogP across states) that
# survive the 5% prevalence filter in at least 2 states. Whole-microglia
# borderline hits filtered out per-state by prevalence (Plac9, Cd276,
# AI504432) are necessarily excluded here.
state_levels <- names(de_snrnaseq_nebula_per_state)
spotlight_pick <- loc_wide |>
  dplyr::mutate(
    n_states_present = rowSums(
      !is.na(dplyr::pick(dplyr::ends_with("_logFC"))))
  ) |>
  dplyr::filter(n_states_present >= 2, is.finite(max_negLogP)) |>
  dplyr::arrange(dplyr::desc(max_negLogP)) |>
  head(6) |>
  dplyr::pull(symbol)
cat(sprintf("Spotlight genes (top 6 max-state-signal, >=2 states): %s.\n",
            paste(spotlight_pick, collapse = ", ")))

spotlight <- per_state_int |>
  dplyr::filter(symbol %in% spotlight_pick) |>
  dplyr::mutate(
    symbol = factor(symbol, levels = spotlight_pick),
    state  = factor(state, levels = state_levels),
    neg_log10p = -log10(pmax(P.Value, 1e-300)),
    direction  = ifelse(is.finite(logFC) & logFC >= 0, "up", "down")
  )

p_p <- ggplot(spotlight,
              aes(state, neg_log10p, fill = direction)) +
  geom_col() +
  facet_wrap(~ symbol, nrow = 1) +
  scale_fill_manual(values = c(up = "#b40426", down = "#3b4cc0"),
                    name = "logFC direction") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             colour = "grey50") +
  labs(x = NULL, y = "-log10(raw P), interaction",
       title = "Per-substate localisation of top interaction hits") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_f <- ggplot(spotlight,
              aes(state, logFC, fill = direction)) +
  geom_col() +
  facet_wrap(~ symbol, nrow = 1) +
  scale_fill_manual(values = c(up = "#b40426", down = "#3b4cc0"),
                    name = "logFC direction") +
  geom_hline(yintercept = 0, colour = "grey50") +
  labs(x = NULL, y = "log2 FC, interaction") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_p / p_f
```

### IFN microglia: asymmetric NLGF response

The per-state summary above shows that IFN microglia mount a 3.8x
stronger NLGF response in the P301S tau background (528 sig at
FDR < 0.05) than in MAPTKI (138 sig). This asymmetry is the mathematical
substrate of the whole-microglia interaction signal, but is much more
striking in IFN than in any other state. Below we project the two
NLGF contrasts within IFN onto each other to characterise which genes
are P301S-specific (sig only in P301S context), MAPTKI-specific, or
shared, and to list the strongest P301S-specific drivers.

```{r snrnaseq-ifn-asymmetry, fig.width = 9, fig.height = 7}
if ("IFN" %in% names(de_snrnaseq_nebula_per_state)) {
  ifn_fit <- de_snrnaseq_nebula_per_state$IFN
  ifn_maptki <- ifn_fit$top$nlgf_in_maptki |>
    dplyr::select(gene, symbol,
                  logFC_maptki = logFC, P_maptki = P.Value,
                  padj_maptki = adj.P.Val) |>
    dplyr::filter(!is.na(symbol))
  ifn_p301s <- ifn_fit$top$nlgf_in_p301s |>
    dplyr::select(gene, symbol,
                  logFC_p301s = logFC, P_p301s = P.Value,
                  padj_p301s = adj.P.Val) |>
    dplyr::filter(!is.na(symbol))
  ifn_join <- dplyr::inner_join(ifn_maptki, ifn_p301s,
                                by = c("gene", "symbol")) |>
    dplyr::mutate(
      sig_maptki = !is.na(padj_maptki) & padj_maptki < 0.05,
      sig_p301s  = !is.na(padj_p301s)  & padj_p301s  < 0.05,
      category   = dplyr::case_when(
        sig_p301s  & !sig_maptki ~ "P301S-specific",
        sig_maptki & !sig_p301s  ~ "MAPTKI-specific",
        sig_p301s  &  sig_maptki ~ "Shared",
        TRUE                     ~ "ns"
      )
    )
  ifn_join$category <- factor(ifn_join$category,
                              levels = c("Shared", "P301S-specific",
                                         "MAPTKI-specific", "ns"))
  cat("IFN microglia, FDR<0.05 categorisation:\n")
  print(table(ifn_join$category))

  label_n <- 12
  label_set <- dplyr::bind_rows(
    ifn_join |> dplyr::filter(category == "P301S-specific") |>
      dplyr::arrange(P_p301s) |> head(label_n),
    ifn_join |> dplyr::filter(category == "MAPTKI-specific") |>
      dplyr::arrange(P_maptki) |> head(label_n),
    ifn_join |> dplyr::filter(category == "Shared") |>
      dplyr::arrange(pmin(P_p301s, P_maptki)) |> head(6)
  )
  p_scatter <- ggplot(ifn_join,
                      aes(logFC_maptki, logFC_p301s, colour = category)) +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey60") +
    geom_point(data = dplyr::filter(ifn_join, category == "ns"),
               alpha = 0.2, size = 0.6) +
    geom_point(data = dplyr::filter(ifn_join, category != "ns"),
               alpha = 0.8, size = 1.1) +
    ggrepel::geom_text_repel(data = label_set,
                             aes(label = symbol),
                             size = 3, colour = "black",
                             max.overlaps = 40) +
    scale_colour_manual(values = c(
      Shared            = "#7e3a86",
      `P301S-specific`  = "#b40426",
      `MAPTKI-specific` = "#3b4cc0",
      ns                = "grey70"
    )) +
    labs(x = "logFC nlgf_in_maptki (IFN microglia)",
         y = "logFC nlgf_in_p301s (IFN microglia)",
         title = "IFN microglia: NLGF response in P301S vs MAPTKI tau context",
         subtitle = "Points off the dashed identity line = tau-context-dependent response") +
    theme_bw()
  print(p_scatter)

  # |logFC|<=10 sanity filter on the displayed top-25: NEBULA can return
  # extreme natural-log -> log2 estimates (e.g. Mamdc2 logFC_maptki=26.6)
  # for genes with near-zero counts in one arm. The raw TSV stays
  # untouched; this only governs the kable.
  ifn_extreme_n <- ifn_join |>
    dplyr::filter(category == "P301S-specific",
                  abs(logFC_p301s) > 10 | abs(logFC_maptki) > 10) |>
    nrow()
  ifn_top_p301s <- ifn_join |>
    dplyr::filter(category == "P301S-specific",
                  abs(logFC_p301s) <= 10,
                  abs(logFC_maptki) <= 10) |>
    dplyr::arrange(P_p301s) |>
    dplyr::transmute(symbol, gene,
                     logFC_maptki, padj_maptki,
                     logFC_p301s,  padj_p301s) |>
    head(25)
  print(knitr::kable(ifn_top_p301s, digits = 3,
                     caption = sprintf(paste0(
                       "Top P301S-specific IFN NLGF responders ",
                       "(FDR<0.05 in P301S, ns in MAPTKI; |logFC|<=10 ",
                       "in both contrasts, %d outliers excluded from display)"),
                       ifn_extreme_n)))
  write_tsv_safe(ifn_join,
                 file.path(params$results_dir,
                           "ifn_nlgf_asymmetry.tsv"))
} else {
  cat("IFN per-state NEBULA fit not available; skipping asymmetry panel.\n")
}
```
