### Per-substate pathway localisation (fgsea)

The whole-microglia fgsea (built in the integration section and Phase B
builder scripts) identifies pathways with interaction-level support
against the four microglia substates collapsed together. To check
whether those signals are driven by a single substate or are distributed
across the four substates (homeostatic, DAM, IFN, proliferative), we
re-run fgsea on the per-state NEBULA `t`-statistics against all six
gene-set collections used elsewhere in the report: GO BP, GO MF, GO CC,
and the three project-specific curated families (custom_microglia_states
from B2, custom_microglia_ad from B3, custom_module_sources from B4).
The GO collections retain the canonical `min_size = 15` cutoff; the
three custom collections use `min_size = 5` because their tightest sets
(homeostatic_core at 15 mouse symbols, IRM at 28, etc.) drop further
after intersecting with each substate's measured-gene universe.

```{r snrnaseq-substate-fgsea}
gobp           <- get_gobp()
gomf           <- get_gomf()
gocc           <- get_gocc()
custom_states  <- get_custom_microglia_states()
custom_ad      <- get_custom_microglia_ad()
custom_modules <- get_custom_module_sources()

substate_fgsea_path <- file.path(params$cache_dir,
                                 "fgsea_per_state_results.rds")
fgsea_per_state <- cache_or_run(substate_fgsea_path, {
  list(
    gobp                    = run_fgsea_per_state(de_snrnaseq_nebula_per_state,
                                                  gobp,
                                                  min_size = 15, max_size = 500),
    gomf                    = run_fgsea_per_state(de_snrnaseq_nebula_per_state,
                                                  gomf,
                                                  min_size = 15, max_size = 500),
    gocc                    = run_fgsea_per_state(de_snrnaseq_nebula_per_state,
                                                  gocc,
                                                  min_size = 15, max_size = 500),
    custom_microglia_states = run_fgsea_per_state(de_snrnaseq_nebula_per_state,
                                                  custom_states,
                                                  min_size = 5,  max_size = 500),
    custom_microglia_ad     = run_fgsea_per_state(de_snrnaseq_nebula_per_state,
                                                  custom_ad,
                                                  min_size = 5,  max_size = 500),
    custom_module_sources   = run_fgsea_per_state(de_snrnaseq_nebula_per_state,
                                                  custom_modules,
                                                  min_size = 5,  max_size = 500)
  )
})

flatten_fgsea_per_state <- function(nested, collection) {
  purrr::imap_dfr(nested, function(per_contrast, state) {
    purrr::imap_dfr(per_contrast, function(tbl, contrast) {
      df <- as.data.frame(tbl)[, c("pathway", "NES", "padj", "size")]
      df$collection <- collection
      df$state      <- state
      df$contrast   <- contrast
      df
    })
  })
}
fgsea_per_state_long <- purrr::imap_dfr(fgsea_per_state,
                                        flatten_fgsea_per_state)
write_tsv_safe(fgsea_per_state_long,
               file.path(params$results_dir,
                         "fgsea_per_state.tsv"))
collection_counts <- table(fgsea_per_state_long$collection)
cat("Per-state fgsea rows by collection:\n")
for (cn in names(collection_counts)) {
  cat(sprintf("  %-26s %d\n", cn, collection_counts[[cn]]))
}
cat(sprintf("Total: %d rows (6 collections x 4 substates x 5 contrasts x per-collection pathway counts).\n",
            nrow(fgsea_per_state_long)))
```

```{r snrnaseq-substate-fgsea-gobp-heatmap, fig.width = 9, fig.height = 9}
hm_int <- fgsea_per_state_long |>
  dplyr::filter(collection == "gobp", contrast == "interaction") |>
  dplyr::select(pathway, state, NES, padj) |>
  tidyr::pivot_wider(id_cols = pathway, names_from = state,
                     values_from = c(NES, padj),
                     names_glue = "{.value}_{state}")
nes_cols  <- grep("^NES_",  names(hm_int), value = TRUE)
padj_cols <- grep("^padj_", names(hm_int), value = TRUE)
sig_any <- apply(as.matrix(hm_int[, padj_cols]) < 0.1, 1,
                 function(r) any(r, na.rm = TRUE))
hm_int <- hm_int[sig_any, , drop = FALSE]
if (nrow(hm_int) == 0) {
  cat("No GO BP pathway with padj<0.1 in any state for the interaction.\n")
} else {
  # GO BP has ~7000 terms; cap the heatmap to the top 40 by max |NES|
  # across states so labels stay legible. The full per-state GO BP table
  # is exported in fgsea_per_state.tsv above.
  mat <- as.matrix(hm_int[, nes_cols])
  rownames(mat) <- gsub("_", " ", sub("^GOBP_", "", hm_int$pathway))
  colnames(mat) <- sub("^NES_", "", colnames(mat))
  mat[!is.finite(mat)] <- 0
  if (nrow(mat) > 40) {
    keep <- order(apply(abs(mat), 1, max, na.rm = TRUE),
                  decreasing = TRUE)[seq_len(40)]
    mat <- mat[keep, , drop = FALSE]
  }
  col_fun <- circlize::colorRamp2(c(-2.5, 0, 2.5),
                                  c("#3a4cc0", "white", "#b40426"))
  ComplexHeatmap::Heatmap(
    mat, name = "NES",
    col = col_fun,
    row_title = "GO BP pathway",
    column_title = sprintf(
      "Per-state GO BP NES (interaction contrast, top %d by max |NES|)",
      nrow(mat)),
    cluster_columns = FALSE,
    row_names_gp = grid::gpar(fontsize = 7),
    column_names_gp = grid::gpar(fontsize = 9),
    border = TRUE
  )
}
```

```{r snrnaseq-substate-fgsea-gobp-top-per-state}
gobp_top_per_state <- fgsea_per_state_long |>
  dplyr::filter(collection == "gobp", contrast == "interaction",
                is.finite(padj)) |>
  dplyr::group_by(state) |>
  dplyr::arrange(padj, .by_group = TRUE) |>
  dplyr::slice_head(n = 10) |>
  dplyr::ungroup() |>
  dplyr::mutate(pathway = gsub("_", " ",
                               sub("^GOBP_", "", pathway))) |>
  dplyr::select(state, pathway, NES, padj, size)
print(knitr::kable(gobp_top_per_state, digits = 3,
                   caption = "Top 10 GO BP interaction terms per state (by padj)"))
```

Comprehensive cross-modality pathway ranking across all gene-set
collections (GO BP, GO MF, GO CC, and three custom curated families) is
reserved for the agnostic pathway survey introduced later in this
report.

### Sparse whole-microglia hits at 1% per-state prevalence

Several whole-microglia top interaction hits (Plac9, Cd276, AI504432)
are expressed in fewer than 5% of cells per substate and therefore
drop out of the canonical 5% per-state NEBULA fit. To check whether
the per-state signal is genuinely absent or merely filtered, we
re-fit NEBULA at the whole-microglia threshold of 1% per state and
look up the missing genes. The 5% fit remains the published per-state
reference (better stable variance estimates per state).

```{r snrnaseq-substate-sparse-1pct}
sparse_cache <- file.path(params$cache_dir,
                          "de_snrnaseq_nebula_per_state_1pct.rds")
if (file.exists(sparse_cache)) {
  per_state_1pct <- readRDS(sparse_cache)

  whole_top <- de_snrnaseq_nebula$top$interaction |>
    dplyr::filter(is.finite(P.Value)) |>
    dplyr::arrange(P.Value) |>
    head(20)

  in_5pct_genes <- unique(unlist(
    purrr::map(de_snrnaseq_nebula_per_state, ~ .x$top$interaction$gene)
  ))
  filtered_out <- whole_top$gene[!whole_top$gene %in% in_5pct_genes]
  in_some_states <- whole_top$gene[whole_top$gene %in% in_5pct_genes &
    !sapply(whole_top$gene, function(g) {
      all(vapply(de_snrnaseq_nebula_per_state,
                 function(fit) g %in% fit$top$interaction$gene,
                 logical(1)))
    })]
  target_sparse <- unique(c(filtered_out, in_some_states))
  cat(sprintf("Whole-microglia top-20 hits affected by the 5%% per-state filter: %d\n",
              length(target_sparse)))

  sparse_long <- purrr::imap_dfr(per_state_1pct, function(fit, st) {
    tbl <- fit$top$interaction
    idx <- match(target_sparse, tbl$gene)
    data.frame(
      gene        = target_sparse,
      symbol      = symbol_map$symbol[match(target_sparse,
                                            symbol_map$ensembl)],
      state       = st,
      logFC       = tbl$logFC[idx],
      P.Value     = tbl$P.Value[idx],
      adj.P.Val   = tbl$adj.P.Val[idx],
      convergence = tbl$convergence[idx]
    )
  })
  sparse_wide <- sparse_long |>
    dplyr::mutate(neg_log10p = -log10(pmax(P.Value, 1e-300))) |>
    dplyr::select(gene, symbol, state, logFC, neg_log10p, adj.P.Val) |>
    tidyr::pivot_wider(id_cols = c(gene, symbol),
                       names_from = state,
                       values_from = c(logFC, neg_log10p, adj.P.Val),
                       names_glue = "{state}_{.value}") |>
    dplyr::left_join(
      whole_top |> dplyr::select(gene, whole_logFC = logFC,
                                 whole_P = P.Value,
                                 whole_padj = adj.P.Val),
      by = "gene"
    )
  print(knitr::kable(sparse_wide, digits = 3,
                     caption = paste0(
                       "Whole-microglia top interaction hits at 1% per-state ",
                       "prevalence: per-state interaction logFC and -log10(P) ",
                       "with the whole-microglia reference. NA entries mean the ",
                       "gene fell below 1% even at the relaxed threshold in that state.")))
  write_tsv_safe(sparse_wide,
                 file.path(params$results_dir,
                           "nebula_per_state_sparse_1pct.tsv"))
} else {
  cat(sprintf("Sparse 1%% cache not found at %s; skipping.\n", sparse_cache))
  cat("  (Build via `Rscript scripts/build_per_state_nebula_1pct.R`.)\n")
}
```
