## snRNAseq unsupervised co-expression modules (hdWGCNA)

The per-substate NEBULA differential-expression analyses above use
*pre-defined* substate labels. A complementary unsupervised view comes
from co-expression modules built directly from the microglia data with
no a priori gene-set or substate definitions.
We use **hdWGCNA** with metacells pooled within state x genotype strata
(k = 25, reduction = harmony), construct a signed network across all
metacells, identify modules via dynamic tree cut, and compute Harmony-
batch-corrected eigengenes per cell. Per-module DE then aggregates MEs
to the same `genotype_batch` pseudosamples used everywhere else (16 ids,
4 per genotype) and fits the project-wide 2x2 factorial limma with
`batch` as a fixed effect. The five canonical contrasts are emitted.

The cache is built outside the Rmd by `scripts/build_hdwgcna.R`; the
chunks below consume the lightweight summary.

```{r hdwgcna-load}
hdw_cache <- file.path(params$cache_dir, "hdwgcna_microglia.rds")
hdw_de_cache <- file.path(params$cache_dir, "hdwgcna_module_de.rds")
hdw <- NULL
hdw_de <- NULL
if (file.exists(hdw_cache) && file.exists(hdw_de_cache)) {
  hdw    <- readRDS(hdw_cache)
  hdw_de <- readRDS(hdw_de_cache)
  cat(sprintf("hdWGCNA cache loaded.  soft_power=%d, n_modules=%d (incl. grey).\n",
              hdw$soft_power, nrow(hdw$module_meta)))
  print(knitr::kable(hdw$module_meta,
                     caption = "Module sizes (genes per module)"))
} else {
  cat(sprintf(paste0("hdWGCNA cache not found at %s and/or %s; skipping.\n",
                     "  Build via `Rscript scripts/build_hdwgcna.R`.\n"),
              hdw_cache, hdw_de_cache))
}
```

```{r hdwgcna-soft-power, fig.width = 9, fig.height = 4}
# Scale-free fit and connectivity as a function of soft-thresholding
# power. The chosen power is the lowest power with R^2 >= 0.8 (default).
if (!is.null(hdw)) {
  pt <- hdw$power_table
  pt$chosen <- pt$Power == hdw$soft_power
  p1 <- ggplot2::ggplot(pt, ggplot2::aes(Power, SFT.R.sq)) +
    ggplot2::geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey60") +
    ggplot2::geom_line(colour = "grey40") +
    ggplot2::geom_point(ggplot2::aes(colour = chosen), size = 2.5) +
    ggplot2::scale_colour_manual(values = c(`TRUE` = "#b40426",
                                            `FALSE` = "black"),
                                 guide = "none") +
    ggplot2::labs(title = "Scale-free topology fit",
                  x = "soft-thresholding power",
                  y = expression(R^2)) +
    ggplot2::theme_bw()
  p2 <- ggplot2::ggplot(pt, ggplot2::aes(Power, mean.k.)) +
    ggplot2::geom_line(colour = "grey40") +
    ggplot2::geom_point(ggplot2::aes(colour = chosen), size = 2.5) +
    ggplot2::scale_colour_manual(values = c(`TRUE` = "#b40426",
                                            `FALSE` = "black"),
                                 guide = "none") +
    ggplot2::scale_y_log10() +
    ggplot2::labs(title = "Mean connectivity",
                  x = "soft-thresholding power",
                  y = "mean k (log10)") +
    ggplot2::theme_bw()
  print(patchwork::wrap_plots(p1, p2, ncol = 2))
}
```

```{r hdwgcna-module-state-heatmap, fig.width = 10, fig.height = 5}
# Mean module eigengene per (state, genotype) cell pool. Tells us which
# modules are state-enriched (DAM/IFN/...) and which respond to genotype.
if (!is.null(hdw)) {
  ME <- as.matrix(hdw$MEs_cells)
  meta <- hdw$cell_meta
  stopifnot(nrow(ME) == nrow(meta))
  mods <- setdiff(colnames(ME), c("grey", "MG-M0"))
  ME <- ME[, mods, drop = FALSE]
  # Reorder modules by their first-appearing number (MG-M1, MG-M2, ...)
  ord <- order(suppressWarnings(as.integer(sub("^MG-M", "", mods))))
  ME <- ME[, ord, drop = FALSE]

  states <- c("homeostatic", "DAM", "IFN", "proliferative")
  genos  <- c("MAPTKI", "P301S", "NLGF_MAPTKI", "NLGF_P301S")
  sg <- paste(meta$state, meta$genotype, sep = " | ")
  agg <- do.call(rbind, lapply(split(seq_len(nrow(ME)), sg), function(idx)
    colMeans(ME[idx, , drop = FALSE])))
  # Order rows: state first, genotype second
  meta_grid <- expand.grid(genotype = genos, state = states,
                           stringsAsFactors = FALSE)
  meta_grid$key <- paste(meta_grid$state, meta_grid$genotype, sep = " | ")
  agg <- agg[match(meta_grid$key, rownames(agg)), , drop = FALSE]
  rownames(agg) <- meta_grid$key

  col_fun <- circlize::colorRamp2(c(-0.05, 0, 0.05),
                                  c("#3a4cc0", "white", "#b40426"))
  row_split <- factor(meta_grid$state, levels = states)
  ComplexHeatmap::Heatmap(
    agg,
    name = "mean ME",
    col = col_fun,
    cluster_rows    = FALSE,
    cluster_columns = TRUE,
    row_split       = row_split,
    row_title_rot   = 0,
    row_names_gp    = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 9),
    column_title    = "Module",
    border          = TRUE,
    heatmap_legend_param = list(legend_height = grid::unit(3, "cm"))
  )
}
```

```{r hdwgcna-module-de}
# Per-module DE under the 2x2 factorial + batch covariate. n_modules is
# small so the multiple-testing burden is trivial. Export long-form TSV.
if (!is.null(hdw_de)) {
  de_long <- dplyr::bind_rows(lapply(names(hdw_de$top), function(ct) {
    dplyr::mutate(hdw_de$top[[ct]], contrast = ct)
  }))
  write_tsv_safe(de_long,
                 file.path(params$results_dir, "hdwgcna_module_de.tsv"))
  hdw_de_summary <- de_long |>
    dplyr::group_by(contrast) |>
    dplyr::summarise(
      n_modules = dplyr::n(),
      n_sig_FDR05 = sum(adj.P.Val < 0.05, na.rm = TRUE),
      min_padj  = suppressWarnings(min(adj.P.Val, na.rm = TRUE)),
      .groups = "drop"
    )
  print(knitr::kable(hdw_de_summary, digits = 4,
                     caption = "Per-module DE summary (FDR<0.05) over five contrasts"))

  # Top hits per contrast (each module's strongest contrast).
  top_per_contrast <- de_long |>
    dplyr::group_by(contrast) |>
    dplyr::arrange(P.Value, .by_group = TRUE) |>
    dplyr::slice_head(n = 3) |>
    dplyr::select(contrast, module, logFC, P.Value, adj.P.Val)
  print(knitr::kable(top_per_contrast, digits = 4,
                     caption = "Top 3 modules per contrast (by P)"))
}
```

```{r hdwgcna-module-de-heatmap, fig.width = 8, fig.height = 4}
# Visual summary: per-module logFC across the five contrasts. Heatmap
# colour = logFC, '*' annotation = FDR<0.05.
if (!is.null(hdw_de)) {
  ct_order <- c("tau_alone", "nlgf_in_maptki", "nlgf_in_p301s",
                "tau_in_nlgf", "interaction")
  fc_mat <- sapply(ct_order, function(ct) {
    setNames(hdw_de$top[[ct]]$logFC, hdw_de$top[[ct]]$module)
  })
  pa_mat <- sapply(ct_order, function(ct) {
    setNames(hdw_de$top[[ct]]$adj.P.Val, hdw_de$top[[ct]]$module)
  })
  ord <- order(suppressWarnings(as.integer(sub("^MG-M", "", rownames(fc_mat)))))
  fc_mat <- fc_mat[ord, , drop = FALSE]
  pa_mat <- pa_mat[ord, , drop = FALSE]

  range_abs <- max(abs(fc_mat[is.finite(fc_mat)]), na.rm = TRUE)
  col_fun <- circlize::colorRamp2(c(-range_abs, 0, range_abs),
                                  c("#3a4cc0", "white", "#b40426"))
  cell_fn <- function(j, i, x, y, width, height, fill) {
    if (!is.na(pa_mat[i, j]) && pa_mat[i, j] < 0.05) {
      grid::grid.text("*", x, y,
                      gp = grid::gpar(fontsize = 10, col = "black"))
    } else if (!is.na(pa_mat[i, j]) && pa_mat[i, j] < 0.1) {
      grid::grid.text(".", x, y,
                      gp = grid::gpar(fontsize = 10, col = "black"))
    }
  }
  ComplexHeatmap::Heatmap(
    fc_mat,
    name           = "logFC",
    col            = col_fun,
    cluster_rows   = FALSE,
    cluster_columns = FALSE,
    column_order   = seq_len(ncol(fc_mat)),
    cell_fun       = cell_fn,
    row_names_gp   = grid::gpar(fontsize = 9),
    column_names_gp = grid::gpar(fontsize = 9),
    column_title   = "Per-module logFC across 5 contrasts ('*' = FDR<0.05, '.' = FDR<0.1)",
    border         = TRUE
  )
}
```

```{r hdwgcna-module-enrichment}
# Module gene-set overlap (hypergeometric) against canonical microglia
# substate markers (homeostatic / DAM / IFN / proliferative).
# hdWGCNA stores module membership in Ensembl space; project gene sets are
# in symbol space. Map Ensembl -> symbol, drop unmapped rows, and run the
# hypergeometric in symbol space with the mapped universe.
if (!is.null(hdw)) {
  modules_sym <- hdw$modules
  modules_sym$symbol <- symbol_map$symbol[
    match(modules_sym$gene_name, symbol_map$ensembl)
  ]
  modules_sym <- modules_sym[
    !is.na(modules_sym$symbol) & nzchar(modules_sym$symbol), , drop = FALSE
  ]
  enr_universe <- unique(modules_sym$symbol)

  state_markers <- canonical_microglia_markers
  gene_sets <- setNames(state_markers, paste0("substate_", names(state_markers)))
  enr <- module_enrichment(modules_sym, gene_sets,
                            gene_col   = "symbol",
                            module_col = "module",
                            universe   = enr_universe)
  write_tsv_safe(enr,
                 file.path(params$results_dir,
                           "hdwgcna_module_enrichment.tsv"))
  cat(sprintf("Enrichment: universe=%d symbols, %d module x set tests.\n",
              length(enr_universe), nrow(enr)))
  enr_sig <- enr |>
    dplyr::filter(P.Value < 0.05) |>
    dplyr::arrange(P.Value) |>
    dplyr::select(module, gene_set, n_overlap, n_module, n_gene_set,
                  odds_ratio, P.Value, adj.P.Val, overlap)
  if (nrow(enr_sig) == 0) {
    cat("No module-vs-set overlap reached P<0.05.\n")
  } else {
    enr_print <- enr_sig
    enr_print$overlap <- vapply(strsplit(enr_print$overlap, ";"), function(g) {
      if (length(g) == 0L) return("")
      if (length(g) <= 8L) paste(g, collapse = ", ")
      else paste0(paste(g[1:8], collapse = ", "), ", ... (+", length(g) - 8L, ")")
    }, character(1))
    print(knitr::kable(head(enr_print, 25), digits = 3,
                       caption = paste0("Top module-vs-gene-set overlaps ",
                                        "(hypergeometric, P<0.05; top 25 shown)")))
  }
}
```

### Per-substate module-eigengene interaction

The whole-microglia MG-M3 collapse (`tau_in_nlgf` logFC -3.80, padj 6.7e-4)
could in principle be carried by a single substate or be uniform across
microglial substates. To distinguish, we pseudobulk the **cell-level
MG-Mx eigengenes per (state, genotype_batch)**, then re-fit the same 2x2
factorial + batch limma within each substate.

```{r hdwgcna-module-de-per-state, fig.width = 9, fig.height = 5}
if (!is.null(hdw) && !is.null(hdw_de)) {
  hdw_de_state <- fit_module_de_per_state(
    MEs_cells = hdw$MEs_cells,
    meta      = hdw$cell_meta,
    verbose   = FALSE
  )

  # Long-form: one row per (state, module, contrast)
  state_long <- dplyr::bind_rows(lapply(names(hdw_de_state), function(st) {
    fit_st <- hdw_de_state[[st]]
    if (is.null(fit_st)) return(NULL)
    dplyr::bind_rows(lapply(names(fit_st$top), function(ct) {
      dplyr::mutate(fit_st$top[[ct]], contrast = ct, state = st,
                    cells_used = fit_st$cells_used)
    }))
  }))
  state_long <- state_long[, c("state", "module", "contrast",
                               "logFC", "t", "P.Value", "adj.P.Val",
                               "cells_used")]
  write_tsv_safe(state_long,
                 file.path(params$results_dir,
                           "hdwgcna_module_de_per_state.tsv"))

  states_ord    <- c("homeostatic", "DAM", "IFN", "proliferative")
  contrasts_ord <- c("tau_alone", "nlgf_in_maptki", "nlgf_in_p301s",
                     "tau_in_nlgf", "interaction")
  modules_ord   <- sort(setdiff(unique(state_long$module),
                                c("grey", "MG-M0")))

  # Focused kable on MG-M3.
  mg3_state <- state_long |>
    dplyr::filter(module == "MG-M3") |>
    dplyr::mutate(
      state    = factor(state, levels = states_ord),
      contrast = factor(contrast, levels = contrasts_ord)
    ) |>
    dplyr::arrange(contrast, state) |>
    dplyr::select(state, contrast, logFC, P.Value, adj.P.Val, cells_used)
  print(knitr::kable(mg3_state, digits = 4,
                     caption = paste0("MG-M3 eigengene DE ",
                                       "per substate (n = 16 pseudo-",
                                       "samples per state, batch ",
                                       "covariate retained).")))

  # Per-state x contrast logFC heatmap, row-split by module.
  hm_long <- state_long |>
    dplyr::filter(module %in% modules_ord) |>
    dplyr::mutate(
      state    = factor(state, levels = states_ord),
      contrast = factor(contrast, levels = contrasts_ord),
      module   = factor(module, levels = modules_ord),
      row_key  = paste(module, state, sep = " | ")
    )
  row_keys <- as.character(unique(hm_long$row_key[order(hm_long$module,
                                                       hm_long$state)]))
  fc_mat <- matrix(NA_real_, nrow = length(row_keys),
                   ncol = length(contrasts_ord),
                   dimnames = list(row_keys, contrasts_ord))
  pa_mat <- fc_mat
  for (i in seq_len(nrow(hm_long))) {
    r <- as.character(hm_long$row_key[i])
    c <- as.character(hm_long$contrast[i])
    fc_mat[r, c] <- hm_long$logFC[i]
    pa_mat[r, c] <- hm_long$adj.P.Val[i]
  }
  row_split <- factor(sub(" \\| .+", "", row_keys), levels = modules_ord)
  rng <- max(abs(fc_mat[is.finite(fc_mat)]), na.rm = TRUE)
  col_fun <- circlize::colorRamp2(c(-rng, 0, rng),
                                  c("#3a4cc0", "white", "#b40426"))
  cell_fn <- function(j, i, x, y, width, height, fill) {
    pa <- pa_mat[i, j]
    if (is.na(pa)) return(invisible(NULL))
    if (pa < 0.05) grid::grid.text("*", x, y,
                                   gp = grid::gpar(fontsize = 10,
                                                   col = "black"))
    else if (pa < 0.1) grid::grid.text(".", x, y,
                                       gp = grid::gpar(fontsize = 10,
                                                       col = "black"))
  }
  short_rows <- sub("^MG-M[0-9]+ \\| ", "", row_keys)
  ComplexHeatmap::Heatmap(
    fc_mat,
    name            = "logFC",
    col             = col_fun,
    cluster_rows    = FALSE,
    cluster_columns = FALSE,
    column_order    = seq_len(ncol(fc_mat)),
    row_split       = row_split,
    row_title_rot   = 0,
    row_labels      = short_rows,
    row_names_gp    = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 9),
    cell_fun        = cell_fn,
    column_title    = paste0("Per-substate module-eigengene DE",
                             " ('*' = FDR<0.05, '.' = FDR<0.1)"),
    border          = TRUE
  )
}
```

### Module hub connectivity vs whole-microglia interaction logFC

If a module captures a coherent transcriptional response, its
**most-connected** members (highest within-module `kME`) should also
carry the **largest per-gene interaction logFCs**. Below we plot, for
every gene in each non-grey module, the module-internal kME against the
whole-microglia NEBULA interaction `logFC`.

```{r hdwgcna-kme-vs-nebula, fig.width = 10, fig.height = 8}
if (!is.null(hdw) && exists("de_snrnaseq_nebula") &&
    !is.null(de_snrnaseq_nebula)) {
  nebula_int <- de_snrnaseq_nebula$top$interaction
  # NEBULA top stores Ensembl `gene` + `symbol`; modules store `gene_name`
  # (Ensembl). Use Ensembl for the join to avoid symbol-ambiguity.
  modules_mg <- hdw$modules
  modules_mg$symbol <- symbol_map$symbol[
    match(modules_mg$gene_name, symbol_map$ensembl)
  ]
  modules_mg <- modules_mg[modules_mg$module != "grey", , drop = FALSE]

  # Long-form: gene x module-of-membership with own-module kME +
  # NEBULA interaction stats.
  kme_long <- do.call(rbind, lapply(unique(as.character(modules_mg$module)),
                                    function(m) {
    own_kME_col <- paste0("kME_", m)
    if (!own_kME_col %in% colnames(modules_mg)) return(NULL)
    sub <- modules_mg[modules_mg$module == m, , drop = FALSE]
    sub$kME_own <- sub[[own_kME_col]]
    sub$module  <- m
    sub[, c("gene_name", "symbol", "module", "kME_own"), drop = FALSE]
  }))
  kme_long <- merge(kme_long, nebula_int[, c("gene", "logFC", "P.Value",
                                              "adj.P.Val")],
                    by.x = "gene_name", by.y = "gene", all.x = TRUE)

  # Per-module Spearman rho between kME and |logFC|; reported alongside.
  cor_tbl <- kme_long |>
    dplyr::filter(!is.na(logFC) & !is.na(kME_own)) |>
    dplyr::group_by(module) |>
    dplyr::summarise(
      n             = dplyr::n(),
      rho_signed    = suppressWarnings(cor(kME_own, logFC,
                                            method = "spearman")),
      rho_absolute  = suppressWarnings(cor(kME_own, abs(logFC),
                                            method = "spearman")),
      .groups       = "drop"
    )
  print(knitr::kable(cor_tbl, digits = 3,
                     caption = paste0("kME (own module) vs whole-microglia",
                                       " NEBULA interaction logFC - ",
                                       "Spearman rho, per module")))

  # Facet plot: one panel per module, x = kME, y = logFC, colour = LE.
  modules_ord <- sort(setdiff(unique(as.character(kme_long$module)),
                              c("grey", "MG-M0")))
  kme_long$module <- factor(kme_long$module, levels = modules_ord)
  p <- ggplot2::ggplot(
    kme_long[!is.na(kme_long$logFC), ],
    ggplot2::aes(x = kME_own, y = logFC)
  ) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed",
                        colour = "grey60") +
    ggplot2::geom_point(colour = "grey50", alpha = 0.5, size = 0.7) +
    ggplot2::geom_smooth(method = "lm", formula = y ~ x,
                         colour = "black", se = FALSE,
                         linewidth = 0.4) +
    ggplot2::facet_wrap(~ module, scales = "free_x") +
    ggplot2::labs(
      title = paste0("Module kME vs whole-microglia NEBULA ",
                     "interaction logFC"),
      x     = "kME (within own module)",
      y     = "NEBULA interaction logFC"
    ) +
    ggplot2::theme_bw()
  print(p)

  # Export full long-form for follow-up.
  write_tsv_safe(kme_long,
                 file.path(params$results_dir,
                           "hdwgcna_kme_vs_nebula_interaction.tsv"))
}
```

```{r hdwgcna-modules-tsv}
# Export full module assignment + kME table for downstream use, plus a
# module-level summary with the top hub genes (as symbols) per module.
# The displayed hub list excludes likely red-blood-cell contamination
# (Hbb-bs etc.) via `flag_contaminant_hubs()`; the raw `hdwgcna_modules.
# tsv` and `hdwgcna_module_hubs.tsv` exports retain the full unfiltered
# assignment so downstream consumers can see the contamination flag.
if (!is.null(hdw)) {
  mod_tbl <- hdw$modules
  mod_tbl$symbol <- symbol_map$symbol[
    match(mod_tbl$gene_name, symbol_map$ensembl)
  ]
  keep_cols <- intersect(c("gene_name", "symbol", "module", "color",
                           grep("^kME_", colnames(mod_tbl), value = TRUE)),
                         colnames(mod_tbl))
  write_tsv_safe(mod_tbl[, keep_cols, drop = FALSE],
                 file.path(params$results_dir, "hdwgcna_modules.tsv"))

  hub_sym <- hdw$hub_genes
  hub_sym$symbol <- symbol_map$symbol[
    match(hub_sym$gene_name, symbol_map$ensembl)
  ]
  hub_sym$symbol <- ifelse(is.na(hub_sym$symbol) | !nzchar(hub_sym$symbol),
                           hub_sym$gene_name, hub_sym$symbol)
  hub_sym <- flag_contaminant_hubs(hub_sym)
  write_tsv_safe(hub_sym,
                 file.path(params$results_dir, "hdwgcna_module_hubs.tsv"))

  n_flagged <- sum(hub_sym$is_likely_rbc)
  if (n_flagged > 0L) {
    flagged_rows <- hub_sym[hub_sym$is_likely_rbc,
                            c("symbol", "module", "kME"), drop = FALSE]
    cat(sprintf(paste0("RBC-contaminant filter flagged %d hub row(s); ",
                        "excluded from the displayed top-5 below.\n"),
                n_flagged))
    print(knitr::kable(flagged_rows, digits = 3,
                       caption = paste0("Hub genes flagged as likely",
                                         " RBC contamination",
                                         " (excluded from display only)")))
  }

  hub_top5 <- hub_sym |>
    dplyr::filter(!is.na(display_rank) & display_rank <= 5L) |>
    dplyr::arrange(module, display_rank) |>
    dplyr::group_by(module) |>
    dplyr::summarise(hubs_top5 = paste(symbol, collapse = ", "),
                     .groups   = "drop")
  module_summary <- hdw$module_meta |>
    dplyr::left_join(hub_top5, by = "module") |>
    dplyr::filter(module != "grey")
  print(knitr::kable(module_summary, digits = 3,
                     caption = paste0("Module summary: size, colour, ",
                                       "top-5 hub genes by kME ",
                                       "(RBC contaminants filtered out)")))
}
```
