## hdWGCNA soft-power sensitivity

The auto-pick rule (lowest power with scale-free R^2 >= 0.8) chose
soft-power 3 for the canonical build. This sub-section stress-tests
that choice by rerunning the entire pipeline at p=6 and p=8 (see
`scripts/build_hdwgcna.R --soft-power N`) and asking whether the
MG-M3 module persists, dissolves into MG-M2 (DAM), or fragments.
The chunks below no-op gracefully if the sensitivity caches are
absent so a default knit on a partially-built tree still succeeds.

```{r hdwgcna-soft-power-load}
# Load p=3 (canonical) + p=6 + p=8 hdWGCNA builds into a single named
# list `hdw_scan`. Only entries with both caches present are kept; any
# missing run is skipped downstream rather than failing the knit.
hdw_scan <- list()
if (!is.null(hdw) && !is.null(hdw_de)) {
  hdw_scan[["p3"]] <- list(hdw = hdw, hdw_de = hdw_de)
}
for (p in c(6L, 8L)) {
  c_micro <- file.path(params$cache_dir,
                       sprintf("hdwgcna_microglia_p%d.rds", p))
  c_de    <- file.path(params$cache_dir,
                       sprintf("hdwgcna_module_de_p%d.rds", p))
  if (file.exists(c_micro) && file.exists(c_de)) {
    hdw_scan[[sprintf("p%d", p)]] <- list(
      hdw    = readRDS(c_micro),
      hdw_de = readRDS(c_de)
    )
  }
}
cat(sprintf("Soft-power scan: %d build(s) loaded (%s).\n",
            length(hdw_scan),
            paste(names(hdw_scan), collapse = ", ")))
if (length(hdw_scan) >= 2L) {
  cat("Chosen soft_power per run:\n")
  # soft_power is stored as integer in override builds (p6/p8) and as
  # double in the auto-pick build (p3); coerce both to a numeric vector.
  print(vapply(hdw_scan, function(x) x$hdw$soft_power, numeric(1L)))
}
```

```{r hdwgcna-soft-power-curve, fig.width = 9, fig.height = 4}
# Scale-free R^2 + mean connectivity vs power, with the chosen-per-run
# power highlighted. The power tables are derived from identical
# metacells/datExpr (seed fixed) so we draw one curve and mark all three
# chosen points on it.
if (length(hdw_scan) >= 2L) {
  pt <- hdw_scan[["p3"]]$hdw$power_table
  chosen <- vapply(hdw_scan, function(x) x$hdw$soft_power, numeric(1L))
  pt$tag <- factor(ifelse(pt$Power %in% chosen,
                          paste0("p=", pt$Power), NA_character_),
                   levels = paste0("p=", sort(unique(chosen))))

  p1 <- ggplot2::ggplot(pt, ggplot2::aes(Power, SFT.R.sq)) +
    ggplot2::geom_hline(yintercept = 0.8, linetype = "dashed",
                        colour = "grey60") +
    ggplot2::geom_line(colour = "grey40") +
    ggplot2::geom_point(size = 2) +
    ggplot2::geom_point(data = subset(pt, !is.na(tag)),
                        ggplot2::aes(colour = tag), size = 3.5) +
    ggplot2::scale_colour_brewer("chosen", palette = "Set1") +
    ggplot2::labs(title = "Scale-free topology fit",
                  x = "soft-thresholding power",
                  y = expression(R^2)) +
    ggplot2::theme_bw()
  p2 <- ggplot2::ggplot(pt, ggplot2::aes(Power, mean.k.)) +
    ggplot2::geom_line(colour = "grey40") +
    ggplot2::geom_point(size = 2) +
    ggplot2::geom_point(data = subset(pt, !is.na(tag)),
                        ggplot2::aes(colour = tag), size = 3.5) +
    ggplot2::scale_colour_brewer("chosen", palette = "Set1") +
    ggplot2::scale_y_log10() +
    ggplot2::labs(title = "Mean connectivity",
                  x = "soft-thresholding power",
                  y = "mean k (log10)") +
    ggplot2::theme_bw()
  print(patchwork::wrap_plots(p1, p2, ncol = 2))
}
```

```{r hdwgcna-soft-power-modules}
# Per-run module count (incl. grey) and size summary. Higher power
# expectations: fewer genes pass the connectivity threshold, more sit
# in grey, surviving modules are tighter.
if (length(hdw_scan) >= 2L) {
  size_summary <- do.call(rbind, lapply(names(hdw_scan), function(run) {
    mm <- hdw_scan[[run]]$hdw$module_meta
    data.frame(
      power         = run,
      soft_power    = hdw_scan[[run]]$hdw$soft_power,
      n_modules_all = nrow(mm),
      n_modules_nongrey = sum(mm$module != "grey"),
      n_genes_total = sum(mm$n_genes),
      n_genes_grey  = sum(mm$n_genes[mm$module == "grey"]),
      n_genes_assigned = sum(mm$n_genes[mm$module != "grey"]),
      largest_module = mm$module[which.max(mm$n_genes - 1e9 * (mm$module == "grey"))],
      largest_module_size = max(mm$n_genes[mm$module != "grey"])
    )
  }))
  print(knitr::kable(size_summary,
                     caption = "Module count and size per soft-power run"))

  per_module <- do.call(rbind, lapply(names(hdw_scan), function(run) {
    mm <- hdw_scan[[run]]$hdw$module_meta
    data.frame(power = run, module = as.character(mm$module),
               n_genes = mm$n_genes)
  }))
  per_module_wide <- tidyr::pivot_wider(per_module, names_from = power,
                                        values_from = n_genes,
                                        values_fill = 0L)
  print(knitr::kable(per_module_wide,
                     caption = "Genes per module across soft-power runs (0 = absent)"))
}
```

```{r hdwgcna-soft-power-overlap, fig.width = 11, fig.height = 4.5}
# Cross-tabulation of module assignment between p=3 and each other run.
# We expect (incl. grey):
#   - diagonal/dominant mapping = module preservation (good for MG-M3 robustness),
#   - off-diagonal smear / dissolution into grey = power destabilises modules.
# Adjusted Rand Index summarises the overall agreement (1 = identical).
ari <- function(a, b) {
  # Pair-counting ARI on two integer/character partitions of equal length.
  # No external package; uses the standard contingency formula.
  tab <- table(a, b)
  n   <- sum(tab)
  ch2 <- function(x) x * (x - 1L) / 2
  Cn  <- ch2(n)
  if (Cn == 0L) return(NA_real_)
  sum_ai <- sum(ch2(rowSums(tab)))
  sum_bj <- sum(ch2(colSums(tab)))
  exp_ix <- sum_ai * sum_bj / Cn
  max_ix <- 0.5 * (sum_ai + sum_bj)
  if (max_ix == exp_ix) return(NA_real_)
  (sum(ch2(tab)) - exp_ix) / (max_ix - exp_ix)
}

if (length(hdw_scan) >= 2L) {
  mods_by_power <- lapply(hdw_scan, function(x)
    x$hdw$modules[, c("gene_name", "module"), drop = FALSE])
  # All runs see the same datExpr (gene_fraction = 0.05, identical metacells)
  # so gene_name sets should match. Inner-join defensively anyway.
  ari_vec <- numeric(0)
  for (run in setdiff(names(hdw_scan), "p3")) {
    j <- dplyr::inner_join(mods_by_power[["p3"]], mods_by_power[[run]],
                            by = "gene_name", suffix = c("_p3", paste0("_", run)))
    a <- as.character(j[[paste0("module_p3")]])
    b <- as.character(j[[paste0("module_", run)]])
    ari_vec[run] <- ari(a, b)
  }
  cat("Adjusted Rand Index vs p=3 (incl. grey, all genes):\n")
  print(round(ari_vec, 3))

  # Cross-tabulation heatmaps (p=3 rows x p=N cols), one panel per
  # comparison run. Row/col order: numeric module name order, grey last.
  mod_order <- function(m) {
    m <- as.character(m)
    nums <- suppressWarnings(as.integer(sub("^MG-M", "", m)))
    nums[is.na(nums)] <- 1e6  # grey to the end
    m[order(nums)]
  }
  panels <- lapply(setdiff(names(hdw_scan), "p3"), function(run) {
    j <- dplyr::inner_join(mods_by_power[["p3"]], mods_by_power[[run]],
                            by = "gene_name", suffix = c("_p3", paste0("_", run)))
    a <- factor(j[[paste0("module_p3")]],
                levels = mod_order(unique(j[[paste0("module_p3")]])))
    b <- factor(j[[paste0("module_", run)]],
                levels = mod_order(unique(j[[paste0("module_", run)]])))
    xt <- as.matrix(unclass(table(a, b)))
    list(run = run, xt = xt)
  })

  # Use a shared colour scale across panels so cross-panel reading is fair.
  max_count <- max(vapply(panels, function(p) max(p$xt), numeric(1L)))
  col_fun <- circlize::colorRamp2(c(0, max_count / 2, max_count),
                                  c("white", "#fec44f", "#993404"))

  # Draw heatmaps separately (rather than ComplexHeatmap's `+`
  # concatenation) to avoid the row-alignment requirement when the
  # comparison runs differ in module count.
  for (panel in panels) {
    ComplexHeatmap::draw(
      ComplexHeatmap::Heatmap(
        panel$xt,
        name = "n_genes",
        col  = col_fun,
        cluster_rows    = FALSE,
        cluster_columns = FALSE,
        row_order       = seq_len(nrow(panel$xt)),
        column_order    = seq_len(ncol(panel$xt)),
        row_title       = "p=3 module",
        column_title    = sprintf("%s module  (ARI vs p=3 = %.3f)",
                                  panel$run, ari_vec[panel$run]),
        cell_fun = function(j, i, x, y, w, h, fill) {
          v <- panel$xt[i, j]
          if (v > 0) grid::grid.text(v, x, y,
                                     gp = grid::gpar(fontsize = 7))
        },
        row_names_gp    = grid::gpar(fontsize = 9),
        column_names_gp = grid::gpar(fontsize = 9),
        border          = TRUE
      )
    )
  }
}
```

```{r hdwgcna-soft-power-mg3-tracking}
# Track MG-M3 (the canonical p=3 module of interest) through p=6 and p=8:
#   (a) Jaccard of MG-M3 vs each non-grey module at the other powers ->
#       identify the "MG-M3-equivalent" module at each power.
#   (b) Map each MG-M3 hub gene to its module at p=6 and p=8; tabulate
#       the destination distribution. If most hubs land in the same
#       non-grey module, MG-M3 persists. If they scatter or sink to grey,
#       the module is power-fragile.
if (length(hdw_scan) >= 2L) {
  mg3_genes_p3 <- mods_by_power[["p3"]]$gene_name[
    mods_by_power[["p3"]]$module == "MG-M3"
  ]
  cat(sprintf("MG-M3 at p=3 has %d genes.\n", length(mg3_genes_p3)))

  pairwise_jaccard <- function(target_genes, mods_other) {
    by_mod <- split(mods_other$gene_name, as.character(mods_other$module))
    sapply(by_mod, function(g) {
      length(intersect(g, target_genes)) /
        max(1L, length(union(g, target_genes)))
    })
  }
  jacc <- lapply(setdiff(names(hdw_scan), "p3"), function(run) {
    j <- pairwise_jaccard(mg3_genes_p3, mods_by_power[[run]])
    data.frame(power = run, module = names(j), jaccard = unname(j))
  })
  jacc_tbl <- do.call(rbind, jacc)
  jacc_tbl <- jacc_tbl[order(jacc_tbl$power, -jacc_tbl$jaccard), , drop = FALSE]
  # Best non-grey module per power
  jacc_nongrey <- jacc_tbl[jacc_tbl$module != "grey", , drop = FALSE]
  best_tbl <- jacc_nongrey |>
    dplyr::group_by(power) |>
    dplyr::slice_max(jaccard, n = 1L, with_ties = FALSE) |>
    dplyr::ungroup() |>
    as.data.frame()
  print(knitr::kable(best_tbl, digits = 3,
                     caption = paste0("Best-matching non-grey module to ",
                                       "p=3 MG-M3 at each power")))
  print(knitr::kable(jacc_tbl, digits = 3,
                     caption = paste0("Per-module Jaccard vs p=3 MG-M3 ",
                                       "at each power (grey shown)")))

  # Hub-level tracking: where do MG-M3 hubs at p=3 land?
  mg3_hubs <- hdw_scan[["p3"]]$hdw$hub_genes
  mg3_hubs <- mg3_hubs[mg3_hubs$module == "MG-M3", , drop = FALSE]
  mg3_hubs$symbol <- symbol_map$symbol[
    match(mg3_hubs$gene_name, symbol_map$ensembl)
  ]
  mg3_hubs$symbol <- ifelse(is.na(mg3_hubs$symbol) | !nzchar(mg3_hubs$symbol),
                            mg3_hubs$gene_name, mg3_hubs$symbol)
  for (run in setdiff(names(hdw_scan), "p3")) {
    mg3_hubs[[paste0("module_", run)]] <- mods_by_power[[run]]$module[
      match(mg3_hubs$gene_name, mods_by_power[[run]]$gene_name)
    ]
  }
  hub_display <- mg3_hubs[, c("symbol", "kME",
                              grep("^module_p", colnames(mg3_hubs),
                                   value = TRUE)),
                          drop = FALSE]
  hub_display <- hub_display[order(-hub_display$kME), , drop = FALSE]
  print(knitr::kable(hub_display, digits = 3,
                     caption = "MG-M3 hub gene destinations at higher powers"))

  hub_dest_summary <- do.call(rbind, lapply(
    setdiff(names(hdw_scan), "p3"),
    function(run) {
      v <- mg3_hubs[[paste0("module_", run)]]
      out <- as.data.frame(table(v, useNA = "ifany"))
      out$power <- run
      out
    }))
  colnames(hub_dest_summary)[1L] <- "destination_module"
  hub_dest_summary <- hub_dest_summary[, c("power", "destination_module", "Freq")]
  print(knitr::kable(hub_dest_summary,
                     caption = "MG-M3 hub destination tally per power"))

  write_tsv_safe(jacc_tbl,
                 file.path(params$results_dir,
                           "hdwgcna_soft_power_mg3_jaccard.tsv"))
  write_tsv_safe(hub_display,
                 file.path(params$results_dir,
                           "hdwgcna_soft_power_mg3_hubs.tsv"))
}
```

```{r hdwgcna-soft-power-de}
# Per-power module DE: how many modules survive FDR<0.05 at each contrast?
# We expect the MG-M3-equivalent module to retain its strong tau_in_nlgf /
# interaction signal if MG-M3 truly represents a coherent transcriptional
# response (and not an artefact of the soft-power choice).
if (length(hdw_scan) >= 2L) {
  de_summary <- do.call(rbind, lapply(names(hdw_scan), function(run) {
    do.call(rbind, lapply(names(hdw_scan[[run]]$hdw_de$top), function(ct) {
      t <- hdw_scan[[run]]$hdw_de$top[[ct]]
      data.frame(
        power     = run,
        contrast  = ct,
        n_modules = nrow(t),
        n_sig     = sum(t$adj.P.Val < 0.05, na.rm = TRUE),
        min_padj  = suppressWarnings(min(t$adj.P.Val, na.rm = TRUE))
      )
    }))
  }))
  de_summary_wide <- tidyr::pivot_wider(
    de_summary[, c("power", "contrast", "n_sig")],
    names_from = power, values_from = n_sig, values_fill = 0L,
    names_prefix = "n_sig_"
  )
  print(knitr::kable(de_summary_wide,
                     caption = paste0("n_sig at FDR<0.05 per contrast, ",
                                       "per soft-power run")))

  # MG-M3-equivalent module DE across the 5 contrasts -- the headline
  # robustness check. Pull MG-M3 (p=3) and the best-Jaccard module per
  # higher power; compare logFC + padj on each contrast.
  best_per_run <- setNames(
    c("MG-M3", best_tbl$module),
    c("p3", as.character(best_tbl$power))
  )
  mg3_de_xpower <- do.call(rbind, lapply(names(best_per_run), function(run) {
    mod <- best_per_run[[run]]
    do.call(rbind, lapply(names(hdw_scan[[run]]$hdw_de$top), function(ct) {
      t <- hdw_scan[[run]]$hdw_de$top[[ct]]
      r <- t[as.character(t$module) == mod, , drop = FALSE]
      if (nrow(r) == 0L) return(NULL)
      data.frame(
        power     = run,
        module    = mod,
        contrast  = ct,
        logFC     = r$logFC,
        P.Value   = r$P.Value,
        adj.P.Val = r$adj.P.Val
      )
    }))
  }))
  print(knitr::kable(mg3_de_xpower, digits = 4,
                     caption = paste0("DE on the MG-M3-equivalent module ",
                                       "across soft-power runs")))

  write_tsv_safe(de_summary,
                 file.path(params$results_dir,
                           "hdwgcna_soft_power_de_summary.tsv"))
  write_tsv_safe(mg3_de_xpower,
                 file.path(params$results_dir,
                           "hdwgcna_soft_power_mg3_de.tsv"))
}
```

```{r hdwgcna-soft-power-summary}
# Plain-English headline: how robust is MG-M3 to soft-power choice?
# The check is binary: (a) does the best-matching module at each higher
# power overlap >= 0.5 with p=3 MG-M3, and (b) does its tau_in_nlgf
# logFC remain in the same direction with comparable magnitude?
if (length(hdw_scan) >= 2L && exists("best_tbl") && exists("mg3_de_xpower")) {
  mg3_p3_tau <- mg3_de_xpower$logFC[
    mg3_de_xpower$power == "p3" & mg3_de_xpower$contrast == "tau_in_nlgf"
  ]
  for (rr in seq_len(nrow(best_tbl))) {
    run  <- as.character(best_tbl$power[rr])
    mod  <- as.character(best_tbl$module[rr])
    jc   <- best_tbl$jaccard[rr]
    tau_lfc <- mg3_de_xpower$logFC[
      mg3_de_xpower$power == run & mg3_de_xpower$contrast == "tau_in_nlgf"
    ]
    tau_padj <- mg3_de_xpower$adj.P.Val[
      mg3_de_xpower$power == run & mg3_de_xpower$contrast == "tau_in_nlgf"
    ]
    same_sign <- length(tau_lfc) == 1L && length(mg3_p3_tau) == 1L &&
      !is.na(tau_lfc) && !is.na(mg3_p3_tau) &&
      sign(tau_lfc) == sign(mg3_p3_tau)
    verdict <- if (isTRUE(jc >= 0.5) && same_sign) {
      "PERSISTS"
    } else if (isTRUE(jc >= 0.3)) {
      "RESHAPES (moderate overlap)"
    } else {
      "DISSOLVES"
    }
    cat(sprintf(paste0("Soft-power %s: MG-M3 -> %s (Jaccard %.3f). ",
                        "tau_in_nlgf logFC %.2f (padj %.2g). [%s]\n"),
                run, mod, jc,
                if (length(tau_lfc)  == 1L) tau_lfc  else NA_real_,
                if (length(tau_padj) == 1L) tau_padj else NA_real_,
                verdict))
  }
}
```
