# GeoMx: normalisation and DE

The GeoMx `RNA` assay carries two layers: `counts` holds essentially
raw integer counts (99.98% of nonzero values are integer; the few
fractional residues come from upstream GeomxTools background steps),
and `data` holds the vendor Q3-normalised values: exactly
`counts / q_norm_qFactors`, where `q_norm_qFactors` lives in
`meta.data` (median 1.04, range 0.10-7.49). The canonical fit below
reads the `counts` layer (coerced to integer storage), so it is a
**raw-counts + voom + edgeR TMM** fit with `slide name` as a
fixed-effect covariate. No Q3 normalisation is applied here.

Two complementary normalisation audits sit alongside this canonical
fit: `rmd/09c_geomx_audit.Rmd` swaps `slide` from a fixed-effect block
to a random effect (via `limma::duplicateCorrelation`), and
`rmd/09d_geomx_q3_audit.Rmd` adds Q3 back into the pipeline via two
variants (Q3 → log2 → `limma-trend`, and Q3 → `voom + TMM`) so the
+OXPHOS NES can be evaluated under orthogonal normalisation choices.

```{r geomx-de}
de_gx_path <- file.path(params$cache_dir, "de_geomx.rds")
de_geomx <- cache_or_run(de_gx_path, {
  DefaultAssay(geomx) <- "RNA"
  cnt <- as.matrix(GetAssayData(geomx, assay = "RNA", layer = "counts"))
  storage.mode(cnt) <- "integer"
  meta <- geomx@meta.data
  meta$genotype <- factor(as.character(meta$genotype), levels = genotype_levels)
  meta$slide <- factor(meta$`slide name`)
  design <- model.matrix(~ 0 + genotype + slide, data = meta)
  colnames(design) <- sub("^genotype", "", colnames(design))
  colnames(design) <- gsub("[^A-Za-z0-9_]", "_", colnames(design))
  cm <- make_contrast_matrix(design)
  fit <- fit_limma_voom(cnt, group = meta$genotype, design = design,
                        contrasts = cm, min_count = 5)
  for (cn in names(fit$top)) {
    fit$top[[cn]] <- dplyr::rename(fit$top[[cn]], symbol = gene)
  }
  list(meta = meta, design = design, contrasts = cm, fit = fit)
})

gx_summary <- purrr::imap_dfr(de_geomx$fit$top, function(tbl, cn) {
  data.frame(
    contrast = cn,
    n_total  = nrow(tbl),
    n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
    n_up     = sum(tbl$adj.P.Val < 0.05 & tbl$logFC > 0, na.rm = TRUE),
    n_down   = sum(tbl$adj.P.Val < 0.05 & tbl$logFC < 0, na.rm = TRUE)
  )
})
knitr::kable(gx_summary, caption = "GeoMx DE: significant gene counts")
```

```{r geomx-volcano, fig.width = 10, fig.height = 4}
(volcano(de_geomx$fit$top$nlgf_in_maptki, "GeoMx: NLGF in MAPTKI") |
 volcano(de_geomx$fit$top$nlgf_in_p301s,  "GeoMx: NLGF in P301S"))
```
