# Bulk proteomics: aggregate to protein groups, normalise, DE

The proteomics TSV is a Spectronaut-style PTM report containing
phospho-, oxidation- and carbamidomethyl-modified peptides from the
TiO2-enriched fraction. We aggregate intensities per `PG.ProteinGroups`
(summing peptide-level intensities across all rows of the same PG), log2,
median-normalise across samples and fit `limma`.

```{r proteomics-de}
prot_path <- file.path(params$cache_dir, "de_proteomics.rds")
de_proteomics <- cache_or_run(prot_path, {
  # Identify intensity columns.
  int_cols <- grep("PTM\\.Quantity$", colnames(prot_raw), value = TRUE)
  meta_map <- match_intensity_columns(int_cols, sample_key)
  use_cols <- meta_map$column[!is.na(meta_map$genotype)]
  meta <- meta_map[!is.na(meta_map$genotype), , drop = FALSE]
  meta$genotype <- factor(meta$genotype, levels = genotype_levels)
  rownames(meta) <- meta$column

  # Aggregate to PG by summing peptide intensities (matching MaxLFQ-style behaviour).
  pg <- prot_raw$PG.ProteinGroups
  mat <- as.matrix(prot_raw[, use_cols])
  mat[mat == 0 | !is.finite(mat)] <- NA_real_

  ag <- rowsum(mat, group = pg, na.rm = TRUE)
  # Restore NAs where all entries were NA after rowsum (rowsum yields 0).
  zero_to_na <- function(m) { m[m == 0] <- NA_real_; m }
  ag <- zero_to_na(ag)
  log_mat <- log2(ag)
  log_mat <- log_mat[rowSums(!is.na(log_mat)) >= 8, ]
  log_mat <- median_normalise(log_mat)
  log_mat <- prevalence_filter(log_mat, meta$genotype, min_present = 3, min_groups = 2)

  design <- model.matrix(~ 0 + genotype, data = meta)
  colnames(design) <- sub("^genotype", "", colnames(design))
  cm <- make_contrast_matrix(design)
  fit <- fit_limma_log(log_mat, group = meta$genotype, design = design,
                       contrasts = cm)
  # Map PG -> gene symbol via the first PG.Genes for that PG.
  pg_gene <- prot_raw |>
    dplyr::distinct(PG.ProteinGroups, PG.Genes) |>
    dplyr::group_by(PG.ProteinGroups) |>
    dplyr::summarise(symbol = dplyr::first(PG.Genes), .groups = "drop")
  for (cn in names(fit$top)) {
    fit$top[[cn]] <- dplyr::left_join(fit$top[[cn]],
      pg_gene, by = c("feature" = "PG.ProteinGroups"))
  }
  list(meta = meta, design = design, contrasts = cm, fit = fit,
       log_matrix = log_mat)
})

cat(sprintf("Proteomics: %d PGs analysed, %d samples\n",
            nrow(de_proteomics$log_matrix), nrow(de_proteomics$meta)))

prot_summary <- purrr::imap_dfr(de_proteomics$fit$top, function(tbl, cn) {
  data.frame(
    contrast = cn,
    n_total  = nrow(tbl),
    n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
    n_up     = sum(tbl$adj.P.Val < 0.05 & tbl$logFC > 0, na.rm = TRUE),
    n_down   = sum(tbl$adj.P.Val < 0.05 & tbl$logFC < 0, na.rm = TRUE)
  )
})
knitr::kable(prot_summary, caption = "Proteomics DE: significant PG counts")
```

```{r proteomics-pca, fig.width = 5, fig.height = 4}
pc <- prcomp(t(de_proteomics$log_matrix |>
                 (function(m) { m[is.na(m)] <- 0; m })()),
             center = TRUE, scale. = FALSE)
pc_df <- data.frame(
  PC1 = pc$x[, 1], PC2 = pc$x[, 2],
  genotype = de_proteomics$meta$genotype,
  label = de_proteomics$meta$label
)
ggplot(pc_df, aes(PC1, PC2, colour = genotype, label = label)) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  scale_colour_manual(values = genotype_colours) +
  theme_bw() + labs(title = "Proteomics PCA (zero-imputed)")
```
