## Phospho-protein stoichiometric correction

Raw phosphosite log2FC mixes two effects: changes in the abundance of the
parent protein and changes in the *stoichiometry* of phosphorylation at
that site. To isolate the latter, we subtract the matched parent-protein
log2-intensity from each phosphosite log2-intensity per sample, then refit
limma. Sites whose parent protein is missing from the proteomics matrix
are dropped. Sample pairing across the two modalities is by `key_idx`
(both datasets cover DIA_01..16 of the same animals).

```{r phospho-protein-correction}
phos_corr_path <- file.path(params$cache_dir, "de_phospho_corrected.rds")

de_phospho_corr <- cache_or_run(phos_corr_path, {
  # Site metadata: map Gene-pSite -> parent PG.ProteinGroups.
  site_meta <- phos_raw |>
    dplyr::distinct(`Gene-pSite`, PG.Genes, PG.ProteinGroups,
                    PTM.SiteAA, PTM.SiteLocation) |>
    dplyr::filter(!is.na(`Gene-pSite`), `Gene-pSite` != "")

  phospho_log <- de_phospho$log_matrix
  protein_log <- de_proteomics$log_matrix

  # Pair phospho columns with proteomics columns by key_idx (sample identity).
  ph_key <- de_phospho$meta$key_idx[match(colnames(phospho_log),
                                          de_phospho$meta$column)]
  pr_key <- de_proteomics$meta$key_idx[match(colnames(protein_log),
                                             de_proteomics$meta$column)]
  pr_for_ph <- match(ph_key, pr_key)
  stopifnot(!any(is.na(pr_for_ph)))                # all 16 phospho samples must match
  protein_log_aligned <- protein_log[, pr_for_ph, drop = FALSE]
  colnames(protein_log_aligned) <- colnames(phospho_log)

  # Lookup parent PG per site; keep only sites with a parent in the proteomics matrix.
  site_to_pg <- site_meta$PG.ProteinGroups[match(rownames(phospho_log),
                                                 site_meta$`Gene-pSite`)]
  has_protein <- !is.na(site_to_pg) & site_to_pg %in% rownames(protein_log_aligned)
  message(sprintf(
    "[phospho-corr] %d / %d sites have a matched parent protein.",
    sum(has_protein), length(has_protein)))

  corr_mat <- phospho_log[has_protein, , drop = FALSE] -
              protein_log_aligned[site_to_pg[has_protein], , drop = FALSE]
  rownames(corr_mat) <- rownames(phospho_log)[has_protein]

  # Re-apply prevalence filter on the corrected matrix (subtraction can introduce
  # NAs where either component was missing).
  corr_mat <- prevalence_filter(corr_mat, de_phospho$meta$genotype,
                                min_present = 3, min_groups = 2)

  meta <- de_phospho$meta
  design <- model.matrix(~ 0 + genotype, data = meta)
  colnames(design) <- sub("^genotype", "", colnames(design))
  cm <- make_contrast_matrix(design)
  fit <- fit_limma_log(corr_mat, group = meta$genotype, design = design,
                       contrasts = cm)
  for (cn in names(fit$top)) {
    fit$top[[cn]] <- dplyr::left_join(fit$top[[cn]], site_meta,
      by = c("feature" = "Gene-pSite")) |>
      dplyr::rename(symbol = PG.Genes)
  }
  list(meta = meta, design = design, contrasts = cm, fit = fit,
       log_matrix = corr_mat,
       n_sites_with_parent = sum(has_protein),
       n_sites_tested      = nrow(corr_mat))
})

cat(sprintf("Phospho-corrected: %d sites with parent protein; %d retained after filter.\n",
            de_phospho_corr$n_sites_with_parent,
            de_phospho_corr$n_sites_tested))

phos_corr_summary <- purrr::imap_dfr(de_phospho_corr$fit$top, function(tbl, cn) {
  data.frame(
    contrast = cn,
    n_total  = nrow(tbl),
    n_sig    = sum(tbl$adj.P.Val < 0.05, na.rm = TRUE),
    n_up     = sum(tbl$adj.P.Val < 0.05 & tbl$logFC > 0, na.rm = TRUE),
    n_down   = sum(tbl$adj.P.Val < 0.05 & tbl$logFC < 0, na.rm = TRUE)
  )
})
knitr::kable(phos_corr_summary,
             caption = "Phospho-corrected (phospho - protein) DE summary")

if (isTRUE(params$publication_mode)) {
  masked <- c("nlgf_in_maptki", "tau_alone", "interaction")
  for (cn in masked) de_phospho_corr$fit$top[[cn]]$logFC <- NA_real_
}
```

```{r phospho-corr-scatter, fig.width = 9, fig.height = 4}
# Compare raw phospho log2FC vs protein-corrected log2FC on the two NLGF contrasts.
compare_corr <- function(cn, title) {
  raw <- de_phospho$fit$top[[cn]] |>
    dplyr::transmute(feature, raw = logFC, raw_p = adj.P.Val)
  cor_tbl <- de_phospho_corr$fit$top[[cn]] |>
    dplyr::transmute(feature, corr = logFC, corr_p = adj.P.Val)
  cmp <- dplyr::inner_join(raw, cor_tbl, by = "feature") |>
    dplyr::filter(is.finite(raw), is.finite(corr))
  rho <- suppressWarnings(cor(cmp$raw, cmp$corr, method = "spearman"))
  ggplot(cmp, aes(raw, corr)) +
    geom_hline(yintercept = 0, colour = "grey70") +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_abline(slope = 1, intercept = 0, colour = "#b40426",
                linetype = "dashed") +
    geom_point(alpha = 0.25, size = 0.6) +
    labs(x = "raw phospho log2FC", y = "corrected (phospho - protein) log2FC",
         title = title,
         subtitle = sprintf("Spearman rho = %.3f, n = %d", rho, nrow(cmp))) +
    theme_bw()
}
compare_corr("nlgf_in_maptki", "NLGF in MAPTKI") |
compare_corr("nlgf_in_p301s",  "NLGF in P301S")
compare_corr("interaction",    "interaction (tau-dependent NLGF)") |
compare_corr("tau_in_nlgf",    "tau in NLGF")
```
