# Cross-modality integration

## Gene-level log2FC table

Per contrast, we collect the log2FC for every dataset, harmonised on gene
symbol. The phosphoproteomics input is aggregated to gene level by taking
the maximum absolute log2FC across sites per gene per contrast.

```{r integration-table}
integ_path <- file.path(params$cache_dir, "integration_table.rds")
# The snRNAseq column sources from the single-cell NEBULA fit; the
# pseudobulk logFC is kept alongside as `snrnaseq_pb` for cross-checking.
integration_tbl <- cache_or_run(integ_path, {
  contrasts <- names(contrast_definitions)
  contrasts <- c(contrasts, "interaction")
  build_one <- function(cn) {
    safe_min <- function(x) {
      if (all(is.na(x))) NA_real_ else suppressWarnings(min(x, na.rm = TRUE))
    }
    sn_pb <- de_snrnaseq$fit$top[[cn]] |>
      dplyr::transmute(symbol, snrnaseq_pb = logFC, snrnaseq_pb_p = adj.P.Val) |>
      dplyr::filter(!is.na(symbol), symbol != "") |>
      dplyr::group_by(symbol) |>
      dplyr::summarise(snrnaseq_pb   = mean(snrnaseq_pb, na.rm = TRUE),
                       snrnaseq_pb_p = safe_min(snrnaseq_pb_p),
                       .groups = "drop")
    sn <- de_snrnaseq_nebula$top[[cn]] |>
      dplyr::transmute(symbol, snrnaseq = logFC, snrnaseq_p = adj.P.Val) |>
      dplyr::filter(!is.na(symbol), symbol != "") |>
      dplyr::group_by(symbol) |>
      dplyr::summarise(snrnaseq   = mean(snrnaseq, na.rm = TRUE),
                       snrnaseq_p = safe_min(snrnaseq_p),
                       .groups = "drop") |>
      dplyr::full_join(sn_pb, by = "symbol")
    gx <- de_geomx$fit$top[[cn]] |>
      dplyr::transmute(symbol, geomx = logFC, geomx_p = adj.P.Val)
    pr <- de_proteomics$fit$top[[cn]] |>
      dplyr::transmute(symbol, proteomics = logFC, proteomics_p = adj.P.Val) |>
      dplyr::filter(!is.na(symbol)) |>
      dplyr::group_by(symbol) |>
      dplyr::summarise(proteomics   = mean(proteomics, na.rm = TRUE),
                       proteomics_p = safe_min(proteomics_p),
                       .groups = "drop")
    aggregate_site_level <- function(df, value_col, p_col) {
      df |>
        dplyr::filter(!is.na(symbol)) |>
        dplyr::group_by(symbol) |>
        dplyr::summarise(
          "{value_col}" := {
            v  <- .data[[value_col]]
            ok <- !is.na(v)
            if (any(ok)) v[ok][which.max(abs(v[ok]))] else NA_real_
          },
          "{p_col}" := suppressWarnings(min(.data[[p_col]], na.rm = TRUE)),
          .groups = "drop"
        ) |>
        dplyr::mutate(
          "{p_col}" := ifelse(is.infinite(.data[[p_col]]), NA_real_, .data[[p_col]])
        )
    }
    ph <- de_phospho$fit$top[[cn]] |>
      dplyr::transmute(symbol, phospho = logFC, phospho_p = adj.P.Val) |>
      aggregate_site_level("phospho", "phospho_p")
    pc <- de_phospho_corr$fit$top[[cn]] |>
      dplyr::transmute(symbol, phospho_corr = logFC,
                       phospho_corr_p = adj.P.Val) |>
      aggregate_site_level("phospho_corr", "phospho_corr_p")
    out <- sn |>
      dplyr::full_join(gx, by = "symbol") |>
      dplyr::full_join(pr, by = "symbol") |>
      dplyr::full_join(ph, by = "symbol") |>
      dplyr::full_join(pc, by = "symbol") |>
      dplyr::mutate(contrast = cn)
    out
  }
  do.call(rbind, lapply(contrasts, build_one))
})

cat(sprintf("Integration table: %d gene-contrast rows.\n", nrow(integration_tbl)))
head(integration_tbl, 5)
write_tsv_safe(integration_tbl,
               file.path(params$results_dir, "integration_logfc_per_contrast.tsv"))
```

## Per-contrast modality concordance

```{r concordance-plots, fig.width = 10, fig.height = 8}
plot_for <- function(cn, title_suffix) {
  sub <- integration_tbl |> dplyr::filter(contrast == cn)
  p1 <- concordance_plot(sub, "snrnaseq", "geomx",
                         label_col = "symbol",
                         x_lab = "snRNAseq log2FC", y_lab = "GeoMx log2FC",
                         title = paste0("snRNAseq vs GeoMx | ", title_suffix))
  p2 <- concordance_plot(sub, "snrnaseq", "proteomics",
                         label_col = "symbol",
                         x_lab = "snRNAseq log2FC", y_lab = "Proteomics log2FC",
                         title = paste0("snRNAseq vs Proteomics | ", title_suffix))
  p3 <- concordance_plot(sub, "geomx", "proteomics",
                         label_col = "symbol",
                         x_lab = "GeoMx log2FC", y_lab = "Proteomics log2FC",
                         title = paste0("GeoMx vs Proteomics | ", title_suffix))
  p4 <- concordance_plot(sub, "proteomics", "phospho",
                         label_col = "symbol",
                         x_lab = "Proteomics log2FC", y_lab = "Phospho log2FC",
                         title = paste0("Proteomics vs Phospho | ", title_suffix))
  (p1 | p2) / (p3 | p4)
}
plot_for("nlgf_in_maptki", "NLGF effect in MAPTKI")
plot_for("nlgf_in_p301s",  "NLGF effect in P301S")
plot_for("interaction",    "interaction (tau-dependent NLGF response)")
```

## Concordance summary

```{r concordance-summary}
modalities <- c("snrnaseq", "geomx", "proteomics", "phospho", "phospho_corr")
pairs <- t(combn(modalities, 2))
concord_tbl <- purrr::map_dfr(unique(integration_tbl$contrast), function(cn) {
  sub <- integration_tbl |> dplyr::filter(contrast == cn)
  apply(pairs, 1, function(p) {
    x <- sub[[p[1]]]; y <- sub[[p[2]]]
    ok <- is.finite(x) & is.finite(y)
    if (sum(ok) < 20) return(c(p[1], p[2], cn, NA, NA, sum(ok)))
    c(p[1], p[2], cn,
      round(cor(x[ok], y[ok], method = "spearman"), 3),
      round(cor(x[ok], y[ok], method = "pearson"), 3),
      sum(ok))
  }) |>
    t() |>
    as.data.frame() |>
    setNames(c("x", "y", "contrast", "spearman", "pearson", "n"))
})
concord_tbl$spearman <- as.numeric(concord_tbl$spearman)
concord_tbl$pearson  <- as.numeric(concord_tbl$pearson)
concord_tbl$n        <- as.integer(concord_tbl$n)
knitr::kable(concord_tbl,
             caption = "Spearman / Pearson concordance of log2FC between modalities")
write_tsv_safe(concord_tbl,
               file.path(params$results_dir, "modality_concordance.tsv"))
```

# Joint pathway analysis (fgsea)

We compute fgsea on the moderated `t`-statistic for the contrasts of
interest, separately per dataset, against MSigDB C5 GO:BP (mouse), then
look for pathways that are concordantly enriched across modalities. GO:BP
is a fine-grained ontology-based collection (~7k pathways), so the
displayed output is restricted to the top hits per contrast per modality;
the full table is persisted to
`storage/results/fgsea_gobp_per_contrast.tsv` for downstream use.

(Additional collections: GO MF, GO CC, and custom curated sets, are
introduced in section 12, the agnostic pathway survey. The GO MF and GO
CC fgsea result caches are built ahead of the knit by
`scripts/build_fgsea_extra_collections.R`, which writes
`storage/cache/fgsea_gomf_results.rds`,
`storage/cache/fgsea_gocc_results.rds`, and the corresponding per-contrast
TSVs to `storage/results/`; that script is idempotent and skips
already-built caches unless `--overwrite` is passed.)

```{r fgsea-setup}
prep_t <- function(top_tbl, gene_col = "symbol") {
  tbl <- top_tbl |> dplyr::filter(!is.na(.data[[gene_col]]),
                                  .data[[gene_col]] != "")
  tbl <- tbl[!duplicated(tbl[[gene_col]]), ]
  tbl
}

run_dataset <- function(de_obj, pathways, gene_col = "symbol",
                        min_size = 10, max_size = 500) {
  contrasts <- c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
                 "tau_alone", "tau_in_nlgf")
  out <- lapply(contrasts, function(cn) {
    tbl <- prep_t(de_obj$fit$top[[cn]], gene_col)
    if (!"t" %in% names(tbl)) tbl$t <- tbl$logFC
    run_fgsea_for_contrast(tbl, pathways, stat_col = "t",
                           gene_col = gene_col,
                           min_size = min_size, max_size = max_size)
  })
  names(out) <- contrasts
  out
}

# The snRNAseq fgsea ranking uses NEBULA's z-statistic (under-powered
# pseudobulk would dampen the interaction signal). NEBULA's output is
# wrapped in a fit-shaped list so `run_dataset` works unchanged.
snrnaseq_nebula_shim <- list(fit = list(top = de_snrnaseq_nebula$top))
```

```{r fgsea-gobp}
gobp <- get_gobp()

gobp_path <- file.path(params$cache_dir, "fgsea_gobp_results.rds")
gsea_gobp <- cache_or_run(gobp_path, {
  list(
    snrnaseq     = run_dataset(snrnaseq_nebula_shim, gobp,
                               min_size = 15, max_size = 500),
    geomx        = run_dataset(de_geomx,             gobp,
                               min_size = 15, max_size = 500),
    proteomics   = run_dataset(de_proteomics,        gobp,
                               min_size = 15, max_size = 500),
    phospho      = run_dataset(de_phospho,           gobp,
                               min_size = 15, max_size = 500),
    phospho_corr = run_dataset(de_phospho_corr,      gobp,
                               min_size = 15, max_size = 500)
  )
})

joint_gobp <- purrr::map_dfr(
  c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
    "tau_alone", "tau_in_nlgf"),
  function(cn) {
    extract <- function(name) {
      tbl <- gsea_gobp[[name]][[cn]] |> as.data.frame()
      tbl[, c("pathway", "NES", "padj")] |>
        setNames(c("pathway", paste0("nes_", name), paste0("padj_", name)))
    }
    Reduce(function(a, b) dplyr::full_join(a, b, by = "pathway"),
           lapply(names(gsea_gobp), extract)) |>
      dplyr::mutate(contrast = cn)
  }
)
write_tsv_safe(joint_gobp,
               file.path(params$results_dir, "fgsea_gobp_per_contrast.tsv"))
cat(sprintf("GO BP joint table: %d pathway-contrast rows.\n", nrow(joint_gobp)))
```

```{r fgsea-gobp-top}
# Rank GO BP terms per contrast by geometric mean of -log10(padj) across
# modalities (require at least 2 finite padj), then keep the top 8 per
# contrast for display.
nes_cols  <- grep("^nes_",  names(joint_gobp), value = TRUE)
padj_cols <- grep("^padj_", names(joint_gobp), value = TRUE)
gobp_ranked <- joint_gobp
gobp_ranked$score <- apply(as.matrix(gobp_ranked[, padj_cols]), 1, function(r) {
  r <- r[is.finite(r) & r > 0]
  if (length(r) < 2) return(NA_real_)
  exp(mean(log(-log10(r))))
})
gobp_ranked$consistent <- apply(as.matrix(gobp_ranked[, nes_cols]), 1, function(r) {
  r <- r[is.finite(r)]
  if (length(r) < 2) return(NA)
  all(sign(r) == sign(r[1]))
})
gobp_top <- gobp_ranked |>
  dplyr::filter(!is.na(score),
                contrast %in% c("nlgf_in_maptki", "nlgf_in_p301s",
                                "interaction",    "tau_in_nlgf")) |>
  dplyr::group_by(contrast) |>
  dplyr::arrange(dplyr::desc(score), .by_group = TRUE) |>
  dplyr::slice_head(n = 8) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    pathway = gsub("_", " ", tolower(sub("^GOBP_", "", pathway)))
  )
knitr::kable(
  gobp_top[, c("contrast", "pathway", nes_cols, "score", "consistent")],
  digits   = 3,
  caption  = "Top GO BP terms per contrast (geomean -log10(padj) across modalities)"
)
```
