# Cell-cell communication: CellChat + MultiNicheNet

So far the divergence story has been told inside the microglia compartment:
how does an amyloid pathology (NLGF) reshape microglia *transcription* and
*pathway activity* differently when introduced into a P301S tau context vs
a MAPTKI tau-null context? The complement to that intracellular view is
**intercellular**: how does the *signalling landscape* between microglia
and the five other cell-type compartments (Astrocyte, Neuronal, OPC,
Oligodendrocyte, Vascular) shift across the four genotypes, and in
particular does the tau-context-dependent NLGF response have a
ligand-receptor (LR) interaction signature?

Two complementary tools are run on the same nine-cell-type partitioning
(`Astrocyte / Microglia_{DAM,IFN,homeostatic,proliferative} / Neuronal /
OPC / Oligodendrocyte / Vascular`):

- **CellChat v2.2.0.9001** ([Jin *et al.* 2024](https://doi.org/10.1038/s41596-024-01045-4)):
  per-condition population-level inference of LR communication
  probability via a curated CellChatDB.mouse (3,379 LR interactions,
  293 pathways). Strengths: population-level pathway aggregation
  (`computeCommunProbPathway`), centrality metrics, comparative
  visualisation across the 4 genotypes via `mergeCellChat`.
- **MultiNicheNet 2.1.0** ([Browaeys *et al.* 2023](https://doi.org/10.1101/2023.06.13.544751)):
  multi-sample, multi-condition LR prioritisation under the same
  *factorial 2×2 + batch* design used everywhere else in this project,
  with the same five contrasts (`tau_alone`, `nlgf_in_maptki`,
  `nlgf_in_p301s`, `tau_in_nlgf`, `interaction`). Strengths: replicates
  the project-wide statistical framework at LR level, integrates with
  NicheNet ligand-activity priors, ranks LR pairs by a unified
  prioritisation score that combines sender DE / receiver DE / fraction
  expressing.

All compute is pre-baked into three caches (`cellchat_per_condition.rds`
~1 GB, `cellchat_merged.rds` ~67 MB, `multinichenet_output.rds` ~277 MB);
this section loads them and produces plots + tables only. See
`scripts/build_cellchat.R` and `scripts/build_multinichenet.R` for the
build recipe; both ran on the symbol-keyed `seurat_full_processed.rds`
to keep gene IDs aligned with CellChatDB and the NicheNet priors.

```{r ccc-load}
suppressPackageStartupMessages({
  library(CellChat)
  library(multinichenetr)
  library(NMF)
})

cellchat_per_condition <- readRDS(
  file.path(params$cache_dir, "cellchat_per_condition.rds"))
cellchat_merged        <- readRDS(
  file.path(params$cache_dir, "cellchat_merged.rds"))
multinichenet_output   <- readRDS(
  file.path(params$cache_dir, "multinichenet_output.rds"))

# Microglia cell-type label set (used for receiver filtering throughout)
microglia_cell_types <- c("Microglia_DAM", "Microglia_IFN",
                          "Microglia_homeostatic", "Microglia_proliferative")

# Project-wide colour palette extended to the 9 cell types. Microglia get
# warm reds (DAM = strongest), other lineages get distinct hues.
cell_type_colours <- c(
  Astrocyte               = "#7FC97F",
  Microglia_homeostatic   = "#FDC086",
  Microglia_DAM           = "#B0344D",
  Microglia_IFN           = "#E08754",
  Microglia_proliferative = "#FFD92F",
  Neuronal                = "#386CB0",
  OPC                     = "#BEAED4",
  Oligodendrocyte         = "#666666",
  Vascular                = "#A6CEE3"
)

# Semantic contrast names (matching the rest of the project) <-> raw
# MultiNicheNet contrast strings (matching scripts/build_multinichenet.R).
mnn_contrast_map <- c(
  tau_alone       = "P301S-MAPTKI",
  nlgf_in_maptki  = "NLGF_MAPTKI-MAPTKI",
  nlgf_in_p301s   = "NLGF_P301S-P301S",
  tau_in_nlgf     = "NLGF_P301S-NLGF_MAPTKI",
  interaction     = "(NLGF_P301S-P301S)-(NLGF_MAPTKI-MAPTKI)"
)
mnn_contrast_levels <- names(mnn_contrast_map)

cat(sprintf("CellChat per-condition list (%d): %s\n",
            length(cellchat_per_condition),
            paste(names(cellchat_per_condition), collapse = ", ")))
cat(sprintf("MultiNicheNet prioritization table: %d rows, %d cols\n",
            nrow(multinichenet_output$prioritization_tables$group_prioritization_tbl),
            ncol(multinichenet_output$prioritization_tables$group_prioritization_tbl)))
```

## (a) Per-condition signalling architecture (CellChat)

The most compact CellChat readout is the **cell-type interaction count
matrix**, for each genotype, how many significant LR pairs link each
ordered (sender, receiver) cell-type pair. Differences in the absolute
volume of communication between conditions are the first-order signal;
shifts in the cell-type wiring pattern at fixed volume are the second-
order signal that the comparative analysis below dissects.

```{r ccc-cellchat-circle, fig.width = 13, fig.height = 13}
# 2x2 grid of weighted aggregated cell-type interaction circles, one per
# genotype. Edge weight = sum of LR-pair communication probabilities for
# every (sender, receiver) cell-type pair. Edge colour = sender; node size =
# cell-type abundance in that condition.
op <- par(mfrow = c(2L, 2L), mar = c(2, 2, 4, 2))
on.exit(par(op), add = TRUE)
for (cond in genotype_levels) {
  cc <- cellchat_per_condition[[cond]]
  groupSize <- as.numeric(table(cc@idents))
  netVisual_circle(
    cc@net$weight,
    vertex.weight = groupSize,
    weight.scale  = TRUE,
    label.edge    = FALSE,
    edge.weight.max = max(sapply(cellchat_per_condition,
                                 function(x) max(x@net$weight))),
    color.use     = cell_type_colours[levels(cc@idents)],
    title.name    = sprintf("%s (n_LR_sig = %d)",
                            cond, nrow(subsetCommunication(cc)))
  )
}
```

```{r ccc-cellchat-count-heatmap, fig.width = 14, fig.height = 7}
# Per-condition cell-type interaction count heatmap, side-by-side. Reads
# more cleanly than the circles for counting-level comparisons. ComplexHeatmap
# is used (rather than CellChat::netVisual_heatmap) to align the colour
# scale across all four conditions.
count_mats <- lapply(cellchat_per_condition, function(cc) cc@net$count)
max_count  <- max(sapply(count_mats, max))
col_fun_ct <- circlize::colorRamp2(c(0, max_count / 2, max_count),
                                   c("#f7fbff", "#6baed6", "#08306b"))

ht_list <- NULL
for (cond in genotype_levels) {
  m <- count_mats[[cond]]
  ht <- ComplexHeatmap::Heatmap(
    m,
    name             = "n_LR",
    col              = col_fun_ct,
    cluster_rows     = FALSE,
    cluster_columns  = FALSE,
    show_heatmap_legend = (cond == "MAPTKI"),
    column_title     = cond,
    row_title        = if (cond == "MAPTKI") "Sender" else NULL,
    row_names_gp     = grid::gpar(fontsize = 7),
    column_names_gp  = grid::gpar(fontsize = 7),
    column_names_rot = 45,
    border           = TRUE,
    cell_fun = function(j, i, x, y, w, h, fill) {
      v <- m[i, j]
      if (v > 0) {
        grid::grid.text(v, x, y, gp = grid::gpar(fontsize = 7))
      }
    }
  )
  ht_list <- if (is.null(ht_list)) ht else (ht_list + ht)
}
ComplexHeatmap::draw(
  ht_list,
  column_title = paste(
    "CellChat per-condition cell-type interaction counts.",
    "Rows = sender, cols = receiver. Numbers = significant LR pairs."
  ),
  column_title_side = "bottom"
)
```

```{r ccc-cellchat-condition-summary}
cellchat_cond_summary <- purrr::imap_dfr(cellchat_per_condition,
  function(cc, cond) {
    tibble::tibble(
      condition       = cond,
      n_cells         = ncol(cc@data.signaling),
      n_pathways      = length(cc@netP$pathways),
      n_LR_sig        = nrow(subsetCommunication(cc)),
      total_strength  = sum(cc@net$weight),
      microglia_send  = sum(cc@net$weight[microglia_cell_types, ]),
      microglia_recv  = sum(cc@net$weight[, microglia_cell_types]),
      microglia_intra = sum(cc@net$weight[microglia_cell_types,
                                          microglia_cell_types])
    )
  }) |>
  dplyr::mutate(condition = factor(condition, levels = genotype_levels)) |>
  dplyr::arrange(condition)

knitr::kable(
  cellchat_cond_summary,
  digits  = 3,
  caption = paste(
    "Per-condition CellChat summary: n significant LR pairs, total",
    "communication weight, and microglia-as-sender / -receiver /",
    "intra-microglia weight subtotals. NLGF arms have ~2x more",
    "significant LR pairs and ~1.5x more microglia-receiver weight",
    "than non-NLGF arms."
  )
)
write_tsv_safe(cellchat_cond_summary,
               file.path(params$results_dir,
                         "ccc_cellchat_condition_summary.tsv"))
```

## (b) Differential pathway ranking across conditions

`CellChat::rankNet` compares the **information-flow** of every signalling
pathway across the merged conditions. We use it twice: once for the two
pairwise NLGF contrasts (`nlgf_in_maptki`, `nlgf_in_p301s`) and once
overall to surface pathways that change at all across the 4 genotypes.

```{r ccc-cellchat-ranknet-nlgf, fig.width = 11, fig.height = 9}
# Two pairwise rankNet panels: one per NLGF contrast. CellChat::rankNet
# returns a ggplot when `do.stat = TRUE`; we extract `gg.obj` and lay them
# out side-by-side with patchwork. Pathways are coloured by which arm
# carries higher info-flow (Mann-Whitney-U test on probability mass).
g_nlgf_maptki <- CellChat::rankNet(
  cellchat_merged,
  mode          = "comparison",
  comparison    = which(names(cellchat_per_condition) %in%
                        c("MAPTKI", "NLGF_MAPTKI")),
  stacked       = FALSE,
  do.stat       = TRUE,
  return.data   = FALSE
)
g_nlgf_p301s <- CellChat::rankNet(
  cellchat_merged,
  mode          = "comparison",
  comparison    = which(names(cellchat_per_condition) %in%
                        c("P301S", "NLGF_P301S")),
  stacked       = FALSE,
  do.stat       = TRUE,
  return.data   = FALSE
)
# Patchwork-combine the underlying ggplots
(g_nlgf_maptki + ggplot2::ggtitle("nlgf_in_maptki")) +
  (g_nlgf_p301s + ggplot2::ggtitle("nlgf_in_p301s"))
```

```{r ccc-cellchat-ranknet-full, fig.width = 8, fig.height = 12}
# Overall rankNet across all 4 conditions: pathways stacked-bar style with
# information-flow contribution per condition. Sorted by total info-flow.
g_all <- CellChat::rankNet(
  cellchat_merged,
  mode        = "comparison",
  comparison  = seq_along(cellchat_per_condition),
  stacked     = TRUE,
  do.stat     = FALSE,
  return.data = FALSE
)
g_all + ggplot2::ggtitle("Information flow per pathway (all 4 conditions)")
```

```{r ccc-cellchat-pathway-table}
# Long-form pathway info-flow table for downstream filtering / quoting.
# Re-uses rankNet's internal computation by calling with return.data = TRUE.
pathway_flow <- CellChat::rankNet(
  cellchat_merged,
  mode        = "comparison",
  comparison  = seq_along(cellchat_per_condition),
  stacked     = FALSE,
  do.stat     = FALSE,
  return.data = TRUE
)$signaling.contribution
pathway_flow$condition <- factor(pathway_flow$group, levels = genotype_levels)

# Wide form: rows = pathway, cols = condition (information flow), plus
# the NLGF-vs-baseline deltas and the divergence column.
pathway_flow_wide <- pathway_flow |>
  dplyr::select(pathway = name, condition, contribution) |>
  tidyr::pivot_wider(names_from = condition, values_from = contribution,
                     values_fill = 0) |>
  dplyr::mutate(
    nlgf_in_maptki = NLGF_MAPTKI - MAPTKI,
    nlgf_in_p301s  = NLGF_P301S  - P301S,
    interaction    = (NLGF_P301S - P301S) - (NLGF_MAPTKI - MAPTKI),
    abs_inter      = abs(interaction)
  ) |>
  dplyr::arrange(dplyr::desc(abs_inter))

knitr::kable(
  head(pathway_flow_wide, 20L) |> dplyr::select(-abs_inter),
  digits  = 3,
  caption = paste(
    "CellChat per-pathway information flow across the 4 genotypes,",
    "with the project's pairwise NLGF contrasts and the divergence",
    "`interaction` column. Sorted by |interaction|. Positive",
    "interaction = NLGF amplifies the pathway more strongly in P301S",
    "than in MAPTKI; negative = the opposite."
  )
)
write_tsv_safe(pathway_flow_wide,
               file.path(params$results_dir,
                         "ccc_cellchat_pathway_information_flow.tsv"))
```

## (c) Pairwise differential interactions (cell-type level)

`netVisual_diffInteraction` on the merged object visualises the **delta**
between two conditions' cell-type interaction count matrices as a single
circle plot (red = up in condition B, blue = up in condition A). We
produce one such circle per project pairwise contrast, then a manually-
computed *interaction* delta heatmap that has no direct CellChat
equivalent (`netVisual_diffInteraction` only takes pairwise comparisons).

```{r ccc-cellchat-diff-interaction, fig.width = 13, fig.height = 13}
# 2x2 grid of pairwise differential-interaction circles.
diff_pairs <- list(
  nlgf_in_maptki = c("MAPTKI",      "NLGF_MAPTKI"),
  nlgf_in_p301s  = c("P301S",       "NLGF_P301S"),
  tau_alone      = c("MAPTKI",      "P301S"),
  tau_in_nlgf    = c("NLGF_MAPTKI", "NLGF_P301S")
)
op <- par(mfrow = c(2L, 2L), mar = c(2, 2, 4, 2))
on.exit(par(op), add = TRUE)
for (cn in names(diff_pairs)) {
  pair <- diff_pairs[[cn]]
  idx  <- match(pair, names(cellchat_per_condition))
  netVisual_diffInteraction(
    cellchat_merged,
    weight.scale = TRUE,
    comparison   = idx,
    measure      = "weight",
    color.use    = cell_type_colours[levels(cellchat_merged@idents$joint)],
    title.name   = sprintf("%s   (%s -> %s)", cn, pair[1L], pair[2L])
  )
}
```

```{r ccc-cellchat-interaction-delta-heatmap, fig.width = 9, fig.height = 7}
# Factorial interaction at the cell-type interaction-weight level:
#   delta_M = (NLGF_P301S - P301S) - (NLGF_MAPTKI - MAPTKI)
# Positive cells = sender-receiver pair carries MORE NLGF-driven communication
# in P301S background than in MAPTKI background (tau-context-amplified
# signalling).
weight_mats <- lapply(cellchat_per_condition, function(cc) cc@net$weight)
delta_int <- (weight_mats$NLGF_P301S - weight_mats$P301S) -
             (weight_mats$NLGF_MAPTKI - weight_mats$MAPTKI)
lim <- max(abs(delta_int))
col_fun_di <- circlize::colorRamp2(c(-lim, 0, lim),
                                   c("#3a4cc0", "white", "#b40426"))
ComplexHeatmap::Heatmap(
  delta_int,
  name             = "Delta\nweight",
  col              = col_fun_di,
  cluster_rows     = FALSE,
  cluster_columns  = FALSE,
  row_title        = "Sender",
  column_title     = paste(
    "Interaction = (NLGF_P301S - P301S) - (NLGF_MAPTKI - MAPTKI).",
    "Positive = tau-context-amplified signalling."
  ),
  row_names_gp     = grid::gpar(fontsize = 8),
  column_names_gp  = grid::gpar(fontsize = 8),
  column_names_rot = 45,
  border           = TRUE,
  cell_fun = function(j, i, x, y, w, h, fill) {
    v <- delta_int[i, j]
    if (abs(v) > lim * 0.25) {
      grid::grid.text(sprintf("%.2f", v), x, y,
                      gp = grid::gpar(fontsize = 7))
    }
  }
)
```

```{r ccc-cellchat-interaction-delta-table}
delta_long <- as.data.frame.table(delta_int, responseName = "delta") |>
  dplyr::rename(sender = Var1, receiver = Var2) |>
  dplyr::arrange(dplyr::desc(abs(delta)))
knitr::kable(
  head(delta_long, 15L),
  digits  = 3,
  caption = paste(
    "Top 15 sender-receiver cell-type pairs by |interaction delta| in",
    "CellChat communication weight. Positive = NLGF amplifies the pair",
    "more strongly in P301S; negative = the opposite."
  )
)
write_tsv_safe(delta_long,
               file.path(params$results_dir,
                         "ccc_cellchat_interaction_delta_weights.tsv"))
```

## (d) Top differential ligand-receptor pairs per contrast (MultiNicheNet)

MultiNicheNet's `prioritization_tables$group_prioritization_tbl` ranks
every LR pair per contrast by a unified `prioritization_score` that
combines sender/receiver DE, ligand activity, and fraction expressing.
We surface the top hits per contrast, first unfiltered, then restricted
to **microglia receivers** (the divergence-narrative receivers).

```{r ccc-multinichenet-top-lr-all}
g_tbl <- multinichenet_output$prioritization_tables$group_prioritization_tbl |>
  dplyr::mutate(contrast_label = names(mnn_contrast_map)[match(contrast, mnn_contrast_map)])

# Top-10 LR pairs per project contrast (any cell type).
top_lr_all <- g_tbl |>
  dplyr::filter(!is.na(contrast_label)) |>
  dplyr::group_by(contrast_label) |>
  dplyr::arrange(dplyr::desc(prioritization_score), .by_group = TRUE) |>
  dplyr::slice_head(n = 10L) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    contrast_label = factor(contrast_label, levels = mnn_contrast_levels)
  ) |>
  dplyr::arrange(contrast_label, dplyr::desc(prioritization_score))

knitr::kable(
  top_lr_all |>
    dplyr::select(contrast_label, sender, receiver, ligand, receptor,
                  lr_interaction, prioritization_score,
                  scaled_lfc_ligand, scaled_lfc_receptor,
                  fraction_expressing_ligand_receptor),
  digits  = 3,
  caption = paste(
    "Top-10 MultiNicheNet ligand-receptor pairs per project contrast",
    "(any sender-receiver pair). `prioritization_score` is a unified",
    "rank combining sender / receiver DE + ligand activity + fraction",
    "expressing. Sorted by descending score within contrast."
  )
)
write_tsv_safe(top_lr_all,
               file.path(params$results_dir,
                         "ccc_multinichenet_top_lr_all.tsv"))
```

```{r ccc-multinichenet-top-lr-microglia}
# Microglia-receiver subset: restricts to the four microglia substates as
# receivers, since the project's divergence narrative is microglia-centred.
top_lr_mg <- g_tbl |>
  dplyr::filter(!is.na(contrast_label),
                receiver %in% microglia_cell_types) |>
  dplyr::group_by(contrast_label) |>
  dplyr::arrange(dplyr::desc(prioritization_score), .by_group = TRUE) |>
  dplyr::slice_head(n = 10L) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    contrast_label = factor(contrast_label, levels = mnn_contrast_levels)
  ) |>
  dplyr::arrange(contrast_label, dplyr::desc(prioritization_score))

knitr::kable(
  top_lr_mg |>
    dplyr::select(contrast_label, sender, receiver, ligand, receptor,
                  lr_interaction, prioritization_score,
                  scaled_lfc_ligand, scaled_lfc_receptor,
                  fraction_expressing_ligand_receptor),
  digits  = 3,
  caption = paste(
    "Top-10 MultiNicheNet ligand-receptor pairs per project contrast,",
    "restricted to microglia substates as receivers."
  )
)
write_tsv_safe(top_lr_mg,
               file.path(params$results_dir,
                         "ccc_multinichenet_top_lr_microglia.tsv"))
```

```{r ccc-multinichenet-interaction-spotlight}
# Focus on the divergence interaction contrast: top-25 microglia-receiver
# LR pairs, with a per-row note about whether the ligand or receptor is
# also a hdWGCNA hub (when those globals are present).
inter_label <- "interaction"
top_int_mg <- g_tbl |>
  dplyr::filter(contrast_label == inter_label,
                receiver %in% microglia_cell_types) |>
  dplyr::arrange(dplyr::desc(prioritization_score)) |>
  dplyr::slice_head(n = 25L)

# Optional cross-flags (graceful if the upstream globals are missing).
hub_syms <- if (exists("hub_tbl") && nrow(hub_tbl) > 0L) {
  unique(hub_tbl$symbol)
} else character(0)

top_int_mg <- top_int_mg |>
  dplyr::mutate(
    ligand_is_hdwgcna_hub   = ligand   %in% hub_syms,
    receptor_is_hdwgcna_hub = receptor %in% hub_syms
  )

knitr::kable(
  top_int_mg |>
    dplyr::select(sender, receiver, ligand, receptor, lr_interaction,
                  prioritization_score,
                  scaled_lfc_ligand, scaled_lfc_receptor,
                  fraction_expressing_ligand_receptor,
                  ligand_is_hdwgcna_hub, receptor_is_hdwgcna_hub),
  digits  = 3,
  caption = paste(
    "Top-25 microglia-receiver LR pairs for the divergence interaction",
    "contrast. `ligand_is_hdwgcna_hub` / `receptor_is_hdwgcna_hub` flag",
    "overlap with the four hdWGCNA module hub sets."
  )
)
write_tsv_safe(top_int_mg,
               file.path(params$results_dir,
                         "ccc_multinichenet_interaction_top25_microglia.tsv"))
cat(sprintf("Hub overlap among top-25 interaction LR pairs: %d ligands, %d receptors\n",
            sum(top_int_mg$ligand_is_hdwgcna_hub),
            sum(top_int_mg$receptor_is_hdwgcna_hub)))
```

## (e) Circos diagrams per contrast (MultiNicheNet)

`make_circos_group_comparison` draws a colour-banded chord diagram of the
top prioritised LR pairs per group, useful as a one-figure-per-paper view
of the LR landscape. We render one circos per project contrast using its
top-20 LR pairs (microglia-receiver filtered).

```{r ccc-multinichenet-circos, fig.width = 10, fig.height = 10, results = "hide"}
# `get_top_n_lr_pairs(rank_per_group = TRUE)` extracts the top-N LR pairs
# per group from the prioritization table. `make_circos_group_comparison`
# then draws a chord diagram coloured by sender (inner) and receiver
# (outer). One panel per contrast; we loop and print each separately so
# Rmd captures them in the document flow.
n_per_contrast <- 20L

# Per-cell-type sender / receiver colour vectors (multinichenet wants
# named vectors keyed by cell-type name).
sender_cols   <- cell_type_colours
receiver_cols <- cell_type_colours

# Build a fresh subset table per contrast (microglia receiver, top-N) and
# pass to make_circos_group_comparison. Because the function uses base
# graphics + circlize, we print captions as cat() calls outside the chunk
# (handled via for-loop printing) and rely on the Rmd's PNG device.
for (lab in mnn_contrast_levels) {
  raw_contrast <- mnn_contrast_map[[lab]]
  sub <- g_tbl |>
    dplyr::filter(contrast == raw_contrast,
                  receiver %in% microglia_cell_types) |>
    dplyr::arrange(dplyr::desc(prioritization_score)) |>
    dplyr::slice_head(n = n_per_contrast)
  if (nrow(sub) == 0L) next
  # `make_circos_group_comparison` requires the standard `group` column;
  # for a single-group subset it still works (renders a one-band chord).
  tryCatch(
    {
      make_circos_group_comparison(
        prioritized_tbl_oi = sub,
        colors_sender      = sender_cols,
        colors_receiver    = receiver_cols
      )
      cat(sprintf("\n\n*Figure: top-%d microglia-receiver LR pairs for contrast '%s'.*\n\n",
                  n_per_contrast, lab))
    },
    error = function(e) {
      message(sprintf("circos for %s failed: %s", lab, conditionMessage(e)))
    }
  )
}
```

## (f) Sanity table: condition-, pathway- and contrast-level summaries

```{r ccc-sanity-cellchat-pathway-condition}
# Per-(condition, pathway) significant LR count using subsetCommunication.
# Restricted to the union of the top-15 pathways by total weight across
# conditions to keep the table compact.
pathway_long <- purrr::imap_dfr(cellchat_per_condition, function(cc, cond) {
  com <- subsetCommunication(cc)  # default returns net data frame
  if (nrow(com) == 0L) return(NULL)
  com$condition <- cond
  com[, c("condition", "pathway_name", "source", "target", "ligand", "receptor")]
})

pathway_n_lr <- pathway_long |>
  dplyr::count(condition, pathway_name, name = "n_LR") |>
  tidyr::pivot_wider(names_from = condition, values_from = n_LR,
                     values_fill = 0L)
pathway_n_lr <- pathway_n_lr[, c("pathway_name",
                                 intersect(genotype_levels,
                                           names(pathway_n_lr)))]
pathway_n_lr$total <- rowSums(pathway_n_lr[, -1L, drop = FALSE])
pathway_n_lr$interaction <- with(pathway_n_lr,
  (NLGF_P301S - P301S) - (NLGF_MAPTKI - MAPTKI))
pathway_n_lr <- pathway_n_lr[order(-abs(pathway_n_lr$interaction)), ]

knitr::kable(
  head(pathway_n_lr, 20L),
  caption = paste(
    "Per-pathway significant LR pair counts across the 4 genotypes,",
    "with the `interaction` column = the factorial divergence in LR",
    "counts (NLGF_P301S - P301S) - (NLGF_MAPTKI - MAPTKI). Sorted by",
    "|interaction|."
  )
)
write_tsv_safe(pathway_n_lr,
               file.path(params$results_dir,
                         "ccc_cellchat_pathway_lr_counts.tsv"))
```

```{r ccc-sanity-multinichenet-by-celltype}
# Per-contrast sender / receiver representation in MultiNicheNet's top-100
# prioritised LR pairs (any group). Surfaces which cell types dominate
# each contrast and where microglia substates sit in the senders/receivers.
top100_per_contrast <- g_tbl |>
  dplyr::filter(!is.na(contrast_label)) |>
  dplyr::group_by(contrast_label) |>
  dplyr::arrange(dplyr::desc(prioritization_score), .by_group = TRUE) |>
  dplyr::slice_head(n = 100L) |>
  dplyr::ungroup()

sender_summary <- top100_per_contrast |>
  dplyr::count(contrast_label, sender, name = "n_top100_as_sender") |>
  tidyr::pivot_wider(id_cols    = sender,
                     names_from = contrast_label,
                     values_from = n_top100_as_sender,
                     values_fill = 0L)
sender_summary <- sender_summary[, c("sender",
                                     intersect(mnn_contrast_levels,
                                               names(sender_summary)))]
sender_summary$total <- rowSums(sender_summary[, -1L, drop = FALSE])
sender_summary <- sender_summary[order(-sender_summary$total), ]

receiver_summary <- top100_per_contrast |>
  dplyr::count(contrast_label, receiver, name = "n_top100_as_receiver") |>
  tidyr::pivot_wider(id_cols    = receiver,
                     names_from = contrast_label,
                     values_from = n_top100_as_receiver,
                     values_fill = 0L)
receiver_summary <- receiver_summary[, c("receiver",
                                         intersect(mnn_contrast_levels,
                                                   names(receiver_summary)))]
receiver_summary$total <- rowSums(receiver_summary[, -1L, drop = FALSE])
receiver_summary <- receiver_summary[order(-receiver_summary$total), ]

knitr::kable(
  sender_summary,
  caption = paste(
    "MultiNicheNet top-100 LR pairs per contrast, counted by SENDER",
    "cell type. Surfaces which cell types push the most prioritised",
    "signalling each contrast."
  )
)
knitr::kable(
  receiver_summary,
  caption = paste(
    "Same top-100 LR pairs, counted by RECEIVER cell type. Surfaces",
    "which cell types receive the most prioritised signalling each",
    "contrast."
  )
)
write_tsv_safe(sender_summary,
               file.path(params$results_dir,
                         "ccc_multinichenet_top100_sender_per_contrast.tsv"))
write_tsv_safe(receiver_summary,
               file.path(params$results_dir,
                         "ccc_multinichenet_top100_receiver_per_contrast.tsv"))
```

```{r ccc-sanity-overlap-cellchat-multinichenet}
# Cross-tool sanity: how many of MultiNicheNet's top-50 microglia-receiver
# LR pairs per contrast also appear in the CellChat significant-LR set
# (across any of the 4 conditions)? Establishes whether the two tools
# converge on the same LR space at the headline level.
#
# Two complementary metrics are reported because the underlying databases
# differ in size: CellChatDB.mouse has ~42 distinct ligands appearing in
# significant LR rows per condition vs NicheNet's ~490 ligands in the
# MNN prioritisation table — so the LR-pair overlap is a hard upper
# bound on what's even representable in CellChat's universe.
#   1. **lr_pair overlap**: exact ligand-receptor pair match. CellChat
#      encodes heteromeric receptor complexes with an underscore
#      (e.g. `Nrp1_Plxna2`) which would never match MNN's single-symbol
#      receptor format; we expand CellChat complex receptors into
#      separate one-ligand-one-receptor rows before pairing.
#   2. **ligand-only overlap**: ligand symbol match (receptor identity
#      ignored). Less strict, more robust to database differences in
#      receptor complex encoding.
mnn_top50_mg <- g_tbl |>
  dplyr::filter(!is.na(contrast_label),
                receiver %in% microglia_cell_types) |>
  dplyr::group_by(contrast_label) |>
  dplyr::arrange(dplyr::desc(prioritization_score), .by_group = TRUE) |>
  dplyr::slice_head(n = 50L) |>
  dplyr::ungroup()

# Build CellChat reference sets across all 4 conditions:
#   - ligand-only set:   union of `ligand` column values
#   - LR-pair set:       explode receptor complexes (split on "_") so
#                        a `Sema3c_NRP1_PLXNA2` row contributes both
#                        `Sema3c_NRP1` and `Sema3c_PLXNA2` to the set.
cc_ref <- purrr::imap_dfr(cellchat_per_condition, function(cc, cond) {
  com <- subsetCommunication(cc)
  if (nrow(com) == 0L) return(NULL)
  tibble::tibble(condition = cond,
                 ligand    = com$ligand,
                 receptor  = com$receptor)
})

cc_ligand_set <- unique(cc_ref$ligand)
cc_lr_set <- unique(unlist(lapply(seq_len(nrow(cc_ref)), function(i) {
  comps <- strsplit(cc_ref$receptor[i], "_", fixed = TRUE)[[1L]]
  paste(cc_ref$ligand[i], comps, sep = "_")
})))

mnn_top50_mg$in_cellchat_lr     <- mnn_top50_mg$lr_interaction %in% cc_lr_set
mnn_top50_mg$in_cellchat_ligand <- mnn_top50_mg$ligand         %in% cc_ligand_set

cross_overlap <- mnn_top50_mg |>
  dplyr::group_by(contrast_label) |>
  dplyr::summarise(
    n_total        = dplyr::n(),
    n_lr_overlap   = sum(in_cellchat_lr),
    n_lig_overlap  = sum(in_cellchat_ligand),
    pct_lr         = round(100 * n_lr_overlap  / n_total, 1),
    pct_ligand     = round(100 * n_lig_overlap / n_total, 1),
    .groups        = "drop"
  ) |>
  dplyr::mutate(contrast_label = factor(contrast_label,
                                        levels = mnn_contrast_levels)) |>
  dplyr::arrange(contrast_label)

knitr::kable(
  cross_overlap,
  caption = paste(
    "Cross-tool LR overlap. For each contrast: of MultiNicheNet's top-50",
    "microglia-receiver LR pairs, how many also appear in the union of",
    "CellChat significant LR pairs across the 4 conditions. `pct_lr`",
    "matches on ligand-receptor pair (CellChat heteromeric receptor",
    "complexes split into components before pairing); `pct_ligand`",
    "matches on ligand symbol only. The two tools' databases differ",
    "substantially in coverage of microglia / DAM signalling:",
    "CellChatDB.mouse does NOT contain Apoe, C1qa, C1qb, Cx3cl1 as",
    "ligands and does NOT contain Trem2 as a receptor, while these are",
    "all prominent in NicheNet and dominate MNN's microglia-receiver",
    "top hits, so the *LR-pair* convergence is necessarily low. The",
    "*ligand-level* convergence is much higher in NLGF arms (88% in",
    "nlgf_in_p301s) than in the baseline tau contrast (10% in",
    "tau_alone), indicating that NLGF activates a signalling programme",
    "where the two databases agree on the ligand vocabulary (e.g.,",
    "Csf1, Tgfb1, Spp1, Tnf families) even when they disagree on the",
    "exact LR pair."
  )
)
write_tsv_safe(cross_overlap,
               file.path(params$results_dir,
                         "ccc_overlap_multinichenet_cellchat.tsv"))
cat(sprintf(
  "CellChat reference universe: %d unique ligands, %d unique LR pairs (after exploding receptor complexes)\n",
  length(cc_ligand_set), length(cc_lr_set)))
```

```{r ccc-sanity-dam-sampling-balance, fig.width = 12, fig.height = 5}
# DAM-sampling sanity check for the CCC interaction-contrast finding.
#
# The top-25 microglia-receiver LR pairs under the divergence interaction
# contrast are DAM-dominated (sender + receiver) and are driven by a
# small set of macrophage-trophic ligands (Csf1, Tgfb1, Spp1, ...). That
# could be a genuine *interaction* — DAM rewiring is non-additive across
# the 2x2 — or it could be confounded by DAM population *abundance*
# differing across genotypes. Prioritisation scores in MultiNicheNet
# weight each pair by `fraction_expressing_ligand_receptor`, which is
# computed *within each sender/receiver cell-type pool*, so a uniform
# DAM expansion shouldn't bias the per-pair score — but it can still
# enlarge the DAM pool's contribution to top-N rankings through reduced
# sampling noise. This chunk quantifies the asymmetry directly.
#
# Strategy:
#   (1) state x genotype contingency + per-genotype substate fractions
#   (2) per-substate Fisher tests for NLGF vs non-NLGF enrichment
#   (3) per-sample DAM fraction across 16 genotype_batch samples, with
#       a 2x2 ANOVA-style test for the non-additive (tau:nlgf) component
#   (4) stacked-bar composition + per-sample DAM-fraction box plot
# Substate labels come from `label_microglia_states()` (same routine the
# per-substate NEBULA cache uses). The chunk reuses `microglia_seurat`
# from the parent knit when it has the `state` column, and reloads +
# relabels from cache otherwise (smoke-test driver). Note that the
# cached `microglia_seurat_processed.rds` is the *pre-QC-pruning* object
# (26104 cells, before `rmd/02a` drops contamination outlier clusters);
# the full knit canonical path operates on the *post-pruning* in-memory
# Seurat (~23.5k cells). Both paths produce the same qualitative DAM
# expansion story; the canonical TSVs land via the full knit.
loaded_locally <- !(exists("microglia_seurat") &&
                    inherits(microglia_seurat, "Seurat") &&
                    "state" %in% names(microglia_seurat@meta.data))
mg_seu <- if (loaded_locally) {
  obj <- readRDS(file.path(params$cache_dir,
                           "microglia_seurat_processed.rds"))
  sm <- if (exists("symbol_map") && is.data.frame(symbol_map)) {
    symbol_map
  } else {
    readRDS(file.path(params$cache_dir, "snrnaseq_symbol_map.rds"))
  }
  label_microglia_states(obj, symbol_map = sm)
} else microglia_seurat

# (1) Population-level substate x genotype contingency
state_geno <- table(state    = mg_seu$state,
                    genotype = mg_seu$genotype)
state_geno_frac <- prop.table(state_geno, margin = 2L)

knitr::kable(
  state_geno,
  caption = paste(
    "Substate x genotype contingency (counts of microglia nuclei).",
    "Headline metric for the audit: the DAM row across the 4 genotype",
    "columns. A markedly larger DAM count in NLGF arms relative to",
    "MAPTKI/P301S is expected (DAM is the amyloid response); the audit",
    "tests whether that expansion is *non-additive* across tau context."
  )
)
knitr::kable(
  round(100 * state_geno_frac, 1),
  digits  = 1,
  caption = paste("Substate composition per genotype (% of microglia).",
                  "Rows sum to 100 within each column.")
)

# (2) Per-substate Fisher tests: NLGF vs non-NLGF enrichment
nlgf_label <- ifelse(grepl("^NLGF_", as.character(mg_seu$genotype)),
                     "NLGF", "non_NLGF")
per_state_fisher <- purrr::map_dfr(levels(mg_seu$state), function(s) {
  in_state <- mg_seu$state == s
  tab <- table(in_state = in_state, arm = nlgf_label)
  ft <- fisher.test(tab)
  tibble::tibble(
    state         = s,
    n_NLGF        = sum(in_state & nlgf_label == "NLGF"),
    n_other       = sum(in_state & nlgf_label == "non_NLGF"),
    frac_NLGF     = sum(in_state & nlgf_label == "NLGF") /
                    sum(nlgf_label == "NLGF"),
    frac_other    = sum(in_state & nlgf_label == "non_NLGF") /
                    sum(nlgf_label == "non_NLGF"),
    odds_NLGF     = unname(ft$estimate),
    ci_lo         = unname(ft$conf.int[1L]),
    ci_hi         = unname(ft$conf.int[2L]),
    pvalue        = ft$p.value
  )
})
per_state_fisher$padj <- p.adjust(per_state_fisher$pvalue, method = "BH")

knitr::kable(
  per_state_fisher,
  digits  = 3,
  caption = paste(
    "Per-substate Fisher tests for NLGF vs non-NLGF enrichment.",
    "odds_NLGF > 1 means the substate is over-represented in the two",
    "NLGF arms relative to MAPTKI/P301S. The DAM row quantifies the",
    "amyloid-driven expansion that the CCC sanity check is interested",
    "in; the homeostatic row should show the mirror-image depletion."
  )
)

# (3) Per-sample DAM fraction across 16 genotype_batch samples,
#     followed by a 2x2 factorial ANOVA on the logit-DAM fraction so the
#     interaction component is testable on the same parameterisation as
#     the rest of the project (tau + nlgf + tau:nlgf).
sample_state <- as.data.frame(
  table(genotype_batch = mg_seu$genotype_batch,
        state          = mg_seu$state),
  responseName = "n"
)
sample_totals <- sample_state |>
  dplyr::group_by(genotype_batch) |>
  dplyr::summarise(total = sum(n), .groups = "drop")
sample_dam <- sample_state |>
  dplyr::filter(state == "DAM") |>
  dplyr::left_join(sample_totals, by = "genotype_batch") |>
  dplyr::mutate(
    frac_DAM = n / total,
    genotype = sub("_batch[0-9]+$", "", as.character(genotype_batch)),
    tau      = factor(ifelse(grepl("P301S", genotype), "P301S", "MAPTKI"),
                      levels = c("MAPTKI", "P301S")),
    nlgf     = factor(ifelse(grepl("^NLGF_", genotype), "NLGF", "noNLGF"),
                      levels = c("noNLGF", "NLGF"))
  )
sample_dam$genotype <- factor(sample_dam$genotype,
                              levels = levels(mg_seu$genotype))

# logit transformation gives an unbounded response amenable to lm.
# Empirical proportions stay away from 0/1 (min DAM frac ~0.05 expected),
# so a simple logit is fine without adjusting for boundaries.
sample_dam$logit_DAM <- qlogis(pmin(pmax(sample_dam$frac_DAM, 1e-3),
                                    1 - 1e-3))
fit_anova <- lm(logit_DAM ~ tau * nlgf, data = sample_dam)
anova_tbl <- as.data.frame(anova(fit_anova))
anova_tbl$term <- rownames(anova_tbl)
anova_tbl <- anova_tbl[, c("term", setdiff(names(anova_tbl), "term"))]
rownames(anova_tbl) <- NULL

knitr::kable(
  anova_tbl,
  digits  = 3,
  caption = paste(
    "Two-way ANOVA on the logit per-sample DAM fraction with the",
    "project's factorial parameterisation (tau, nlgf, tau:nlgf).",
    "A significant `tau:nlgf` row means the amyloid-driven DAM",
    "expansion differs between the two tau backgrounds, i.e. the",
    "*sampling* level already carries an interaction signal, which",
    "would attribute the CCC interaction-contrast DAM dominance to a",
    "non-additive population shift rather than per-cell rewiring. A",
    "non-significant `tau:nlgf` row implies the DAM expansion is",
    "additive at the population level, so the CCC DAM dominance is",
    "more likely to reflect genuine LR rewiring."
  )
)

# (4) Visuals: stacked-bar composition + DAM-fraction box per genotype
substate_palette <- c(
  homeostatic   = "#FDC086",
  DAM           = "#B0344D",
  IFN           = "#E08754",
  proliferative = "#FFD92F"
)
bar_long <- as.data.frame(state_geno_frac, responseName = "frac")
names(bar_long)[1:2] <- c("state", "genotype")

p_bar <- ggplot2::ggplot(bar_long,
                         ggplot2::aes(x = genotype, y = frac, fill = state)) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::scale_fill_manual(values = substate_palette,
                             name   = "Substate") +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::labs(x = NULL, y = "Fraction of microglia",
                title = "Substate composition per genotype") +
  ggplot2::theme_classic(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 25, hjust = 1))

p_box <- ggplot2::ggplot(sample_dam,
                         ggplot2::aes(x = genotype, y = frac_DAM,
                                      colour = genotype)) +
  ggplot2::geom_boxplot(outlier.shape = NA, width = 0.6) +
  ggplot2::geom_jitter(width = 0.12, size = 1.8, alpha = 0.85) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::guides(colour = "none") +
  ggplot2::labs(x = NULL, y = "DAM fraction per sample",
                title = "Per-sample DAM fraction (4 batches/genotype)") +
  ggplot2::theme_classic(base_size = 11) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 25, hjust = 1))

print(patchwork::wrap_plots(p_bar, p_box, nrow = 1L,
                            widths = c(1.0, 1.05)))

# (5) Exports
state_geno_df <- as.data.frame(state_geno, responseName = "n")
names(state_geno_df)[1:2] <- c("state", "genotype")
state_geno_df <- state_geno_df |>
  dplyr::left_join(
    as.data.frame(state_geno_frac, responseName = "frac"),
    by = c("state", "genotype")
  )
write_tsv_safe(state_geno_df,
               file.path(params$results_dir,
                         "ccc_dam_sampling_balance_state_geno.tsv"))
write_tsv_safe(per_state_fisher,
               file.path(params$results_dir,
                         "ccc_dam_sampling_balance_fisher.tsv"))
write_tsv_safe(sample_dam,
               file.path(params$results_dir,
                         "ccc_dam_sampling_balance_sample_dam.tsv"))
write_tsv_safe(anova_tbl,
               file.path(params$results_dir,
                         "ccc_dam_sampling_balance_anova.tsv"))

# (6) Free heavy object iff we loaded it (full knit keeps the global)
if (loaded_locally) {
  rm(mg_seu); invisible(gc(verbose = FALSE))
}
```

```{r ccc-sanity-hub-hypergeometric}
# hdWGCNA hub hypergeometric test against the MNN top-25 microglia-
# receiver LR pairs (interaction contrast).
#
# The per-row hub flags on the spotlight table (`ligand_is_hdwgcna_hub`,
# `receptor_is_hdwgcna_hub`) report raw overlap counts. This chunk asks
# the formal question: is that overlap enriched relative to chance,
# given the operational universe of all ligand/receptor symbols
# MultiNicheNet considered for *any* contrast?
#
# Operational universe:
#   `mnn_universe <- unique(c(g_tbl$ligand, g_tbl$receptor))` over the
#   full prioritisation table -> ~829 unique gene symbols. This is the
#   correct null because MultiNicheNet only ranks genes inside its LR
#   database that were also expressed; hub genes that aren't in this
#   universe could never have appeared in the top-25 regardless of
#   their hdWGCNA status.
#
# Module sets tested (4 per panel):
#   set_kind = "hubs"          -> top-25 kME hub genes per module
#                                 (RBC-contaminant rows filtered).
#   set_kind = "module_genes"  -> the full module assignment table
#                                 from hdwgcna_modules.tsv (570-850
#                                 genes per non-grey module).
# Sample sides tested (3 per panel):
#   sample_kind = "ligand_receptor" -> union of top-25 ligand+receptor
#                                      symbols (after de-duplication).
#   sample_kind = "ligand"           -> top-25 ligand symbols only.
#   sample_kind = "receptor"         -> top-25 receptor symbols only.
# All restricted to the MNN universe before hypergeometric.
#
# Hypergeometric (phyper, upper tail):
#   q = overlap - 1, m = |module_set ∩ universe|,
#   n = |universe| - m, k = |sample ∩ universe|.
# BH-adjusted across the 4 modules x 2 set_kinds x 3 sample_kinds = 24
# tests in a single panel.
mnn_universe <- unique(c(g_tbl$ligand, g_tbl$receptor))

# Reuse hub_sym from the hdWGCNA child if it's already in memory;
# otherwise reload + rebuild with the same RBC contaminant filter.
hub_local <- if (exists("hub_sym") && is.data.frame(hub_sym) &&
                 all(c("symbol", "module", "is_likely_rbc") %in%
                     names(hub_sym))) {
  hub_sym
} else {
  hdw_local <- readRDS(file.path(params$cache_dir,
                                 "hdwgcna_microglia.rds"))
  sm <- if (exists("symbol_map") && is.data.frame(symbol_map)) {
    symbol_map
  } else {
    readRDS(file.path(params$cache_dir, "snrnaseq_symbol_map.rds"))
  }
  hs <- hdw_local$hub_genes
  hs$symbol <- sm$symbol[match(hs$gene_name, sm$ensembl)]
  hs$symbol <- ifelse(is.na(hs$symbol) | !nzchar(hs$symbol),
                      hs$gene_name, hs$symbol)
  flag_contaminant_hubs(hs)
}

# Full module-genes table (symbols) — needed for the "module_genes"
# set_kind. Rebuild from the hdWGCNA cache regardless of in-memory
# availability so the symbol column is guaranteed populated.
hdw_for_mods <- if (exists("hdw") && is.list(hdw) &&
                    !is.null(hdw$modules)) hdw else
  readRDS(file.path(params$cache_dir, "hdwgcna_microglia.rds"))
sm_for_mods <- if (exists("symbol_map") && is.data.frame(symbol_map)) {
  symbol_map
} else {
  readRDS(file.path(params$cache_dir, "snrnaseq_symbol_map.rds"))
}
mod_full <- hdw_for_mods$modules
mod_full$symbol <- sm_for_mods$symbol[match(mod_full$gene_name,
                                            sm_for_mods$ensembl)]
mod_full$symbol <- ifelse(is.na(mod_full$symbol) | !nzchar(mod_full$symbol),
                          mod_full$gene_name, mod_full$symbol)

# Sample-side gene lists (top_int_mg is built earlier in section (d))
sample_lists <- list(
  ligand_receptor = unique(c(top_int_mg$ligand, top_int_mg$receptor)),
  ligand          = unique(top_int_mg$ligand),
  receptor        = unique(top_int_mg$receptor)
)
modules_to_test <- setdiff(unique(hub_local$module), "grey")

hypergeo_one <- function(module, set_kind, sample_kind) {
  module_set <- if (set_kind == "hubs") {
    unique(hub_local$symbol[hub_local$module == module &
                            !hub_local$is_likely_rbc])
  } else {
    unique(mod_full$symbol[mod_full$module == module])
  }
  module_in_uni <- intersect(module_set, mnn_universe)
  sample_in_uni <- intersect(sample_lists[[sample_kind]], mnn_universe)
  overlap_syms  <- intersect(module_in_uni, sample_in_uni)
  N <- length(mnn_universe)
  m <- length(module_in_uni)
  k <- length(sample_in_uni)
  q <- length(overlap_syms)
  pval <- if (m == 0L || k == 0L) NA_real_ else
    phyper(q - 1L, m, N - m, k, lower.tail = FALSE)
  expected <- if (N == 0L) NA_real_ else (m * k) / N
  list(
    module        = module,
    set_kind      = set_kind,
    sample_kind   = sample_kind,
    module_size   = length(module_set),
    module_in_uni = m,
    sample_in_uni = k,
    overlap       = q,
    expected      = expected,
    enrichment    = if (is.na(expected) || expected == 0)
                      NA_real_ else q / expected,
    pvalue        = pval,
    overlap_syms  = paste(sort(overlap_syms), collapse = ",")
  )
}

hypergeo_grid <- expand.grid(
  module      = modules_to_test,
  set_kind    = c("hubs", "module_genes"),
  sample_kind = names(sample_lists),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
)
hub_hyper <- purrr::pmap_dfr(hypergeo_grid, hypergeo_one)
hub_hyper$padj <- p.adjust(hub_hyper$pvalue, method = "BH")
hub_hyper <- hub_hyper[order(hub_hyper$set_kind,
                             hub_hyper$sample_kind,
                             hub_hyper$module), ]

knitr::kable(
  hub_hyper[, c("module", "set_kind", "sample_kind",
                "module_size", "module_in_uni", "sample_in_uni",
                "overlap", "expected", "enrichment",
                "pvalue", "padj")],
  digits  = 3,
  caption = paste(
    "Hypergeometric over-representation of the top-25 microglia-receiver",
    "LR pairs (interaction contrast) in each hdWGCNA module set, against",
    sprintf("the operational universe of %d unique MultiNicheNet",
            length(mnn_universe)),
    "ligand/receptor symbols. `set_kind = hubs` uses the top-25 kME hub",
    "genes per module (RBC contaminants filtered out); `module_genes`",
    "uses the full module assignment. `sample_kind` partitions the test",
    "set into ligand-only, receptor-only, and ligand+receptor union.",
    "`enrichment` = overlap / expected; `padj` is BH across the 24",
    "tests displayed."
  )
)
write_tsv_safe(hub_hyper,
               file.path(params$results_dir,
                         "ccc_hdwgcna_hub_hypergeometric.tsv"))

# Plain-English headline: which (module, set_kind) combinations are
# significantly enriched in the ligand_receptor union after FDR.
sig_lr <- hub_hyper |>
  dplyr::filter(sample_kind == "ligand_receptor", !is.na(padj)) |>
  dplyr::arrange(padj)
if (any(sig_lr$padj < 0.05, na.rm = TRUE)) {
  hits <- sig_lr[sig_lr$padj < 0.05 & !is.na(sig_lr$padj), ]
  cat(sprintf("Significantly enriched (padj < 0.05) in the ligand+receptor union: %d / %d.\n",
              nrow(hits), nrow(sig_lr)))
  for (i in seq_len(nrow(hits))) {
    cat(sprintf("  %s (%s): overlap=%d, expected=%.2f, enrich=%.2fx, padj=%.3g\n",
                hits$module[i], hits$set_kind[i], hits$overlap[i],
                hits$expected[i], hits$enrichment[i], hits$padj[i]))
  }
} else {
  cat("No (module, set_kind) combination reaches padj < 0.05 in the ligand+receptor union.\n")
}

# Free the helper copy if we loaded it ourselves (full knit keeps `hdw`
# and `hub_sym` globals untouched).
rm(mod_full, hdw_for_mods, sm_for_mods)
if (!exists("hub_sym")) rm(hub_local)
invisible(gc(verbose = FALSE))
```

```{r ccc-cleanup}
# Free the heavy CCC caches so they don't sit in memory during the
# session-info / manifest chunk.
rm(cellchat_per_condition, cellchat_merged, multinichenet_output,
   g_tbl, top100_per_contrast, pathway_long, pathway_n_lr)
invisible(gc(verbose = FALSE))
```
