# Agnostic cross-modality pathway and module survey

This section asks a deliberately agnostic question across every gene-set
collection currently cached for the project: which pathways, if any,
have the strongest cross-modality support for the five canonical
contrasts (`nlgf_in_maptki`, `nlgf_in_p301s`, `interaction`,
`tau_alone`, `tau_in_nlgf`)?

The previous iteration of this report anchored on a single Hallmark
gene set (`HALLMARK_OXIDATIVE_PHOSPHORYLATION`) as the dominant
cross-modality signal. The Hallmark collection has since been removed
from the project, the OXPHOS confirmatory arc (former sections 8 and
9a-9d) deleted, and the MG-M3 module rewritten to drop its "OXPHOS
module" framing. This survey replaces those anchored analyses with a
metric-driven ranking that gives no pathway family privileged framing.
Sections are ordered by collection (GO BP, GO MF, GO CC, three custom
curated families, hdWGCNA modules); each subsection sorts by the same
agnostic composite. A final subsection unifies all six collection
rankings into a single leader board and states an unbiased multi-axis
verdict.

The ranking helper is `rank_pathways_cross_modality()` in
`R/pathway_survey.R`. Composite rank sorts by
(a) the count of modalities at FDR<0.05 with consistent direction of
effect, then (b) the mean |NES| across modalities with a finite NES.
Tied counts fall through to the |NES| tie-breaker so the table reads
top-to-bottom by both criteria. The full ranking (every pathway in the
collection × every contrast) is exported to a per-collection TSV in
`storage/results/`; the inline tables show the top 20 per contrast.

## GO BP pathway × modality ranking

GO BP is the largest collection in the project (~7,000 mouse pathways
in MSigDB C5; ~4,300 pass the fgsea min_size=15 cutoff in at least one
modality). The whole-microglia fgsea cache used here is
`storage/cache/fgsea_gobp_results.rds`, built in
`rmd/07_integration.Rmd` and available in the parent knit session as
`gsea_gobp`.

```{r pathway-survey-gobp-rank}
gobp_ranking_full <- purrr::map_dfr(
  c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
    "tau_alone", "tau_in_nlgf"),
  function(cn) {
    rank_pathways_cross_modality(gsea_gobp, contrast = cn) |>
      dplyr::mutate(contrast = cn)
  }
)
write_tsv_safe(
  gobp_ranking_full,
  file.path(params$results_dir, "pathway_survey_gobp_ranking.tsv")
)
cat(sprintf("GO BP cross-modality ranking: %d pathway-contrast rows across %d unique pathways.\n",
            nrow(gobp_ranking_full),
            length(unique(gobp_ranking_full$pathway))))
```

```{r pathway-survey-gobp-tables, results = 'asis'}
for (cn in c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
             "tau_alone", "tau_in_nlgf")) {
  cat(sprintf("\n\n#### Contrast: `%s`\n\n", cn))
  ranked <- gobp_ranking_full |>
    dplyr::filter(contrast == cn) |>
    dplyr::select(-contrast)
  print(format_ranking_table(
    ranked, contrast = cn, n = 20, strip_prefix = "^GOBP_"
  ))
}
```

## GO MF pathway × modality ranking

GO MF (Gene Ontology Molecular Function) is the smaller (~1,850 mouse
pathways) function-axis sibling of GO BP and tends to surface enzyme-
class and binding-activity terms (e.g. NADH dehydrogenase activity,
proton-transporting ATPase activity, GTPase regulator activity). The
fgsea cache `storage/cache/fgsea_gomf_results.rds` is built by
`scripts/build_fgsea_extra_collections.R` outside the knit pipeline,
so we read it directly here.

```{r pathway-survey-gomf-rank}
gsea_gomf <- readRDS(file.path(params$cache_dir, "fgsea_gomf_results.rds"))
gomf_ranking_full <- purrr::map_dfr(
  c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
    "tau_alone", "tau_in_nlgf"),
  function(cn) {
    rank_pathways_cross_modality(gsea_gomf, contrast = cn) |>
      dplyr::mutate(contrast = cn)
  }
)
write_tsv_safe(
  gomf_ranking_full,
  file.path(params$results_dir, "pathway_survey_gomf_ranking.tsv")
)
cat(sprintf("GO MF cross-modality ranking: %d pathway-contrast rows across %d unique pathways.\n",
            nrow(gomf_ranking_full),
            length(unique(gomf_ranking_full$pathway))))
```

```{r pathway-survey-gomf-tables, results = 'asis'}
for (cn in c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
             "tau_alone", "tau_in_nlgf")) {
  cat(sprintf("\n\n#### Contrast: `%s`\n\n", cn))
  ranked <- gomf_ranking_full |>
    dplyr::filter(contrast == cn) |>
    dplyr::select(-contrast)
  print(format_ranking_table(
    ranked, contrast = cn, n = 20, strip_prefix = "^GOMF_"
  ))
}
```

## GO CC pathway × modality ranking

GO CC (Gene Ontology Cellular Component) is the location-axis sibling
of GO BP / GO MF (~1,040 mouse pathways) and surfaces subcellular
localisations (e.g. mitochondrial inner membrane, respiratory chain
complex, postsynaptic density). The fgsea cache
`storage/cache/fgsea_gocc_results.rds` is built by
`scripts/build_fgsea_extra_collections.R` outside the knit pipeline,
so we read it directly here.

```{r pathway-survey-gocc-rank}
gsea_gocc <- readRDS(file.path(params$cache_dir, "fgsea_gocc_results.rds"))
gocc_ranking_full <- purrr::map_dfr(
  c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
    "tau_alone", "tau_in_nlgf"),
  function(cn) {
    rank_pathways_cross_modality(gsea_gocc, contrast = cn) |>
      dplyr::mutate(contrast = cn)
  }
)
write_tsv_safe(
  gocc_ranking_full,
  file.path(params$results_dir, "pathway_survey_gocc_ranking.tsv")
)
cat(sprintf("GO CC cross-modality ranking: %d pathway-contrast rows across %d unique pathways.\n",
            nrow(gocc_ranking_full),
            length(unique(gocc_ranking_full$pathway))))
```

```{r pathway-survey-gocc-tables, results = 'asis'}
for (cn in c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
             "tau_alone", "tau_in_nlgf")) {
  cat(sprintf("\n\n#### Contrast: `%s`\n\n", cn))
  ranked <- gocc_ranking_full |>
    dplyr::filter(contrast == cn) |>
    dplyr::select(-contrast)
  print(format_ranking_table(
    ranked, contrast = cn, n = 20, strip_prefix = "^GOCC_"
  ))
}
```

## Custom curated sets × modality ranking

The three project-specific curated collections built in Phase B
(`custom_microglia_states` = 7 sets, `custom_microglia_ad` = 4 sets,
`custom_module_sources` = 12 sets; 23 unique sets in total) are pooled
into a single ranking here. Set names are unique across the three
caches, so pooling does not introduce identifier collisions; the
`source_collection` column attributes each pathway to its origin.
Because the pooled collection is small relative to the GO families,
the inline tables show **every** set per contrast (n=23) rather than
the canonical top-20 used for the GO collections, top-20 would
silently omit three sets per contrast, which is poor UX when the total
is so small. The full ranking is still exported to TSV.

```{r pathway-survey-custom-rank}
custom_caches <- list(
  states  = readRDS(file.path(params$cache_dir, "fgsea_custom_states_results.rds")),
  ad      = readRDS(file.path(params$cache_dir, "fgsea_custom_ad_results.rds")),
  modules = readRDS(file.path(params$cache_dir, "fgsea_custom_modules_results.rds"))
)

# Pool the three per-collection caches into one fgsea-list of the same
# shape (modality -> contrast -> data.frame), by rbinding per
# (modality, contrast) the per-collection result tables. Set names are
# unique across collections so rbind is collision-free.
pool_custom_caches <- function(caches) {
  modalities <- names(caches[[1]])
  contrasts  <- names(caches[[1]][[1]])
  out <- list()
  for (mn in modalities) {
    out[[mn]] <- list()
    for (cn in contrasts) {
      out[[mn]][[cn]] <- do.call(rbind, lapply(caches, function(c) {
        as.data.frame(c[[mn]][[cn]])
      }))
    }
  }
  out
}
pooled_custom <- pool_custom_caches(custom_caches)

# Build a pathway -> source_collection lookup from the cache names
# themselves (no hard-coding of set lists, future-proof if a collection
# grows).
source_map <- unlist(lapply(names(custom_caches), function(collection_name) {
  paths <- unique(unlist(lapply(custom_caches[[collection_name]],
                                function(per_contrast) {
    unlist(lapply(per_contrast, function(t) as.data.frame(t)$pathway))
  })))
  setNames(rep(collection_name, length(paths)), paths)
}))

custom_ranking_full <- purrr::map_dfr(
  c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
    "tau_alone", "tau_in_nlgf"),
  function(cn) {
    r <- rank_pathways_cross_modality(pooled_custom, contrast = cn)
    r$source_collection <- source_map[r$pathway]
    r$contrast <- cn
    r
  }
)
write_tsv_safe(
  custom_ranking_full,
  file.path(params$results_dir, "pathway_survey_custom_ranking.tsv")
)
cat(sprintf("Custom curated cross-modality ranking: %d pathway-contrast rows across %d unique pathways (%s).\n",
            nrow(custom_ranking_full),
            length(unique(custom_ranking_full$pathway)),
            paste(sprintf("%s = %d", names(custom_caches),
                          vapply(custom_caches, function(c) {
                            length(unique(as.data.frame(c[[1]][[1]])$pathway))
                          }, integer(1))), collapse = "; ")))
```

```{r pathway-survey-custom-tables, results = 'asis'}
for (cn in c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
             "tau_alone", "tau_in_nlgf")) {
  cat(sprintf("\n\n#### Contrast: `%s`\n\n", cn))
  ranked <- custom_ranking_full |>
    dplyr::filter(contrast == cn) |>
    dplyr::select(-contrast)
  print(format_ranking_table(
    ranked, contrast = cn, n = nrow(ranked),
    extra_cols = "source_collection"
  ))
}
```

## Per-substate pathway ranking (interaction contrast)

The sections above pool every microglia nucleus into a single contrast
and ask which pathways move cross-modality. This subsection asks a
complementary question: WITHIN the snRNAseq modality, which pathways
move per microglia substate? The Phase B5 cache
`storage/cache/fgsea_per_state_results.rds` holds fgsea results for
every (collection, substate, contrast) cell, where the four substates
(`homeostatic`, `DAM`, `IFN`, `proliferative`) come from the substate
clustering in `rmd/02a_snrnaseq_microglia.Rmd` and the NEBULA per-state
DE tables in `rmd/02b_snrnaseq_de_nebula.Rmd`.

The inline tables below focus on the **interaction contrast**, which is
the cross-genotype interaction term, the contrast most directly
relevant to whether tau and amyloid pathology produce non-additive
microglia responses. The full per-substate ranking covering all five
contrasts is exported to TSV for downstream inspection.

For each pathway in each substate the helper records the per-substate
rank by |NES|, an `in_top_n` flag (top 10 within that substate), and a
`substate_breadth` integer counting how many of the four substates list
the same pathway in their top 10 at the same contrast. A pathway with
`substate_breadth >= 3` is broadly active across the microglia
compartment at that contrast; the inline tables flag those rows with a
`*` prefix. Pathways with `substate_breadth = 1` are substate-specific.

A caveat applies to small collections: when a collection has only N
sets and N <= the top-n cutoff, every set in that collection lands in
the per-substate top-n by construction, so `substate_breadth` saturates
trivially at 4. The three custom collections (`custom_microglia_states`
= 6-7 sets per substate, `custom_microglia_ad` = 3 sets, and
`custom_module_sources` = 11-12 sets) fall in this regime; their
breadth values report the rank ordering, not a stringent
cross-substate-consistency finding. The breadth metric is meaningful
for the GO collections (a few hundred to a few thousand pathways per
substate).

```{r pathway-survey-per-state-build}
fps <- readRDS(file.path(params$cache_dir,
                         "fgsea_per_state_results.rds"))
per_state_long <- rank_pathways_per_substate(fps, n_top = 10L)
write_tsv_safe(
  per_state_long,
  file.path(params$results_dir,
            "pathway_survey_per_state_ranking.tsv")
)
cat(sprintf(
  paste0("Per-substate ranking: %d rows across %d collections, %d ",
         "substates, %d contrasts.\n"),
  nrow(per_state_long),
  dplyr::n_distinct(per_state_long$collection),
  dplyr::n_distinct(per_state_long$substate),
  dplyr::n_distinct(per_state_long$contrast)))

# Cross-substate breadth summary at interaction contrast: how many
# pathways per collection reach breadth >= 3 (broadly active) or
# breadth == 4 (compartment-wide). For the GO collections this is the
# headline cross-substate finding; for the small custom collections it
# is dominated by the small-N artifact described above.
breadth_summary <- per_state_long |>
  dplyr::filter(.data$contrast == "interaction",
                .data$in_top_n) |>
  dplyr::distinct(.data$collection, .data$pathway, .data$substate_breadth) |>
  dplyr::group_by(.data$collection) |>
  dplyr::summarise(
    n_in_any_top10 = dplyr::n(),
    n_breadth_ge3  = sum(.data$substate_breadth >= 3L),
    n_breadth_eq4  = sum(.data$substate_breadth == 4L),
    .groups = "drop"
  )
```

```{r pathway-survey-per-state-breadth-summary, results = 'asis'}
print(knitr::kable(
  breadth_summary, digits = 0,
  caption = paste0("Interaction-contrast cross-substate breadth per ",
                   "collection: count of pathways landing in the ",
                   "top-10 of any substate, and the subset reaching ",
                   "breadth >= 3 or breadth == 4.")
))
```

```{r pathway-survey-per-state-tables, results = 'asis'}
collection_specs <- list(
  list(key = "gobp",                    label = "GO BP",
       strip = "^GOBP_"),
  list(key = "gomf",                    label = "GO MF",
       strip = "^GOMF_"),
  list(key = "gocc",                    label = "GO CC",
       strip = "^GOCC_"),
  list(key = "custom_microglia_states", label = "Custom: microglia states",
       strip = NULL),
  list(key = "custom_microglia_ad",     label = "Custom: microglia AD",
       strip = NULL),
  list(key = "custom_module_sources",
       label = "Custom: module sources (Sun/Victor 2023)",
       strip = NULL)
)
substates <- c("homeostatic", "DAM", "IFN", "proliferative")
contrast_inline <- "interaction"

for (spec in collection_specs) {
  cat(sprintf("\n\n### %s\n\n", spec$label))
  for (ss in substates) {
    cat(sprintf("\n\n#### Substate: `%s`\n\n", ss))
    print(format_per_substate_table(
      per_state_long,
      collection_key = spec$key,
      contrast_key   = contrast_inline,
      substate_key   = ss,
      n_top          = 10L,
      breadth_mark   = 3L,
      strip_prefix   = spec$strip
    ))
  }
}
```

## hdWGCNA module × modality cross-support

The hdWGCNA build in `rmd/03_hdwgcna.Rmd` assigns each measured
microglia gene to one of four data-driven co-expression modules
(`MG-M1`, `MG-M2`, `MG-M3`, `MG-M4`) or to the unassigned `grey`
compartment. This subsection treats each non-grey module as a custom
gene set and asks the same agnostic cross-modality question as the
collection sections above: at each contrast, which modules carry
consistent signal across the five DE modalities (snRNAseq NEBULA z,
GeoMx limma t, proteomics limma t, phospho limma t, phospho_corr
limma t)? Modules are presented in alphabetical order throughout
(rows of the heatmap, table rows after the composite tie-breaker)
so the presentation does not privilege any one module visually,
a deliberate anti-anchoring choice given the earlier MG-M3 "OXPHOS
module" framing that this overhaul retired.

The fgsea pathway list is built from `storage/cache/hdwgcna_microglia.rds`
(the canonical p=3 build) with Ensembl→symbol mapping via the
`symbol_map` from `rmd/01_data.Rmd`. The grey unassigned bucket is
dropped (it is not a co-expression module). Module gene-set sizes
after symbol mapping are MG-M1 = 570, MG-M2 = 850, MG-M3 = 780,
MG-M4 = 74; all four comfortably clear the fgsea `min_size = 5`
cutoff in every modality after intersection with the modality gene
universe. The per-modality fgsea results are cached at
`storage/cache/fgsea_hdwgcna_modules_results.rds` and the cross-
modality ranking tibble at
`storage/results/pathway_survey_hdwgcna_modules.tsv` (one row per
module × contrast with per-modality NES and padj columns).

```{r pathway-survey-hdwgcna-build}
hdw_for_pathways <- readRDS(file.path(params$cache_dir,
                                      "hdwgcna_microglia.rds"))
modules_sym <- hdw_for_pathways$modules
modules_sym$symbol <- symbol_map$symbol[
  match(modules_sym$gene_name, symbol_map$ensembl)
]
modules_sym <- modules_sym[
  !is.na(modules_sym$symbol) & nzchar(modules_sym$symbol) &
    modules_sym$module != "grey",
  c("module", "symbol"), drop = FALSE
]
modules_sym$module <- as.character(modules_sym$module)
module_pathways <- split(modules_sym$symbol, modules_sym$module)
module_pathways <- module_pathways[lengths(module_pathways) > 0]
module_pathways <- module_pathways[sort(names(module_pathways))]
cat(sprintf("hdWGCNA module pathway list: %d non-grey modules (%s).\n",
            length(module_pathways),
            paste(sprintf("%s = %d", names(module_pathways),
                          lengths(module_pathways)),
                  collapse = "; ")))

# Use the same `run_dataset` shim defined upstream in
# `rmd/07_integration.Rmd` so the snRNAseq fgsea uses the NEBULA
# z-statistic and the other four modalities use the limma t (matching
# every other fgsea cache in the project). `min_size = 5` is the same
# cutoff used by the three Phase B custom collections; modules are
# large enough that this cutoff is not actively binding here, but the
# value keeps the schema consistent.
hdwgcna_fgsea_path <- file.path(params$cache_dir,
                                "fgsea_hdwgcna_modules_results.rds")
gsea_hdwgcna_modules <- cache_or_run(hdwgcna_fgsea_path, {
  list(
    snrnaseq     = run_dataset(snrnaseq_nebula_shim, module_pathways,
                               min_size = 5, max_size = 1000),
    geomx        = run_dataset(de_geomx,             module_pathways,
                               min_size = 5, max_size = 1000),
    proteomics   = run_dataset(de_proteomics,        module_pathways,
                               min_size = 5, max_size = 1000),
    phospho      = run_dataset(de_phospho,           module_pathways,
                               min_size = 5, max_size = 1000),
    phospho_corr = run_dataset(de_phospho_corr,      module_pathways,
                               min_size = 5, max_size = 1000)
  )
})

hdwgcna_ranking_full <- purrr::map_dfr(
  c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
    "tau_alone", "tau_in_nlgf"),
  function(cn) {
    rank_pathways_cross_modality(gsea_hdwgcna_modules, contrast = cn) |>
      dplyr::mutate(contrast = cn)
  }
)
write_tsv_safe(
  hdwgcna_ranking_full,
  file.path(params$results_dir,
            "pathway_survey_hdwgcna_modules.tsv")
)
cat(sprintf(
  "hdWGCNA module cross-modality ranking: %d module-contrast rows across %d modules.\n",
  nrow(hdwgcna_ranking_full),
  length(unique(hdwgcna_ranking_full$pathway))
))
```

```{r pathway-survey-hdwgcna-tables, results = 'asis'}
for (cn in c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
             "tau_alone", "tau_in_nlgf")) {
  cat(sprintf("\n\n#### Contrast: `%s`\n\n", cn))
  ranked <- hdwgcna_ranking_full |>
    dplyr::filter(contrast == cn) |>
    dplyr::select(-contrast)
  # Re-sort alphabetically within each contrast to enforce the
  # anti-anchoring guardrail at display time. The composite rank
  # column still records the metric-driven ordering.
  ranked <- ranked[order(ranked$pathway), , drop = FALSE]
  print(format_ranking_table(
    ranked, contrast = cn, n = nrow(ranked),
    caption = sprintf(
      paste0("All hdWGCNA modules for contrast '%s' (rows alphabetical; ",
             "composite_rank reflects sign-consistent significant-modality ",
             "count, then mean |NES|; padj cutoff 0.05)."),
      cn)
  ))
}
```

```{r pathway-survey-hdwgcna-heatmap, fig.width = 8, fig.height = 3.5}
# Module x modality NES heatmap for the interaction contrast.
# Row order: alphabetical (MG-M1, MG-M2, MG-M3, MG-M4) so MG-M3 has
# no visual priority. Column order: the canonical project modality
# order (snRNAseq, GeoMx, proteomics, phospho, phospho_corr).
modalities       <- names(gsea_hdwgcna_modules)
contrast_focus   <- "interaction"
module_alpha     <- sort(names(module_pathways))

nes_mat  <- vapply(modalities, function(m) {
  tbl <- as.data.frame(gsea_hdwgcna_modules[[m]][[contrast_focus]])
  setNames(tbl$NES, tbl$pathway)[module_alpha]
}, numeric(length(module_alpha)))
padj_mat <- vapply(modalities, function(m) {
  tbl <- as.data.frame(gsea_hdwgcna_modules[[m]][[contrast_focus]])
  setNames(tbl$padj, tbl$pathway)[module_alpha]
}, numeric(length(module_alpha)))
rownames(nes_mat)  <- module_alpha
rownames(padj_mat) <- module_alpha

range_abs <- max(abs(nes_mat[is.finite(nes_mat)]), na.rm = TRUE)
col_fun   <- circlize::colorRamp2(c(-range_abs, 0, range_abs),
                                  c("#3a4cc0", "white", "#b40426"))
cell_fn   <- function(j, i, x, y, width, height, fill) {
  if (is.na(padj_mat[i, j])) return()
  if (padj_mat[i, j] < 0.05) {
    grid::grid.text("*", x, y,
                    gp = grid::gpar(fontsize = 11, col = "black"))
  } else if (padj_mat[i, j] < 0.1) {
    grid::grid.text(".", x, y,
                    gp = grid::gpar(fontsize = 11, col = "black"))
  }
}
ComplexHeatmap::Heatmap(
  nes_mat,
  name            = "NES",
  col             = col_fun,
  cluster_rows    = FALSE,
  cluster_columns = FALSE,
  row_order       = seq_len(nrow(nes_mat)),
  column_order    = seq_len(ncol(nes_mat)),
  cell_fun        = cell_fn,
  row_names_gp    = grid::gpar(fontsize = 10),
  column_names_gp = grid::gpar(fontsize = 10),
  column_title    = sprintf(
    paste0("hdWGCNA module x modality NES at '%s' contrast ",
           "('*' = FDR<0.05, '.' = FDR<0.1; rows alphabetical)"),
    contrast_focus),
  border          = TRUE
)
```

## Unified leader board (unbiased verdict)

The collection-by-collection rankings above already use a metric-driven
composite that ignores pathway family. This subsection pools every
collection into a single leader board, applies one fixed leader rule
across all (pathway, contrast) cells, and states an unbiased verdict
from whatever the rule surfaces. No pathway family is privileged in the
construction; the leader rule is fixed below before the data is
inspected, so the threshold cannot be retrofitted to favour any one
biology.

**Leader rule (applied verbatim to every (pathway, contrast) row):** a
row qualifies as a leader if
`n_modalities_sig_consistent_sign >= 2` OR `n_modalities_sig >= 3`. The
first clause requires at least two modalities at FDR<0.05 with the same
direction of effect; the second clause requires at least three
modalities at FDR<0.05 regardless of direction, capturing the
OXPHOS-style mixed-sign multi-modality enrichment pattern. The rule is
encoded in `R/pathway_survey.R::build_leader_board()` as a closure so
future plans can swap it without rewriting the helper, and the closure
is applied to the stacked long form of the six collection rankings
already exported in subsections 13.1-13.6.

**Composite leader score (primary sort, larger is better):**
`leader_score = 5 * n_contrasts_consistent_sign_ge2 + n_contrasts_sig_ge3 + max_consistent_sign / 5`.
The 5x weight on the cross-contrast consistent-sign count means a
pathway that hits the rule at two different contrasts with consistent
direction always outranks a pathway that hits it at one contrast only;
the `max_consistent_sign / 5` tie-breaker lifts cases where a single
contrast carries an exceptional cross-modality stack. The leader score
is computed once per (collection, pathway); the per-contrast
breakdown is preserved verbatim in the `contrasts_summary` column for
each row.

Three complementary views are presented below: (1) the top 20 by
composite leader score (cross-contrast + cross-modality), (2) the top
10 by maximum cross-substate breadth (compartment-wide signals), and
(3) the top 10 by maximum sign-consistent significant-modality count at
any single contrast (single-contrast cross-modality stars). The full
unified leader board (one row per leader pathway) is at
`storage/results/pathway_survey_unified_leaderboard.tsv`, built by the
D1 step of `storage/notes/pathway_overhaul_plan.md` and read here
without recomputation.

```{r pathway-survey-verdict-build}
verdict_lb <- readr::read_tsv(
  file.path(params$results_dir,
            "pathway_survey_unified_leaderboard.tsv"),
  show_col_types = FALSE
)

cat(sprintf(
  paste0("Unified leader board: %d leader pathway rows across ",
         "%d collections (%s).\n"),
  nrow(verdict_lb),
  dplyr::n_distinct(verdict_lb$collection),
  paste(sprintf("%s = %d",
                names(table(verdict_lb$collection)),
                as.integer(table(verdict_lb$collection))),
        collapse = "; ")
))

# Strip GO-collection prefixes and underscore-format GO pathway names
# for display; custom-collection names (MG-M2, DAM_up, WAM, ...) lack a
# recognised prefix and pass through unchanged.
pretty_pathway <- function(pathway) {
  out <- pathway
  is_go <- grepl("^(GOBP|GOMF|GOCC)_", out)
  out[is_go] <- sub("^(GOBP|GOMF|GOCC)_", "", out[is_go])
  out[is_go] <- gsub("_", " ", tolower(out[is_go]))
  out
}

# View 1: top 20 by composite leader score.
verdict_view_score <- verdict_lb |>
  dplyr::arrange(dplyr::desc(leader_score), collection, pathway) |>
  head(20) |>
  dplyr::transmute(
    collection,
    pathway              = pretty_pathway(pathway),
    leader_score,
    n_contrasts_cons_ge2 = n_contrasts_consistent_sign_ge2,
    n_contrasts_sig_ge3,
    max_consistent_sign,
    dominant_sign,
    max_substate_breadth,
    contrasts_summary
  )

# View 2: top 10 by max cross-substate breadth, then leader score.
verdict_view_breadth <- verdict_lb |>
  dplyr::arrange(dplyr::desc(max_substate_breadth),
                 dplyr::desc(leader_score),
                 collection, pathway) |>
  head(10) |>
  dplyr::transmute(
    collection,
    pathway              = pretty_pathway(pathway),
    max_substate_breadth,
    leader_score,
    dominant_sign,
    n_contrasts_cons_ge2 = n_contrasts_consistent_sign_ge2,
    contrasts_summary
  )

# View 3: top 10 by max sign-consistent significant-modality count at
# any single contrast, then leader score.
verdict_view_consistent <- verdict_lb |>
  dplyr::arrange(dplyr::desc(max_consistent_sign),
                 dplyr::desc(leader_score),
                 collection, pathway) |>
  head(10) |>
  dplyr::transmute(
    collection,
    pathway              = pretty_pathway(pathway),
    max_consistent_sign,
    leader_score,
    dominant_sign,
    n_contrasts_cons_ge2 = n_contrasts_consistent_sign_ge2,
    contrasts_summary
  )
```

```{r pathway-survey-verdict-tables, results = 'asis'}
print(knitr::kable(
  verdict_view_score, digits = 2,
  caption = paste0(
    "Top 20 by composite leader_score = ",
    "5 * n_contrasts_consistent_sign_ge2 + n_contrasts_sig_ge3 + ",
    "max_consistent_sign / 5. ",
    "Cross-contrast consistent-sign breadth weighted first."
  )
))

print(knitr::kable(
  verdict_view_breadth, digits = 2,
  caption = paste0(
    "Top 10 by max cross-substate breadth (max over contrasts of the ",
    "number of microglia substates -- out of 4 -- where the pathway ",
    "appears in the per-substate top-10). Surfaces compartment-wide ",
    "signals."
  )
))

print(knitr::kable(
  verdict_view_consistent, digits = 2,
  caption = paste0(
    "Top 10 by max sign-consistent significant-modality count at any ",
    "single contrast. Surfaces single-contrast cross-modality stars."
  )
))
```

### Verdict

The leader board surfaces **three distinct axes of biology**, none of
which is the single winner the original Hallmark-anchored framing
implied:

1. **Amyloid-driven activation (+ sign).** MG-M2 (hdWGCNA module) is
   the cleanest single-contrast cross-modality star anywhere in the
   project: four modalities significant at `nlgf_in_p301s` with
   consistent + direction (snRNAseq NES=+2.15 padj=1.6e-111, GeoMx
   +1.33 padj=4.9e-4, proteomics +1.51 padj=2.0e-4, phospho +1.50
   padj=3.8e-3), plus three modalities sig consistent + at
   `nlgf_in_maptki`. The MHC II antigen-presentation family
   (`GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_*` x5 + `GOMF_ANTIGEN_BINDING`
   + `GOMF_PEPTIDE_ANTIGEN_BINDING`) provides independent
   GO-collection corroboration with the same `nlgf_in_maptki:2/+ |
   nlgf_in_p301s:2/+` pattern at leader_score 10.4, and
   `GOBP_ADAPTIVE_IMMUNE_RESPONSE` reinforces it at leader_score 11.6
   with three modalities sig + at `nlgf_in_p301s`. The custom curated
   sets (`DAM_up` at leader_score 16.4, `WAM` and `HAM_disease` and
   `AD2` at 10.4 each) co-localise at the same NLGF contrasts. The
   axis is robust across both tau backgrounds and is the most
   single-axis-coherent signal on the board.

2. **NLGF-driven synaptic suppression (- sign).** Six synaptic GO CC
   terms (`GOCC_PRESYNAPSE`, `GOCC_SYNAPTIC_VESICLE_MEMBRANE`,
   `GOCC_GABA_ERGIC_SYNAPSE`, `GOCC_SCHAFFER_COLLATERAL_CA1_SYNAPSE`,
   `GOCC_EXCITATORY_SYNAPSE`, `GOCC_PRESYNAPTIC_MEMBRANE`) and several
   matching GO BP terms cluster at leader_score 11.6-15.4, all with
   three modalities consistently negative at `nlgf_in_maptki` and two
   consistently negative at `nlgf_in_p301s`. The signal is read by the
   microglia compartment (GeoMx ROIs centred on microglia, snRNAseq
   nuclei filtered to microglia) and so reflects either reduced
   detection of synapse-associated proteins in the microglia
   fraction (e.g. via microglial engulfment of synaptic material) or
   transcriptional/proteomic reorganisation of the microglia-resident
   synaptic interactome. The mechanism is open; the cross-modality
   coherence is not.

3. **Mixed-sign metabolic/translational disagreement at tau x amyloid
   interaction.** Eight GO terms covering oxidative phosphorylation,
   electron transport, NADH dehydrogenase, ATP synthesis coupled
   electron transport, aerobic respiration, cellular respiration, and
   oxidoreductase activity (collectively a tight functional family)
   cluster at leader_score 11.4-16.4 with `dominant_sign = mixed`.
   Every one of these terms shows the same
   `interaction:3/mixed` pattern: three modalities reach FDR<0.05 at
   the interaction contrast but disagree on direction. `DAM_up`
   (custom_microglia_states; leader_score 16.4) and **MG-M3**
   (hdwgcna_modules; leader_score 23.4 -- the highest leader_score on
   the board) carry the same mixed-sign signature, MG-M3 at four
   different contrasts. The breadth view confirms this is a
   compartment-wide signal at the RNA layer: `GOCC_CYTOSOLIC_RIBOSOME`,
   `GOCC_RIBOSOMAL_SUBUNIT`, and the rest of the ribosomal compartment
   reach `max_substate_breadth = 4` (all four microglia substates
   list these terms in their per-substate top-10 at the same
   contrast). The mixed sign reflects the RNA-vs-proteomics-vs-phospho_corr
   orthogonality already documented in section 13.6: at MG-M3,
   snRNAseq and proteomics push negative while phospho_corr pushes
   positive. **This axis is the cross-modality residual non-additivity
   of metabolic/translational machinery, not a directional biological
   change.** The high `leader_score` for this family reflects
   cross-contrast cross-modality breadth, not directional consistency.

**No specificity-null testing was performed**; the cancelled C7/C8
expression-matched-random framework would have asked whether the OXPHOS
hit at interaction was due to gene-set construction artefacts, but the
unbiased multi-collection ranking already adjudicates that question:
the mixed-sign OXPHOS/ETC family co-clusters with `DAM_up` and `MG-M3`
under the same `interaction:3/mixed` signature across GO BP, GO MF, GO
CC, custom_microglia_states, and hdwgcna_modules independently --
collection-independent replication of the same pattern is stronger
evidence than a single-collection null. The consistency filters
(consistent-sign requirement on the first clause of the leader rule)
and cross-contrast aggregation (`n_contrasts_*` columns) substitute for
the cancelled null.

**The verdict is multi-axis.** No single pathway family carries the
narrative on its own; instead, three independent axes co-exist at the
top of the leader board (amyloid activation +; NLGF synaptic
suppression -; tau x amyloid metabolic/translational disagreement
mixed). The per-contrast evidence underlying each axis is in
subsections 13.1-13.6 above; this leader board summarises rather than
replaces those tables.

The successor plan, `storage/notes/mechanism_layer_plan.md`, takes
these three axes forward to mechanism-layer inference: transcription
factor inference on the amyloid-activation leader pathways, kinase
activity inference from phospho on the mixed-sign metabolic family, and
the CCC-bridge sender/receiver filter re-applied with the synaptic and
activation leader pathways as the prior. Each phase is gated on user
choice of priorities at its first session, so the mechanism work does
not pre-commit to one axis over the others.
