# TF activity inference

This section is the first layer of the mechanism-layer arc described in
`storage/notes/mechanism_layer_plan.md` (Phase E). The agnostic pathway
survey at section 12 surfaced **three biological axes** in the cross-
modality leader board: amyloid-driven activation, NLGF-driven synaptic
suppression, and mixed-sign metabolic/translational disagreement at
`tau x amyloid` interaction. The pathway-level evidence answers
*"which functional gene programmes shift"* but not *"which upstream
regulators drive those shifts"*. TF inference begins to answer the
latter.

The mechanism layer derives every regulator candidate from the same
DE caches the pathway survey used; no new data ingestion and no
re-anchoring on a pre-committed regulator family. The TF-target prior
is `CollecTRI` mouse via `decoupleR::get_collectri()` -- a curated
literature + ChIP-seq prior covering 1,114 TFs over 6,110 targets in
mouse symbol space. Phase F (kinase inference) and Phase G (CCC
re-analysis) will follow the same pattern with OmniPath kinase and
ligand-receptor priors.

## Cross-modality TF activity ranking

This subsection ranks TFs by cross-modality agreement at each contrast
using the same scoring conventions as section 12's pathway survey, so
the two layers are read side-by-side with parallel mental models.

**Inference threshold (stated before applying to data).** Per-modality
significance uses the `ulm` linear-model F-test p-value (decoupleR's
canonical univariate statistic), BH-adjusted within each
`(modality, contrast)` cell, with `padj < 0.10` as the per-modality
sig call. The threshold is locked at 0.10 (looser than the
pathway-survey's 0.05) because TF activity inference has sparser
per-TF evidence than fgsea -- CollecTRI TF target sets range from 5
to a few hundred genes vs. fgsea minimum 15 -- and the looser
threshold compensates for the lower per-modality power. Per-TF
ranking and sign attribution use the `consensus` score (decoupleR's
recommended primary ranking metric, signed, 87% sign-agreement with
ulm, 0.87 correlation). The two statistics serve two roles: ulm
arbitrates significance; consensus arbitrates magnitude and direction.
This split is the decoupleR vignette convention.

**Primary modalities and substate breadth axis.** Cross-modality
counts (`n_modalities_sig`, max 3) cover snRNAseq whole-microglia +
GeoMx + bulk proteomics. The four microglia substates (homeostatic,
DAM, IFN, proliferative) feed a separate substate-breadth axis
(max 4), mirroring the pathway-survey's treatment of substates as a
breadth axis rather than additional cross-modality columns. Phospho
is excluded from Phase E by design: phospho-site DE statistics need
protein-level collapse and signed re-mapping before they can serve as
TF-target evidence, which is the work of Phase F.

**Leader rule (verbatim from pathway-survey).** A `(TF, contrast)`
row is a leader if
`n_modalities_sig_consistent_sign >= 2` OR `n_modalities_sig >= 3`.
The first clause requires at least two modalities at FDR<0.10 with
the same sign of activity score; the second requires all three at
FDR<0.10 regardless of sign. The composite `leader_score` formula is
identical to the pathway-survey:
`5 * n_contrasts_consistent_sign_ge2 + n_contrasts_sig_ge3 + max_consistent_sign / 5`.

```{r tf-inference-build}
tf_activity <- readRDS(file.path(params$cache_dir,
                                 "tf_activity_decoupler.rds"))

contrasts5 <- c("nlgf_in_maptki", "nlgf_in_p301s",
                "interaction", "tau_alone", "tau_in_nlgf")
primary_mods <- c("snrnaseq", "geomx", "proteomics")

tf_ranking_per_contrast <- lapply(contrasts5, function(cn) {
  rank_tfs_cross_modality(tf_activity, cn,
                          primary_modalities = primary_mods)
})
names(tf_ranking_per_contrast) <- contrasts5

tf_ranking_long <- dplyr::bind_rows(lapply(contrasts5, function(cn) {
  r <- tf_ranking_per_contrast[[cn]]
  r$contrast <- cn
  r
}))

tf_substate_breadth_long <- compute_tf_substate_breadth(tf_activity,
                                                       contrasts = contrasts5)

tf_leaderboard <- build_tf_leader_board(
  tf_ranking_long,
  substate_breadth_long = tf_substate_breadth_long
)

readr::write_tsv(
  tf_leaderboard,
  file.path(params$results_dir, "tf_activity_unified_leaderboard.tsv")
)

cat(sprintf(
  paste0("Cross-modality TF ranking: %d unique TFs covered across ",
         "primary modalities (%s) at the 5 contrasts.\n",
         "Unified leader board: %d leader TFs at FDR<0.10 with the ",
         "default rule (>=2 consistent sign OR >=3 sig across the 3 ",
         "primary modalities at any single contrast).\n"),
  length(unique(tf_ranking_long$source)),
  paste(primary_mods, collapse = ", "),
  nrow(tf_leaderboard)
))
```

### Per-contrast cross-modality top-10 TFs

The five tables below show the top 10 TFs per contrast by the same
composite rank used in the pathway survey (sign-consistent significant-
modality count desc, then mean `|consensus score|` desc). Contrasts
with no sig TFs at any modality still produce a ranking by absolute
consensus score; rows with `n_modalities_sig == 0` are sorted by mean
`|consensus score|` and do not contribute to the leader board.

```{r tf-inference-per-contrast-tables, results = 'asis'}
for (cn in contrasts5) {
  print(format_tf_ranking_table(tf_ranking_per_contrast[[cn]],
                                contrast = cn, n = 10))
}
```

The qualitative pattern across the five tables: TFs reach
cross-modality significance almost exclusively at the two NLGF
contrasts (`nlgf_in_maptki`, `nlgf_in_p301s`). The interaction
contrast yields a thin signal (1 TF at the leader rule's all-three-sig
clause); `tau_alone` and `tau_in_nlgf` have essentially no
cross-modality sig signal. This is the empirical reflection of the
TF inference layer being lower-powered than gene-set enrichment at
small effect sizes -- the interaction and tau-only contrasts already
showed weaker cross-modality pathway signal in section 12, and the
TF layer inherits that limitation.

### Unified TF leader board

```{r tf-inference-leader-table, results = 'asis'}
tf_leader_view <- tf_leaderboard |>
  dplyr::transmute(
    source,
    leader_score,
    n_contrasts_cons_ge2 = n_contrasts_consistent_sign_ge2,
    n_contrasts_sig_ge3,
    max_consistent_sign,
    dominant_sign,
    max_substate_breadth,
    contrasts_summary
  )

print(knitr::kable(
  tf_leader_view, digits = 2,
  caption = paste0(
    "Unified TF leader board (all leader rows). leader_score = ",
    "5 * n_contrasts_consistent_sign_ge2 + n_contrasts_sig_ge3 + ",
    "max_consistent_sign / 5; max_substate_breadth is the count out ",
    "of 4 microglia substates where the TF reaches FDR<0.10 on ",
    "consensus p-value at any contrast."
  )
))
```

### TF x modality activity heatmap (nlgf_in_maptki contrast)

The heatmap below shows the top leader TFs (alphabetical row order so
no TF is visually privileged) by signed consensus score across the
three primary modalities at the `nlgf_in_maptki` contrast -- the
contrast with the largest cross-modality TF leader population. Cell
text marks per-modality significance: `*` = FDR<0.05, `.` = FDR<0.10.
NA cells (TF dropped by `minsize=5` in that modality's CollecTRI
target overlap) are blank.

```{r tf-inference-heatmap, fig.width = 7, fig.height = 4.5}
heatmap_contrast <- "nlgf_in_maptki"
tf_alpha <- sort(tf_leaderboard$source)

ranking_for_heatmap <- tf_ranking_per_contrast[[heatmap_contrast]]
score_cols <- paste0("score_", primary_mods)
padj_cols  <- paste0("padj_",  primary_mods)

score_mat <- as.matrix(
  ranking_for_heatmap[match(tf_alpha, ranking_for_heatmap$source),
                      score_cols, drop = FALSE]
)
padj_mat  <- as.matrix(
  ranking_for_heatmap[match(tf_alpha, ranking_for_heatmap$source),
                      padj_cols, drop = FALSE]
)
rownames(score_mat) <- tf_alpha
rownames(padj_mat)  <- tf_alpha
colnames(score_mat) <- primary_mods
colnames(padj_mat)  <- primary_mods

range_abs <- max(abs(score_mat[is.finite(score_mat)]), na.rm = TRUE)
if (!is.finite(range_abs) || range_abs == 0) range_abs <- 1
col_fun <- circlize::colorRamp2(c(-range_abs, 0, range_abs),
                                c("#3a4cc0", "white", "#b40426"))
cell_fn <- function(j, i, x, y, width, height, fill) {
  if (is.na(padj_mat[i, j])) return()
  if (padj_mat[i, j] < 0.05) {
    grid::grid.text("*", x, y,
                    gp = grid::gpar(fontsize = 11, col = "black"))
  } else if (padj_mat[i, j] < 0.10) {
    grid::grid.text(".", x, y,
                    gp = grid::gpar(fontsize = 11, col = "black"))
  }
}
ComplexHeatmap::Heatmap(
  score_mat,
  name            = "consensus score",
  col             = col_fun,
  cluster_rows    = FALSE,
  cluster_columns = FALSE,
  row_order       = seq_len(nrow(score_mat)),
  column_order    = seq_len(ncol(score_mat)),
  cell_fun        = cell_fn,
  row_names_gp    = grid::gpar(fontsize = 10),
  column_names_gp = grid::gpar(fontsize = 10),
  column_title    = sprintf(
    paste0("Leader TFs x modality (consensus score) at '%s' contrast ",
           "('*' = FDR<0.05, '.' = FDR<0.10; rows alphabetical)"),
    heatmap_contrast),
  na_col          = "grey90",
  border          = TRUE
)
```

### Interpretive notes for this subsection

The cross-modality TF leader board concentrates almost entirely on
the **amyloid-driven activation axis** identified in section 12:
every leader TF has `dominant_sign = "+"`, and contrasts at which
leaders qualify are the two NLGF contrasts. The synaptic suppression
axis (negative signal at NLGF contrasts on synaptic GO CC terms) and
the mixed-sign metabolic/translational interaction axis do not
surface at the TF layer with this prior and threshold. Two
mechanistic readings of that absence: (a) synaptic suppression in
section 12 is most likely driven by detection-level engulfment /
trans-synaptic protein reorganisation rather than transcriptional
regulation, so a TF-target prior would not be expected to surface
it; (b) the interaction-axis OXPHOS / ribosomal signal is a
metabolic compartment shift that may be governed by post-
transcriptional regulators (kinases, miRNAs) not represented in a
TF-target prior. Phase F will revisit the interaction axis with the
OmniPath kinase prior. The full per-axis treatment, including this
absence as a *finding*, lives in subsection 13.3 (verdict).

The leader TFs themselves are biologically face-valid for the
amyloid-activation axis. Nr1h3 (LXR alpha) is the canonical
microglial cholesterol-handling and DAM-priming regulator; Spi1
(PU.1) is the master microglial lineage TF and an established AD
GWAS locus; Irf1 mediates interferon-driven activation that overlaps
with the IFN microglial substate. Usf2 is the only TF with
`max_substate_breadth = 4` (active in every microglia substate at
some contrast), making it the closest TF analogue of the compartment-
wide patterns surfaced in the pathway survey's per-substate analysis.
Two CollecTRI complex sources (`A0A979HLR9`, `A0A087WSP5`) appear in
the leader board with UniProt-accession names rather than gene
symbols; these are CollecTRI's "complex" entries (typically
heterodimer-resolved). E4 will decide whether to filter them out of
axis-restricted analyses or annotate them and retain.

## Axis-restricted TF rankings

Subsection 14.1 ranked TFs by their *global* cross-modality activity
profile and surfaced almost exclusively amyloid-activation drivers at
the two NLGF contrasts. Subsection 14.2 asks the complementary
question: for each of the three biological axes named in section
13's D2 verdict (amyloid-driven activation; NLGF-driven synaptic
suppression; mixed-sign metabolic/translational at the `tau x
amyloid` interaction), which TFs carry the strongest activity
*within that axis's gene universe*? The answer is computed by
restricting the CollecTRI TF--target graph to TFs whose target sets
overlap the axis universe at >= 5 distinct symbols, and averaging the
existing E2 cache scores across the (axis-contrast x primary-modality)
cells.

**Axis classification rule (locked at E4).** A leader row from
`storage/results/pathway_survey_unified_leaderboard.tsv` is assigned
to axis `amyloid_activation` if its `contrasts_summary` matches the
regex `nlgf_in_[a-z0-9]+:[0-9]+/[+]` (an NLGF contrast with positive
consensus sign); to `synaptic_suppression` if it matches
`nlgf_in_[a-z0-9]+:[0-9]+/-` (NLGF, negative sign); to
`interaction_metabolic` if it matches `interaction:[0-9]+/mixed`
(the interaction contrast with mixed-sign consensus). Multi-label
is permitted -- a row may contribute to more than one axis universe,
reflecting that single pathways may bridge axes (e.g. `DAM_up`
contributes to both axes 1 and 3 in the live leader board). Orphan
rows (matching no axis) are dropped. The rule is strictly directional
and so may differ from the predecessor pathway-survey prose in
borderline cases; notably `AD2`, which the section 13 verdict
grouped with the amyloid-activation axis by curatorial provenance,
carries `nlgf_in_maptki:2/-` in the leader board and so lands in
`synaptic_suppression` by the directional rule. Directional grouping
is preferred at the mechanism layer because the axis universe should
reflect the biological direction the axis represents -- mixing
direction within an axis would dilute the per-TF inference signal.

**Interpretation A (locked at E4).** Per-TF activity scores are read
from the existing `tf_activity_decoupler.rds` cache (computed against
each modality's full universe in E2). E4 restricts the TF *set* by
universe-overlap (CollecTRI targets >= `min_targets = 5` distinct
symbols in the axis universe), then averages the cached consensus
scores across the (axis-contrast x primary-modality) cells where the
TF has a finite score. The alternative ("Interpretation B" -- re-run
decoupleR with each modality's CollecTRI network clipped to the axis
universe) was considered and rejected: clipping shrinks per-TF
target counts inside the run, costing statistical power that the
small per-modality universes already cannot afford. Mean is taken
signed (not absolute); ranking within axis is by `|mean|` so
strong-signal TFs surface regardless of sign. The locked choice is
recorded in the `R/tf_inference.R` module header so a future session
can revisit on demand.

```{r tf-axis-restricted-build}
leaderboard <- readr::read_tsv(
  file.path(params$results_dir,
            "pathway_survey_unified_leaderboard.tsv"),
  show_col_types = FALSE
)

# Gene-set caches keyed by the leaderboard's `collection` column.
# msigdb_* are named lists `pathway -> character` of mouse gene
# symbols; custom_microglia_* are likewise. hdWGCNA modules need
# Ensembl -> symbol translation (the `modules` slot uses Seurat
# rownames i.e. ENSMUSG IDs and CollecTRI is symbol-keyed).
gobp     <- readRDS(file.path(params$cache_dir, "msigdb_gobp_mouse.rds"))
gocc     <- readRDS(file.path(params$cache_dir, "msigdb_gocc_mouse.rds"))
gomf     <- readRDS(file.path(params$cache_dir, "msigdb_gomf_mouse.rds"))
cstates  <- readRDS(file.path(params$cache_dir,
                              "custom_microglia_states.rds"))
cad      <- readRDS(file.path(params$cache_dir,
                              "custom_microglia_ad.rds"))

hwg        <- readRDS(file.path(params$cache_dir,
                                "hdwgcna_microglia.rds"))
symbol_map <- readRDS(file.path(params$cache_dir,
                                "snrnaseq_symbol_map.rds"))
hwg_modules <- hwg$modules
hwg_modules$gene_name <- symbol_map$symbol[
  match(hwg_modules$gene_name, symbol_map$ensembl)]
hwg_modules <- hwg_modules[
  !is.na(hwg_modules$gene_name) & nzchar(hwg_modules$gene_name) &
    as.character(hwg_modules$module) != "grey", , drop = FALSE]

gene_sets <- list(
  gobp                    = gobp,
  gocc                    = gocc,
  gomf                    = gomf,
  custom_microglia_states = cstates,
  custom_microglia_ad     = cad,
  hdwgcna_modules         = hwg_modules
)

axis_universes <- build_axis_gene_universe(leaderboard, gene_sets)

# CollecTRI mouse prior. OmnipathR-cached locally from E2; near-
# instantaneous on the second call. split_complexes = FALSE retains
# heterodimer-resolved TF complexes (UniProt-accession sources).
collectri_mouse <- decoupleR::get_collectri(
  organism        = "mouse",
  split_complexes = FALSE
)

tf_axis_scores <- score_tf_per_axis(
  tf_cache           = tf_activity,
  axis_universes     = axis_universes,
  network            = collectri_mouse,
  primary_modalities = primary_mods,
  min_targets        = 5L,
  score_statistic    = "consensus"
)

readr::write_tsv(
  tf_axis_scores,
  file.path(params$results_dir, "tf_activity_axis_restricted.tsv")
)

axis_summary <- vapply(names(axis_universes), function(ax) {
  n_uni <- length(axis_universes[[ax]]$universe)
  n_ctr <- length(axis_universes[[ax]]$axis_contrasts)
  n_tf  <- sum(tf_axis_scores$axis == ax)
  sprintf("  %-22s  %5d universe genes  %d axis contrasts  %d TFs scored",
          ax, n_uni, n_ctr, n_tf)
}, character(1))
cat(sprintf(
  "Axis-restricted TF inference: %d axes (%s).\n",
  length(axis_universes),
  paste(names(axis_universes), collapse = ", ")))
cat(paste(axis_summary, collapse = "\n"), "\n", sep = "")
```

### Amyloid-activation axis (axis 1)

The amyloid-activation universe is the union of leader-pathway gene
sets whose `contrasts_summary` carries at least one positive NLGF
contrast: the gene programmes that section 13's D2 verdict identified
as upregulated at amyloid onset (MG-M2, MHC II antigen presentation,
adaptive immune response, the activated DAM_up / WAM / HAM_disease /
IRM cores). The per-TF mean is the average of the consensus score
across (NLGF contrasts x primary modalities) cells in the E2 cache;
the signed mean carries direction (positive: TF activity rises with
amyloid in this universe; negative: TF activity falls).

```{r tf-axis-table-amyloid, results = 'asis'}
print(format_axis_restricted_table(tf_axis_scores,
                                   "amyloid_activation",
                                   n = 15))
```

```{r tf-axis-lollipop-amyloid, fig.width = 7, fig.height = 4.5}
plot_axis_lollipop(tf_axis_scores, "amyloid_activation", n = 12)
```

### Synaptic-suppression axis (axis 2)

The synaptic-suppression universe is the union of leader-pathway gene
sets whose `contrasts_summary` carries at least one negative NLGF
contrast: six synaptic GO CC terms (synapse, presynaptic / postsynaptic
membranes, plus matching compartments) and parallel synapse GO BP
terms. Axes 1 and 2 share the same set of `axis_contrasts` (both
NLGF contrasts) in the per-TF mean by construction of the directional
rule, so the *ranking by mean* is structurally similar to axis 1
modulo which TFs survive the universe-overlap filter. The biological
differentiator lives in `n_targets_in_axis_universe`, which counts
CollecTRI targets of each TF that fall inside the synaptic-suppression
universe; contrast with the same column under axis 1 for any shared
TF to see the per-axis target-density story.

```{r tf-axis-table-synaptic, results = 'asis'}
print(format_axis_restricted_table(tf_axis_scores,
                                   "synaptic_suppression",
                                   n = 15))
```

```{r tf-axis-lollipop-synaptic, fig.width = 7, fig.height = 4.5}
plot_axis_lollipop(tf_axis_scores, "synaptic_suppression", n = 12)
```

### Interaction metabolic/translational axis (axis 3)

The interaction universe is the union of leader-pathway gene sets
whose `contrasts_summary` carries the `interaction` contrast with
mixed-sign consensus: the eight OXPHOS / ETC GO terms, the ribosomal
/ translational compartment, and the `MG-M3` / `DAM_up` interaction
reading. Unlike axes 1 and 2, axis 3's `axis_contrasts` reduces to
the single `interaction` contrast, so the per-TF mean is
structurally independent of the NLGF-driven means above. The signed
mean here is expected to lean negative for proliferation /
cell-fate TFs whose target programmes include OXPHOS or ribosomal
biogenesis.

```{r tf-axis-table-interaction, results = 'asis'}
print(format_axis_restricted_table(tf_axis_scores,
                                   "interaction_metabolic",
                                   n = 15))
```

```{r tf-axis-lollipop-interaction, fig.width = 7, fig.height = 4.5}
plot_axis_lollipop(tf_axis_scores, "interaction_metabolic", n = 12)
```

### Interpretive notes for this subsection

The axis-restricted TF rankings deliver three biological readings,
two of which are tightly coupled by construction and one of which
is genuinely independent.

**Axis 1 (amyloid-activation)** broadens the cross-modality leader
board from 14.1: Nr1h3 (LXR alpha, microglial cholesterol / DAM
priming), Spi1 (PU.1, master microglial lineage TF + AD GWAS locus),
Irf1 (interferon-driven activation) all surface among the top
mean-activity TFs, alongside canonical inflammatory amplifiers
Nfkb1 / Rela / Sp3 / Usf1 / Usf2. The signed mean is predominantly
positive across the top of the ranking (10 of the top 15 TFs;
expected, given the amyloid axis is defined as positive at NLGF);
the five negative entries (Rreb1, Sin3a, Pa2g4, Klf7, Snai2) are
TFs whose targets are coordinately *repressed* in the amyloid
universe -- a real signal worth retaining rather than filtering on
sign alone. Spi1's CollecTRI target set is heavily skewed towards
the amyloid universe (184 in-universe targets) vs the synaptic
universe (84 in-universe targets) -- the live values in the
`n_targets_in_axis_universe` column for the rendered tables --
confirming the bookkeeping correctly attributes Spi1's primary axis
as amyloid.

**Axis 2 (synaptic-suppression)** does **not** deliver a separate set
of "synaptic-suppression TFs" by the mean-activity ranking. By the
Interpretation-A formula the per-TF mean is identical to axis 1 for
any TF passing both universe-overlap filters: the contrasts averaged
over are the same (both NLGF contrasts) and the cached scores are
computed against full per-modality universes, not against
axis-restricted networks. The differentiator is structural -- which
TFs survive the synaptic-universe filter, and how many of their
CollecTRI targets fall in that universe. This is the documented
property of Interpretation A; an Interpretation-B re-run (one
decoupleR call per axis with the network clipped to the axis
universe) would deliver genuinely axis-specific scores at the cost
of per-modality statistical power. The choice was deliberate and
recorded.

The biological reading of axis 2 should therefore be: "these are TFs
whose target genes overlap the synaptic-suppression universe at >=5
CollecTRI edges *and* are active at NLGF contrasts" -- i.e. TFs
whose activity is upregulated by amyloid while having synaptic-
program targets. Whether their NLGF activity is *causally driving*
the synaptic suppression or merely co-occurring with it cannot be
resolved at the TF layer alone; the E5 verdict subsection will
revisit with sign-aware interpretation.

**Axis 3 (interaction metabolic/translational)** is the genuinely
independent axis at this layer. Its `axis_contrasts` reduces to the
single `interaction` contrast, so the per-TF mean diverges sharply
from axes 1/2. Top TFs by `|mean|` are canonical
metabolic/proliferation regulators -- Myc (drives ribosomal
biogenesis and translation), Creb1, Tp53, Jun, Fos, Hif1a, Ctnnb1
-- and the signed mean leans **negative** for 14 of the top 15 TFs
(Hdac1 the lone positive exception, with only 9 in-universe targets;
its rank is more a function of low denominator noise than a robust
opposing signal). This is the upstream signal one would predict to
drive the negative-NES end of the OXPHOS / ribosome gene-set
readings at the `interaction` contrast in section 13: a coordinated
reduction in Myc-driven biosynthetic activity at the `tau x amyloid`
interaction is the most parsimonious mechanism for a compartment
shift away from the high-translation / high-OXPHOS state. The
mixed-sign character of the gene-set readings at the pathway layer
reflects internal heterogeneity inside the OXPHOS / ribosome sets,
but the TF layer reads a coherent direction.

A diagnostic cross-axis observation: Rela (the NF-kB p65 subunit)
appears in *both* axis 1 (rank 7, mean +2.19) and axis 3 (rank 13,
mean -1.36), with opposite signs. The amyloid contrasts activate
Rela targets (inflammatory response on); the interaction contrast
suppresses them (compartment-shift away from inflammation as part
of the metabolic re-tuning at the `tau x amyloid` interface).
A0A087WSP5 (the second CollecTRI complex source) shows the same
sign-reversal pattern (axis 1 mean +2.42; axis 3 mean -1.59). These
sign reversals are not bookkeeping artefacts -- they are the
expected signature of a coherent biological transition between the
two contrast settings. The E5 verdict will weigh whether the axis-3
signal is robust to the more demanding sign-aware aggregation.

**Resolution of the open question from 14.1.4.** The two CollecTRI
complex sources `A0A979HLR9` and `A0A087WSP5` are **retained** in
the axis-restricted analyses, not filtered out. They are
heterodimer-resolved CollecTRI entries (`split_complexes = FALSE`),
both surface repeatedly with high `|mean|` scores in axes 1 and 3,
and both have substantial target counts in those axis universes.
Filtering them out would obscure substantive biology. A future
annotation pass (deferred beyond Phase E) can resolve them to
specific subunits via UniProt; for now they are reported with their
accession names and treated as compound TF entities.

The E5 verdict subsection will (a) take the per-axis top TFs above
and produce a sign-aware narrative per axis, mirroring the D2
verdict shape from section 13; (b) cross-check that the three axes
remain visibly distinct after the verdict-level aggregation, per the
anti-anchoring guardrails; and (c) explicitly list any axis whose
TF-layer reading does not corroborate the section-13 pathway-layer
verdict, naming the inconsistency rather than smoothing it.

## Verdict (sign-aware TF synthesis)

This subsection synthesises the cross-modality TF ranking (14.1) and
the axis-restricted TF ranking (14.2) into a per-axis verdict, mirroring
the shape of the section-13 pathway-survey verdict. The mechanism-layer
reading is *not* a re-derivation of the pathway-survey axes -- those
axes are accepted as inputs -- it is an answer to the question *which
upstream regulators carry the strongest signal within each axis's
biology, and which axis is best read by the TF layer at all?*

**Inference thresholds (re-stated before the verdict so the bar is
explicit).** Per-modality significance is FDR<0.10 on the decoupleR
`ulm` p-value (BH-adjusted within each `(modality, contrast)` cell);
ranking and sign use the `consensus` score. The cross-modality leader
rule (subsection 14.1) is `n_modalities_sig_consistent_sign >= 2` OR
`n_modalities_sig >= 3` across snRNAseq whole-microglia + GeoMx + bulk
proteomics. The axis-restricted ranking (subsection 14.2) filters TFs
to those with `>= 5` CollecTRI targets in the axis gene universe and
ranks within axis by `|mean consensus score|`; the verdict displays the
top 5 TFs per axis.

```{r tf-inference-verdict-build}
axis_scores_tbl <- readr::read_tsv(
  file.path(params$results_dir,
            "tf_activity_axis_restricted.tsv"),
  show_col_types = FALSE
)

tf_verdict_evidence <- list(
  amyloid_activation = paste0(
    "Top 5 axis-restricted TFs are 4-positive / 1-negative; signed mean ",
    "range [-2.57, +4.32] across NLGF contrasts. Hybrid per-contrast view: ",
    "nlgf_in_maptki:[-2.09, +5.40]; nlgf_in_p301s:[-3.06, +3.23] (maptki ",
    "carries the stronger positive tail; p301s the stronger negative tail) ",
    "-- a real per-contrast asymmetry the axis-mean smooths over. ",
    "2 of the top 5 (A0A979HLR9, Spi1) also reach the cross-modality ",
    "leader rule at the two NLGF contrasts (subsection 14.1). Reading: ",
    "Spi1 (PU.1, master microglial lineage TF + AD GWAS locus), ",
    "Nfkb1 / Sp3 (canonical inflammatory amplifiers), and the CollecTRI ",
    "complex source A0A979HLR9 are activated at NLGF onset; Rreb1 is the ",
    "lone repressor in the top 5. Corroborates the section-13 axis-1 ",
    "verdict (MG-M2 module, MHC II antigen-presentation family, ",
    "GOBP_ADAPTIVE_IMMUNE_RESPONSE, custom DAM_up / WAM / HAM_disease ",
    "sets) at the mechanism layer."),
  synaptic_suppression = paste0(
    "Top 5 axis-restricted TFs are identical to axis 1 by Interpretation ",
    "A construction (both axes share their NLGF contrasts and the ",
    "per-TF mean is computed from full-universe scores). Hybrid per-",
    "contrast view is therefore also identical to axis 1 ",
    "(nlgf_in_maptki:[-2.09, +5.40]; nlgf_in_p301s:[-3.06, +3.23]); the ",
    "axis 1 / axis 2 entanglement propagates to the per-contrast layer ",
    "as a structural consequence of the shared contrast set, not as an ",
    "information loss specific to the axis-mean. The structural ",
    "differentiator is in-universe target counts (e.g. Spi1: 184 ",
    "amyloid-universe targets vs 84 synaptic-universe targets). Reading: ",
    "the TF layer does NOT independently identify drivers of synaptic ",
    "suppression at this resolution; the synaptic-axis top TFs read as ",
    "'amyloid-activated TFs whose target programmes also include ",
    "synaptic genes', not as 'TFs that drive synaptic gene downregulation'. ",
    "This is an honest non-finding -- the absence of a distinct TF ",
    "signature for synaptic suppression is consistent with the ",
    "engulfment-mediated mechanism hypothesis raised in 14.1.4 and ",
    "14.2.4."),
  interaction_metabolic = paste0(
    "Top 5 axis-restricted TFs are uniformly negative; signed mean range ",
    "[-4.85, -1.81] at the single interaction contrast. Hybrid per-",
    "contrast view collapses to interaction:[-4.85, -1.81], identical to ",
    "the axis-mean range by construction (only one axis_contrast at axis ",
    "3); the hybrid columns are preserved in the TSV for schema ",
    "consistency with axes 1 and 2. 1 of the top 5 (A0A979HLR9) appears ",
    "in the cross-modality leader board, but its leader status comes ",
    "from the NLGF contrasts -- no TF reached the cross-modality leader ",
    "rule at the interaction contrast in the current build. Reading: ",
    "Myc / Creb1 / Tp53 / Jun, the canonical metabolic and proliferation ",
    "regulators of ribosomal biogenesis and biosynthetic activity, are ",
    "coordinately suppressed at the tau x amyloid interaction. The TF ",
    "layer reads a COHERENT NEGATIVE direction here, in contrast to the ",
    "mixed-sign signal at the gene-set layer of section 13. The most ",
    "parsimonious upstream mechanism for the section-13 axis-3 biology ",
    "(MG-M3 module + DAM_up at interaction + the OXPHOS/ETC GO family + ",
    "ribosomal compartment) is a Myc-led reduction in biosynthetic ",
    "throughput at the tau x amyloid interface; the mixed-sign reading ",
    "at the gene-set layer reflects RNA / proteomics / phospho-corrected ",
    "orthogonality already documented in section 13, not internal ",
    "disagreement on direction at the TF layer.")
)

tf_verdict <- build_tf_verdict_table(
  axis_tbl           = axis_scores_tbl,
  tf_leaderboard     = tf_leaderboard,
  n_top_per_axis     = 5L,
  evidence_summaries = tf_verdict_evidence
)

readr::write_tsv(
  tf_verdict,
  file.path(params$results_dir, "tf_activity_verdict.tsv")
)

cat(sprintf(
  "TF inference verdict: %d axes summarised (%s), top-%d TFs per axis.\n",
  nrow(tf_verdict),
  paste(tf_verdict$axis, collapse = ", "),
  5L))
```

### Verdict summary table

The summary table below has one row per D2 axis. `top_TFs` lists the
top 5 TFs per axis sorted by `|mean consensus score|` (the axis-
restricted ranking from 14.2); `top_TF_signs` carries the signed
direction of each TF's mean (`+` = positive activator,
`-` = negative repressor at the axis's contrast set); `mean_score_range`
gives the [min, max] of the signed mean across those 5 TFs;
`per_contrast_summary` is the locked **hybrid** companion -- it gives
the [min, max] of the cross-modality mean at each axis-contrast in
isolation, so a reader can see whether the axis-mean smooths over
heterogeneous per-contrast signal or is reading a coherent direction.
`n_top_in_cross_modality_leaderboard` counts how many of the top 5
also reached the cross-modality leader rule (subsection 14.1);
`evidence_summary` is a prose distillation of the per-axis reading.

The hybrid display (mean range + per-contrast ranges side-by-side) was
locked at the F4 decision gate on 2026-05-24 (user directive: "Do the
hybrid. Have other analyses do the hybrid as well for consistency") and
retrofitted to this E5 verdict layer for consistency with the F4
kinase verdict that follows in subsection 15.2.

```{r tf-inference-verdict-table, results = 'asis'}
verdict_view <- tf_verdict |>
  dplyr::transmute(
    axis,
    top_TFs,
    top_TF_signs,
    mean_score_range,
    per_contrast_summary,
    n_top_in_cm_lb = n_top_in_cross_modality_leaderboard
  )
print(knitr::kable(
  verdict_view, digits = 2,
  caption = paste0(
    "Per-axis TF verdict summary (top 5 axis-restricted TFs per axis). ",
    "Hybrid view: `mean_score_range` gives the [min, max] of the ",
    "axis-mean across the top 5; `per_contrast_summary` gives ",
    "per-axis-contrast [min, max] of the cross-modality mean across ",
    "the same top 5 (exposes per-contrast asymmetry that the axis-mean ",
    "smooths over). `n_top_in_cm_lb` = count of top-5 TFs that also ",
    "reached the cross-modality leader rule in subsection 14.1."
  )
))
```

The `evidence_summary` column is omitted from the inline kable to keep
the row heights tractable; the full per-axis prose lives in subsections
14.3.1-14.3.3 below, and the column is preserved verbatim in the
exported TSV (`storage/results/tf_activity_verdict.tsv`). The TSV
also carries `per_contrast_score_range` (the [min, max] across the
top-5 x axis-contrasts cells, useful for at-a-glance comparison to
`mean_score_range`).

### Amyloid-activation axis (axis 1)

**Top 5 TFs by axis-restricted |mean|:** A0A979HLR9 (+4.32), Spi1
(+2.81), Rreb1 (-2.57), Sp3 (+2.50), Nfkb1 (+2.47). Four positive,
one negative; signed mean range [-2.57, +4.32]. Two of the five
(A0A979HLR9 and Spi1) also reached the cross-modality leader rule at
the two NLGF contrasts in subsection 14.1, with `dominant_sign = "+"`
in both cases.

**Per-contrast hybrid view (locked F4).** The cross-modality mean
ranges of the same top 5 are
`nlgf_in_maptki:[-2.09, +5.40]; nlgf_in_p301s:[-3.06, +3.23]`. The
maptki contrast (NLGF on the tau-loss background) carries the
stronger positive activator tail (+5.40 > axis-mean upper +4.32 >
p301s upper +3.23); the p301s contrast (NLGF on the tau-gain
background) carries the stronger negative repressor tail (-3.06 <
axis-mean lower -2.57 < maptki lower -2.09). The two NLGF contrasts
therefore agree on direction for most top TFs but disagree on the
magnitude of the extremes -- a real asymmetry the axis-mean
[-2.57, +4.32] smooths over. The hybrid display is intentional: a
reader concerned that "axis 1 reads strongly positive" might mask a
mixed-direction signal can check the per-contrast row and see it does
not. The asymmetry is consistent with the literature reading that
the tau-loss MAPTKI background releases a stronger Spi1 / Nfkb1
amyloid-on response (no tau-driven attenuation), and the tau-gain
P301S background runs a tighter activator window but a wider
repressor window (consistent with NF-kB-family cross-regulation
detuning under tauopathy).

**Biological reading.** Spi1 (PU.1) is the master microglial lineage
TF and a well-established AD GWAS locus; its activation at NLGF
onset is the prototypical microglial-program-on signal at the
amyloid--tau interface. Nfkb1 (NF-kB p50) and Sp3 are canonical
inflammatory amplifiers whose joint activation is a hallmark of
microglial disease-associated states. A0A979HLR9 is a CollecTRI
heterodimer-resolved complex source retained per the 14.2.4 decision
to keep complex sources in axis-restricted analyses (footnote: a
future UniProt annotation pass can resolve the subunit identity).
Rreb1 is a Ras-responsive transcription repressor; its negative
signal at NLGF contrasts indicates coordinated downregulation of a
small Rreb1-target programme in the amyloid universe (only 19
in-axis-universe targets) -- a real but minor counter-signal worth
preserving rather than filtering.

**Cross-reference to section-13 axis 1.** Section 13's axis-1
verdict named MG-M2 (hdWGCNA module), the MHC II
antigen-presentation family (`GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_*`
x5 + `GOMF_ANTIGEN_BINDING` + `GOMF_PEPTIDE_ANTIGEN_BINDING`),
`GOBP_ADAPTIVE_IMMUNE_RESPONSE`, and the custom `DAM_up`, `HAM_disease`,
`WAM` sets as the leader-pathway evidence. The TF-layer reading
above corroborates that evidence with a parsimonious mechanism: the
Spi1 / Nfkb1 / Sp3 / Rela ensemble are the canonical drivers of
microglial inflammatory activation, antigen-processing-and-presentation
machinery, and DAM-state induction in the published literature. The
mechanism layer therefore reads axis 1 *strongly*.

### Synaptic-suppression axis (axis 2)

**Top 5 TFs by axis-restricted |mean|:** identical to axis 1
(A0A979HLR9, Spi1, Rreb1, Sp3, Nfkb1) by Interpretation A
construction. The per-TF mean is computed from full-modality-universe
consensus scores averaged over the two NLGF contrasts; both NLGF
axes share that contrast set, so the mean ranking does not
differentiate them.

**Per-contrast hybrid view (locked F4).** The per-contrast summary is
likewise identical to axis 1
(`nlgf_in_maptki:[-2.09, +5.40]; nlgf_in_p301s:[-3.06, +3.23]`)
because the same contrast pair is averaged over the same top 5 TFs.
This is the expected hybrid behaviour at axis 2 under Interpretation
A and is preserved unsmoothed in the verdict so a future reader can
see that the entanglement between axes 1 and 2 propagates to the
hybrid layer as well -- it is a structural property of sharing the
contrast set, not an information loss specific to the axis-mean.

**The structural differentiator is in-universe target count, not
mean.** Spi1's CollecTRI target set has 184 hits in the
amyloid-activation universe but only 84 hits in the synaptic-
suppression universe; A0A979HLR9 (the complex source) has 442 vs
383. The synaptic axis is therefore best interpreted as a
*target-overlap filter*: which amyloid-activated TFs have target
programmes that *also* touch the synaptic-suppression universe? Not
as a list of TFs that *drive* synaptic suppression.

**This is the honest non-finding.** The mechanism layer does not
identify TF-level drivers of synaptic suppression at this resolution
with this prior. Two non-mutually-exclusive interpretations of the
absence: (a) synaptic suppression in the microglia compartment is
detection-level engulfment -- microglia phagocytose synaptic
material at NLGF onset and so apparent downregulation of synaptic
transcripts/proteins inside the microglia fraction is a
mass-balance signal, not a transcriptional one. CollecTRI cannot
surface a TF for a mechanism that is not transcriptional. (b)
Transcriptional repression of synaptic genes in microglia, if it
occurs, may be driven by repressor complexes or epigenetic
regulators not well-represented in the CollecTRI literature-curated
prior (e.g. PRC2, NuRD, REST). A future plan could test (b) with a
chromatin-prior tool (e.g. ChEA3) before concluding the absence is
biological rather than methodological.

**Cross-reference to section-13 axis 2.** Section 13's axis-2
verdict named six synaptic GO CC terms (`GOCC_EXCITATORY_SYNAPSE`,
`GOCC_GABA_ERGIC_SYNAPSE`, `GOCC_PRESYNAPSE`, `GOCC_PRESYNAPTIC_MEMBRANE`,
`GOCC_SCHAFFER_COLLATERAL_CA1_SYNAPSE`, `GOCC_SYNAPTIC_VESICLE_MEMBRANE`,
listed alphabetically) plus matching synapse GO BP terms, all
consistently negative at the two NLGF contrasts. The TF-layer reading
above does *not* corroborate that pathway-layer evidence with a
distinct mechanism: the verdict for axis 2 is therefore an explicit
inconsistency call-out -- the pathway-layer signal is real but the
TF layer cannot name its upstream driver. Phase F (kinase inference)
and Phase G (CCC sender-receiver) will revisit.

### Interaction metabolic/translational axis (axis 3)

**Top 5 TFs by axis-restricted |mean|:** A0A979HLR9 (-4.85), Myc
(-3.25), Creb1 (-2.28), Tp53 (-2.15), Jun (-1.81). All five
negative; signed mean range [-4.85, -1.81] at the single
interaction contrast. One of the five (A0A979HLR9) appears in the
cross-modality leader board, but its leader status is earned at
the NLGF contrasts; no TF reached the cross-modality leader rule at
the interaction contrast itself in the current build.

**Per-contrast hybrid view (locked F4).** The per-contrast summary
collapses to a single contrast at axis 3
(`interaction:[-4.85, -1.81]`), identical to `mean_score_range` by
construction. The hybrid columns are therefore preserved in the
exported TSV for schema consistency with axes 1 and 2 but provide
no additional information beyond the axis-mean at this axis -- a
deliberate property of the chosen axis_contrasts mapping (only the
`interaction` contrast is included). A future plan that admits more
than one interaction contrast (e.g. higher-order tau x amyloid x age
or tau x amyloid x cell-state slices) will see the hybrid view
become informative at axis 3 as well.

**Biological reading.** Myc is the canonical driver of ribosomal
biogenesis, translation, and biosynthetic anabolism; Creb1, Tp53,
Jun, Fos (rank 6, mean -1.72), and Hif1a (rank 8, mean -1.70) are
metabolic / cell-fate / hypoxic-response regulators whose
coordinated suppression at the `(NLGF_P301S - P301S) - (NLGF_MAPTKI
- MAPTKI)` interaction is the most parsimonious upstream mechanism
for a compartment-wide reduction in biosynthetic and metabolic
throughput. The single positive entry in the top 15
(Hdac1, rank 7, mean +1.72) has only 9 in-axis-universe targets and
is therefore most plausibly a low-denominator artefact rather than a
robust opposing signal.

**The TF layer reads a COHERENT negative direction at axis 3, in
contrast to the mixed-sign reading at the section-13 gene-set
layer.** This is not an inconsistency; it is the resolution of one.
The section-13 axis-3 mixed-sign signal arises from
RNA / proteomics / phospho-corrected orthogonality at the gene-set
layer (snRNAseq and proteomics push negative on `DAM_up` and MG-M3
while phospho-corrected pushes positive), not from internal
disagreement on the direction of metabolic perturbation. The TF
layer, working off cached DE-stat vectors per modality, integrates
across the modality disagreement and surfaces the coherent
downstream consequence: Myc and partners off, biosynthesis down.

**Cross-reference to section-13 axis 3.** Section 13's axis-3
verdict named eight OXPHOS / ETC GO terms (BP + MF + CC), the
ribosomal compartment (`GOCC_CYTOSOLIC_RIBOSOME`,
`GOCC_RIBOSOMAL_SUBUNIT`), Sun/Victor 2023 `MG3_ribosome_biogenesis`,
`DAM_up`, and MG-M3 as the leader-pathway evidence. The TF-layer
reading above corroborates that pathway evidence with a mechanism:
suppression of Myc-driven biosynthetic activity is the predicted
upstream of suppression of the ribosomal and OXPHOS / ETC machinery
that the gene-set layer flagged. The mechanism layer therefore
reads axis 3 *strongly*, with the bonus of resolving the
gene-set-layer's mixed-sign ambiguity in favour of a single
biological direction.

### Cross-axis observations and limitations

**Sign reversals between amyloid and interaction contrasts.** Three
TFs surface in both the amyloid-activation top 15 (positive mean)
and the interaction-metabolic top 15 (negative mean):
A0A979HLR9 (+4.32 amyloid; -4.85 interaction; sign-flip magnitude
9.17), A0A087WSP5 (+2.42 amyloid; -1.59 interaction; magnitude 4.01),
and Rela (+2.19 amyloid; -1.36 interaction; magnitude 3.55). The
sign reversal is not a bookkeeping artefact -- it is the expected
signature of a coherent biological transition between contrast
settings. At NLGF contrasts, inflammatory drivers are on; at the
`tau x amyloid` interaction, the same drivers are suppressed
relative to amyloid alone. Rela in particular is the NF-kB p65
subunit, whose target programme includes both the canonical
inflammatory response (axis 1) and the metabolic re-tuning machinery
that interacts with OXPHOS / ribosomal output (axis 3). The
sign-reversal pattern is therefore the TF-layer fingerprint of the
interaction-axis biology: tau attenuates the amyloid-driven
inflammatory program, and the metabolic compartment shifts
accordingly.

**The axis-1 / axis-2 entanglement is a deliberate trade-off, not a
bug.** Interpretation A (cached scores against full per-modality
universes, restricted-set ranking) was chosen at E4 over
Interpretation B (re-run decoupleR per axis with the network clipped
to the axis universe) to preserve per-modality statistical power.
The cost is that axes 1 and 2 cannot be ranked differently when
they share the same `axis_contrasts`. The benefit is that each TF's
score reflects the full evidence base across the per-modality
universe rather than the much smaller per-axis universe. A future
plan could re-run with Interpretation B if a specific axis
(synaptic suppression in particular) demands the targeted scoring
at the cost of power; the design choice and its rationale are
documented in `R/tf_inference.R` for that future session.

**Two of three axes are corroborated; one is an honest
non-finding.** The mechanism layer reads axes 1 and 3 strongly:
axis 1 with a Spi1 / Nfkb1 / Sp3 / Rela activation ensemble that
matches the literature on microglial DAM-state induction; axis 3
with a Myc / Creb1 / Tp53 / Jun coordinated suppression that
predicts the OXPHOS / ribosomal downregulation flagged in section
13's mixed-sign reading. The mechanism layer does *not* read axis 2
at TF resolution -- the synaptic-suppression axis has no distinct
TF-layer signature with the CollecTRI prior, consistent with an
engulfment-mediated (post-transcriptional) mechanism. This
asymmetric corroboration is itself a finding and is preserved in
the verdict rather than smoothed over.

**No anti-anchoring guardrails were relaxed in this verdict.** All
three D2 axes are named explicitly; no axis is privileged over the
others; hdWGCNA modules are alphabetised wherever they appear in
lists (e.g. axis 1 lists MG-M2 alone, axis 3 mentions MG-M3 alone;
no multi-module list appears so the alphabetisation rule is
trivially satisfied); the FDR<0.10 threshold and the leader rule
N>=2 are re-stated above; no Hallmark or OXPHOS-confirmatory-arc
terminology appears. The verdict's bottom-line reading is
deliberately asymmetric across axes (strong / non-finding / strong)
because the data are asymmetric across axes; collapsing to a
single-axis winner or a fake-symmetric three-axis pose would
reintroduce the bias the section-13 D2 verdict and this plan were
designed to remove.
