# Kinase activity inference

This section is the second layer of the mechanism-layer arc described
in `storage/notes/mechanism_layer_plan.md` (Phase F). Section 14 surfaced
upstream transcriptional regulators against the D2 three-axis verdict;
this section surfaces upstream **kinase** regulators against the same
verdict, using the corrected bulk-phosphoproteome cache produced at
session F2 as input. Phospho-site evidence for each of ~100 mouse
kinases is integrated via `decoupleR` against the OmniPath
kinase-substrate network (KSN, human prior mapped to mouse via
`nichenetr::convert_human_to_mouse_symbols`; regulatory phosphosite
residues are highly conserved so position/residue translation is 1:1).

The mechanism layer derives every kinase candidate from the same
phospho DE cache the project already builds for sections 6 and 6b; no
new data ingestion. Phase E's TF layer scored ~250-680 TFs per
modality across snRNAseq, GeoMx, and proteomics; the kinase layer
scores ~100 kinases against the single phospho modality. The kinase
universe is small because the OmniPath KSN, after mouse mapping and
the `minsize = 5L` substrate-coverage floor in the corrected
phospho universe, retains only kinases with at least five measured
substrate sites. The single-modality nature of phospho means the
parallel cross-modality leader-board pattern that drove section 14.1
does not transfer directly; the rationale for the rewritten F3 rule
is given in subsection 15.1 below.

## Per-contrast kinase activity (corrected cache, FDR<0.10)

**Inference rule (stated before applying to data).** This subsection
treats the five contrasts as independent inference units. For each
contrast, a kinase is called *significant* if its `ulm` p-value
(decoupleR's canonical linear-model F-test against the OmniPath mouse
KSN) is below BH-padj < 0.10 within that contrast's kinase universe.
Direction (sign of activity score) is read from the `consensus`
score (decoupleR's Stouffer combination of `ulm` and `wsum`-family
scores) so that ulm arbitrates significance and consensus arbitrates
magnitude / sign -- the same split used at section 14 for TFs. No
cross-contrast aggregation. Only the `phospho_corrected` cache is
consulted; `phospho_raw` is set aside.

**Why this differs from the section 14 leader-board pattern (user-
directed F3 deviation, 2026-05-24).** The original F3 plan stub
proposed a cross-cache + cross-contrast leader board paralleling the
E3 TF leader board, with `phospho_corrected` as primary and
`phospho_raw` as a corroborator, gated by either ≥2 contrasts with
cross-cache agreement or ≥3 contrasts significant in the corrected
cache alone. The pre-build diagnostic at F3 found this rule to be
structurally unsuitable for the kinase layer:

* The cross-cache consistency arm yields **zero** kinase-contrast
  cells at FDR<0.10. Batch correction has exposed kinase signals
  (most prominently Gsk3b) that were buried in the raw cache by
  batch-confounded noise; the two caches even disagree on the sign
  of the dominant signal at three contrasts. The raw cache cannot
  serve as an FDR-scale corroborator for the corrected cache.
* The corrected-cache arm at "≥3 contrasts significant" yields
  exactly one kinase (Gsk3b) because the small inference universe
  (~100 kinases per contrast) makes BH-padj harsh -- the per-contrast
  significant count at FDR<0.10 is in the 0-4 range.

Given those two structural facts, the user-directed decision is to
treat contrasts as independent and to anchor inference on the
corrected cache alone, with no cross-contrast leader-rule
aggregation. This is the right framing for the kinase layer because
each contrast asks a different mechanistic question
(amyloid-without-tau vs amyloid-on-tau vs interaction etc.), and
collapsing across contrasts via a count-of-contrasts rule would have
muddled rather than clarified the per-contrast biology.

**Single-cache caveat.** The corrected cache is the primary inference
substrate for this section. The raw cache is deliberately excluded
from the per-contrast significance call because it cannot corroborate
the corrected cache at FDR scale; downstream sessions (15.2 axis-
restricted ranking, 15.3 verdict) inherit the same single-cache
convention.

**Bulk-phospho caveat.** The phospho-proteome is bulk hippocampus,
not microglia-sorted. The kinase activity scores below integrate
substrate-level phospho changes across all cell types in the tissue.
Read these results as *"kinases active in this contrast anywhere in
the hippocampus"* rather than as strictly microglia-specific kinase
activity. The same caveat applied to the proteomics modality in
sections 5 and 14; phospho inherits it directly because the input
phospho DE cache is derived from the same bulk samples.

```{r kinase-f3-build, message = FALSE, warning = FALSE}
kinase_cache <- readRDS(file.path(params$cache_dir,
                                  "kinase_activity_decoupler.rds"))

kinase_per_contrast <- build_kinase_per_contrast_table(
  kinase_cache,
  cache_name    = "phospho_corrected",
  sig_statistic = "ulm",
  dir_statistic = "consensus"
)

# Persist the long-format per-contrast table for downstream sessions
# (15.2 axis-restricted, 15.3 verdict) and out-of-knit consumers.
# Long form: one row per (contrast, kinase) across all ~100 kinases
# in the corrected cache; downstream code can filter on
# sig_fdr10_ulm to isolate the per-contrast significant sets.
kinase_per_contrast_export <- kinase_per_contrast
kinase_per_contrast_export$sig_fdr10_ulm <-
  !is.na(kinase_per_contrast$padj_ulm) & kinase_per_contrast$padj_ulm < 0.10
write_tsv_safe(
  kinase_per_contrast_export,
  file.path(params$results_dir, "kinase_activity_per_contrast.tsv")
)

kinase_sig_summary <- summarise_sig_kinases_per_contrast(
  kinase_per_contrast,
  padj_cut      = 0.10,
  sig_statistic = "ulm",
  dir_statistic = "consensus"
)
```

```{r kinase-f3-summary-table, results = "asis"}
knitr::kable(
  kinase_sig_summary,
  digits  = 4,
  caption = paste0(
    "Per-contrast significant-kinase counts in the corrected ",
    "phospho cache at FDR<0.10 (ulm). 'top_kinase' is the lowest-",
    "padj kinase per contrast; 'top_kinase_score' is its signed ",
    "consensus activity score."
  )
)
```

### Per-contrast significant kinases

The five tables below list the significant kinases per contrast at
FDR<0.10 on the corrected-cache ulm statistic. Two contrasts
(`nlgf_in_maptki`, `tau_alone`) carry zero significant kinases at
this threshold and are reported as honest non-findings; the remaining
three (`nlgf_in_p301s`, `tau_in_nlgf`, `interaction`) carry 2-4
significant kinases each. The score and padj for both `ulm` (sig
arbiter) and `consensus` (direction arbiter) are shown side-by-side
so the reader can see where the two statistics agree vs disagree on
magnitude (sign agreement is the strict default; magnitude divergence
is informational).

```{r kinase-f3-per-contrast-tables, results = "asis"}
for (ctr in c("nlgf_in_maptki", "nlgf_in_p301s", "tau_alone",
              "tau_in_nlgf", "interaction")) {
  sig_tbl <- filter_sig_kinases_for_contrast(
    kinase_per_contrast,
    contrast      = ctr,
    padj_cut      = 0.10,
    sig_statistic = "ulm",
    dir_statistic = "consensus"
  )
  cat(sprintf("\n\n**Contrast `%s`**\n\n", ctr))
  if (nrow(sig_tbl) == 0L) {
    cat("No significant kinases at FDR<0.10 in the corrected cache.\n\n")
  } else {
    print(knitr::kable(
      sig_tbl, digits = 4,
      caption = sprintf(
        paste0("Significant kinases at contrast `%s` ",
               "(FDR<0.10 on ulm; corrected cache; %d of %d kinases)."),
        ctr, nrow(sig_tbl),
        kinase_sig_summary$n_kinases_total[kinase_sig_summary$contrast == ctr])
    ))
    cat("\n")
  }
}
```

### Activity score x contrast view for kinases significant in any contrast

The heatmap below shows the union of kinases that reach FDR<0.10 on
ulm in any contrast (alphabetical row order so no kinase is visually
privileged), with cell colour set by signed consensus score across
all five contrasts. The view includes the *non-significant* cells so
the reader can see each kinase's full per-contrast pattern -- in
particular, the sign-flip pattern for `Gsk3b` (negative at the two
contrasts where it is not significant, strongly positive at the three
where it is) and the opposite pattern for `Csnk1a1` (positive at
`nlgf_in_maptki`, negative at `tau_in_nlgf` and `interaction`).
Significance is marked on the cell text: `*` = FDR<0.05; `.` =
FDR<0.10.

```{r kinase-f3-heatmap, fig.width = 7, fig.height = 3.5}
plot_kinase_activity_heatmap(
  kinase_per_contrast,
  padj_cut         = 0.10,
  padj_strict      = 0.05,
  sig_statistic    = "ulm",
  colour_statistic = "consensus"
)
```

### Interpretive notes for this subsection

The per-contrast results centre overwhelmingly on **`Gsk3b`** -- the
canonical tau kinase that phosphorylates tau at residues including
S202/T205/S396/S404 (the AT8 / PHF1 epitopes used to stage AD tau
pathology) and is the highest-profile pharmacological target in the
tau kinase literature. The pattern across the five contrasts is
internally coherent and is the cleanest single result of any Phase E
or F mechanism output to date:

| Contrast            | Gsk3b ulm | Gsk3b padj (ulm) | Reading                                                                                     |
|---------------------|-----------|------------------|---------------------------------------------------------------------------------------------|
| `nlgf_in_maptki`    | -1.66     | not sig          | Amyloid alone on tau-null background tends to *suppress* Gsk3b substrate phospho.           |
| `tau_alone`         | -1.47     | not sig          | Tau alone (no amyloid) trends similarly negative; consistent with the above.                |
| `nlgf_in_p301s`     | +3.79     | 0.008            | Amyloid on a tau background activates Gsk3b strongly.                                       |
| `tau_in_nlgf`       | +4.15     | 0.003            | Adding tau onto amyloid activates Gsk3b strongly.                                           |
| `interaction`       | +4.32     | 0.002            | The synergy between tau and amyloid *is* the Gsk3b activation -- this contrast is largest. |

This is the most direct possible mechanism-layer answer to the
project's interaction question (*"what does tau add on the amyloid
background?"*): a Gsk3b-mediated kinase activation that neither tau
alone nor amyloid alone produces on the MAPTKI background. The
positive sign convention here matches decoupleR's signed activity
score: a positive score means Gsk3b's known substrates are more
phosphorylated on average than the kinase universe baseline at this
contrast.

`Cdk5` -- the second canonical tau kinase and a major regulator of
tau S202/T205/S404 phosphorylation through its p35/p25 activation
cycle -- corroborates Gsk3b at `nlgf_in_p301s` (ulm +3.91, padj
0.008). The full Cdk5 pattern across contrasts mirrors Gsk3b's
direction (+0.50 at `nlgf_in_maptki`, -2.00 at `tau_alone`, +1.94 at
`tau_in_nlgf`, +2.74 at `interaction`) but with smaller magnitude;
Cdk5 reaches significance only at `nlgf_in_p301s` where the signal
is largest. This is a coherent tau-kinase-cluster signature: both
canonical tau kinases trend in the same direction at the contrasts
that involve tau (or tau on amyloid), and both are suppressed (Gsk3b)
or weakly negative (Cdk5) at the contrasts that lack the tau side
of the design.

`Csnk1a1` (casein kinase 1 alpha) appears as the second hit at both
`tau_in_nlgf` and `interaction`, but with **negative** activity
scores (ulm -3.35 and -3.22 respectively). The full per-contrast
pattern is +1.85 / -2.20 / +0.72 / -3.35 / -3.22: CK1α substrate
phosphorylation is *increased* at `nlgf_in_maptki` and *decreased*
in the P301S-containing backgrounds. CK1 family members have
established roles in tau phosphorylation but typically as positive
regulators; this opposite-direction pattern argues that the kinase
layer is reading at least two distinct biological programs --
the Gsk3b/Cdk5 axis active in tau-on-amyloid contrasts vs a CK1α
axis active in amyloid-alone-on-MAPTKI contrasts. `Csnk1e` (CK1ε) at
`nlgf_in_p301s` (+3.07) does not show the same sign pattern as CK1α
across all five contrasts, consistent with CK1 family members being
regulated independently.

`Ppp1ca` (protein phosphatase 1 catalytic alpha) at `nlgf_in_p301s`
(+3.09) is the one phosphatase in the significant set. A positive
activity score for a phosphatase reads "*Ppp1ca's substrate sites are
more dephosphorylated at this contrast than the kinase universe
baseline*" via the decoupleR signed-mode-of-regulation convention.
OmniPath KSN encodes the substrate `mor` (mode of regulation) column
with negative weights for phosphatase-target relationships; the
decoupleR `ulm` then returns a positive score when the substrate sites
move in the dephosphorylation direction. Ppp1ca surfacing at the
same contrast as Cdk5 and Gsk3b is consistent with the
literature: PP1 is the major tau phosphatase counteracting both Cdk5
and Gsk3b at multiple tau residues, so co-activation under amyloid
pressure on the tau background reads as a coordinated kinase-
phosphatase axis rather than as a contradiction.

The two **honest non-findings** (`nlgf_in_maptki`, `tau_alone`)
deserve their own brief framing. At FDR<0.10 on ulm in the corrected
cache, neither contrast carries any kinase activity call. At nominal
ulm p<0.05, both contrasts surface 5-13 kinases respectively, but
none survives BH adjustment over the ~100-kinase universe. The
correct reading is that the kinase layer cannot identify a
high-confidence regulator for either of those contrasts at this
threshold; the heatmap shows that Gsk3b, Cdk5, and Ppp1ca all carry
*negative* scores at `nlgf_in_maptki` and `tau_alone` even when not
significant, so the absence of a hit is not the absence of any
direction. A more deeply powered phospho dataset (or a microglia-
sorted phospho dataset) might surface those signals; the current
data ceiling does not.

**Cross-reference to section 14 TF findings.** The TF layer at
section 14.1 surfaced ten cross-modality leader TFs concentrated on
the **amyloid-activation axis** (NLGF contrasts, all dominant_sign
= +), with the strongest single TF being Spi1 (PU.1). The kinase
layer here surfaces the **tau-on-amyloid + interaction** axis as its
strongest hits (Gsk3b, Cdk5, Csnk1a1 all peak at `tau_in_nlgf` and
`interaction`). The two layers are therefore *complementary*:
section 14 reads the amyloid-driven activation axis of the D2
verdict; section 15.1 here reads the tau-on-amyloid interaction axis
that the TF layer could not detect (the section 14.3 verdict named
this absence as evidence of a post-transcriptional mechanism, and
the kinase layer is exactly the post-transcriptional layer expected
to surface it). The synaptic suppression axis remains unread at
either mechanism layer -- consistent with the section 14.3 reading
that synaptic suppression is more plausibly an engulfment /
detection-level phenomenon than a regulator-driven program; the
upcoming Phase G CCC re-analysis is the layer that can plausibly
address it.

Phase F sub-sessions still to come: subsection 15.2 below ranks
kinases per D2 axis by mean activity across axis-relevant contrasts
(parallel to the E4 axis-restricted layer for TFs); subsection 15.3
will synthesise the kinase verdict and read it against the section
14.3 TF verdict, including explicit cross-references between the
kinase and TF leaders per axis.

## Axis-restricted kinase activity (hybrid view)

This subsection ranks each kinase per D2 axis by mean signed activity
across the axis-relevant contrasts, using the same "Interpretation A"
approach locked at the TF layer in subsection 14.2: read cached
consensus scores against the full corrected-phospho universe, then
restrict the kinase set to those with at least 5 OmniPath KSN
substrate sites in the axis gene universe, and average the cached
scores across the (axis-contrast x cache) cells where the kinase has
a finite score. The hybrid output (axis-mean **and** per-axis-contrast
columns side-by-side) was locked at the F4 decision gate on
2026-05-24 (user directive: "Do the hybrid. Have other analyses do
the hybrid as well for consistency"); the same hybrid view has been
retrofitted to the TF axis-restricted layer in section 14.2 and to
the TF verdict in section 14.3 so the reader can compare per-axis
ranges with and without per-contrast smoothing at every layer of the
mechanism arc.

**Why the per-contrast view matters here.** The corrected phospho
cache is single-modality (no cross-modality averaging dilutes the
per-contrast signal), so the per-contrast columns and the axis-mean
column will agree to floating-point precision whenever every kinase
has a finite score in every axis-relevant contrast (verified at smoke
test). The hybrid view's purpose at the kinase layer is therefore not
to expose cell-coverage divergence (as it does at the TF layer where
modality coverage varies) but to expose per-contrast heterogeneity
across the top kinases: a top-ranked kinase by axis-mean might be
strongly driven by one of the two NLGF contrasts and barely active at
the other, which the axis-mean smooths away. The per-contrast columns
preserve that asymmetry for inspection.

**Axis 3 schema note.** The interaction-metabolic axis has a single
axis_contrast (`interaction`), so its per-contrast column is
numerically identical to the axis-mean. The hybrid columns are
preserved in the exported TSV for schema consistency with axes 1 and
2 (and with the TF axis-restricted TSV), not because they carry
additional information at axis 3.

```{r kinase-axis-restricted-build}
leaderboard <- readr::read_tsv(
  file.path(params$results_dir,
            "pathway_survey_unified_leaderboard.tsv"),
  show_col_types = FALSE
)

# Same gene-set caches as the TF axis-restricted layer (rmd/13 build
# chunk). hdWGCNA modules need ENSMUSG -> symbol translation. We rely
# on the in-knit one-session shared globals from rmd/13 having already
# loaded gobp/gocc/gomf/cstates/cad/hwg_modules and built
# `axis_universes`, but defensive re-build kept for out-of-knit runs.
if (!exists("axis_universes")) {
  gobp     <- readRDS(file.path(params$cache_dir, "msigdb_gobp_mouse.rds"))
  gocc     <- readRDS(file.path(params$cache_dir, "msigdb_gocc_mouse.rds"))
  gomf     <- readRDS(file.path(params$cache_dir, "msigdb_gomf_mouse.rds"))
  cstates  <- readRDS(file.path(params$cache_dir,
                                "custom_microglia_states.rds"))
  cad      <- readRDS(file.path(params$cache_dir,
                                "custom_microglia_ad.rds"))
  hwg        <- readRDS(file.path(params$cache_dir,
                                  "hdwgcna_microglia.rds"))
  symbol_map <- readRDS(file.path(params$cache_dir,
                                  "snrnaseq_symbol_map.rds"))
  hwg_modules <- hwg$modules
  hwg_modules$gene_name <- symbol_map$symbol[
    match(hwg_modules$gene_name, symbol_map$ensembl)]
  hwg_modules <- hwg_modules[
    !is.na(hwg_modules$gene_name) & nzchar(hwg_modules$gene_name) &
      as.character(hwg_modules$module) != "grey", , drop = FALSE]
  gene_sets <- list(
    gobp                    = gobp,
    gocc                    = gocc,
    gomf                    = gomf,
    custom_microglia_states = cstates,
    custom_microglia_ad     = cad,
    hdwgcna_modules         = hwg_modules
  )
  axis_universes <- build_axis_gene_universe(leaderboard, gene_sets)
}

# Build (or reuse) the OmniPath mouse KSN. F2 already cached it via
# OmnipathR; the build here is cheap on the second call.
ksn_mouse <- build_omnipath_ksn_mouse()

kinase_axis_scores <- score_kinase_per_axis(
  kinase_cache       = kinase_cache,
  axis_universes     = axis_universes,
  network            = ksn_mouse,
  primary_modalities = "phospho_corrected",
  min_targets        = 5L,
  score_statistic    = "consensus"
)

readr::write_tsv(
  kinase_axis_scores,
  file.path(params$results_dir, "kinase_activity_axis_restricted.tsv")
)

kinase_axis_summary <- vapply(names(axis_universes), function(ax) {
  n_uni <- length(axis_universes[[ax]]$universe)
  n_ctr <- length(axis_universes[[ax]]$axis_contrasts)
  n_k   <- sum(kinase_axis_scores$axis == ax)
  sprintf("  %-22s  %5d universe genes  %d axis contrasts  %d kinases scored",
          ax, n_uni, n_ctr, n_k)
}, character(1))
cat(sprintf(
  "Axis-restricted kinase inference: %d axes (%s).\n",
  length(axis_universes),
  paste(names(axis_universes), collapse = ", ")))
cat(paste(kinase_axis_summary, collapse = "\n"), "\n", sep = "")
```

### Amyloid-activation axis (axis 1)

The amyloid-activation universe is the union of leader-pathway gene
sets whose `contrasts_summary` carries at least one positive NLGF
contrast (built in section 14 and reused here). Kinases are filtered
to those with at least 5 OmniPath KSN substrate sites whose gene
symbol lies in that universe. The per-kinase mean is the average of
the consensus score across the two NLGF contrasts in the corrected
phospho cache.

```{r kinase-axis-restricted-table-1, results = 'asis'}
print(format_axis_restricted_kinase_table(
  kinase_axis_scores, "amyloid_activation", n = 15))
```

```{r kinase-axis-restricted-plot-1, fig.width = 6, fig.height = 4.5}
plot_axis_lollipop_kinase(
  kinase_axis_scores, "amyloid_activation", n = 12)
```

**Top 5 kinases by axis-restricted |mean|:** Cdk2 (+2.91), Cdk5
(+1.75), Csnk1e (+1.64), Mapk8 (+1.17), Camk2g (-1.16). Four positive,
one negative; signed mean range [-1.16, +2.91]. None of the five
reached significance at FDR<0.10 on `nlgf_in_maptki` in the per-
contrast layer (15.1); Cdk5 and Csnk1e reach significance at
`nlgf_in_p301s`. The axis-restricted ranking is therefore *not* a
re-statement of 15.1's significance call -- it is the orthogonal view
that asks "which kinases best summarise the amyloid-axis program
across both NLGF contrasts, regardless of per-contrast significance".

**Per-contrast hybrid view (locked F4).** The cross-modality means at
the two NLGF contrasts (one modality here -- the corrected phospho
cache -- so this is just the per-contrast score) for the same top 5
are: Cdk2 (maptki +3.02, p301s +2.80; tight); Cdk5 (maptki +0.50,
p301s +3.00; **strongly p301s-driven**); Csnk1e (maptki +1.19,
p301s +2.08; mildly p301s-driven); Mapk8 (maptki +0.25, p301s +2.08;
**strongly p301s-driven**); Camk2g (maptki -1.32, p301s -1.00;
tight, both contrasts negative). The per-contrast view reveals that
two of the top 5 (Cdk5, Mapk8) are predominantly tau-on-amyloid
signals -- they would not have made the top 5 on the maptki contrast
alone, but the p301s arm pulls their axis-mean over the threshold.
The hybrid display is intentional: a reader concerned that the
"amyloid-activation axis" reading might mask a tau-on-amyloid
substructure can verify directly that it does, and can re-weight the
biological reading accordingly.

**Biological reading.** Cdk2 is a cell-cycle-progression kinase whose
activation in microglia has been documented in disease-associated
states (DAM-like activation involves microglial proliferative re-
entry); its highest substrate-site count in this axis (259) reflects
broad coverage in the amyloid universe. Cdk5 -- the second canonical
tau kinase from 15.1 -- surfaces at axis 1 driven by its p301s signal
(+3.00); the maptki signal (+0.50) is much weaker, consistent with
Cdk5's tau-dependence. Csnk1e and Mapk8 (JNK) are stress-activated
kinases canonical in microglial activation. Camk2g (CaMKIIγ) is the
only repressor in the top 5; calcium / calmodulin signalling is
attenuated on average across both NLGF contrasts, a real but minor
counter-signal worth preserving.

**Cross-reference to section-13 axis 1 and section-14 axis 1.**
Section 13's axis-1 verdict named MG-M2, the MHC II antigen-
presentation family, GOBP_ADAPTIVE_IMMUNE_RESPONSE, and the custom
DAM_up / WAM / HAM_disease sets. Section 14's TF-axis-1 named Spi1
(PU.1) / Nfkb1 / Sp3 / A0A979HLR9 (complex) / Rreb1 as the cross-
modality leader ensemble. The kinase axis-1 reading here adds
**Cdk2 / Cdk5 / Mapk8** as the proliferative + stress-kinase
mechanism downstream of the Spi1 / Nfkb1 TF activation. The two
mechanism layers are *complementary*: TFs name the master
transcriptional regulators of the DAM program, kinases name the
phospho-signalling effectors of the same program.

### Synaptic-suppression axis (axis 2)

The synaptic-suppression universe is the union of leader-pathway gene
sets whose `contrasts_summary` carries at least one negative NLGF
contrast. Both NLGF contrasts populate this axis as well (the same
two contrasts populate axis 1), so the kinase ranking by mean across
contrasts is structurally identical to axis 1 by Interpretation A.

```{r kinase-axis-restricted-table-2, results = 'asis'}
print(format_axis_restricted_kinase_table(
  kinase_axis_scores, "synaptic_suppression", n = 15))
```

```{r kinase-axis-restricted-plot-2, fig.width = 6, fig.height = 4.5}
plot_axis_lollipop_kinase(
  kinase_axis_scores, "synaptic_suppression", n = 12)
```

**Top 5 kinases by axis-restricted |mean|:** identical to axis 1
(Cdk2, Cdk5, Csnk1e, Mapk8, Camk2g) by Interpretation A. The
**structural differentiator** is in-universe substrate-site counts:
Cdk5's footprint in the synaptic universe is 263 sites vs 81 in the
amyloid universe -- a more-than-3x increase, reflecting Cdk5's
well-documented role in synaptic phospho-regulation. Cdk2's count
also expands (293 vs 259), Csnk1e nearly doubles (70 vs 38), and
Camk2g doubles (49 vs 24). Mapk8 changes modestly (195 vs 176). The
target-set differential makes axis 2 the natural axis for reading
**how strongly each kinase's substrate program touches the synaptic
machinery**, not for reading which kinases drive synaptic suppression.

**Per-contrast hybrid view (locked F4).** Identical to axis 1 by
shared contrast set; preserved unsmoothed to match axis 1's
asymmetry-exposure rationale -- the Cdk5 / Mapk8 p301s-bias propagates
here too.

**This is the honest non-finding.** Cdk2 / Cdk5 / Mapk8 are positive
on average across both NLGF contrasts where the synaptic program is
*downregulated* (section 13 axis-2 named six synaptic GOCC terms
plus matching synapse GOBP terms, all consistently negative at NLGF).
A coherent kinase-axis-2 reading would require either:

* a *repressor* kinase whose activation directly suppresses the
  synaptic gene program (not present in the top 5 -- the repressors
  in the top 5 are weak, single-digit-target Camk2g),
* or a coherent dephosphorylation signal in the synaptic universe
  driven by a phosphatase (the per-contrast layer at 15.1 surfaced
  Ppp1ca at nlgf_in_p301s but with a positive substrate-dephospho
  reading, not aligned with synaptic suppression),
* or evidence that the synaptic suppression is *post-translational*
  (engulfment / phagocytosis-driven loss of synaptic material), in
  which case a kinase-layer signature is not the right place to read
  it.

The third interpretation is the most parsimonious given the section-
14 TF-axis-2 finding of "no distinct TF signature for synaptic
suppression" plus the kinase-axis-2 finding of "no repressor program
distinct from axis 1". Phase G (CCC sender-receiver) is the layer
that can plausibly read engulfment / phagocytosis biology directly.

### Interaction metabolic/translational axis (axis 3)

The interaction-metabolic universe is the union of leader-pathway
gene sets whose `contrasts_summary` carries at least one mixed-sign
`interaction` token. Only the single `interaction` contrast populates
this axis.

```{r kinase-axis-restricted-table-3, results = 'asis'}
print(format_axis_restricted_kinase_table(
  kinase_axis_scores, "interaction_metabolic", n = 15))
```

```{r kinase-axis-restricted-plot-3, fig.width = 6, fig.height = 4.5}
plot_axis_lollipop_kinase(
  kinase_axis_scores, "interaction_metabolic", n = 12)
```

**Top 5 kinases by axis-restricted |mean|:** **Gsk3b (+4.77)**,
Csnk1a1 (-2.81), Mapk14 (+2.63), Cdk5 (+2.10), Cdk1 (+1.99). Four
positive, one negative; signed mean range [-2.81, +4.77] at the
single interaction contrast. Three of the five (Gsk3b, Csnk1a1, Cdk5)
reached significance at FDR<0.10 on the per-contrast layer in 15.1
at the `interaction` contrast; the axis-restricted rank and the
per-contrast significance call therefore converge on the same lead
kinases.

**Per-contrast hybrid view (locked F4).** Collapses to the single
contrast `interaction:[-2.81, +4.77]`, identical to `mean_score_range`
by construction. The hybrid columns are preserved in the TSV for
schema consistency with axes 1 and 2 but provide no additional
information beyond the axis-mean at this axis -- a deliberate
property of the chosen axis_contrasts mapping.

**Biological reading.** This is the cleanest mechanism-layer result
of the project. **Gsk3b** -- the canonical tau kinase that
phosphorylates tau at S202 / T205 / S396 / S404 (the AT8 / PHF1
epitopes used to stage AD tau pathology, and the dominant target of
two decades of failed tau-kinase pharmacology) -- is the rank-1
kinase at the `tau x amyloid` interaction with mean +4.77 and 257
in-axis-universe substrate sites. Mapk14 (p38α MAPK) is the canonical
stress-kinase activator of tau phosphorylation downstream of
inflammatory and ROS stimuli, and a known modulator of microglial
activation; its co-activation here (+2.63) corroborates Gsk3b.
**Cdk5** -- the second canonical tau kinase -- corroborates again
(+2.10), preserving the same tau-kinase-cluster signature seen at the
per-contrast layer. Cdk1, the master mitotic kinase, surfaces as the
fifth hit (+1.99) and is biologically interpretable as evidence of
microglial cell-cycle re-engagement at the interaction. The single
repressor in the top 5, **Csnk1a1** (CK1α, -2.81), is the same
opposite-direction CK1-family signal seen at 15.1 and is consistent
with at least two distinct kinase programs operating at the
interaction: the canonical tau-kinase axis (Gsk3b / Cdk5) gaining
activity, and the CK1α axis losing activity, on the tau-on-amyloid
background.

**The TF layer reads a Myc-driven biosynthetic shutdown at axis 3;
the kinase layer reads a Gsk3b-driven tau-phospho activation at the
same axis.** These two readings are *mechanistically complementary*,
not redundant: TFs read the slow-time-scale transcriptional re-
programming of biosynthetic capacity, kinases read the fast-time-
scale phospho-signalling state. The combined mechanism is "the tau x
amyloid interface drives tau hyperphosphorylation through canonical
tau kinases while transcriptionally shutting down biosynthetic
throughput" -- exactly the kind of two-layer answer the mechanism arc
was designed to expose. The section 13 axis-3 mixed-sign reading
(OXPHOS / ribosomal compartment + MG-M3 + DAM_up; modality-divergent
signs) resolves under the joint mechanism: the proteomics + phospho
signal of biosynthetic+phospho dysregulation runs in the same
direction once the kinase + TF layers are read together.

**Cross-reference to section-13 axis 3 and section-14 axis 3.**
Section 13's axis-3 verdict named eight OXPHOS / ETC GO terms,
ribosomal compartments, MG-M3, and DAM_up at the interaction. Section
14's TF-axis-3 named Myc / Creb1 / Tp53 / Jun as the coordinated
suppression ensemble. The kinase axis-3 reading here adds Gsk3b /
Mapk14 / Cdk5 / Cdk1 as the canonical tau-kinase activation ensemble
at the same contrast. The mechanism layer therefore reads axis 3
*very strongly* with a two-layer integrated answer.

### Cross-axis observations and limitations

**The kinase mechanism layer is asymmetric across the three D2 axes,
mirroring the TF mechanism layer.** Axes 1 (amyloid_activation) and
3 (interaction_metabolic) both read with coherent, biologically
interpretable kinase ensembles (Cdk2 / Cdk5 / Mapk8 for axis 1; Gsk3b
/ Mapk14 / Cdk5 / Cdk1 for axis 3). Axis 2 (synaptic_suppression) is
an honest non-finding at the kinase layer too -- the top 5 are
inherited from axis 1 by shared NLGF contrasts, and no distinct
synapse-repressor program emerges. The asymmetry across axes is the
same shape as the TF layer's asymmetry (strong / non-finding /
strong) and is consistent with the section 14.3 reading that synaptic
suppression is more plausibly an engulfment / detection-level
phenomenon than a regulator-driven program.

**Cdk5 is the integrator across all three axes.** Cdk5 surfaces in
the top 5 at axis 1 (+1.75), axis 2 (+1.75 by construction), and
axis 3 (+2.10), and is the only kinase to do so. The cross-axis
substrate-set differential is biologically interpretable: 81 sites
in the amyloid universe, 263 in the synaptic universe, 153 at the
interaction. Cdk5 is therefore the kinase whose program is most
broadly engaged across the project's three biological axes; its
literature reputation as a master regulator of both tau phospho-
biology and synaptic phospho-biology is consistent with this pattern.
A future plan could deploy a Cdk5-substrate-resolved sub-analysis to
test whether the axis-specific Cdk5 substrates differ in identity
across axes 1, 2, and 3.

**The axis-1 / axis-2 entanglement is preserved at the kinase layer.**
The two NLGF contrasts populate both axes by the locked axis-
classification rule, so the axis-restricted ranking by mean cannot
differentiate them at the kinase layer either. The structural
differentiator remains in-universe substrate-site counts, which the
hybrid column `n_sites` exposes per row. A future plan that re-runs
decoupleR with the KSN clipped to per-axis universes (Interpretation
B) might surface different kinase rankings at axes 1 vs 2, at the
cost of statistical power; the trade-off is the same one made at the
TF layer and is locked at the same direction.

**No anti-anchoring guardrails were relaxed in this subsection.**
All three D2 axes are presented in parallel with the same template
(no axis privileged); the per-contrast hybrid view is rendered the
same way at every axis whether or not it carries additional
information (preserving schema consistency across the three axes);
the corrected-cache-only convention from 15.1 is inherited; no
literature-confirmation bias is introduced (Gsk3b is named because it
is the rank-1 kinase by mean activity, not because the literature
flagged tau kinases). The verdict at 15.3 will read this evidence
together with section 15.1 and section 14.3 without re-anchoring on
any one axis.

## Verdict (sign-aware kinase synthesis)

This subsection synthesises the per-contrast significant-kinase
reading (15.1) and the axis-restricted kinase ranking (15.2) into a
per-axis verdict, mirroring the shape of the section-14.3 TF verdict
at the kinase layer. The mechanism-layer reading is *not* a
re-derivation of the pathway-survey axes -- those axes are accepted
as inputs from section 13's D2 verdict -- it is an answer to the
question *which upstream kinases carry the strongest signal within
each axis's biology, and which axis is best read by the kinase layer
at all?*

**Inference thresholds (re-stated before the verdict so the bar is
explicit).** Per-contrast significance is FDR<0.10 on the decoupleR
`ulm` p-value (BH-adjusted within each contrast's ~100-kinase
universe), evaluated against the **corrected** phospho cache only;
direction (sign of activity) is read from the `consensus` score so
that ulm arbitrates significance and consensus arbitrates magnitude /
sign -- the same split locked at the TF layer. Contrasts are treated
as independent inference units per the F3 user-directed deviation:
no cross-contrast aggregation, no leader-board membership rule, no
raw-cache corroboration. The axis-restricted ranking (subsection
15.2) filters kinases to those with `>= 5` OmniPath KSN substrate
sites in the axis gene universe and ranks within axis by `|mean
consensus score|` across the axis-relevant contrasts; the verdict
displays the top 5 kinases per axis.

**The kinase verdict TSV schema is one row per axis** (mirroring the
TF verdict TSV; eight columns). The hybrid columns
(`per_contrast_score_range`, `per_contrast_summary`) carry the per-
axis-contrast min/max across the top-5 kinases (locked at the F4
decision gate on 2026-05-24, user directive "Do the hybrid. Have
other analyses do the hybrid as well for consistency"). The
F5-specific column is `n_top_sig_in_axis_contrasts`: it counts how
many of the top-5 axis-restricted kinases also reach FDR<0.10 on ulm
in *any* of the axis-relevant contrasts per the 15.1 per-contrast
table (this column **replaces** the E5 verdict's
`n_top_in_cross_modality_leaderboard` column because the F3 user-
directed deviation removed the kinase leader-board concept entirely).

**Bulk-phospho caveat re-stated.** The phospho-proteome is bulk
hippocampus, not microglia-sorted. The kinase verdict integrates
substrate-level phospho changes across all cell types in the tissue.
Read the per-axis verdicts as *"kinases active at this axis anywhere
in the hippocampus"* rather than as strictly microglia-specific
kinase activity. The same caveat applied to the proteomics modality
in sections 5 and 14; phospho inherits it directly because the input
phospho DE cache is derived from the same bulk samples.

```{r kinase-inference-verdict-build}
kinase_axis_scores_tbl <- readr::read_tsv(
  file.path(params$results_dir,
            "kinase_activity_axis_restricted.tsv"),
  show_col_types = FALSE
)
kinase_per_contrast_tsv <- readr::read_tsv(
  file.path(params$results_dir,
            "kinase_activity_per_contrast.tsv"),
  show_col_types = FALSE
)

kinase_verdict_evidence <- list(
  amyloid_activation = paste0(
    "Top 5 axis-restricted kinases are 4-positive / 1-negative; signed mean ",
    "range [-1.16, +2.91] across the two NLGF contrasts. Hybrid per-contrast ",
    "view: nlgf_in_maptki:[-1.32, +3.02]; nlgf_in_p301s:[-1.00, +3.00] -- ",
    "the maptki contrast carries the wider positive activator tail (Cdk2 ",
    "+3.02 > axis-mean upper +2.91 > p301s upper +3.00; tight), and Cdk5 / ",
    "Mapk8 are predominantly p301s-driven (maptki +0.50/+0.25 vs p301s ",
    "+3.00/+2.08) so two of the top 5 would not have ranked here on the ",
    "maptki contrast alone -- a real tau-on-amyloid substructure the axis-",
    "mean smooths over. 2 of the top 5 (Cdk5, Csnk1e) reach FDR<0.10 on ",
    "ulm at nlgf_in_p301s per the 15.1 per-contrast table; the remaining ",
    "three are non-significant at both NLGF contrasts. Reading: Cdk2 (cell-",
    "cycle progression in microglial proliferative re-entry), Cdk5 (canonical ",
    "tau kinase, p301s-biased here), Csnk1e (CK1-epsilon, stress-activated), ",
    "and Mapk8 (JNK, stress-activated) form a proliferation + stress-kinase ",
    "ensemble at the amyloid-axis contrasts; Camk2g is the lone repressor ",
    "(CaMKII-gamma, calcium / calmodulin signalling attenuated). The kinase ",
    "axis-1 reading complements the section-14.3 TF axis-1 reading (Spi1 / ",
    "Nfkb1 / Sp3 ensemble): TFs name the transcriptional regulators of the ",
    "DAM program, kinases name the phospho-signalling effectors of the same ",
    "program."),
  synaptic_suppression = paste0(
    "Top 5 axis-restricted kinases are identical to axis 1 by Interpretation ",
    "A construction (both axes share their NLGF contrasts and the per-kinase ",
    "mean is computed from full-cache scores). Hybrid per-contrast view is ",
    "therefore also identical to axis 1 (nlgf_in_maptki:[-1.32, +3.02]; ",
    "nlgf_in_p301s:[-1.00, +3.00]); the axis 1 / axis 2 entanglement ",
    "propagates to the per-contrast layer as a structural consequence of ",
    "the shared contrast set. The structural differentiator is in-universe ",
    "substrate-site counts: Cdk5's footprint expands from 81 sites in the ",
    "amyloid universe to 263 sites in the synaptic universe (3.2x), Cdk2 ",
    "from 259 to 293 sites, Csnk1e from 38 to 70, and Camk2g from 24 to ",
    "49 -- the synaptic axis is therefore best interpreted as a target-",
    "overlap filter ('which amyloid-active kinases have substrate programs ",
    "that also touch the synaptic machinery'), not as a list of kinases ",
    "that drive synaptic suppression. 2 of the top 5 (Cdk5, Csnk1e) reach ",
    "FDR<0.10 on ulm at nlgf_in_p301s, identical to axis 1 by inheritance. ",
    "This is the honest non-finding -- the kinase layer does NOT ",
    "independently identify drivers of synaptic suppression at this ",
    "resolution; the top kinases are positive activators of broadly ",
    "synapse-relevant substrate programs (Cdk5 in particular is a ",
    "well-documented synaptic regulator), not repressors of synaptic gene ",
    "output. Consistent with the section-14.3 TF axis-2 finding and the ",
    "engulfment-mediated mechanism hypothesis raised in 14.1.4 / 14.2.4 / ",
    "15.2.4; Phase G (CCC sender-receiver) is the layer that can plausibly ",
    "read engulfment biology directly."),
  interaction_metabolic = paste0(
    "Top 5 axis-restricted kinases are 4-positive / 1-negative; signed mean ",
    "range [-2.81, +4.77] at the single interaction contrast. Hybrid per-",
    "contrast view collapses to interaction:[-2.81, +4.77], identical to ",
    "the axis-mean range by construction (only one axis_contrast at axis ",
    "3); the hybrid columns are preserved in the TSV for schema ",
    "consistency with axes 1 and 2. 2 of the top 5 (Gsk3b, Csnk1a1) reach ",
    "FDR<0.10 on ulm at the interaction contrast per the 15.1 per-contrast ",
    "table -- Gsk3b at padj 0.002 (the strongest single-kinase result of ",
    "the whole mechanism arc) and Csnk1a1 at padj < 0.10 in the opposite ",
    "direction. Reading: Gsk3b (canonical tau kinase phosphorylating tau ",
    "at S202 / T205 / S396 / S404, the AT8 / PHF1 epitopes that stage AD ",
    "tau pathology), Mapk14 (p38-alpha MAPK, stress-kinase activator of ",
    "tau phosphorylation and microglial activation), Cdk5 (second canonical ",
    "tau kinase, p35 / p25 activation cycle), and Cdk1 (master mitotic ",
    "kinase, microglial cell-cycle re-engagement) form a canonical tau-",
    "kinase activation ensemble at the tau x amyloid interaction; Csnk1a1 ",
    "(CK1-alpha) is the opposite-direction signal whose substrate program ",
    "is suppressed on the tau-on-amyloid background. Combined with the ",
    "section-14.3 TF axis-3 reading (Myc / Creb1 / Tp53 / Jun coordinated ",
    "suppression of biosynthetic activity), the joint mechanism at axis 3 ",
    "is 'the tau x amyloid interface drives tau hyperphosphorylation ",
    "through canonical tau kinases (Gsk3b lead) while transcriptionally ",
    "shutting down biosynthetic throughput (Myc lead)' -- exactly the kind ",
    "of two-layer answer the mechanism arc was designed to expose. The ",
    "section-13 axis-3 mixed-sign reading (OXPHOS / ribosomal compartment ",
    "+ MG-M3 + DAM_up at interaction; modality-divergent signs) resolves ",
    "under this joint mechanism: the proteomics + phospho signal of ",
    "biosynthetic + phospho dysregulation runs in the same direction once ",
    "the kinase + TF layers are read together.")
)

kinase_verdict <- build_kinase_verdict_table(
  axis_tbl           = kinase_axis_scores_tbl,
  per_contrast_tbl   = kinase_per_contrast_tsv,
  n_top_per_axis     = 5L,
  evidence_summaries = kinase_verdict_evidence
)

readr::write_tsv(
  kinase_verdict,
  file.path(params$results_dir, "kinase_activity_verdict.tsv")
)

cat(sprintf(
  "Kinase inference verdict: %d axes summarised (%s), top-%d kinases per axis.\n",
  nrow(kinase_verdict),
  paste(kinase_verdict$axis, collapse = ", "),
  5L))
```

### Verdict summary table

The summary table below has one row per D2 axis. `top_kinases` lists
the top 5 kinases per axis sorted by `|mean consensus score|` (the
axis-restricted ranking from 15.2); `top_kinase_signs` carries the
signed direction of each kinase's mean (`+` = positive substrate-
phosphorylation, `-` = negative substrate-phosphorylation at the
axis's contrast set, applied to the decoupleR signed-mode-of-
regulation convention so phosphatases and dephosphorylation edges
are sign-flipped as appropriate); `mean_score_range` gives the
[min, max] of the signed mean across those 5 kinases;
`per_contrast_summary` is the locked **hybrid** companion -- it
gives the [min, max] of the consensus score at each axis-contrast
in isolation, so a reader can see whether the axis-mean smooths over
heterogeneous per-contrast signal or is reading a coherent direction.
`n_top_sig_in_axis_contrasts` counts how many of the top 5 also
reach FDR<0.10 on ulm in *any* of the axis-relevant contrasts per
the 15.1 per-contrast table (this column **replaces** the E5
verdict's `n_top_in_cross_modality_leaderboard` because the F3
user-directed deviation removed the kinase leader-board concept);
`evidence_summary` is a prose distillation of the per-axis reading.

The hybrid display (mean range + per-contrast ranges side-by-side)
matches the section-14.3 TF verdict layout so the kinase verdict
and the TF verdict can be read in parallel without schema friction.

```{r kinase-inference-verdict-table, results = 'asis'}
kinase_verdict_view <- kinase_verdict |>
  dplyr::transmute(
    axis,
    top_kinases,
    top_kinase_signs,
    mean_score_range,
    per_contrast_summary,
    n_top_sig_in_axis_contrasts
  )
print(knitr::kable(
  kinase_verdict_view, digits = 2,
  caption = paste0(
    "Per-axis kinase verdict summary (top 5 axis-restricted kinases ",
    "per axis). Hybrid view: `mean_score_range` gives the [min, max] ",
    "of the axis-mean across the top 5; `per_contrast_summary` gives ",
    "per-axis-contrast [min, max] of the consensus score across the ",
    "same top 5 (exposes per-contrast asymmetry that the axis-mean ",
    "smooths over). `n_top_sig_in_axis_contrasts` = count of top-5 ",
    "kinases that also reach FDR<0.10 on ulm in any of the axis-",
    "relevant contrasts per the 15.1 per-contrast table."
  )
))
```

The `evidence_summary` column is omitted from the inline kable to
keep the row heights tractable; the full per-axis prose lives in
subsections 15.3.2-15.3.4 below, and the column is preserved verbatim
in the exported TSV (`storage/results/kinase_activity_verdict.tsv`).
The TSV also carries `per_contrast_score_range` (the [min, max]
across the top-5 x axis-contrasts cells, useful for at-a-glance
comparison to `mean_score_range`).

### Amyloid-activation axis (axis 1)

**Top 5 kinases by axis-restricted |mean|:** Cdk2 (+2.91), Cdk5
(+1.75), Csnk1e (+1.64), Mapk8 (+1.17), Camk2g (-1.16). Four
positive, one negative; signed mean range [-1.16, +2.91]. Two of the
five (Cdk5 and Csnk1e) reach FDR<0.10 on ulm at `nlgf_in_p301s` per
the 15.1 per-contrast table; the remaining three (Cdk2, Mapk8,
Camk2g) are non-significant at both NLGF contrasts but reach the
top-5 axis-mean ranking through coherent direction across the two
contrasts.

**Per-contrast hybrid view (locked F4).** The consensus score min /
max at the two NLGF contrasts across the same top 5 is
`nlgf_in_maptki:[-1.32, +3.02]; nlgf_in_p301s:[-1.00, +3.00]`. The
maptki contrast (NLGF on the tau-loss background) carries the
slightly wider positive tail (+3.02, driven by Cdk2 maptki +3.02 /
p301s +2.80; tight); Cdk5 and Mapk8 are predominantly p301s-driven
(Cdk5 maptki +0.50 vs p301s +3.00; Mapk8 maptki +0.25 vs p301s
+2.08) so two of the top 5 would not have ranked here on the maptki
contrast alone -- a real tau-on-amyloid substructure that the
axis-mean [-1.16, +2.91] smooths over. The hybrid display is
intentional: a reader concerned that "axis 1 reads as proliferation +
stress" might mask a tau-on-amyloid bias can check the per-contrast
row and see that Cdk5 / Mapk8 in particular *do* carry that bias,
and re-weight the biological reading accordingly.

**Biological reading.** Cdk2 is a cell-cycle-progression kinase
whose activation in microglia has been documented in disease-
associated states (DAM-like activation involves microglial
proliferative re-entry); its 259 in-axis-universe substrate sites is
the highest of the top 5 and reflects broad coverage in the amyloid
universe. Cdk5 -- the second canonical tau kinase from 15.1 --
surfaces at axis 1 driven by its p301s signal (+3.00); the maptki
signal (+0.50) is much weaker, consistent with Cdk5's tau-dependence
documented in the per-contrast layer. Csnk1e (CK1-epsilon) and Mapk8
(JNK) are stress-activated kinases canonical in microglial activation
and inflammatory signalling. Camk2g (CaMKII-gamma) is the lone
repressor in the top 5; calcium / calmodulin signalling is
attenuated on average across both NLGF contrasts, a real but minor
counter-signal worth preserving rather than filtering.

**Cross-reference to section-13 axis 1 and section-14.3 axis 1.**
Section 13's axis-1 verdict named MG-M2 (hdWGCNA module), the MHC II
antigen-presentation family, `GOBP_ADAPTIVE_IMMUNE_RESPONSE`, and the
custom `DAM_up` / `HAM_disease` / `WAM` sets as leader-pathway
evidence. Section 14.3's TF axis-1 named the Spi1 / Nfkb1 / Sp3 /
A0A979HLR9 (complex) / Rreb1 ensemble as the cross-modality leader
TFs. The kinase axis-1 reading here adds **Cdk2 / Cdk5 / Mapk8** as
the proliferative + stress-kinase mechanism downstream of the Spi1 /
Nfkb1 TF activation. The mechanism layer reads axis 1 across two
complementary layers: TFs name the master transcriptional regulators
of the DAM program; kinases name the phospho-signalling effectors of
the same program. The mechanism layer therefore reads axis 1
*strongly*.

### Synaptic-suppression axis (axis 2)

**Top 5 kinases by axis-restricted |mean|:** identical to axis 1
(Cdk2, Cdk5, Csnk1e, Mapk8, Camk2g) by Interpretation A construction.
The per-kinase mean is computed from full-cache consensus scores
averaged over the two NLGF contrasts; both NLGF axes share that
contrast set, so the mean ranking does not differentiate them.

**Per-contrast hybrid view (locked F4).** The per-contrast summary is
likewise identical to axis 1
(`nlgf_in_maptki:[-1.32, +3.02]; nlgf_in_p301s:[-1.00, +3.00]`)
because the same contrast pair is averaged over the same top 5
kinases. This is the expected hybrid behaviour at axis 2 under
Interpretation A and is preserved unsmoothed in the verdict so a
future reader can see that the entanglement between axes 1 and 2
propagates to the hybrid layer as well -- it is a structural property
of sharing the contrast set, not an information loss specific to the
axis-mean.

**The structural differentiator is in-universe substrate-site count,
not mean.** Cdk5's KSN substrate-site count expands from 81 in the
amyloid universe to 263 in the synaptic universe (a 3.2x increase
that reflects Cdk5's well-documented role in synaptic phospho-
regulation); Cdk2 expands from 259 to 293, Csnk1e from 38 to 70,
Camk2g from 24 to 49, and Mapk8 changes modestly from 176 to 195.
The synaptic axis is therefore best interpreted as a *target-overlap
filter*: which amyloid-active kinases have substrate programs that
*also* touch the synaptic machinery? Not as a list of kinases that
*drive* synaptic suppression. Two of the top 5 (Cdk5, Csnk1e) reach
FDR<0.10 on ulm at `nlgf_in_p301s`, identical to axis 1 by
inheritance.

**This is the honest non-finding.** The kinase layer does NOT
independently identify drivers of synaptic suppression at this
resolution. The top kinases are positive activators of broadly
synapse-relevant substrate programs (Cdk5 in particular is a
canonical synaptic phospho regulator), not repressors of synaptic
gene output. Two non-mutually-exclusive interpretations of the
absence: (a) synaptic suppression at NLGF onset in the microglia
compartment is detection-level engulfment -- microglia phagocytose
synaptic material and so apparent downregulation of synaptic
transcripts/proteins is a mass-balance signal, not a phospho-
signalling one. OmniPath KSN cannot surface a kinase for a mechanism
that is not phospho-regulatory. (b) Phospho-regulatory repression of
synaptic gene output in microglia, if it occurs, may be driven by
phosphatase complexes or by kinases whose substrate set is poorly
represented in OmniPath's literature-curated prior (e.g. Camk1
family, Prkcz family). A future plan could test (b) with a
broader-prior tool or with a microglia-sorted phospho-proteome.

**Cross-reference to section-13 axis 2 and section-14.3 axis 2.**
Section 13's axis-2 verdict named six synaptic GO CC terms plus
matching synapse GO BP terms, all consistently negative at the two
NLGF contrasts. Section 14.3's TF axis-2 was an honest non-finding
("TF layer does NOT independently identify drivers of synaptic
suppression at this resolution"). The kinase axis-2 verdict here
*also* reads as an honest non-finding, paralleling the TF reading
shape -- two mechanism layers agree that the synaptic-suppression
axis is not a regulator-driven program at the resolution available
to either layer. Phase G (CCC sender-receiver) is the layer that can
plausibly read the engulfment / phagocytosis biology directly.

### Interaction metabolic/translational axis (axis 3)

**Top 5 kinases by axis-restricted |mean|:** **Gsk3b (+4.77)**,
Csnk1a1 (-2.81), Mapk14 (+2.63), Cdk5 (+2.10), Cdk1 (+1.99). Four
positive, one negative; signed mean range [-2.81, +4.77] at the
single `interaction` contrast. Two of the five (Gsk3b, Csnk1a1) reach
FDR<0.10 on ulm at the `interaction` contrast itself per the 15.1
per-contrast table -- **Gsk3b at padj 0.002, the strongest single-
kinase result of the whole mechanism arc**, and Csnk1a1 at padj <
0.10 in the opposite (negative) direction.

**Per-contrast hybrid view (locked F4).** The per-contrast summary
collapses to a single contrast at axis 3
(`interaction:[-2.81, +4.77]`), identical to `mean_score_range` by
construction. The hybrid columns are therefore preserved in the
exported TSV for schema consistency with axes 1 and 2 but provide
no additional information beyond the axis-mean at this axis -- a
deliberate property of the chosen axis_contrasts mapping (only the
`interaction` contrast is included at axis 3). A future plan that
admits more than one interaction contrast (e.g. higher-order tau x
amyloid x age or tau x amyloid x cell-state slices) will see the
hybrid view become informative at axis 3 as well.

**Biological reading.** This is the cleanest mechanism-layer result
of the project. **Gsk3b** -- the canonical tau kinase that
phosphorylates tau at S202 / T205 / S396 / S404 (the AT8 / PHF1
epitopes used to stage AD tau pathology, and the dominant target of
two decades of pharmacological development) -- is the rank-1 kinase
at the tau x amyloid interaction with mean +4.77 and 257 in-axis-
universe substrate sites. The Gsk3b activation peaks at the
`interaction` contrast itself; the per-contrast layer at 15.1 shows
Gsk3b's sign pattern across all five contrasts is -1.66 / -1.47 /
+3.79 / +4.15 / +4.32 -- amyloid alone on tau-null background and tau
alone both *suppress* Gsk3b substrate phospho; the combination of tau
and amyloid drives the activation, peaking at the interaction
contrast. The synergy *is* the Gsk3b activation; this reads as the
most direct possible mechanism-layer answer to the project's
interaction question ("what does tau add on the amyloid
background?").

Mapk14 (p38-alpha MAPK, +2.63) is the canonical stress-kinase
activator of tau phosphorylation downstream of inflammatory and ROS
stimuli and a documented modulator of microglial activation; its
co-activation here corroborates Gsk3b at the interaction. **Cdk5**
(+2.10) -- the second canonical tau kinase -- corroborates again,
preserving the same tau-kinase-cluster signature seen at the per-
contrast layer. Cdk1 (+1.99), the master mitotic kinase, surfaces as
the fifth hit and is biologically interpretable as evidence of
microglial cell-cycle re-engagement at the interaction. The single
repressor in the top 5, **Csnk1a1** (CK1-alpha, -2.81), is the same
opposite-direction CK1-family signal seen at 15.1 and is consistent
with at least two distinct kinase programs operating at the
interaction: the canonical tau-kinase axis (Gsk3b / Cdk5) gaining
activity, and the CK1-alpha axis losing activity, on the tau-on-
amyloid background.

**Joint mechanism reading across TF + kinase layers at axis 3.** The
TF layer at section 14.3.4 reads a Myc / Creb1 / Tp53 / Jun
coordinated suppression of biosynthetic activity at the interaction
(all five top TFs negative; signed mean range [-4.85, -1.81]). The
kinase layer at axis 3 reads a Gsk3b / Mapk14 / Cdk5 / Cdk1 canonical
tau-kinase activation at the same contrast. These two readings are
*mechanistically complementary*, not redundant: TFs read the slow-
time-scale transcriptional re-programming of biosynthetic capacity;
kinases read the fast-time-scale phospho-signalling state. The
combined mechanism at axis 3 is "the tau x amyloid interface drives
tau hyperphosphorylation through canonical tau kinases while
transcriptionally shutting down biosynthetic throughput" -- exactly
the kind of two-layer answer the mechanism arc was designed to
expose. **The section-13 axis-3 mixed-sign reading resolves under
this joint mechanism**: the proteomics + phospho signal of
biosynthetic + phospho dysregulation at axis 3 was previously called
"mixed-sign" because RNA / proteomics push negative on `DAM_up` and
MG-M3 while phospho-corrected pushes positive; the joint TF + kinase
reading shows the two layers run in the *same biological direction*
(biosynthesis off, tau-kinase substrate program on) with the apparent
sign disagreement at the gene-set layer arising from RNA / proteomics
/ phospho-corrected orthogonality, not internal disagreement on the
direction of metabolic + phospho perturbation.

**Cross-reference to section-13 axis 3 and section-14.3 axis 3.**
Section 13's axis-3 verdict named eight OXPHOS / ETC GO terms (BP +
MF + CC), the ribosomal compartment, MG-M3, and `DAM_up` at the
interaction. Section 14.3's TF axis-3 named the Myc / Creb1 / Tp53 /
Jun coordinated-suppression ensemble. The kinase axis-3 reading here
adds Gsk3b / Mapk14 / Cdk5 / Cdk1 as the canonical tau-kinase
activation ensemble at the same contrast. The mechanism layer reads
axis 3 *very strongly* with a two-layer integrated answer.

### Cross-axis observations and limitations

**The kinase mechanism layer is asymmetric across the three D2 axes,
mirroring the TF mechanism layer.** Axes 1 (amyloid_activation) and
3 (interaction_metabolic) both read with coherent, biologically
interpretable kinase ensembles (Cdk2 / Cdk5 / Mapk8 for axis 1;
Gsk3b / Mapk14 / Cdk5 / Cdk1 for axis 3). Axis 2 (synaptic_suppression)
is an honest non-finding at the kinase layer -- the top 5 are
inherited from axis 1 by the shared NLGF contrast set, and no
distinct synapse-repressor program emerges. The asymmetry across
axes is the same shape as the TF layer's asymmetry (strong /
non-finding / strong) and is consistent with the section-14.3
reading that synaptic suppression is more plausibly an engulfment /
detection-level phenomenon than a regulator-driven program.

**Cdk5 is the integrator across all three axes at the kinase layer.**
Cdk5 surfaces in the top 5 at axis 1 (+1.75 mean, p301s-driven),
axis 2 (+1.75 by Interpretation A inheritance), and axis 3 (+2.10
mean at the interaction contrast). It is the only kinase to do so.
The cross-axis substrate-set differential is biologically
interpretable: 81 sites in the amyloid universe, 263 in the synaptic
universe, 153 at the interaction. Cdk5 is therefore the kinase whose
program is most broadly engaged across the project's three biological
axes; its literature reputation as a master regulator of both tau
phospho-biology and synaptic phospho-biology is consistent with this
pattern. A future plan could deploy a Cdk5-substrate-resolved sub-
analysis to test whether the axis-specific Cdk5 substrates differ in
identity across axes 1, 2, and 3 (the same future-work hook recorded
at 15.2.4).

**Two-layer mechanism reading per axis (preferred summary).** The
combined section-14.3 TF verdict and section-15.3 kinase verdict
read each D2 axis with the following per-axis mechanism:

* Axis 1 (amyloid_activation): TFs name Spi1 / Nfkb1 / Sp3 / Rela
  (cross-modality leaders A0A979HLR9 and Spi1 reach the cross-modality
  leader rule); kinases name Cdk2 / Cdk5 / Mapk8 / Csnk1e (Cdk5 +
  Csnk1e reach FDR<0.10 at nlgf_in_p301s). Joint reading: the canonical
  microglial DAM-state induction at NLGF onset, with the
  transcriptional axis driving the program and the kinase axis
  delivering the phospho-signalling effector layer. Both layers
  corroborate.
* Axis 2 (synaptic_suppression): TFs read as honest non-finding
  (top 5 inherited from axis 1 by Interpretation A); kinases read
  as honest non-finding (top 5 inherited from axis 1 by Interpretation
  A). Joint reading: the synaptic-suppression axis is not a
  regulator-driven program at the resolution available to either
  mechanism layer; most parsimonious mechanism is an engulfment /
  detection-level phenomenon. Phase G CCC is the layer expected to
  surface it.
* Axis 3 (interaction_metabolic): TFs name Myc / Creb1 / Tp53 / Jun
  (coordinated suppression of biosynthetic activity); kinases name
  Gsk3b / Mapk14 / Cdk5 / Cdk1 (canonical tau-kinase activation), with
  Gsk3b at padj 0.002 the strongest single-kinase result. Joint
  reading: at the tau x amyloid interaction the cell suppresses
  biosynthetic capacity transcriptionally while activating canonical
  tau-kinase substrate phosphorylation; both readings run in the
  same biological direction and jointly resolve the section-13
  axis-3 mixed-sign call to a single coherent mechanism. Both layers
  corroborate strongly.

**The axis-1 / axis-2 entanglement is preserved at the kinase verdict
layer.** Both NLGF contrasts populate axes 1 and 2 by the locked
axis-classification rule, so the axis-restricted ranking by mean
cannot differentiate them at the kinase layer either; the same trade-
off was accepted at the TF layer at E4 / E5. The structural
differentiator remains in-universe substrate-site count, exposed per
row in the F4 axis-restricted TSV. A future plan that re-runs
decoupleR with the KSN clipped to per-axis universes (Interpretation
B) might surface different kinase rankings at axes 1 vs 2 at the
cost of statistical power; the trade-off is the same one made at the
TF layer and is locked at the same direction.

**No anti-anchoring guardrails were relaxed in this verdict.** All
three D2 axes are named explicitly; no axis is privileged over the
others; the F3 FDR<0.10 ulm threshold and the F4 `min_targets = 5`
filter are re-stated above; no Hallmark or OXPHOS-confirmatory-arc
terminology appears; hdWGCNA modules are alphabetised where they
appear (the verdict mentions MG-M2 at axis 1 and MG-M3 at axis 3
individually only, so the alphabetisation rule is trivially
satisfied here). The verdict's bottom-line reading is deliberately
asymmetric across axes (strong / non-finding / strong) because the
data are asymmetric across axes; collapsing to a single-axis winner
or a fake-symmetric three-axis pose would reintroduce the bias the
section-13 D2 verdict, the section-14.3 TF verdict, and this kinase
verdict were jointly designed to remove.

**Phase F COMPLETE.** With this verdict subsection, the mechanism
layer arc has produced two complementary verdicts (TFs at 14.3,
kinases at 15.3) against the same three D2 axes from section 13.
The next active session in `storage/notes/mechanism_layer_plan.md`
is Phase G1 -- the CCC re-analysis approach decision gate. CCC is
expected to be the layer that can plausibly surface the synaptic-
suppression axis (axis 2) where both TF and kinase layers read as
honest non-finding.
