# Integrated biological model

This section is the synthesis chapter for the project. Sections 1-16
emitted per-modality DE, per-axis pathway / TF / kinase / LR verdicts,
and the three mechanism-layer verdicts that closed the Phase G arc.
None of those sections compute an integrated reading: a reader who
finishes section 16 has 14 columns of TSV evidence per axis but no
explicit answer to "which competing biological model does the
project's evidence actually favour, and how strongly?". This chapter
closes that gap as a **structured hypothesis-comparison synthesis**
rather than a free narrative, the stance was locked at the start of
Phase H (`storage/notes/biological_model_plan.md` §H0) to insulate the
synthesis from training-data drift toward canonical AD-microglia
stories.

The synthesis proceeds in five movements. Subsection 17.1 names the
eleven competing biological entities adjudicated by this chapter (six
per-axis contest models in three pairwise contests, one cross-axis
integrator slot, and four cross-axis themes). Subsection 17.2 renders
the 92-row claims ledger (75 Phase H rows + 6 Phase I rows from the
per-state NF-kB attenuation test reported in section 18, 2 Phase J
rows from the causal signalling-network reconstruction reported in
section 20, 2 Phase K rows from the SCENIC data-driven regulon
recovery reported in section 21, 2 Phase L rows from the GeoMx
spatial-deconvolution tissue-composition layer reported in section 22,
2 Phase M rows from the snRNAseq microglial activation-pseudotime
trajectory layer reported in section 23, 2 Phase N rows from the
cross-cell-type interaction-specificity layer reported in section 24,
plus 1 Phase O row from the gene-level differential-dynamics
interaction layer also reported in section 23)
that records every atomic evidence claim the chapter draws on. Subsection 17.3 renders the H3 adjudication
arithmetic in two parts: per-axis-contest favoured-model verdicts and
cross-axis entity support tallies. Subsection 17.4 lists the testable
predictions that would upgrade Suggestive claims, resurrect
contradicted models, or strengthen low-support themes. Subsection 17.5
states the integrated biological reading the project's evidence
currently supports, phrased as a working synthesis grounded in the
ledger, not as an authoritative conclusion.

No new computation is performed in this section. Every claim, every
adjudication column, and every verdict is derived from the H2 ledger
(`storage/results/biological_model_claims_ledger.tsv`) and the H3
adjudication tables (`storage/results/biological_model_adjudication.tsv`
and `storage/results/biological_model_contest_verdicts.tsv`), which
were emitted by `scripts/build_biological_model_ledger.R` and
`scripts/build_biological_model_adjudication.R` respectively. The
chapter is a structured reading of those three TSVs.

```{r biological-model-load-tsvs}
# H2 ledger (built outside the knit by scripts/build_biological_model_ledger.R;
# readRDS-equivalent explicit load per the cache-loading convention in
# CLAUDE.md for caches built outside the knit).
biomodel_ledger <- readr::read_tsv(
  file.path(params$results_dir, "biological_model_claims_ledger.tsv"),
  show_col_types = FALSE
)

# H3 adjudication (per-entity table and per-contest verdicts; both built by
# scripts/build_biological_model_adjudication.R outside the knit).
biomodel_adjudication <- readr::read_tsv(
  file.path(params$results_dir, "biological_model_adjudication.tsv"),
  show_col_types = FALSE
)
biomodel_verdicts <- readr::read_tsv(
  file.path(params$results_dir, "biological_model_contest_verdicts.tsv"),
  show_col_types = FALSE
)

cat(sprintf(
  "Loaded biological-model synthesis tables: ledger %d rows x %d cols; adjudication %d rows x %d cols; verdicts %d rows x %d cols.\n",
  nrow(biomodel_ledger), ncol(biomodel_ledger),
  nrow(biomodel_adjudication), ncol(biomodel_adjudication),
  nrow(biomodel_verdicts), ncol(biomodel_verdicts)
))
```

## Competing biological models being adjudicated

The eleven entities below are the targets of the adjudication. The six
**per-axis contest models** are paired head-to-head within their axis
(Hyp-1A vs Hyp-1B at amyloid_activation, Hyp-2A vs Hyp-2B at synaptic_suppression,
Hyp-3A vs Hyp-3B at interaction_metabolic) so that each contest yields a
single favoured-model verdict in subsection 17.3.1. The remaining five
entities (Hyp-0 plus four cross-axis themes) are adjudicated as
stand-alone support tallies in subsection 17.3.2, the four themes
intentionally overlap with the per-axis contest models (e.g.
T-Tau-attenuates subsumes Hyp-1B; T-Synergy subsumes Hyp-3B), so a single
claim row commonly populates both a per-axis model column and the
theme column it inherits. The hybrid scheme was locked at the H1
decision gate (`storage/notes/biological_model_plan.md` §H1
deliverable 1) over the flat-themes-only and per-axis-only
alternatives because the per-axis contests cleanly inherit the axis
structure of the project's verdicts, the themes surface cross-axis
patterns that no single contest would, and Hyp-0 captures the Cdk5
top-5-at-all-three-axes integrator finding without forcing it into a
per-axis contest where it does not naturally belong.

```{r biological-model-entity-catalogue, results = "asis"}
# Build the structured entity catalogue from the adjudication TSV's
# id / type / contest_id / name columns, plus hand-authored
# entity_statement + key_prediction cells distilled from the H1
# deliverable-1 prose (storage/notes/biological_model_plan.md
# lines 161-269; the verbatim model definitions). Statements and
# predictions are paraphrased here for compactness; the long-form
# versions are in the plan body.
entity_prose <- tibble::tribble(
  ~entity_id, ~entity_statement, ~key_prediction,

  "Hyp-1A",
  "Amyloid drives a canonical DAM-state inflammatory program; the program is qualitatively identical in MAPTKI and P301S backgrounds; tau contributes no inflammatory modulator at axis 1.",
  "Matched-direction matched-magnitude TF / kinase / LR / pathway signals at `nlgf_in_maptki` and `nlgf_in_p301s` with no informative interaction-contrast signal at axis 1.",

  "Hyp-1B",
  "Amyloid drives NF-kB-led inflammation in MAPTKI; tau on the amyloid background SUPPRESSES this program. Rela sign-reversal is the load-bearing signature.",
  "Rela score sign-flip (positive at `nlgf_in_maptki` / `nlgf_in_p301s`, negative at `interaction`); Nfkb1 score-attenuation at interaction; coordinated Spi1 / Nfkb1 / Rela asymmetric reading.",

  "Hyp-2A",
  "NLGF-driven synaptic suppression is mediated by canonical C1q / C3 / Mertk-Pros1 / Tyro3-Axl complement-tagging plus receptor-mediated engulfment.",
  "Classical complement LR pairs (C1q-family, C3, Mertk-Pros1, Tyro3-Axl) at the top of the axis-2 cell-cell communication ranking.",

  "Hyp-2B",
  "NLGF-driven synaptic suppression is mediated by TREM2-mediated clearance plus APP-fragment uptake plus synaptic-adhesion modulation, not by classical complement.",
  "Apoe-Trem2, App-Cd74, Cd200-Cd200r1, Vcan-Cd44, Ncam1-ligand pairs at the top of axis-2 rankings instead of classical complement.",

  "Hyp-3A",
  "The interaction contrast is arithmetically the difference of single-insult arms; no qualitatively new mechanism emerges at the tau x amyloid interface.",
  "Mechanism-layer signals at the interaction are smaller-than-amyloid-alone in magnitude or absent; no interaction-specific TF / kinase / LR signature.",

  "Hyp-3B",
  "A qualitatively new mechanism emerges at the interaction: Gsk3b activation, coordinated TF biosynthetic suppression (Myc / Creb1 / Tp53 / Jun), and axon-guidance LR rewiring at the LR layer.",
  "All three layers carry interaction-contrast-specific sign or magnitude patterns that are not inferrable from either single-insult arm.",

  "Hyp-0",
  "Cdk5 acts as a cross-axis kinase integrator: it ranks in the kinase top 5 at all three axes (ranks 2 / 2 / 4) and its KSN substrate footprint scales with the axis universe (81 / 263 / 153 sites).",
  "Cdk5-substrate enrichment in claims at every axis; CDK5R1 / CDK5R2 expression continuity across genotypes; cross-axis TF or pathway concordance traceable to Cdk5 substrates.",

  "T-Inflammation",
  "A unified inflammatory program drives both axes 1 and 2; amyloid and tau act additively on a canonical DAM inflammatory state, qualitatively the same regardless of tau background.",
  "Axis-1 TF / kinase activations replicate at axis-2 contrasts; no qualitative interaction-specific mechanism at axis 3. Theme subsumes Hyp-1A and Hyp-2A jointly.",

  "T-Compartment-suppression",
  "NLGF suppresses the synaptic compartment via classical complement-mediated pruning at axis 2.",
  "Axis-2 evidence shows synaptic GO term suppression AND classical complement LR pairs at the top of axis-2 rankings. Theme subsumes Hyp-2A.",

  "T-Tau-attenuates",
  "Tau modifies the amyloid program rather than amplifying it; specifically, tau suppresses amyloid-driven NF-kB at the interaction contrast (Rela sign-reversal).",
  "Positive Rela at NLGF arms, negative Rela at interaction; coordinated NF-kB-family asymmetric reading across contrasts. Theme subsumes Hyp-1B.",

  "T-Synergy",
  "A qualitatively new mechanism emerges at the tau x amyloid interface that is not present in either single-insult arm.",
  "(a) Gsk3b activation specifically at interaction; (b) coordinated TF suppression (Myc / Creb1 / Tp53 / Jun) at interaction; (c) interaction-specific LR rewiring (axon-guidance, Gpc3-Unc5c, Sema6a-Plxna4 sign-flip). Theme subsumes Hyp-3B."
)

structural_type_lookup <- c(
  contest_model = "Per-axis contest model",
  m0            = "Cross-axis integrator",
  theme         = "Cross-axis theme"
)
axis_or_scope <- function(entity_type, contest_id) {
  ifelse(entity_type == "contest_model", contest_id, "cross-axis")
}

entity_catalogue <- biomodel_adjudication |>
  dplyr::select(entity_id, entity_type, contest_id, entity_name) |>
  dplyr::mutate(
    structural_type     = unname(structural_type_lookup[entity_type]),
    axis_or_scope       = axis_or_scope(entity_type, contest_id)
  ) |>
  dplyr::left_join(entity_prose, by = "entity_id") |>
  dplyr::select(
    structural_type, entity_id, axis_or_scope, entity_name,
    entity_statement, key_prediction
  ) |>
  dplyr::arrange(
    factor(structural_type, levels = unname(structural_type_lookup)),
    entity_id
  )

print(knitr::kable(
  entity_catalogue,
  col.names = c("Structural type", "ID", "Axis or scope",
                "Entity name", "Biological statement", "Key prediction"),
  caption = paste0(
    "The 11 competing biological entities adjudicated by this ",
    "chapter. Per-axis contest models (Hyp-1A/Hyp-1B, Hyp-2A/Hyp-2B, Hyp-3A/Hyp-3B) ",
    "are paired head-to-head within their axis and yield a binary ",
    "favoured-model verdict in subsection 17.3.1; the cross-axis ",
    "integrator (Hyp-0) and the four cross-axis themes are adjudicated ",
    "as stand-alone support tallies in subsection 17.3.2. Themes ",
    "intentionally subsume per-axis models (e.g. T-Tau-attenuates ",
    "subsumes Hyp-1B); ledger rows can support both simultaneously."
  )
))
```

## Claims ledger

The H2 ledger records `r nrow(biomodel_ledger)` atomic evidence claims
across the project's modalities, axes, and mechanism layers. Each row
carries an `effect_size` cell with a concrete number string and a
`primary_evidence_source` pointer at the TSV row or section the claim
distils. The `supports_models` and `contradicts_models` cells tag each
claim against zero or more of the 11 entities defined in 17.1, which
is what makes the H3 groupby arithmetic in 17.3 unambiguous. The
confidence grade is one of Strong / Moderate / Suggestive per the
H1-locked three-tier rules (Strong = >=2-modality consistent sign +
primary-tool FDR<0.10 + explicit cross-axis or cross-layer
corroboration; Moderate = primary-tool FDR<0.10 but lacks one of those
replication arms; Suggestive = top-N of primary tool without
significance, or with a structural caveat such as Interpretation A
entanglement or n_cells_used = 1).

```{r biological-model-ledger-distribution, results = "asis"}
# A compact distribution summary so the reader sees what is in the
# ledger before scrolling its interactive view below.
grade_dist  <- as.list(table(biomodel_ledger$confidence_grade))
axis_dist   <- as.list(table(biomodel_ledger$axis))
layer_dist  <- as.list(table(biomodel_ledger$layer))
fmt_dist <- function(x) paste(sprintf("%s=%d", names(x), unlist(x)), collapse = ", ")
cat(sprintf(
  paste0("Ledger composition: confidence-grade %s; axis %s; layer %s. ",
         "The ledger is rendered below as a `DT::datatable` so that ",
         "the reader can filter or sort live (e.g. set ",
         "confidence_grade = Strong, set supports_models to contain ",
         "Hyp-3B, sort by axis) without re-running the chapter.\n\n"),
  fmt_dist(grade_dist), fmt_dist(axis_dist), fmt_dist(layer_dist)
))
```

```{r biological-model-ledger-datatable}
# DT::datatable for the 92-row ledger (75 Phase H rows + 6 Phase I rows
# added 2026-05-25 per the section 18 per-state NF-kB attenuation test,
# 2 Phase J rows added 2026-06-04 per the section 20 causal
# signalling-network reconstruction, 2 Phase K rows added
# 2026-06-05 per the section 21 SCENIC data-driven regulon recovery,
# 2 Phase L rows added 2026-06-05 per the section 22 GeoMx
# spatial-deconvolution tissue-composition layer, 2 Phase M rows
# added 2026-06-06 per the section 23 snRNAseq activation-pseudotime
# trajectory layer, 2 Phase N rows added 2026-06-06 per the
# section 24 cross-cell-type interaction-specificity layer, plus 1 Phase O
# row added 2026-06-08 per the section 23 gene-level differential-dynamics
# interaction layer). Default visible columns are the
# narrative essentials; the audit-trail columns (corroborating_evidence,
# notes, n_replicates_in_modalities, n_replicates_in_layers) are
# loaded but hidden by default and toggleable via the Column visibility
# button. Confidence-grade and supports/contradicts columns are
# filterable via the search panes at column footers.
ledger_view <- biomodel_ledger |>
  dplyr::select(
    claim_id, axis, layer, claim, direction, effect_size,
    confidence_grade, supports_models, contradicts_models,
    primary_evidence_source, corroborating_evidence,
    n_replicates_in_modalities, n_replicates_in_layers, notes
  )

DT::datatable(
  ledger_view,
  rownames  = FALSE,
  filter    = "top",
  extensions = c("Buttons", "FixedHeader"),
  options   = list(
    pageLength    = 15,
    lengthMenu    = c(10, 15, 25, 50, 92),
    autoWidth     = FALSE,
    scrollX       = TRUE,
    fixedHeader   = TRUE,
    dom           = "Bfrtip",
    buttons       = list(
      "copy", "csv",
      list(extend = "colvis", text = "Column visibility")
    ),
    columnDefs    = list(
      # Hide audit-trail columns by default; user can re-show them via
      # the Column visibility button. Indices are 0-based.
      list(visible = FALSE,
           targets = c(9, 10, 11, 12, 13))
    )
  ),
  caption = "H2 biological-model claims ledger (filterable, sortable). Click the Column visibility button to surface the hidden audit-trail columns (`primary_evidence_source`, `corroborating_evidence`, `n_replicates_in_modalities`, `n_replicates_in_layers`, `notes`)."
)
```

## Model adjudication

The adjudication arithmetic was emitted by H3 as a single-pass groupby
on the H2 ledger's `supports_models` and `contradicts_models` membership
cells. Two tables surface the result: per-axis contest verdicts (one
row per contest, a binary favoured-model verdict against the head-to-
head competitor) and per-entity stand-alone tallies (one row per
entity, including the Hyp-0 integrator and the four cross-axis themes,
which do not compete head-to-head with each other).

### Per-axis contest verdicts

```{r biological-model-contest-verdicts-table, results = "asis"}
verdict_view <- biomodel_verdicts |>
  dplyr::transmute(
    contest_id,
    model_a = sprintf("%s - %s", model_a_id, model_a_name),
    a_support   = model_a_support_count,
    a_contradict = model_a_contradict_count,
    a_net       = model_a_net_support,
    a_strong    = model_a_n_strong_supports,
    model_b = sprintf("%s - %s", model_b_id, model_b_name),
    b_support   = model_b_support_count,
    b_contradict = model_b_contradict_count,
    b_net       = model_b_net_support,
    b_strong    = model_b_n_strong_supports,
    favoured_model,
    margin     = favoured_by_margin,
    tie_break  = tie_break_rule_invoked
  )
print(knitr::kable(
  verdict_view, digits = 0,
  col.names = c("Contest", "Model A", "A sup", "A con", "A net",
                "A Strong", "Model B", "B sup", "B con", "B net",
                "B Strong", "Favoured", "Margin", "Tie-break"),
  caption = paste0(
    "Per-axis contest favoured-model verdicts. `sup` / `con` / ",
    "`net` are support_count / contradict_count / net_support ",
    "derived from the H2 ledger; `Strong` is the number of ledger ",
    "rows supporting the model that carry a Strong confidence grade. ",
    "`Margin` is the favoured model's net_support minus the loser's ",
    "net_support; `Tie-break` is `no_tie_break_needed` for all three ",
    "contests (every verdict resolves at the first arithmetic level)."
  )
))
```

**Axis 1 (amyloid_activation): Hyp-1B beats Hyp-1A by net-support margin 18
(34 - 16) after the Phase I lift and the Phase K + Phase L + Phase M + Phase N Suggestive rows, up from
margin 6 (24 - 18) at the end of Phase H.** The decisive evidence for Hyp-1B is the Rela sign-reversal
across contrasts (positive at `nlgf_in_maptki` / `nlgf_in_p301s`,
negative at `interaction`; H2-006 Strong with cross-axis corroboration
at H2-072) plus the coordinated NF-kB-family asymmetric reading that
Hyp-1A does not predict; nine ledger rows now actively contradict Hyp-1A
(the original three at Phase H were H2-007 per-contrast asymmetric
reading, H2-028 ifn_nlgf_asymmetry 8x P301S-specific gene differential,
and H2-072 cross-axis Rela sign-reversal; the six Phase I rows I-001
through I-006 each contradict Hyp-1A by per-substate replication of the
interaction sign-reversal plus the formal compositional adjudication).
Hyp-1A retains 25 supporting rows (the matched-direction TF / kinase
/ pathway claims at the two NLGF arms plus the Phase K Spi1 regulon-
existence row K-001, the Phase L DAM tissue-expansion row L-001,
the Phase M amyloid-progression row M-001, and the Phase N
cross-cell-type microglia-led row N-001,
which each support BOTH amyloid models and so lift each net by one
without moving the margin; Hyp-1B does not contradict them because
matched-direction support and asymmetric-reading support are
non-exclusive evidence) but its single 1-row inflammatory baseline
cannot account for the sign-flip, and the per-substate replication in
Phase I rules out the alternative explanation that the sign-flip is a
whole-microglia aggregation artefact. The 11 Strong supports for Hyp-1B
vs 6 for Hyp-1A break in Hyp-1B's favour at the strong-evidence level
even more decisively after Phase I (Hyp-1B gained 4 Strong rows from
the Phase I set: I-001 / I-002 / I-005 / I-006).

**Axis 2 (synaptic_suppression): Hyp-2B beats Hyp-2A by net-support margin
12 (8 - (-4)).** The decisive evidence for Hyp-2B is the cluster of
axis-restricted LR rows (Apoe-Trem2 ranks 9 + 10; App-Cd74 ranks 12 +
20; the 49/100 axis-2 microglia-involving fraction) plus the absence
of classical complement pairs (C1q-family, C3, Mertk-Pros1, Tyro3-Axl)
in the top 800 axis-restricted leader cells. The H2 ledger flags an
explicit scope ambiguity: while Hyp-2A is contradicted on the
axis-restricted view (rows H2-035 / H2-036 / H2-037 / H2-038 / H2-039
/ H2-045), Pros1-Mertk does appear strongly in the cross-tool leader
board (`ccc_lr_cross_tool_leaderboard.tsv` leader_score 15.4, 6
microglia-microglia rows), H2-040 carries the cross-tool Pros1-Mertk
row as the sole Moderate support for Hyp-2A, narrowing the verdict to
"Hyp-2B wins on axis-restricted scope; Hyp-2A's classical-complement
mechanism survives narrowly as a cross-tool efferocytosis signal at a
specific microglia-microglia node". The contest still resolves in
Hyp-2B's favour but the scope qualification is important and is named
explicitly in 17.5.

**Axis 3 (interaction_metabolic): Hyp-3B beats Hyp-3A by net-support margin
55 (28 - (-27)).** Hyp-3A is unanimously contradicted: every axis-3
ledger row that supports Hyp-3B (Gsk3b activation padj 0.00152 at
H2-051; Mapk14 / Cdk5 / Cdk1 supporting kinase rows; Myc / Creb1 /
Tp53 / Jun coordinated TF suppression; OXPHOS / ETC family, ribosomal,
MG-M3 module pathway support; Sema6a-Plxna4 sole 5/5-contrast G3
leader sign-flip at H2-057; Gpc3-Unc5c / Entpd1-Adora1 / L1cam-Ephb2
LR rows) also contradicts Hyp-3A by mutual exclusion of "no
interaction-specific mechanism". This is the largest margin of the
three contests and reflects how decisively axis 3 falls to the synergy
mechanism, the Gsk3b padj 0.00152 is the strongest single-kinase
result of the entire project, and 8 of the 28 Hyp-3B-supporting rows
carry the Strong confidence grade (17 Moderate, 3 Suggestive). The
1-row gap between Hyp-3B's support_count of 28 and Hyp-3A's
contradict_count of 27 is the LR-layer row H2-063 ("51 of the top 100
axis-3 LR pairs involve a microglia sender or receiver") that supports
Hyp-3B but leaves its `contradicts_models` cell empty, a
microglia-involving-rate context claim that documents substantial
microglia-mediated LR signal at the interaction axis without being
directly model-discriminating against Hyp-3A's "no interaction-specific
mechanism".

### Cross-axis entity support tallies

```{r biological-model-entity-tally, results = "asis"}
entity_tally_view <- biomodel_adjudication |>
  dplyr::select(
    entity_id, entity_type, contest_id, entity_name,
    support_count, contradict_count, net_support,
    n_strong_supports, n_moderate_supports, n_suggestive_supports
  ) |>
  dplyr::arrange(dplyr::desc(net_support), dplyr::desc(n_strong_supports))

print(knitr::kable(
  entity_tally_view,
  col.names = c("ID", "Type", "Contest", "Entity name",
                "Sup", "Con", "Net", "Strong", "Mod", "Sug"),
  caption = paste0(
    "Per-entity stand-alone support tallies, sorted by net_support ",
    "descending then by n_strong_supports descending. All 11 entities ",
    "are included; per-axis contest models are listed alongside ",
    "themes and Hyp-0 to make the cross-entity ranking explicit. ",
    "`Sup` / `Con` / `Net` = support_count / contradict_count / ",
    "net_support; Strong / Mod / Sug = confidence-grade decomposition ",
    "of the supporting rows."
  )
))
```

```{r biological-model-axis-anchoring-table, results = "asis"}
# Per-axis breakdown of supports for the cross-axis entities (Hyp-0 +
# 4 themes). This is the data pattern the H3 completion note flagged
# as the single-axis anchoring observation: each theme draws bulk
# support from one axis; Hyp-0 is the most evenly distributed entity.
axis_anchor_view <- biomodel_adjudication |>
  dplyr::filter(entity_type %in% c("integrator", "theme")) |>
  dplyr::transmute(
    entity_id,
    entity_name,
    amyloid_activation     = support_at_amyloid_activation,
    synaptic_suppression   = support_at_synaptic_suppression,
    interaction_metabolic  = support_at_interaction_metabolic,
    cross_axis             = support_at_cross_axis,
    total                  = support_count,
    primary_axis_share     = sprintf(
      "%d/%d (%.0f%%)",
      pmax(support_at_amyloid_activation,
           support_at_synaptic_suppression,
           support_at_interaction_metabolic,
           support_at_cross_axis),
      support_count,
      100 * pmax(support_at_amyloid_activation,
                 support_at_synaptic_suppression,
                 support_at_interaction_metabolic,
                 support_at_cross_axis) / support_count
    )
  ) |>
  dplyr::arrange(dplyr::desc(total))

print(knitr::kable(
  axis_anchor_view,
  col.names = c("ID", "Entity name", "Axis 1", "Axis 2", "Axis 3",
                "Cross", "Total", "Primary-axis share"),
  caption = paste0(
    "Per-axis support breakdown for the 5 cross-axis entities ",
    "(Hyp-0 + 4 themes). The `Primary-axis share` column gives the ",
    "fraction of total support drawn from the single axis with the ",
    "most rows, the diagnostic for single-axis anchoring."
  )
))
```

The headline pattern is **single-axis anchoring** of three of the four
cross-axis themes: T-Inflammation draws 28 of 31 supports from axis 1
(90%); T-Compartment-suppression draws 16 of 17 supports from axis 2
(94%); T-Synergy draws 27 of 30 supports from axis 3 (90%). Each
"cross-axis" theme is, on the project's evidence, predominantly the
cross-axis-friendly framing of a single-axis finding rather than a
genuinely distributed signal across axes. The exceptions are
**T-Tau-attenuates** (1 / 0 / 2 / 9 split across the 4 axis slots
after the Phase I and Phase J lifts, heavily cross_axis-anchored at
75% because the 6 Phase I per-state rows plus the Phase J NF-kB-route
row J-002 are all tagged cross_axis, given they test the
axis-1-to-axis-3 sign-reversal at per-substate resolution and probe
its upstream signalling route)
and **Hyp-0** (1 / 1 / 2 / 3, the most evenly distributed entity
of all 11, with 5 Strong-graded rows distributed across axes plus
the cross-axis section). Hyp-0's evenness vindicates the H1 decision
to admit it as a stand-alone integrator slot rather than forcing it
into a per-axis contest, and it remains the most-distributed
cross-axis claim the project's evidence supports even after Phase I.

The cross-entity ranking puts **Hyp-1B at the top (34 supports, 11
Strong)** after the Phase I, Phase K, Phase L, Phase M, and Phase N lifts, ahead of T-Inflammation
(31 supports, 6 Strong), T-Synergy (30 supports, 9 Strong), Hyp-3B (28 supports, 8
Strong) and the remaining adjudicated entities (T-Compartment-
suppression 17, T-Tau-attenuates 12, Hyp-2B 8, Hyp-0 7). The Phase N
N-001 row lifts T-Inflammation just past T-Synergy on raw count, but
T-Synergy retains more Strong-graded support (9 vs 6), so the new order
is a single Suggestive scope-validation nudge, not a confidence reordering. The
contradicted contest models (Hyp-1A 16 net, Hyp-2A -4 net, Hyp-3A
-27 net) sit at the bottom; Hyp-3A's net of -27 reflects the unanimous
mutual-exclusion contradiction described in the axis-3 verdict above,
and Hyp-1A's net of 16 (down from 18 at Phase H, the dip held by the six
Phase I rows that each contradict the qualitatively-identical-amyloid-
response prediction by per-substate replication of the interaction
sign-reversal, then nudged up one by the margin-neutral Phase N N-001
support).

## Open questions and testable predictions

The ledger's `notes` column on each Moderate / Suggestive row sketches
what additional evidence would upgrade the row to Strong, what
experiment would resurrect a contradicted model, or what observation
would lift a low-support theme. The numbered list below distils those
into the project's most informative outstanding questions, grouped by
the structural slot each addresses.

1. **Axis-2 TF + kinase non-finding: positive route to closure.** The
   axis-2 TF + kinase non-finding rows (H2-029 / H2-030 / H2-031 at
   the TF layer; H2-032 / H2-033 / H2-074 at the kinase layer) are
   honest absences at FDR<0.10 in the axis-restricted view. The
   non-finding currently supports T-Compartment-suppression by
   *absence* (no transcriptional or kinase driver explains the
   synaptic GO term suppression at this resolution; the mechanism is
   plausibly intercellular only). Predictions that would upgrade these
   rows from Suggestive to Strong, or open a new contest: (a) a
   curated synaptic-clearance prior collection (similar to the KSN /
   CollecTRI priors used here but oriented at synaptic adhesion +
   engulfment ligands) applied to the same axis-2 axis-restricted
   universe; (b) a Cdk5-substrate-restricted GSEA at the axis-2
   contrasts (would directly test the Hyp-0 integrator framing at axis
   2 against the synaptic suppression evidence); (c) a per-substate
   NEBULA fit restricted to MG_DAM cells at axis-2 contrasts to
   surface within-substate transcriptional drivers that whole-
   microglia NEBULA dilutes.
2. **Hyp-2A scope ambiguity: Pros1-Mertk vs the axis-restricted absence.**
   The Hyp-2A vs Hyp-2B verdict resolves in Hyp-2B's favour on axis-restricted
   scope, but the H2-040 cross-tool Pros1-Mertk row keeps Hyp-2A
   partially alive as a TAM-kinase efferocytosis signal at specific
   microglia-microglia nodes. Predictions that would settle the scope
   ambiguity: (a) a cell-type-resolved Pros1-Mertk validation in
   GeoMx microglia AOIs (would confirm whether the cross-tool signal
   is genuinely microglia-microglia or driven by a different sender);
   (b) a Mertk-knockout NLGF cross to directly test whether classical
   complement / TAM efferocytosis is functionally required for the
   synaptic suppression phenotype; (c) a re-analysis of the axis-2
   leader board with the cross-tool leader rule (`n_tools_sig_consistent_sign
   >= 2` in any genotype at any axis-2-relevant contrast) rather than
   the axis-restricted average score, to see whether Pros1-Mertk
   surfaces in the top 20 under the cross-tool scoring.
3. **Hyp-3A resurrection: what would the no-interaction hypothesis need
   to look like?** Hyp-3A is unanimously contradicted (0 supports, 26
   contradicts). The only way to resurrect Hyp-3A on the project's
   evidence would be (a) a re-pseudobulking under a different design
   matrix that eliminates the Gsk3b interaction signal (e.g. a
   genotype:batch interaction term that absorbs the interaction
   variance); (b) replication of the interaction signal in an
   independent cohort showing that the Gsk3b / Myc / axon-guidance
   pattern is cohort-specific rather than biology; (c) demonstration
   that the interaction-restricted gene universe is a poor proxy for
   true interaction biology (e.g. via a permutation null on the axis
   3 universe). The current evidence is too consistent across layers
   for Hyp-3A to be a viable alternative without one of these.
4. **T-Tau-attenuates support: predictions (a) and (b) RESOLVED by the
   Phase I per-state NF-kB attenuation test (section 18).** The theme
   stood at 5 supports / 3 Strong at the end of Phase H. Phase I
   executed predictions (a) and (b) jointly: (a) an NF-kB-target GSEA
   restricted to the union of CollecTRI's Rela / Nfkb1 / Nfkb2 / Rel /
   Relb / A0A979HLR9 target genes at the interaction contrast, and
   (b) a per-substate analysis covering ALL 4 microglia substates (not
   just MG_homeostatic). The verdict tier resolved to **Global
   attenuation**: 4 of 4 microglia substates independently carry NF-kB
   suppression at the interaction under the cross-layer gate (TF
   complex consensus < 0 AND GSEA union NES < 0 AND >=1 layer at
   FDR<0.10 within state), with MG_DAM as the lead carrier (consensus
   p=7e-11, GSEA padj=0.027) and MG_homeostatic close behind
   (consensus p=5e-8, GSEA padj=0.091). The section 11 substate-
   composition ANOVA (tau:nlgf F=1.03 p=0.330) formally ruled out
   "state-driven-by-abundance" as the alternative explanation,
   establishing that the per-state signal is transcriptional rather
   than compositional. The Phase I rows (I-001..I-006) lifted
   T-Tau-attenuates from 5 supports / 3 Strong to 11 / 7. Prediction
   (c), the tau-knockout-on-NLGF arm, remains unresolved because
   the project does not have that cohort; it remains the highest-
   value experimental follow-up for confirming tau's causal role in
   the attenuation rather than ruling out other genotype confounders.
5. **Hyp-0 integrator framing: Cdk5-substrate enrichment as the load-
   bearing test.** Hyp-0 is the strongest cross-axis claim with 5
   Strong rows distributed across axes (the most evenly distributed
   entity of all 11). The integrator framing predicts that Cdk5
   substrates (as defined by the KSN substrate set; 81 / 263 / 153
   sites at the three axes) are enriched in the top-N differentially
   expressed genes / phosphorylated sites at every axis. A direct
   test would be a Cdk5-substrate-restricted GSEA at all three axes
   (predicted significant enrichment at every axis if Hyp-0 is correct;
   significant at axis 3 only if Hyp-0 is an axis-3-anchored Gsk3b /
   tau-kinase covariate rather than a true integrator). The H2
   ledger does not currently emit such a test as a claim row; a
   future H-phase round could add it.

## Integrated model statement

The project's evidence currently favours the following integrated
biological reading of how amyloid reshapes microglia under different
tau backgrounds. Each clause is grounded in the per-axis contest
verdict (17.3.1) and cross-axis theme tally (17.3.2); none of it is
asserted beyond what the H2 ledger and H3 adjudication support.

**Per-axis verdicts.** At the amyloid-activation axis (axis 1), the
project favours **Hyp-1B over Hyp-1A** by net-support margin 18 (lifted
from 6 by the Phase I per-state NF-kB attenuation test reported in
section 18): amyloid drives a robust NF-kB-led inflammatory program
in MAPTKI, and tau on the amyloid background SUPPRESSES (rather than
amplifies) that program at the interaction contrast, with Rela
sign-reversal as the load-bearing signature now corroborated by
per-substate replication in all four microglia substates. At the
synaptic-suppression axis (axis 2),
the project favours **Hyp-2B over Hyp-2A** on axis-restricted scope by
net-support margin 12: NLGF-driven synaptic suppression is mediated
by TREM2-mediated clearance plus APP-fragment uptake plus synaptic-
adhesion modulation, NOT by canonical C1q / C3 complement-tagging,
although the cross-tool Pros1-Mertk signal narrowly keeps a TAM-
kinase efferocytosis route alive at specific microglia-microglia
nodes. At the interaction axis (axis 3), the project favours **Hyp-3B
over Hyp-3A** by net-support margin 55: a qualitatively new mechanism
emerges at the tau x amyloid interface that is not present in either
single-insult arm, anchored by the project's strongest single-kinase
result (Gsk3b activation, padj 0.00152) and reinforced by coordinated
TF biosynthetic suppression (Myc / Creb1 / Tp53 / Jun) and axon-
guidance LR rewiring at the LR layer.

**Cross-axis integration.** Layered on top of the three per-axis
verdicts, the cross-axis themes rank **T-Inflammation (31 supports, 6
Strong) > T-Synergy (30 supports, 9 Strong) > T-Compartment-
suppression (17 supports, 8 Strong) > T-Tau-attenuates (12 supports,
7 Strong)** as the project's currently best-supported cross-axis
readings, though T-Inflammation's narrow lead over T-Synergy on raw
count rests on the single Suggestive Phase N scope-validation row
(N-001) while T-Synergy retains more Strong-graded support (9 vs 6), so
the two remain effectively co-leading. T-Tau-attenuates lifted from its Phase H position of 5
supports / 3 Strong to 11 / 7 after the Phase I per-state NF-kB
attenuation test (section 18, rows I-001..I-006) demonstrated that the
whole-microglia Rela sign-reversal that anchored Phase H is
independently carried by all four microglia substates (MG_DAM as the
lead carrier, MG_homeostatic close behind, MG_IFN and MG_proliferative
weaker but sign-consistent), with the section 11 substate-composition
ANOVA (tau:nlgf F=1.03 p=0.330) formally ruling out the
compositional-artefact alternative. The Phase J causal-topology probe
(section 20, row J-002) adds one further Suggestive support, bringing
T-Tau-attenuates to 12 / 7 by crediting the recovered Gsk3b-rooted
upstream route to the NF-kB regulatory module. The Phase K SCENIC
data-driven regulon layer (section 21, rows K-001 / K-002) adds two
further Suggestive supports to T-Inflammation (lifting it to 28 / 6) by
recovering Spi1 and Rel as de novo motif-supported microglial regulons
independently of the section-14 CollecTRI prior; both rows are margin-
neutral by construction (they corroborate regulon EXISTENCE, not
activity direction, and the controlled prior-swap is mostly discordant),
so the three locked contest verdicts are unchanged. The Phase L GeoMx
spatial-deconvolution tissue-composition layer (section 22, rows
L-001 / L-002) then adds one further Suggestive support to
T-Inflammation (lifting it to 29 / 6) plus one honest non-corroboration
that feeds no model. L-001 records that the DAM microglial compartment
expands as a fraction of deconvolved tissue under amyloid in both tau
backgrounds (within-microglia DAM fraction 0.18/0.18 -> 0.79/0.93;
DAM abundance logFC +0.40 at nlgf_in_maptki, FDR 3.5e-5, and +0.53 at
nlgf_in_p301s, FDR 2.4e-7) -- a composition-scoped existence claim,
deliberately distinct from the section-8 snRNAseq differential
expression, that like K-001 supports BOTH amyloid models by
construction and so is margin-neutral (the amyloid margin holds at 18).
L-002 records the NULL composition interaction (0 of 10 cell types
FDR<0.10 at the interaction, best Astrocyte adj.P 0.25; Microglia_DAM
interaction logFC +0.13, t=0.98, p=0.33) as an honest non-corroboration
with empty supports / contradicts -- NOT as evidence for any model --
and is the compositional counterpart to the section-11 substate-
composition ANOVA, reinforcing that the interaction signal is
transcriptional rather than compositional. Both Phase L rows are graded
Suggestive rather than Strong because the composition estimate is
model-dependent (a sorting-biased microglial reference, no histology
ground truth, kappa=35 substate collinearity that suppresses the IFN
and homeostatic interaction terms, whole-tissue non-plaque ROIs), so
the three locked contest verdicts remain unchanged. The Phase M
snRNAseq activation-pseudotime trajectory layer (section 23, rows
M-001 / M-002) then adds one further Suggestive support to
T-Inflammation (lifting it to 30 / 6) and one to T-Synergy (lifting it
to 30 / 9), introducing a new `dynamics` evidence class -- how far
each microglion sits along the homeostatic->DAM activation continuum
(Slingshot on the harmony embedding, root = homeostatic) -- that
neither the discrete substate expression DE nor the discrete tissue
composition (Phase L) can resolve. M-001 records that the amyloid main
effect advances this progression under NLGF in both tau backgrounds
(mean homeostatic->DAM pseudotime logFC +7.87 at nlgf_in_maptki, FDR
8.9e-7, and +10.32 at nlgf_in_p301s, FDR 3.7e-8; fraction past the
DAM-onset threshold 4%/4% -> 40%/59%), and -- like K-001 and L-001 --
supports BOTH amyloid models by construction (a shared
amyloid-activation premise that does not discriminate the
tau-modulation contest), so the amyloid margin holds at 18. M-002
records the project's first POSITIVE and significant interaction in a
non-expression layer: the P301S background supra-additively amplifies
the amyloid-driven progression (interaction on mean pseudotime logFC
+2.46, FDR 0.077; fraction-past-threshold interaction +0.19, FDR
0.013), and the load-bearing progression-vs-composition decomposition
attributes it to genuine within-state PROGRESSION (progression channel
+2.30, FDR 0.077, ~94% of the total) rather than the discrete
composition reshuffling that was null at L-002 (composition channel
+0.13, FDR 0.81) -- with even homeostatic-labelled cells advancing at
the interaction (within_homeostatic +2.68, FDR 0.077), which cannot be
a composition artefact. Per the user-approved M5 gate (option B) this
interaction is recorded THEMATICALLY against T-Synergy only -- not
scored against the Hyp-3A / Hyp-3B contest -- because pseudotime is
mechanism-agnostic (it evidences synergy in progression RATE, not the
specific Gsk3b / Myc / axon-guidance mechanism Hyp-3B names), so the
interaction margin holds at 55. Both Phase M rows are graded
Suggestive rather than Strong because pseudotime is a single inferred
ordering partly confounded with composition (guardrail #1): the
asserted homeostatic root is NOT the most-potent state by any
root-free proxy (RNA entropy 3/4, n_genes 3/4, CytoTRACE2 4/4), so the
axis is an ACTIVATION ordering rather than developmental potency or
time; and 1 of 3 cross-check methods (scanpy DPT, rho -0.09) disagrees
while the two curve / fate methods agree (nested Slingshot rho 0.98,
CellRank2 rho 0.57). The three locked contest verdicts therefore
remain unchanged at 18 / 12 / 55. The Phase N cross-cell-type
interaction-specificity layer (section 24, rows N-001 / N-002) then adds
one further Suggestive support to T-Inflammation (lifting it to 31 / 6)
plus one honest reframing that feeds no model, introducing a new
`specificity` evidence class -- the tau x amyloid factorial refit in all
six broad cell-type units (Astrocyte, Microglia-pooled, Neuronal,
Oligodendrocyte, OPC, Vascular) x two estimators (pseudobulk +
single-cell NEBULA) x two power regimes (NATIVE / cell-count-MATCHED at
K=289 per unit x replicate), the project's first cross-cell-type readout
where every prior row scored microglia only. N-001 records that the
amyloid MAIN effect is microglia-LED even under cell-count matching
(MATCHED nlgf_in_maptki NEBULA Microglia 625 vs next-highest Astrocyte
218; pseudobulk 232 vs 29; microglia lead in both tau backgrounds and
both estimators DESPITE a 2.64x median-UMI depth disadvantage), a
comparative validation that the microglia-scoped ledger never had; like
K-001 / L-001 / M-001 it supports BOTH amyloid models by construction
(margin-neutral, amyloid holds at 18) and is the row that lifts
T-Inflammation past T-Synergy on raw count. N-002 records the load-
bearing reframing: the tau x amyloid INTERACTION is NOT
microglia-specific -- at MATCHED power the gene-level interaction
collapses to ~0 in all six units including microglia (sole survivor
Oligodendrocyte, 6 genes NEBULA, a NON-microglial type; the 4 NATIVE
microglia interaction genes do not survive matching), the per-unit
interaction profiles are mutually distinct (Microglia-vs-unit Spearman
0.08-0.15) rather than a shared programme, and the broadest pathway-
level interaction response is Neuronal, not Microglia. Per the
user-approved N5 gate (option C, feed both halves) N-002 FEEDS NO MODEL
(empty supports / contradicts, mirroring L-002): it is NOT scored
against Hyp-3A because the full-power microglia interaction the axis-3
contest rests on stays real (the MATCHED ~0 is a comparability statement
at the deliberately severe K=289 downsample, not a positive
no-interaction result), NOT for Hyp-3B because it is mechanism-agnostic
and about other cell types, and NOT for T-Synergy because the distinct
per-unit profiles mean the breadth is not a shared supra-additive
programme (tagging it would overclaim cross-cell-type mechanistic unity,
unlike M-002's genuine within-microglia progression synergy). Both Phase
N rows are graded Suggestive because MATCHED equalises cell COUNT not
transcriptome DEPTH (the Neuronal pathway dominance is read only as
"broadest outside microglia", never a quantitative neuronal-magnitude
claim) and the readout is single-dataset; the reframing is that the
amyloid response validates the microglia focus while the divergence
readout does not, so the three locked contest verdicts remain unchanged
at 18 / 12 / 55. Finally, the Phase O gene-level
differential-dynamics layer (section 23, row O-001) resolves the M-002
aggregate interaction to genes: a tradeSeq NB-GAM difference-of-differences
Wald along the homeostatic→DAM pseudotime finds 110 genes (FDR<0.10) whose
expression-trajectory *shape* is tau × amyloid-modulated: the project's first
gene-level positive interaction, and a non-redundant `dynamics_de` evidence
class that the section-24 matched-power static interaction cannot detect (0 of
110 reach FDR<0.10 in either estimator). Per the user-approved O5 gate (ALT-1,
always-margin-neutral) O-001 FEEDS NO MODEL (empty supports / contradicts,
mirroring L-002 / N-002): the pre-registered overlap test, locked before the
genes were interpreted, found Hyp-3B's named effectors absent (Gsk3b and Myc
both missing; canonical axon-guidance 1 of 110, at chance), so the
scored-contest branch was not triggered, and because O-001 is the gene-level
companion of M-002 from the same tradeSeq fit, scoring T-Synergy would
double-count that theme. It is held at Suggestive by a quality caveat, roughly
a quarter of the 110 genes are ambient (7) or ribosomal / translation (20)
confounds, so no clean microglial mechanism is asserted, and the three locked
contest verdicts once more remain unchanged at 18 / 12 / 55. The per-axis-anchoring analysis
(17.3.2) flags that all three of the highest-support themes draw
90-94% of their support from a single axis, so the cross-axis framing
is mostly a re-statement of the per-axis verdicts in theme-overlay
language, each theme is the cross-axis-friendly framing of its
dominant axis's favoured-model verdict. The genuine cross-axis claim is **Hyp-0
(Cdk5 cross-axis integrator)**, which is the most evenly distributed
entity in the adjudication (1 / 1 / 2 / 3 split across axes and the
cross-axis section, 5 Strong rows total). Hyp-0's strongest evidence is
the Cdk5 top-5-at-all-three-axes ranking with substrate footprint
that scales with the axis universe size; its testable prediction
(Cdk5-substrate enrichment at every axis) is not yet directly tested
by a claim row and is the project's highest-value follow-up.

**Integrated reading.** Putting these together: in MAPTKI mice,
amyloid drives a canonical DAM inflammatory program with NF-kB at its
transcriptional core (Hyp-1A's matched-direction support is genuine; the
program is real). On top of that, tau pathology in the P301S
background reshapes how amyloid acts on microglia in two qualitatively
distinct ways. First, tau ATTENUATES the amyloid-driven NF-kB
inflammation specifically at the interaction (T-Tau-attenuates,
anchored at axis 1's Hyp-1B sign-reversal verdict, with the per-state
test in section 18 establishing that this attenuation is GLOBAL across
microglia substates rather than localised to one substate or driven by
a substate-composition shift). Second, at the
interaction itself, a qualitatively new mechanism emerges that is not
the arithmetic sum of single-insult arms: Gsk3b activation,
biosynthetic / translational TF suppression, axon-guidance LR
rewiring, mitochondrial / ribosomal pathway reorganisation
(T-Synergy, anchored at axis 3's Hyp-3B verdict). Independently, the
synaptic compartment is suppressed by NLGF via a TREM2 / APP-fragment-
mediated route rather than classical complement (T-Compartment-
suppression, anchored at axis 2's Hyp-2B verdict, with a residual
TAM-kinase route partially surviving at specific microglia-microglia
nodes). Cdk5 ties the three axes together as a candidate cross-axis
kinase integrator (Hyp-0), most plausibly because Cdk5 substrates
participate in all three axis biologies (DAM inflammatory regulation,
synaptic adhesion / engulfment, tau / cytoskeletal phosphorylation).
This reading is what the project's evidence currently supports; it is
NOT an authoritative scientific conclusion, and the 17.4 testable
predictions name the experiments that would either harden, refine, or
overturn it.
