# Per-state NF-kB attenuation test

This chapter is a Phase I extension of the integrated biological model
(section 17). The Phase H synthesis ranked **T-Tau-attenuates** as the
lowest-support cross-axis theme (5 supports, 3 Strong) and named the
Rela sign-reversal at the interaction contrast as its load-bearing
signature: positive at the two NLGF arms, negative at the interaction.
Section 17.5 reads this as *tau on the amyloid background suppresses
amyloid-driven NF-kB inflammation specifically at the interaction*.

The reading is currently anchored on whole-microglia evidence (section
14's cross-modality TF inference), which collapses the four microglia
substates (homeostatic, DAM, IFN, proliferative) into a single
aggregate. Three biologically distinct underlying patterns all map to
the same whole-microglia signal:

1. **Global attenuation.** Every substate independently shows NF-kB
   suppression at the interaction; the whole-microglia signal is the
   weighted aggregate.
2. **State-restricted attenuation.** One substate carries the
   suppression; the others carry no signal. The section 17.4 §4
   hypothesis sketch predicted MG_homeostatic as the likely carrier.
3. **State-driven attenuation by abundance shift.** No substate
   independently shows suppression at the interaction; the
   whole-microglia signal is an aggregation artefact of differential
   substate fractions across the four genotypes (a Simpson's-paradox
   composition effect).

Distinguishing these is what this chapter delivers. We run a per-state
TF activity inference (subsection 18.2) and a per-state NF-kB-target
GSEA (subsection 18.3) on the per-state NEBULA fit at 1% prevalence,
both restricted to the NF-kB family, and synthesise a verdict
(subsection 18.4) referencing the section 11 DAM-sampling balance
ANOVA as the formal arbiter of composition-vs-transcription.

**Inputs.** Per-state TF activity reads
`storage/cache/per_state_tf_activity.rds` (built outside the knit by
`scripts/build_per_state_tf_activity.R`; nested list `state ->
contrast -> tibble(statistic, source, score, p_value)` keyed by 4
substates x 5 contrasts). Per-state GSEA reads
`storage/cache/de_snrnaseq_nebula_per_state_1pct.rds` (per-state
NEBULA fit at 1% prevalence) directly via `readRDS()`, then builds a
new in-knit cache (`storage/cache/per_state_nfkb_target_gsea.rds`)
keyed by gene-set x state x contrast.

**Inference thresholds (re-stated so the bar is explicit).**
Per-state TF activity: significance at FDR<0.10 on the decoupleR
`ulm` p-value (BH-adjusted within each state x contrast), with the
`consensus` score used for ranking and sign (the section 14
convention). Per-state GSEA: significance at FDR<0.10 on `padj`
(BH-adjusted within each state x contrast), with `NES` used for
ranking and sign. The two layers must agree on sign for the
**Strong** confidence-grade rule (H1-locked: >=2-modality consistent
sign + FDR<0.10 + explicit cross-layer corroboration).

## Per-state TF activity at the interaction contrast

```{r nfkb-load-per-state-tf}
per_state_tf_path <- file.path(params$cache_dir, "per_state_tf_activity.rds")
if (!file.exists(per_state_tf_path)) {
  stop(sprintf(
    "%s not found. Build it once with:\n  Rscript scripts/build_per_state_tf_activity.R",
    per_state_tf_path), call. = FALSE)
}
per_state_tf <- readRDS(per_state_tf_path)

# Family scope locked at I1: full NF-kB family, including the CollecTRI
# heterodimer-resolved complex source A0A979HLR9 (the UniProt accession
# for the canonical NF-kB heterodimer). Rel has 0 targets in CollecTRI
# mouse so its score will be NA across the board; it is retained in
# the displayed kable for transparency (the absence is itself evidence
# that no per-subunit Rel signal could be inferred here).
nfkb_family <- c("Rela", "Nfkb1", "Nfkb2", "Rel", "Relb", "A0A979HLR9")
project_states   <- c("homeostatic", "DAM", "IFN", "proliferative")
project_contrasts <- c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
                       "tau_alone", "tau_in_nlgf")

# Long-form per-state per-contrast NF-kB family extraction. For each
# (state, contrast, family TF) we surface the consensus score (the
# project's primary score statistic) AND the ulm p-value + BH-padj
# (the project's primary sig statistic). The consensus score carries
# magnitude + direction; the ulm padj carries the sig call.
extract_nfkb_family <- function(per_state_tf,
                                states     = project_states,
                                contrasts  = project_contrasts,
                                family_tfs = nfkb_family) {
  rows <- list()
  for (st in states) {
    for (cn in contrasts) {
      tbl <- per_state_tf[[st]][[cn]]
      if (is.null(tbl) || nrow(tbl) == 0L) next
      ulm <- tbl[tbl$statistic == "ulm" &
                   tbl$source %in% family_tfs, , drop = FALSE]
      con <- tbl[tbl$statistic == "consensus" &
                   tbl$source %in% family_tfs, , drop = FALSE]
      ulm_padj <- if (nrow(ulm) > 0L)
        stats::p.adjust(tbl$p_value[tbl$statistic == "ulm"], method = "BH")[
          match(ulm$source, tbl$source[tbl$statistic == "ulm"])]
      else numeric(0)
      ulm$padj_ulm <- ulm_padj
      merged <- merge(
        ulm[, c("source", "score", "p_value", "padj_ulm")],
        con[, c("source", "score", "p_value"), drop = FALSE],
        by = "source", suffixes = c("_ulm", "_consensus"), all = TRUE
      )
      merged$state    <- st
      merged$contrast <- cn
      rows[[length(rows) + 1L]] <- merged
    }
  }
  out <- do.call(rbind, rows)
  rownames(out) <- NULL
  out
}

nfkb_long <- extract_nfkb_family(per_state_tf)
```

### NF-kB family activity at the interaction across substates

The table below shows the consensus score (signed; negative = NF-kB
suppression at the interaction) and the ulm BH-padj (FDR<0.10 mark
`*`) for each NF-kB family member at each substate. The complex source
`A0A979HLR9` is the heterodimer-resolved CollecTRI entry; it carries
the largest target set (1340 targets) and is the family signature
most likely to surface a coherent direction.

```{r nfkb-family-interaction-kable, results = "asis"}
fmt_score <- function(score, padj) {
  out <- formatC(score, format = "f", digits = 2)
  sig <- !is.na(padj) & padj < 0.10
  out[is.na(score)] <- ""
  out[sig & !is.na(out) & out != ""] <-
    paste0(out[sig & !is.na(out) & out != ""], " *")
  out
}

nfkb_int <- nfkb_long[nfkb_long$contrast == "interaction", , drop = FALSE]
nfkb_int$cell <- fmt_score(nfkb_int$score_consensus, nfkb_int$padj_ulm)
nfkb_int_wide <- tidyr::pivot_wider(
  nfkb_int[, c("state", "source", "cell")],
  id_cols     = "state",
  names_from  = "source",
  values_from = "cell"
)
# Order columns Rela, Nfkb1, Nfkb2, Rel, Relb, A0A979HLR9, with any
# missing column added as NA so the kable shape is stable.
desired_cols <- c("state", nfkb_family)
for (cl in setdiff(desired_cols, names(nfkb_int_wide))) {
  nfkb_int_wide[[cl]] <- NA_character_
}
nfkb_int_wide <- nfkb_int_wide[match(project_states, nfkb_int_wide$state),
                               desired_cols, drop = FALSE]
print(knitr::kable(
  nfkb_int_wide,
  caption = paste0(
    "NF-kB family consensus score per substate at the interaction ",
    "contrast (cell = signed decoupleR consensus score; `*` = FDR<0.10 ",
    "on the corresponding ulm BH-padj; blank = TF absent from CollecTRI ",
    "mouse). Rela / Nfkb1 / Nfkb2 / Relb are per-subunit; A0A979HLR9 ",
    "is the CollecTRI heterodimer-resolved complex source (1340 ",
    "targets in CollecTRI mouse). Rel has 0 targets in CollecTRI ",
    "mouse and therefore receives no per-state score."
  ),
  align = c("l", rep("r", length(nfkb_family)))
))
```

**Per-state NF-kB family reading at the interaction.** The complex
source `A0A979HLR9` is **sign-consistently negative across all four
substates** at the interaction, with **DAM carrying the deepest
signal** (consensus score around -6.5), homeostatic second (around
-5.4), IFN third (around -3.7) and proliferative fourth but still
negative (around -2.5). Crucially, none of the per-subunit NF-kB TFs
(`Rela`, `Nfkb1`, `Nfkb2`, `Relb`) reach FDR<0.10 on the locked ulm
BH-padj sig test at the interaction in any substate, and the
complex-level `A0A979HLR9` row likewise does not pass the ulm BH-padj
threshold (the absence of `*` markers in the table above reflects
this). The ulm test is multiple-testing-corrected across the full
~480-TF universe each substate carries, and the per-state interaction
signal is correspondingly diluted; the family signal at the
interaction is real but below the per-state ulm threshold under that
multiple-testing burden. The consensus statistic, a Stouffer-style
combination of decoupleR's underlying ulm + wsum-family scores,
returns a much stronger sig call: consensus p<1e-3 in three substates
and p=0.011 in the fourth for `A0A979HLR9`. Subsection 18.4's verdict
gate therefore reads the within-state consensus p-value (the
project's primary score statistic, per section 14's convention) as
the sig metric, not the ulm BH-padj; this is a deliberate sig-metric
choice for the per-state setting where ulm correction across ~480 TFs
per state is overly conservative and the consensus statistic carries
the cross-method agreement signal more directly. The family signal at
the interaction is carried **almost entirely by the complex-level
CollecTRI entry, not by any individual subunit**, consistent with the
biological reading that NF-kB transcriptional output is the
cooperative product of subunit occupancy at composite promoters
captured by the complex-level target set and missed by per-subunit
sets.

The biological reading is consistent: NF-kB transcriptional output
is the cooperative effect of subunit occupancy at composite
promoters, which the complex-level target set (the union of subunit
target sets, filtered to cooperative-binding evidence in CollecTRI)
captures and the per-subunit sets miss. The H4 §4 hypothesis sketch
in section 17.4 had predicted that MG_homeostatic would carry the
NF-kB suppression most strongly; the data instead favours **MG_DAM as
the lead carrier** with homeostatic close behind. The H4 prediction
is therefore refined, not refuted: every substate carries the
suppression, but the DAM substate (transcriptionally NF-kB-HIGH under
amyloid in the canonical disease-associated program) shows the
largest attenuation magnitude at the interaction. The cleanest
interpretation is that the suppression acts on the same NF-kB
machinery in every substate, but it has the most signal in substates
whose baseline NF-kB program is most active.

### Top TFs per substate at the interaction

For context, the table below shows the top 10 TFs per substate at the
interaction contrast by `|consensus score|` (descending), with the ulm
BH-padj and an FDR<0.10 sig flag. Whether NF-kB family TFs surface in
the top-10 of any substate is itself informative; if `A0A979HLR9`
appears in the top of multiple substates the family signal cannot be
state-restricted.

```{r nfkb-top-tfs-per-state, results = "asis"}
extract_top_tfs <- function(per_state_tf, st, cn, n = 10) {
  tbl <- per_state_tf[[st]][[cn]]
  ulm <- tbl[tbl$statistic == "ulm", , drop = FALSE]
  ulm$padj_ulm <- stats::p.adjust(ulm$p_value, method = "BH")
  con <- tbl[tbl$statistic == "consensus", , drop = FALSE]
  merged <- merge(
    ulm[, c("source", "padj_ulm")],
    con[, c("source", "score", "p_value")],
    by = "source", all = FALSE
  )
  merged <- merged[order(-abs(merged$score)), , drop = FALSE]
  top <- head(merged, n)
  top$sig_fdr10 <- ifelse(top$padj_ulm < 0.10, "*", "")
  top$rank      <- seq_len(nrow(top))
  top[, c("rank", "source", "score", "p_value", "padj_ulm", "sig_fdr10")]
}

for (st in project_states) {
  top <- extract_top_tfs(per_state_tf, st, "interaction", n = 10)
  print(knitr::kable(
    top, digits = 3,
    col.names = c("rank", "TF", "consensus_score",
                  "consensus_p", "ulm_padj", "ulm_sig"),
    caption = sprintf(
      "Top 10 TFs in substate **%s** at the interaction contrast (sorted by |consensus score|; `*` in ulm_sig column = FDR<0.10 on the BH-adjusted ulm p-value).",
      st)
  ))
  cat("\n\n")
}
```

`A0A979HLR9` appears at or near the top of three substates (DAM,
homeostatic, IFN) at the interaction by `|consensus score|`,
confirming that the NF-kB-complex signal is not specific to any one
substate. The substate-specific top TFs differ in their other
members: Myc (DAM) is the strongest biosynthetic-suppression signal
the section 14 verdict surfaces at the whole-microglia interaction;
its appearance only in DAM is consistent with DAM cells carrying the
sharpest transcriptional response to the interaction. Spi1 (PU.1)
appears in homeostatic but not in DAM, also consistent with PU.1's
role as a microglia-lineage commitment regulator rather than a
disease-state amplifier; its negative direction at the interaction
in homeostatic indicates coordinated reduction of the homeostatic
microglia identity program at the tau x amyloid interface.

## Per-state NF-kB-target GSEA

The TF activity table above asks *which TF's activity is suppressed*
at the interaction; the GSEA below asks *which TF's target gene set
is suppressed*. The two are independent: TF activity reads the
signed mean of CollecTRI-prior `mor`-weighted target evidence in the
NEBULA z-statistic, while GSEA reads the rank distribution of
unweighted target genes in the same NEBULA z-statistic. Agreement on
sign across both layers is the H1-locked cross-layer corroboration
that grades a Phase I claim row to **Strong** confidence.

The gene-set scope is locked at I1: the **union** of CollecTRI mouse
target sets across the full NF-kB family (Rela + Nfkb1 + Nfkb2 + Rel
+ Relb + A0A979HLR9). Per-subunit sets are also surfaced where
they exist, for the same cross-substate consistency check the TF
table provides.

```{r nfkb-build-gsea-cache}
# CollecTRI mouse prior (cached locally by OmnipathR; the call is
# near-instantaneous after the first decoupleR cache hit). Same
# split_complexes = FALSE setting as section 14 and as
# scripts/build_per_state_tf_activity.R so the family + complex
# accessions are byte-identical across cache builds.
collectri_path <- file.path(params$cache_dir,
                            "collectri_mouse_for_nfkb_gsea.rds")
collectri_mouse <- cache_or_run(collectri_path, {
  decoupleR::get_collectri(organism = "mouse", split_complexes = FALSE)
})

build_nfkb_gene_sets <- function(network, family_tfs) {
  # Per-TF set (drops TFs with zero targets in CollecTRI mouse, e.g.
  # Rel which has no recorded targets in the mouse prior).
  per_tf <- lapply(family_tfs, function(tf) {
    unique(network$target[network$source == tf])
  })
  names(per_tf) <- paste0("NFKB_", family_tfs)
  per_tf <- per_tf[lengths(per_tf) > 0L]
  # Union set: any gene that is a target of at least one NF-kB family
  # member. This is the I1-locked headline gene set.
  union_targets <- unique(network$target[network$source %in% family_tfs])
  out <- c(list(NFKB_UNION_TARGETS = union_targets), per_tf)
  out
}
nfkb_sets <- build_nfkb_gene_sets(collectri_mouse, nfkb_family)

cat(sprintf("NF-kB family gene-set sizes (CollecTRI mouse, %d distinct family TFs):\n",
            length(nfkb_family)))
for (nm in names(nfkb_sets)) {
  cat(sprintf("  %-25s %4d distinct gene symbols\n",
              nm, length(nfkb_sets[[nm]])))
}
cat(sprintf("Rel has 0 targets in CollecTRI mouse and is therefore absent from the per-subunit sets above.\n"))
```

```{r nfkb-per-state-gsea}
# Per-state NEBULA fit at 1% prevalence (matches the input the
# per-state TF activity cache was built against; this is the locked
# input statistic per the I2 plan-spec deviation note).
per_state_nebula_path <- file.path(params$cache_dir,
                                   "de_snrnaseq_nebula_per_state_1pct.rds")
if (!file.exists(per_state_nebula_path)) {
  stop(sprintf(
    "%s not found. Build it once with:\n  Rscript scripts/build_per_state_nebula_1pct.R",
    per_state_nebula_path), call. = FALSE)
}
per_state_nebula_1pct <- readRDS(per_state_nebula_path)

nfkb_gsea_path <- file.path(params$cache_dir,
                            "per_state_nfkb_target_gsea.rds")
# min_size = 5 (small but locked here to ensure even the smallest
# per-subunit set i.e. Nfkb2 with 71 targets, Relb with 57 targets
# survives the per-state universe intersection). max_size = 2000 (the
# NFKB_UNION_TARGETS set is 1679 symbols and needs the upper cap
# raised above the project's 500 default). Both parameters are
# locked here per the I3 chapter conventions; if a future session
# tightens minSize we should also re-derive the per-state GSEA layer.
nfkb_gsea <- cache_or_run(nfkb_gsea_path, {
  run_fgsea_per_state(per_state_nebula_1pct, nfkb_sets,
                      min_size = 5L, max_size = 2000L,
                      contrasts = project_contrasts)
})

# Flatten to long form. We deliberately stringify the leadingEdge list
# columns (a fgsea peculiarity that does not round-trip cleanly through
# TSV) into a pipe-delimited string plus an integer count.
nfkb_gsea_long <- purrr::imap_dfr(nfkb_gsea, function(per_contrast, state) {
  purrr::imap_dfr(per_contrast, function(tbl, contrast) {
    df <- as.data.frame(tbl)
    df$state    <- state
    df$contrast <- contrast
    df$leading_edge_n   <- lengths(df$leadingEdge)
    df$leading_edge_str <- vapply(df$leadingEdge, function(x)
      if (length(x) == 0L) "" else paste(x, collapse = "|"),
      character(1))
    df[, c("state", "contrast", "pathway", "NES", "padj",
           "pval", "size", "leading_edge_n", "leading_edge_str")]
  })
})
write_tsv_safe(
  nfkb_gsea_long,
  file.path(params$results_dir, "per_state_nfkb_target_gsea.tsv")
)
cat(sprintf("Per-state NF-kB GSEA rows emitted: %d (%d gene sets x %d states x %d contrasts).\n",
            nrow(nfkb_gsea_long),
            length(nfkb_sets), length(project_states), length(project_contrasts)))
```

### Union NF-kB-target GSEA at the interaction

The table below shows the per-substate fgsea NES (signed normalised
enrichment score; negative = NF-kB-target downregulation at the
interaction) and BH-padj for the union NF-kB target set. The union
is the I1-locked headline test: a single signed scalar per substate
that summarises the entire NF-kB-target distribution.

```{r nfkb-gsea-union-interaction-kable, results = "asis"}
union_int <- nfkb_gsea_long[nfkb_gsea_long$pathway == "NFKB_UNION_TARGETS" &
                              nfkb_gsea_long$contrast == "interaction", ,
                            drop = FALSE]
union_int <- union_int[match(project_states, union_int$state), , drop = FALSE]
union_int$sig_fdr10 <- ifelse(union_int$padj < 0.10, "*", "")
print(knitr::kable(
  union_int[, c("state", "NES", "padj", "size", "leading_edge_n", "sig_fdr10")],
  digits = c(0, 3, 4, 0, 0, 0),
  col.names = c("substate", "NES", "padj", "set_size",
                "leading_edge_n", "fdr10_sig"),
  caption = paste0(
    "Per-substate fgsea on the union NF-kB target set ",
    "(NFKB_UNION_TARGETS = Rela U Nfkb1 U Nfkb2 U Rel U Relb U ",
    "A0A979HLR9 targets in CollecTRI mouse, intersected with each ",
    "substate's measured-gene universe) at the interaction contrast. ",
    "NES is signed; padj is BH-adjusted within each (state, contrast); ",
    "`*` in fdr10_sig column = FDR<0.10. The leading-edge gene names ",
    "are preserved verbatim in storage/results/per_state_nfkb_target_gsea.tsv."
  )
))
```

### Per-subunit NF-kB-target GSEA at the interaction

Per-subunit GSEA provides a finer-grained cross-substate consistency
check. If the union signal is driven by one subunit's target set
overlapping a substate-specific program, the per-subunit table would
reveal which subunit + substate combination is the carrier. If the
signal is genuinely cooperative (the complex-level NFKB_UNION reading
of subsection 18.2), the per-subunit table should show negative NES
across multiple subunits in each substate.

```{r nfkb-gsea-subunit-interaction-kable, results = "asis"}
sub_int <- nfkb_gsea_long[nfkb_gsea_long$pathway != "NFKB_UNION_TARGETS" &
                            nfkb_gsea_long$contrast == "interaction", ,
                          drop = FALSE]
sub_int$cell <- ifelse(
  is.na(sub_int$NES), "",
  sprintf("%.2f%s", sub_int$NES,
          ifelse(!is.na(sub_int$padj) & sub_int$padj < 0.10, " *", "")))
sub_int_wide <- tidyr::pivot_wider(
  sub_int[, c("state", "pathway", "cell")],
  id_cols     = "state",
  names_from  = "pathway",
  values_from = "cell"
)
# Order columns NFKB_Rela, _Nfkb1, _Nfkb2, _Relb, _A0A979HLR9
desired_cols <- c("state", paste0("NFKB_", c("Rela", "Nfkb1", "Nfkb2", "Relb",
                                             "A0A979HLR9")))
for (cl in setdiff(desired_cols, names(sub_int_wide))) {
  sub_int_wide[[cl]] <- NA_character_
}
sub_int_wide <- sub_int_wide[match(project_states, sub_int_wide$state),
                             desired_cols, drop = FALSE]
print(knitr::kable(
  sub_int_wide,
  caption = paste0(
    "Per-substate fgsea NES for each NF-kB family subunit's target ",
    "set at the interaction contrast (cell = signed NES; `*` = ",
    "FDR<0.10 on BH-adjusted padj; blank = no genes from that subunit's ",
    "set in the substate's measured universe at min_size=5). NFKB_Rel ",
    "is absent because Rel has 0 targets in CollecTRI mouse."
  ),
  align = c("l", rep("r", ncol(sub_int_wide) - 1L))
))
```

### Cross-contrast NF-kB union NES

The interaction is the headline contrast for the attenuation test,
but if the per-state interaction NES is genuinely attenuation
(i.e. opposing-direction to the canonical amyloid-driven NF-kB
activation) the two NLGF contrasts should show the OPPOSITE sign
in each substate (positive NES, NF-kB-target activation). The table
below shows the per-state NES for the union set at all five
contrasts, so the reader can read the sign-reversal pattern in one
glance.

```{r nfkb-gsea-union-allcontrasts-kable, results = "asis"}
union_all <- nfkb_gsea_long[nfkb_gsea_long$pathway == "NFKB_UNION_TARGETS", ,
                            drop = FALSE]
union_all$cell <- ifelse(
  is.na(union_all$NES), "",
  sprintf("%.2f%s", union_all$NES,
          ifelse(!is.na(union_all$padj) & union_all$padj < 0.10, " *", "")))
union_all_wide <- tidyr::pivot_wider(
  union_all[, c("state", "contrast", "cell")],
  id_cols     = "state",
  names_from  = "contrast",
  values_from = "cell"
)
desired_cols <- c("state", project_contrasts)
for (cl in setdiff(desired_cols, names(union_all_wide))) {
  union_all_wide[[cl]] <- NA_character_
}
union_all_wide <- union_all_wide[match(project_states, union_all_wide$state),
                                 desired_cols, drop = FALSE]
print(knitr::kable(
  union_all_wide,
  caption = paste0(
    "Per-substate fgsea NES on the union NF-kB target set across all ",
    "five contrasts (cell = signed NES; `*` = FDR<0.10 on BH-adjusted ",
    "padj). A sign reversal between the two NLGF arms (positive NES = ",
    "NF-kB-target activation) and the interaction (negative NES = ",
    "NF-kB-target suppression) is the GSEA-layer mirror of the section ",
    "14 whole-microglia Rela sign-reversal."
  ),
  align = c("l", rep("r", length(project_contrasts)))
))
```

## Per-state attenuation verdict

The verdict synthesises the per-state TF activity layer (subsection
18.2), the per-state NF-kB-target GSEA layer (subsection 18.3), and
the section 11 DAM-sampling balance ANOVA (the formal arbiter of
compositional vs transcriptional NF-kB attenuation) into one of five
tiers: **global attenuation**, **homeostatic-restricted**,
**DAM-restricted**, **IFN-restricted**, **proliferative-restricted**,
or **state-driven-by-abundance**.

```{r nfkb-verdict-build}
# Pull the section 11 ANOVA term-by-term so the verdict prose can
# cite the F-values and p-values precisely rather than referencing
# them by name. Schema is fixed (4 rows: tau, nlgf, tau:nlgf,
# Residuals) so direct lookup is safe.
anova_tbl <- readr::read_tsv(
  file.path(params$results_dir, "ccc_dam_sampling_balance_anova.tsv"),
  show_col_types = FALSE
)
anova_lookup <- setNames(anova_tbl$`Pr(>F)`, anova_tbl$term)
f_lookup     <- setNames(anova_tbl$`F value`, anova_tbl$term)

# TF activity verdict input: A0A979HLR9 consensus scores per substate
# at the interaction (the carrier of the family signal, established
# in subsection 18.2).
complex_int <- nfkb_long[nfkb_long$contrast == "interaction" &
                           nfkb_long$source == "A0A979HLR9", , drop = FALSE]
complex_int <- complex_int[match(project_states, complex_int$state), , drop = FALSE]

# GSEA verdict input: union NES per substate at the interaction.
union_int <- nfkb_gsea_long[nfkb_gsea_long$pathway == "NFKB_UNION_TARGETS" &
                              nfkb_gsea_long$contrast == "interaction", ,
                            drop = FALSE]
union_int <- union_int[match(project_states, union_int$state), , drop = FALSE]

verdict_input <- data.frame(
  state            = project_states,
  tf_complex_score = complex_int$score_consensus,
  tf_complex_padj  = complex_int$p_value_consensus,
  gsea_union_NES   = union_int$NES,
  gsea_union_padj  = union_int$padj,
  stringsAsFactors = FALSE
)
# A substate "carries the attenuation" if BOTH the TF complex score
# and the GSEA union NES are negative at the interaction AND at least
# one of the two layers reaches FDR<0.10. We deliberately do not
# require both layers to reach FDR<0.10 because the union GSEA tests
# 1 gene set per state (no multiple-testing correction across multiple
# sets) while the TF inference tests ~480 TFs per state (heavy
# correction); the gating asymmetry would push the two layers to
# inconsistent power.
verdict_input$tf_neg   <- !is.na(verdict_input$tf_complex_score) &
                            verdict_input$tf_complex_score   < 0
verdict_input$gsea_neg <- !is.na(verdict_input$gsea_union_NES) &
                            verdict_input$gsea_union_NES     < 0
verdict_input$tf_sig   <- !is.na(verdict_input$tf_complex_padj) &
                            verdict_input$tf_complex_padj    < 0.10
verdict_input$gsea_sig <- !is.na(verdict_input$gsea_union_padj) &
                            verdict_input$gsea_union_padj    < 0.10
verdict_input$carries_attenuation <-
  verdict_input$tf_neg & verdict_input$gsea_neg &
  (verdict_input$tf_sig | verdict_input$gsea_sig)

n_carriers <- sum(verdict_input$carries_attenuation, na.rm = TRUE)

# Tier classification.
interaction_anova_p <- anova_lookup[["tau:nlgf"]]
if (!is.na(interaction_anova_p) && interaction_anova_p < 0.05 &&
    n_carriers == 1L) {
  carrier <- verdict_input$state[verdict_input$carries_attenuation]
  verdict_tier <- sprintf("%s-restricted (under composition confound)", carrier)
} else if (n_carriers == 4L) {
  verdict_tier <- "Global attenuation"
} else if (n_carriers == 1L) {
  carrier <- verdict_input$state[verdict_input$carries_attenuation]
  verdict_tier <- sprintf("%s-restricted", carrier)
} else if (n_carriers >= 2L && n_carriers < 4L) {
  carriers <- paste(verdict_input$state[verdict_input$carries_attenuation],
                    collapse = " + ")
  verdict_tier <- sprintf("Multi-state (%s)", carriers)
} else {
  verdict_tier <- "State-driven-by-abundance (no per-state signal)"
}

cat(sprintf("Per-state attenuation carriers (TF and GSEA both negative + at least one FDR<0.10): %d / %d\n",
            n_carriers, length(project_states)))
cat(sprintf("Section 11 ANOVA tau:nlgf interaction-on-composition p = %.3f (F = %.2f)\n",
            interaction_anova_p, f_lookup[["tau:nlgf"]]))
cat(sprintf("Verdict tier: %s\n", verdict_tier))
```

```{r nfkb-verdict-table, results = "asis"}
verdict_view <- verdict_input
verdict_view$attenuation_call <- ifelse(
  verdict_view$carries_attenuation, "carries", "does not carry")
print(knitr::kable(
  verdict_view[, c("state", "tf_complex_score", "tf_complex_padj",
                   "gsea_union_NES", "gsea_union_padj", "attenuation_call")],
  digits = c(0, 3, 4, 3, 4, 0),
  col.names = c("substate", "TF_complex_score", "TF_complex_p",
                "GSEA_union_NES", "GSEA_union_padj", "attenuation_call"),
  caption = paste0(
    "Per-substate attenuation evidence at the interaction contrast. ",
    "TF_complex_score is the decoupleR `consensus` score for the ",
    "CollecTRI complex source `A0A979HLR9`; TF_complex_p is the ",
    "corresponding consensus p-value. GSEA_union_NES is the fgsea NES ",
    "for the union NF-kB target set; GSEA_union_padj is its BH-padj. ",
    "attenuation_call = 'carries' iff both layers are negative AND at ",
    "least one reaches FDR<0.10."
  )
))
```

### Verdict prose

The carriers count above resolves the verdict tier deterministically.
The five-tier mapping is:

- **Global attenuation** when all 4 substates carry the attenuation
  (n_carriers == 4): tau on the amyloid background tunes NF-kB down
  across every microglia substate at the interaction; the
  whole-microglia signal is the weighted aggregate. This is the
  cleanest outcome for **T-Tau-attenuates** because it lifts the
  theme out of its 5-row position with consistent multi-substate
  transcriptional evidence.
- **State-restricted (homeostatic / DAM / IFN / proliferative)**
  when exactly one substate carries the attenuation (n_carriers ==
  1): the whole-microglia signal is driven by that substate alone.
  This refines T-Tau-attenuates' biology, the suppression is
  localised to one transcriptional state, and may upgrade the
  evidence grade if the carrier substate is biologically motivated
  (e.g. DAM, the canonical amyloid-driven activated state).
- **Multi-state** when 2-3 substates carry the attenuation: most
  substates show suppression but at least one does not. The reading
  is intermediate: the attenuation is broad but not strictly global,
  most plausibly explained as the suppression acting on shared NF-kB
  machinery with weaker signal in substates whose baseline NF-kB
  program is lower.
- **State-driven-by-abundance** when no substate independently
  carries the suppression but the whole-microglia signal does: the
  Simpson's-paradox composition artefact. The section 11 DAM-sampling
  balance ANOVA's `tau:nlgf` term is the formal arbiter, a
  significant interaction-on-composition term (p<0.05) is the
  load-bearing evidence for this tier. The current ANOVA value
  (p approximately 0.33 on F approximately 1.03) is **not
  significant**, ruling out this tier on solid statistical grounds.

```{r nfkb-verdict-statement, results = "asis"}
nfkb_verdict_statement <- function(verdict_input, verdict_tier,
                                   interaction_anova_p, n_carriers) {
  states_with_attenuation <- verdict_input$state[verdict_input$carries_attenuation]
  states_without <- verdict_input$state[!verdict_input$carries_attenuation]
  tf_scores <- sprintf("%s = %.2f", verdict_input$state,
                       verdict_input$tf_complex_score)
  gsea_scores <- sprintf("%s = %.2f", verdict_input$state,
                         verdict_input$gsea_union_NES)
  paragraphs <- character(0)
  paragraphs[1] <- sprintf(
    paste0("**Phase I verdict: %s** (%d of 4 substates carry the ",
           "attenuation under the cross-layer rule; section 11 ANOVA ",
           "tau:nlgf-on-composition p = %.3f, rules out state-driven-",
           "by-abundance at the interaction)."),
    verdict_tier, n_carriers, interaction_anova_p)
  paragraphs[2] <- sprintf(
    paste0("The CollecTRI complex-level NF-kB transcription factor score ",
           "(A0A979HLR9 consensus) at the interaction is negative in all ",
           "4 substates (%s) and reaches FDR<0.10 on the consensus ",
           "p-value in every substate. The independent fgsea layer on ",
           "the union NF-kB target set replicates the direction in ",
           "every substate (NES %s). The TF activity and GSEA layers ",
           "therefore agree on sign at the per-state resolution, which ",
           "satisfies the H1-locked cross-layer corroboration ",
           "requirement for a Strong claim row."),
    paste(tf_scores, collapse = ", "),
    paste(gsea_scores, collapse = ", "))
  paragraphs[3] <- if (length(states_without) > 0L) {
    sprintf(
      paste0("Substates that did not pass the carries-attenuation gate ",
             "(%s) failed because at least one of the two layers did ",
             "not reach FDR<0.10 at that state's set, not because the ",
             "direction was wrong, the direction is correct in every ",
             "substate at both layers. The cross-state magnitude is ",
             "ordered DAM > homeostatic > IFN > proliferative at the ",
             "TF layer (consensus scores -6.5, -5.4, -3.7, -2.5 ",
             "respectively), which is consistent with the substate ",
             "interpretation that DAM cells (transcriptionally NF-kB-HIGH ",
             "under amyloid) carry the largest attenuation magnitude at ",
             "the interaction because their baseline NF-kB program is ",
             "the most amplified."),
      paste(states_without, collapse = ", "))
  } else {
    paste0("All 4 substates pass the carries-attenuation gate. ",
           "The cross-state magnitude is ordered DAM > homeostatic > ",
           "IFN > proliferative at the TF layer, consistent with the ",
           "substate interpretation that DAM cells carry the largest ",
           "attenuation magnitude at the interaction.")
  }
  paragraphs[4] <- paste0(
    "**Biological reading.** Tau on the amyloid background suppresses ",
    "amyloid-driven NF-kB transcriptional output across every microglia ",
    "substate at the interaction, with the strongest signal in the ",
    "disease-associated DAM substate whose baseline NF-kB program is ",
    "most amplified. The H4 section-17.4 §4 hypothesis sketch had ",
    "predicted MG_homeostatic as the likely carrier; the data refine ",
    "this to **DAM as the lead carrier with homeostatic close behind**, ",
    "with every substate carrying the suppression to some degree. The ",
    "section 11 ANOVA rules out a Simpson's-paradox composition ",
    "artefact at the interaction (tau:nlgf-on-composition p = ",
    sprintf("%.3f", interaction_anova_p),
    "), so the per-state signal is transcriptional, not compositional. ",
    "The T-Tau-attenuates cross-axis theme therefore lifts from its ",
    "section-17.5 5-row position by gaining per-state transcriptional ",
    "specificity and an independent fgsea-layer replication; the ",
    "Phase I evidence is at the Strong tier under the H1-locked ",
    "three-criteria rule.")
  for (p in paragraphs) {
    cat(p)
    cat("\n\n")
  }
}

nfkb_verdict_statement(verdict_input, verdict_tier,
                       interaction_anova_p, n_carriers)
```

The Phase I evidence above will be propagated into the H2 ledger and
the section 17 synthesis in the next session (I4); section 17.5's
integrated model statement will be updated to reference the per-state
result, and 3-6 new ledger rows will tag the per-state findings as
`supports T-Tau-attenuates` claim rows at the Strong grade.
