# Causal signalling network reconstruction

This is the final mechanism-layer section and the topological capstone of
narrative arc **J** (`storage/notes/causal_network_plan.md`). The
mechanism layer produced *endpoint activities* in isolation: kinases
(section 15; Gsk3b strongest, interaction axis), transcription factors
(section 14; Spi1 / Nfkb1 / Sp3 on the amyloid axis, Myc / Creb1 / Tp53 /
Jun on the interaction axis), and ligand-receptor pairs (section 16). None
of those layers said anything about the **wiring between** the upstream
kinase signal and the downstream transcriptional signal. This section
reconstructs, per contrast, the directed signed signalling subnetwork that
connects them, solved as an integer linear program (ILP) over the OmniPath
prior-knowledge network by CARNIVAL.

Two headline questions motivate the reconstruction:

* **(a)** is the interaction-axis Gsk3b kinase signal topologically wired
  to the interaction-axis transcription factors (Myc and the other
  section-14.3 axis-3 leaders)?
* **(b)** does the tau-driven NF-&kappa;B attenuation read out at section
  18 have a recoverable upstream signalling path?

**The two endpoints (no new inference).** This section re-uses caches
already on disk. The **upstream** seeds are kinase activities from the
corrected bulk-phosphoproteome (`kinase_activity_decoupler.rds`, the
`phospho_corrected` modality from section 15); the **downstream**
measurements are TF activities aggregated across snRNAseq + GeoMx +
proteomics (`tf_activity_decoupler_split.rds`, the split-complex
re-derivation built for this layer so CollecTRI complexes resolve to
single-protein symbols). A kinase enters as a signed perturbation
(`inputObj`) and a TF enters as a signed measurement (`measObj`); CARNIVAL
finds the smallest signed subnetwork of the prior network that is
consistent with both.

**Thresholds, stated before any result is read.**

* Both endpoints are thresholded at **FDR < 0.10** on the decoupleR `ulm`
  p-value (BH-adjusted within modality &times; contrast); direction is the
  sign of the `consensus` score &mdash; the same significance/direction
  split locked at the TF and kinase layers.
* The prior network is the **OmniPath mouse signed-directed PKN**
  (`organism = 10090`; directed edges with stimulation XOR inhibition),
  **pruned to microglia-expressed symbols** (snRNAseq microglia subset),
  giving 10,713 edges over 3,779 nodes.
* Because the full microglia-expressed PKN is ILP-intractable, each
  contrast's network is further restricted to nodes lying on a directed
  seed&rarr;target walk of **at most `L = 3` hops** before the solve. `L`
  is a mechanistic horizon matched to the 1&ndash;3-hop kinase&rarr;TF
  reachability observed at J1, **not** an exhaustive search: any path
  longer than three hops is excluded by construction. This is recorded in
  every network's `meta$max_path`.
* The ILP is solved by **cbc** (single-threaded; multi-threaded cbc
  segfaults at this build) with sparsity penalty `betaWeight = 0.2`. Node
  activities and edge weights are CARNIVAL's **consensus across all optimal
  solutions** (its built-in multi-solution averaging), so a reported edge
  is one that survives averaging over alternative optima, not an arbitrary
  single solution.

**Cross-compartment bridge (restated wherever a kinase-seeded path is
read).** The kinase seeds are **bulk hippocampus** phospho, not
microglia-sorted (a standing data lock); the TF measurements are
microglial. Every recovered network therefore bridges bulk-tissue
signalling to microglial transcription &mdash; an explicit
cross-compartment inference, not a within-microglia signalling map.

**Anti-anchoring.** An omitted edge is a finding. The verdict reports what
the ILP connects **and** what it fails to connect; a path the solver did
not return is never asserted. All five contrasts and all three D2 axes are
presented in parallel with no privileged axis.

## Per-contrast reconstructed networks

This subsection reconstructs one network per contrast and reports coverage
honestly. The summary table below carries, per contrast: the solve status,
the number of kinase seeds and TF measurements that survived the
microglia-prune and the `L = 3` reachability-prune to enter the ILP, the
number of measured TFs recovered as active, the number of inferred
intermediates the solver introduced, and the active-edge count.

```{r causal-net-build}
causal_net <- readRDS(file.path(params$cache_dir, "causal_network.rds"))

cn_contrasts <- c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
                  "tau_alone", "tau_in_nlgf")

# Per-contrast summary via the helper, augmented with the count of kinase
# seeds offered to each ILP (meta$n_pert_in) so the coverage story is
# legible: seeds_offered -> seeds_active, TFs_offered -> TFs_recovered.
cn_summary <- dplyr::bind_rows(lapply(cn_contrasts, function(ct) {
  s <- summarise_network(causal_net[[ct]])
  m <- causal_net[[ct]]$meta
  s$kinase_seeds_in <- m$n_pert_in %||% 0L
  s
}))
cn_summary_view <- cn_summary |>
  dplyr::transmute(
    contrast,
    status,
    kinase_seeds_in,
    kinase_seeds_active = n_input_active,
    TF_meas_in          = n_meas_in,
    TF_recovered        = n_measured_recovered,
    frac_TF_recovered   = frac_meas_recovered,
    intermediates       = n_inferred_active,
    active_edges        = n_edges
  )

# Long-format node and edge exports (active nodes/edges only, all
# contrasts) for out-of-knit consumers and the section-20.2 / 20.3 layers.
cn_nodes_tsv <- dplyr::bind_rows(lapply(cn_contrasts, function(ct) {
  net <- causal_net[[ct]]
  an  <- net$nodes[net$nodes$activity != 0, , drop = FALSE]
  if (nrow(an) == 0L) return(NULL)
  tibble::tibble(contrast = ct, node = an$node, node_type = an$node_type,
                 activity = an$activity, up = an$up, down = an$down,
                 zero = an$zero)
}))
cn_edges_tsv <- dplyr::bind_rows(lapply(cn_contrasts, function(ct) {
  e <- causal_net[[ct]]$edges
  if (nrow(e) == 0L) return(NULL)
  tibble::tibble(contrast = ct, source = e$source,
                 interaction = e$interaction, target = e$target,
                 weight = e$weight)
}))
readr::write_tsv(cn_nodes_tsv,
                 file.path(params$results_dir, "causal_network_nodes.tsv"))
readr::write_tsv(cn_edges_tsv,
                 file.path(params$results_dir, "causal_network_edges.tsv"))
```

```{r causal-net-summary-table, results = "asis"}
knitr::kable(
  cn_summary_view, digits = 2,
  caption = paste0(
    "Per-contrast CARNIVAL reconstruction summary. `kinase_seeds_in` / ",
    "`TF_meas_in` are the seeds / measurements that entered the ILP after ",
    "the microglia-prune and L=3 reachability-prune; `TF_recovered` is ",
    "the number of those measured TFs the solver placed on an active path; ",
    "`intermediates` is the count of inferred (unmeasured) nodes the solver ",
    "introduced; `active_edges` is the consensus active-edge count. Two ",
    "contrasts are honestly empty (no kinase seed at FDR<0.10)."
  ))
```

Three of the five contrasts reconstruct a non-empty network
(`nlgf_in_p301s`, `interaction`, `tau_in_nlgf`); two are **honestly empty**.
`nlgf_in_maptki` and `tau_alone` carry **zero** kinase activities at
FDR < 0.10 in the corrected phospho cache (the same two non-findings
reported at section 15.1), so there is no perturbation to seed the ILP and
the reachability-prune returns an empty network before any solve. Per the
anti-anchoring guardrail these are recorded as findings, not dropped: the
kinase layer simply provides no high-confidence upstream signal for the
amyloid-on-tau-null and tau-alone contrasts at this threshold.

### Contrast `nlgf_in_p301s` (amyloid on the P301S background)

This is the best-covered reconstruction: three active kinase seeds
(Cdk5, Gsk3b, Ppp1ca) and **18 of 22** measured TFs recovered. The figure
shows two distinct seed-rooted modules. **Gsk3b** activates Nfkb1, Map2k1,
Usf1 and Usf2, and inhibits Mapk14; the Gsk3b&rarr;Nfkb1 activation edge is
the prior-network support for the NF-&kappa;B reading and is examined under
question (b) at section 20.3. **Cdk5** activates Dnmt1, Mapk7 and Src. The
two modules meet at a Map2k1&rarr;Mapk1 hub that fans out to a block of
biosynthetic / cell-cycle TFs (Cdk2, Mafa, Sp3, Srebf1, Srebf2, Stat5a).
Spi1 &mdash; the section-14.3 amyloid-axis lead TF &mdash; is recovered
three hops downstream of Mapk14 via an inferred Gata2 node (Mapk14 &rarr;
Gata2 &dashv; Spi1). The third seed, Ppp1ca, is active but isolated: the
solver placed no active edge incident to it, so it contributes no path.

```{r causal-net-fig-p301s, fig.width = 8.5, fig.height = 6.5}
plot_causal_network(causal_net[["nlgf_in_p301s"]],
                    title = "nlgf_in_p301s")
```

### Contrast `interaction` (the tau &times; amyloid divergence)

The headline contrast reconstructs a compact eight-edge network: a single
kinase seed (Gsk3b) and **all five** measured TFs recovered. The recovered
wiring **directly answers question (a)**: Gsk3b inhibits an inferred Mapk14
node (Gsk3b &dashv; Mapk14, with a Mapk14 &dashv; Gsk3b feedback edge),
and Mapk14 activates Myc and Foxo3 and an inferred Csnk2b that in turn
activates Tbp; Gsk3b additionally inhibits Clock and Sfpq directly. The
interaction-axis Gsk3b kinase is therefore topologically connected to the
interaction-axis TFs &mdash; Myc, the section-14.3 axis-3 lead, sits two
hops downstream of Gsk3b through Mapk14. The second kinase nominated at
this contrast (Csnk1a1) dropped out of the microglia-pruned PKN and so
could not be seeded.

```{r causal-net-fig-interaction, fig.width = 8, fig.height = 6}
plot_causal_network(causal_net[["interaction"]],
                    title = "interaction")
```

### Contrast `tau_in_nlgf` (adding tau on the amyloid background)

A single Gsk3b seed reconstructs a 27-edge network recovering **all 13**
measured TFs. Gsk3b is broadly inhibitory here &mdash; it inhibits Ar,
Clock, Hif1a, Ikbkg, Mapk14 and Pias1 and activates only Map2k1. The
NF-&kappa;B-regulatory module appears through an inferred path
**Gsk3b &dashv; Ikbkg (NEMO) &rarr; Nfkbib (I&kappa;B&beta;)** &mdash;
examined under question (b) at section 20.3. Myc is a convergence point
reached three independent ways (Mapk14 &rarr; Myc, Cdk2 &rarr; Myc,
Creb1 &rarr; Myc), and a PKA-family arm (Akt1 &rarr; Prkaca &rarr; Atf1 /
Creb1 / Srf) supplies the cyclic-nucleotide TFs.

```{r causal-net-fig-tau-in-nlgf, fig.width = 8.5, fig.height = 6.5}
plot_causal_network(causal_net[["tau_in_nlgf"]],
                    title = "tau_in_nlgf")
```

## Axis-restricted networks

This subsection reads the reconstructed networks against the three D2 axes
locked at the pathway-survey layer (`amyloid_activation`,
`synaptic_suppression`, `interaction_metabolic`). Two facts shape how the
axis lens is applied here, and both are themselves findings.

**First, the axes are contrast groupings.** The amyloid-activation and
synaptic-suppression axes both map to the two NLGF contrasts
(`nlgf_in_maptki`, `nlgf_in_p301s`); the interaction-metabolic axis maps to
the single `interaction` contrast. Since `nlgf_in_maptki` is empty, the only
solved network for *both* NLGF axes is `nlgf_in_p301s` &mdash; the same
amyloid/synaptic entanglement reported at the kinase layer (section 15.2),
where the two axes share their contrast set and cannot be differentiated by
the per-contrast signal alone. The `tau_in_nlgf` and `tau_alone` contrasts
sit outside the three-axis framework (consistent with sections 14 and 15).

**Second, gene-universe restriction does not cleanly partition the wiring.**
The natural way to "restrict to an axis" would be to keep only edges whose
*both* endpoints lie in the axis's leader-pathway gene universe (the
`induced` restriction). That yields **zero edges** for the interaction
network against the interaction universe, because the recovered regulators
scatter across all three universes rather than sitting inside any one:
of the eight active interaction-network nodes, only **Gsk3b** is in the
interaction-metabolic universe, while Mapk14 / Myc / Foxo3 fall in the
amyloid universe and Gsk3b / Mapk14 / Clock / Sfpq in the synaptic
universe. This is expected &mdash; the axis universes are defined by broad
GO / module pathway membership, not by regulator identity &mdash; and it
means the recovered topology is a **cross-axis bridge**, not an
axis-internal module. The table below therefore reports gene-universe
membership as an *annotation* (how many active nodes of each axis's solved
network fall in that axis's universe) alongside the `incident`-restricted
edge count (edges with at least one endpoint in the universe), rather than
forcing the degenerate `induced` subgraph.

```{r causal-net-axis-build}
# Reuse the in-knit axis_universes global built at section 14; rebuild
# defensively for out-of-knit runs (mirrors rmd/14's defensive block).
if (!exists("axis_universes")) {
  .lb <- readr::read_tsv(
    file.path(params$results_dir,
              "pathway_survey_unified_leaderboard.tsv"),
    show_col_types = FALSE)
  .gobp <- readRDS(file.path(params$cache_dir, "msigdb_gobp_mouse.rds"))
  .gocc <- readRDS(file.path(params$cache_dir, "msigdb_gocc_mouse.rds"))
  .gomf <- readRDS(file.path(params$cache_dir, "msigdb_gomf_mouse.rds"))
  .cst  <- readRDS(file.path(params$cache_dir,
                             "custom_microglia_states.rds"))
  .cad  <- readRDS(file.path(params$cache_dir, "custom_microglia_ad.rds"))
  .hwg  <- readRDS(file.path(params$cache_dir, "hdwgcna_microglia.rds"))
  .sm   <- readRDS(file.path(params$cache_dir, "snrnaseq_symbol_map.rds"))
  .hm   <- .hwg$modules
  .hm$gene_name <- .sm$symbol[match(.hm$gene_name, .sm$ensembl)]
  .hm <- .hm[!is.na(.hm$gene_name) & nzchar(.hm$gene_name) &
               as.character(.hm$module) != "grey", , drop = FALSE]
  axis_universes <- build_axis_gene_universe(
    .lb,
    list(gobp = .gobp, gocc = .gocc, gomf = .gomf,
         custom_microglia_states = .cst, custom_microglia_ad = .cad,
         hdwgcna_modules = .hm))
}

# For each axis, take its solved contrast networks and report (a) active
# nodes in the axis universe, (b) incident- vs induced-restricted edge
# counts. Amyloid/synaptic resolve to nlgf_in_p301s; interaction to
# interaction (the empty NLGF/tau contrasts contribute nothing).
cn_axis_rows <- dplyr::bind_rows(lapply(names(axis_universes), function(ax) {
  uni  <- axis_universes[[ax]]$universe
  ctrs <- axis_universes[[ax]]$axis_contrasts
  dplyr::bind_rows(lapply(ctrs, function(ct) {
    net <- causal_net[[ct]]
    if (is.null(net) || nrow(net$edges) == 0L) return(NULL)
    an   <- net$nodes[net$nodes$activity != 0, , drop = FALSE]
    inU  <- intersect(an$node, uni)
    rind <- restrict_network_to_axis(net, uni, mode = "induced")
    rinc <- restrict_network_to_axis(net, uni, mode = "incident")
    tibble::tibble(
      axis             = ax,
      contrast         = ct,
      universe_genes   = length(uni),
      active_nodes     = nrow(an),
      nodes_in_universe = length(inU),
      induced_edges    = nrow(rind$edges),
      incident_edges   = nrow(rinc$edges),
      in_universe      = paste(sort(inU), collapse = ", ")
    )
  }))
}))
```

```{r causal-net-axis-table, results = "asis"}
knitr::kable(
  cn_axis_rows |> dplyr::select(-in_universe), digits = 0,
  caption = paste0(
    "Axis-restricted view of each axis's solved contrast network(s). ",
    "`nodes_in_universe` = active nodes whose symbol lies in the axis ",
    "leader-pathway gene universe; `induced_edges` keeps only edges with ",
    "BOTH endpoints in the universe; `incident_edges` keeps edges with at ",
    "least one endpoint in the universe. The gap between induced and ",
    "incident edge counts quantifies the cross-axis bridging -- it is ",
    "widest for the interaction network (induced 0 vs incident 4), where ",
    "only Gsk3b is an interaction-universe member."))
```

For the interaction axis the `induced` count is zero against an `incident`
count of four (the four Gsk3b-incident edges): the recovered Gsk3b &rarr;
Mapk14 &rarr; Myc / Foxo3 path is anchored by the interaction-axis seed
(Gsk3b, the sole interaction-universe member) but resolves onto TFs whose
pathway annotations belong to the amyloid and synaptic universes. The
interaction-axis signal is therefore wired, but it is wired **across**
axes &mdash; the kinase that the interaction contrast nominates drives TFs
that the broader pathway survey files under the amyloid and synaptic
programs. The amyloid and synaptic axes share `nlgf_in_p301s` and so report
the same underlying network differentiated only by which of its active
nodes each universe annotates (15 of 32 in the amyloid universe, 13 of 32
in the synaptic universe).

## Verdict (sign-aware topology synthesis)

This subsection synthesises the reconstruction into a per-axis verdict,
answering the two headline questions and reporting absence as evidence. The
thresholds are restated once: standard per-contrast CARNIVAL, kinase
`inputObj` &rarr; TF `measObj`, both at FDR < 0.10; OmniPath mouse PKN
pruned to microglia-expressed symbols and to an `L = 3` reachability
horizon; cbc, `betaWeight = 0.2`; consensus across optimal solutions.

**Question (a) &mdash; is the interaction Gsk3b wired to the interaction
TFs? Yes.** The `interaction` network recovers
**Gsk3b &dashv; Mapk14 &rarr; Myc** (and &rarr; Foxo3), placing the
section-15 interaction-axis lead kinase two hops upstream of the
section-14.3 interaction-axis lead TF through a single inferred Mapk14
(p38&alpha;) intermediate, with all five measured interaction TFs recovered.
This is the cleanest topological result of the section and the direct
join between the project's kinase and TF mechanism layers at the divergence
contrast.

**Question (b) &mdash; does the section-18 NF-&kappa;B attenuation have a
recoverable upstream path? Partially, and contrast-dependently.** Of the
three solved networks, two touch the NF-&kappa;B machinery and one does
not. In `tau_in_nlgf` the solver recovers
**Gsk3b &dashv; Ikbkg (NEMO) &rarr; Nfkbib (I&kappa;B&beta;)** &mdash; a
candidate upstream route to the NF-&kappa;B regulatory module in which
Gsk3b inhibition of NEMO is direction-consistent with attenuated canonical
NF-&kappa;B signalling. In `nlgf_in_p301s` the only NF-&kappa;B node
recovered is **Nfkb1** itself, reached by a Gsk3b &rarr; Nfkb1 **activation**
edge &mdash; the opposite direction, on the amyloid-on-tau contrast rather
than a tau-attenuation contrast. The `interaction` network recovers **no**
NF-&kappa;B node at all (no NF-&kappa;B TF reached FDR < 0.10 to seed a
measurement, and none was inferred). The section-18 attenuation therefore
has a plausible recovered upstream path on the `tau_in_nlgf` contrast but
**not** on the `interaction` contrast itself &mdash; an honest, mixed
finding rather than a clean confirmation.

**Per-axis topology.** The verdict table records one row per axis. The
**amyloid-activation** axis reads strongly: `nlgf_in_p301s` recovers 18 of
22 TFs including the section-14 amyloid leads (Nfkb1 directly off Gsk3b;
Spi1 three hops downstream via Gata2). The **synaptic-suppression** axis
shares that single solved network and is differentiated only by
gene-universe annotation &mdash; it contributes **no distinct
synaptic-suppression wiring**, the same honest non-finding the kinase
(15.2) and TF (14.3) layers reported for this axis. The
**interaction-metabolic** axis reads through the compact Gsk3b &rarr;
Mapk14 &rarr; Myc / Foxo3 module answering question (a).

**Caveats carried into the verdict.** The kinase seeds are bulk
hippocampus phospho, not microglia-sorted, so every recovered path is a
bulk-signalling-to-microglial-transcription bridge. The `L = 3` horizon
excludes any path longer than three hops by construction, so an absent edge
between two well-separated nodes may reflect the horizon rather than a true
topological gap. Two of five contrasts are empty for lack of a kinase seed.
None of these is a defect to be corrected; each is a stated boundary of the
reconstruction.

```{r causal-net-verdict-build}
cn_verdict <- tibble::tibble(
  axis = c("amyloid_activation", "synaptic_suppression",
           "interaction_metabolic"),
  axis_contrasts = c("nlgf_in_maptki, nlgf_in_p301s",
                     "nlgf_in_maptki, nlgf_in_p301s",
                     "interaction"),
  n_solved_networks = c(1L, 1L, 1L),
  tf_recovered = c("18/22 (nlgf_in_p301s)", "18/22 (nlgf_in_p301s)",
                   "5/5 (interaction)"),
  key_recovered_path = c(
    "Gsk3b -> Nfkb1 (activation); Mapk14 -> Gata2 -| Spi1; Map2k1 -> Mapk1 hub",
    "shares nlgf_in_p301s network; no distinct synaptic-suppression wiring",
    "Gsk3b -| Mapk14 -> Myc / Foxo3; Csnk2b -> Tbp"),
  nfkb_path = c(
    "Gsk3b -> Nfkb1 activation (amyloid-on-tau direction)",
    "shares nlgf_in_p301s network",
    "no NF-kB node recovered"),
  topology_strength = c("strong", "non-finding (shared/entangled)",
                         "strong (headline path recovered)"),
  evidence_summary = c(
    paste0("Amyloid axis reads strongly through nlgf_in_p301s: 18 of 22 ",
           "measured TFs recovered across two seed-rooted modules (Gsk3b, ",
           "Cdk5). Gsk3b activates Nfkb1 directly and the section-14 ",
           "amyloid lead Spi1 is recovered three hops downstream via an ",
           "inferred Gata2. A Map2k1->Mapk1 hub fans out to a ",
           "biosynthetic/cell-cycle TF block (Sp3, Srebf1/2, Stat5a, ",
           "Cdk2, Mafa). Bulk-phospho seed caveat applies."),
    paste0("Synaptic-suppression axis shares its only solved network ",
           "(nlgf_in_p301s) with the amyloid axis and is differentiated ",
           "solely by gene-universe annotation (13 of 32 active nodes in ",
           "the synaptic universe vs 15 in the amyloid universe). No ",
           "synaptic-suppression-specific wiring is recovered -- the same ",
           "honest non-finding the kinase (15.2) and TF (14.3) layers ",
           "reported for this axis, consistent with an engulfment / ",
           "post-translational rather than regulator-driven mechanism."),
    paste0("Interaction axis recovers the headline path: Gsk3b -| Mapk14 ",
           "-> Myc and -> Foxo3 (all 5 measured TFs recovered), wiring the ",
           "section-15 interaction-axis lead kinase to the section-14.3 ",
           "interaction-axis lead TF through a single inferred Mapk14 ",
           "(p38-alpha). The recovered regulators bridge across axis ",
           "universes (only Gsk3b is an interaction-universe member; ",
           "Mapk14/Myc/Foxo3 annotate to the amyloid universe). No ",
           "NF-kB node is recovered at this contrast.")
  ))

readr::write_tsv(cn_verdict,
                 file.path(params$results_dir, "causal_network_verdict.tsv"))
```

```{r causal-net-verdict-table, results = "asis"}
knitr::kable(
  cn_verdict |>
    dplyr::select(axis, axis_contrasts, tf_recovered, nfkb_path,
                  topology_strength),
  caption = paste0(
    "Per-axis causal-topology verdict. `tf_recovered` and ",
    "`key_recovered_path` summarise the axis's solved network(s); ",
    "`nfkb_path` records the NF-kB result for question (b); ",
    "`topology_strength` is the qualitative read. Full prose and the ",
    "`key_recovered_path` / `evidence_summary` columns are preserved in ",
    "`storage/results/causal_network_verdict.tsv`."))
```

The causal layer closes the mechanism arc: the kinase and TF endpoints that
sections 14 and 15 surfaced in isolation are, at the interaction contrast,
**topologically joined** by a Gsk3b &rarr; Mapk14 &rarr; Myc module, while
the NF-&kappa;B attenuation of section 18 finds a candidate upstream route
only on the `tau_in_nlgf` contrast. The reconstruction neither manufactures
connectivity where the data lacks a seed (two empty contrasts) nor asserts
a synaptic-suppression program the topology does not contain.
