# SCENIC data-driven regulon inference

This section adds the project's first **data-driven** gene-regulatory
layer and closes narrative arc **K**
(`storage/notes/scenic_regulons_plan.md`). Every preceding mechanism
layer is *prior-based*: the transcription-factor activities of section 14
read from the CollecTRI literature network, the kinase activities of
section 15 from the OmniPath kinase-substrate prior, the ligand-receptor
calls of section 16 from curated interaction priors, and the causal
topology of section 20 from the OmniPath prior-knowledge network. SCENIC
instead infers regulons *de novo* from this dataset's own microglial
co-expression: GRNBoost2 over the 26,104-nucleus microglia matrix learns
TF&rarr;target adjacencies, cisTarget prunes them to motif-supported,
signed regulons against the mm10 v10 motif collection, and AUCell scores
per-cell regulon activity. The motivating question is narrow and
falsifiable: do the section-14 prior-based TF verdicts survive a method
that does **not** lean on a literature prior? The layer is framed
throughout as orthogonal corroboration or honest discordance &mdash;
never as a re-run of section 14. It feeds two margin-neutral Suggestive
rows to the section-17 claims ledger (K-001 Spi1, K-002 Rel: the
data-driven *recovery* of those regulons, scoped to existence and not
activity direction), and so it does not perturb the locked mouse
2&times;2 contest margins.

**The consensus regulon set, fixed before any contrast was read.** The
GRNBoost2 / cisTarget pipeline was run independently over ten random
seeds, and the consensus was defined *before* inspecting which TFs
survived (the load-bearing anti-anchoring lock). A TF&rarr;target edge is
high-confidence iff it recurs in **&ge; 8 of 10** runs; a consensus
regulon is a TF that retains **&ge; 5** such high-confidence targets
(`min_targets`, mirroring section 14's decoupleR floor). This yields
**51 activating regulons** over 10,711
high-confidence edges, drawn from the 994 candidate regulators
(the genes that survived the &ge; 1%-of-cells expression filter
intersected with the mouse TF list). Every consensus regulon carries the
sign `+`: the repressing modules never reached the &ge; 8/10 bar, an
anticipated outcome since SCENIC's correlation-derived `(-)` regulons are
the less stable half.

**The controlled network swap (primary readout).** To isolate the effect
of the *network* from the effect of the *scoring*, readout 1 feeds the
SCENIC regulons to the section-14 decoupleR machinery **verbatim**: the
same NEBULA *z* statistic (`de_snrnaseq_nebula$top[[contrast]]`), the same
`ulm` significance and cross-method `consensus` direction, the same
`min_targets = 5`, the same FDR < 0.10. Only the prior network changes
&mdash; data-driven SCENIC regulons in place of CollecTRI. Any difference
in TF activity is therefore attributable to the network alone. Readout 2
scores the native AUCell activity through the locked 2&times;2 factorial;
readout 3 mirrors the section-18 per-substate analysis.

**Thresholds, stated before any result is read.** Significance everywhere
is BH-FDR < 0.10 (the project standard). Regulon `min_targets` &ge; 5;
recurrence bar &ge; 8/10; AUCell pseudobulk variance-stabilised by
`qlogis` (logit); cisTarget motif-enrichment at the pyscenic defaults
(`rank_threshold` 5000, `auc_threshold` 0.05, NES &ge; 3.0,
`min_genes` 20).

**Caveats carried throughout.** (a) The inference is **microglia-only**,
so these are microglial-**state** regulons, not cell-type-identity
regulons. (b) The input is single-nucleus (intronic reads included);
SCENIC handles snRNA but the co-expression is nuclear. (c) The working
microglia identity is **inherited** from the section-14
`broad_annotations`, not re-derived here; the same modest ambient
neuronal/oligodendrocyte signal noted at the data-export step is present.
(d) Because the consensus is **all-activating**, the SCENIC decoupleR
network carries no repressing edges; a section-14 *repressing* call that
has no SCENIC counterpart is a structural consequence of that, **not** a
discordance, and is flagged as such wherever signs are compared.

**Anti-anchoring.** All 51 regulons are reported even-handedly,
including the large novel ones absent from section 14. Non-recovery of a
prior TF is reported as evidence that the prior footprint lacks
microglial co-expression support, not hidden. No axis is privileged. The
&ge; 8/10 bar is **not** relaxed to rescue sub-threshold near-misses.

```{r scenic-load}
scenic   <- readRDS(file.path(params$cache_dir, "scenic_summary.rds"))
sc_reg   <- scenic$regulons          # long TF/sign/target/recurrence/importance
sc_net   <- scenic$scenic_net        # decoupleR net (source/target/mor)
sc_h2h   <- scenic$headtohead        # SCENIC vs CollecTRI activity, paired
sc_act   <- scenic$scenic_activity   # decoupleR SCENIC net on NEBULA z (long)
sc_auc   <- scenic$aucell_contrasts  # native AUCell 2x2 factorial (long)
sc_aucpb <- scenic$aucell_pb         # list(mat = regulon x 16, meta)
sc_sub   <- scenic$substate_int      # per-substate decoupleR at interaction
sc_over  <- scenic$target_overlap    # SCENIC-vs-CollecTRI target Jaccard
sc_rec   <- scenic$recovery          # recovery ladder (60 rows)
sc_cen   <- scenic$census            # comparison-TF census (12 rows)
sc_par   <- scenic$params

n_reg  <- sc_par$n_consensus_regulons
n_cand <- sc_par$candidate_tfs_n

# Headline scalars used inline so prose never drifts from the cache.
med_jac   <- median(sc_over$jaccard)
max_jac   <- max(sc_over$jaccard)
max_jac_tf<- sc_over$TF[which.max(sc_over$jaccard)]
n_shared  <- nrow(sc_over)
spi1_ov   <- sc_over[sc_over$TF == "Spi1", ]
n_auc_sig <- length(unique(sc_auc$regulon[sc_auc$sig]))
n_act_sig <- length(unique(sc_act$source[sc_act$sig]))
n_int_aucell <- sum(sc_auc$sig[sc_auc$contrast == "interaction"])

contrast_order <- c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
                    "tau_alone", "tau_in_nlgf")
substate_order <- c("homeostatic", "DAM", "IFN", "proliferative")
diverging <- c("#3a4cc0", "white", "#b40426")   # negative .. zero .. positive
```

## Regulon recovery census

The first question is purely about presence: of the section-14 TF verdict
leads and the section-18 NF-&kappa;B family, which form a motif-supported,
co-expression-recurrent regulon in this dataset's own microglia? The
recovery ladder below classifies **every** TF &mdash; the `r n_reg`
consensus regulons plus the section-14/18 comparison set &mdash; by why it
did or did not enter the &ge; 8/10 consensus, separating structural
non-recovery (below the expression floor, never a candidate regulator)
from motif pruning (a candidate with adjacencies but no motif-supported
regulon) from stochastic near-misses. The class definitions are
data-derived: candidacy is recomputed live from the filtered-matrix
&cap; TF-list intersection, not hand-assigned.

```{r scenic-recovery-ladder, results = "asis"}
class_lab <- c(
  recovered_novel              = "Recovered (novel; not a section-14/18 TF)",
  recovered_comparison         = "Recovered (section-14/18 comparison TF)",
  near_miss                    = "Near-miss (6-7/10 runs, below the >=8 bar)",
  present_targets_subthreshold = "Regulon formed but < 5 stable targets",
  weak                         = "Weak (1-5/10 runs)",
  motif_pruned                 = "Motif-pruned (candidate, 0 motif-supported regulon)",
  expression_floor             = "Expression floor (below the >=1% filter)")
ladder <- as.data.frame(table(sc_rec$recovery_class), stringsAsFactors = FALSE)
names(ladder) <- c("recovery_class", "n_TFs")
ladder$comparison_TFs <- vapply(ladder$recovery_class, function(cl) {
  tf <- sc_rec$TF[sc_rec$recovery_class == cl & sc_rec$is_comparison_tf]
  if (length(tf) == 0L) "-" else paste(sort(tf), collapse = ", ")
}, character(1))
ladder$class <- class_lab[ladder$recovery_class]
ladder <- ladder[order(-ladder$n_TFs), c("class", "n_TFs", "comparison_TFs")]
knitr::kable(
  ladder, row.names = FALSE,
  col.names = c("Recovery class", "n TFs", "Comparison TFs in class"),
  caption = paste0(
    "Recovery ladder over all ", nrow(sc_rec), " classified TFs (",
    n_reg, " consensus regulons + the section-14/18 comparison set). ",
    "`comparison TFs in class` names only the section-14/18 leads; the ",
    "49 novel recovered regulons are >=8/10 by construction. Full table ",
    "in storage/results/scenic_regulon_recovery.tsv."))
```

The section-14/18 comparison TFs alone are tabulated next, with their
per-seed presence and consensus outcome. This is the headline of the
section, reported exactly as computed.

```{r scenic-census-table, results = "asis"}
cen <- sc_cen
cen$recovered <- ifelse(cen$in_consensus == 1L, "yes", "no")
cen_view <- cen[, c("axis", "TF", "max_runs_present", "present_signs",
                    "recovered", "consensus_signs_ntargets")]
cen_view <- cen_view[order(cen_view$axis, -cen_view$max_runs_present), ]
knitr::kable(
  cen_view, row.names = FALSE,
  col.names = c("Axis", "TF", "max runs present", "signs (n runs)",
                "in consensus", "consensus sign:n targets"),
  caption = paste0(
    "Recovery census for the section-14 verdict TFs (amyloid_activation: ",
    "Spi1/Nfkb1/Sp3; interaction_metabolic: Myc/Creb1/Tp53/Jun) and the ",
    "section-18 NF-kB family (Rela/Nfkb1/Nfkb2/Relb/Rel). `max runs ",
    "present` = the most runs in which the TF formed any motif-supported ",
    "regulon; `in consensus` requires >=8/10 edge recurrence AND >=5 ",
    "stable targets."))
```

Two comparison TFs cross the bar into the consensus and no more: **Spi1**
&mdash; the section-14 amyloid-axis lead &mdash; recovers cleanly
(present in all ten runs, 15 high-confidence targets), and **Rel** (an
NF-&kappa;B-family member) likewise (10/10, 9 targets). The remaining
calls are sub-threshold, and their *causes differ*, which is the
informative part:

* **Near-misses (6-7/10):** **Nfkb1** and **Sp3** (both 7/10) and
  **Creb1** (6/10) form recurrent regulons but fall just short of the
  locked &ge; 8/10 bar. These are reported at 6-7/10 honestly; the bar is
  not lowered to admit them.
* **Regulon formed but too few stable targets:** **Relb** is present in
  all ten runs yet retains fewer than five high-confidence targets, so it
  is a regulon that exists but does not meet `min_targets`.
* **Motif-pruned:** **Nfkb2** is a *candidate* regulator (it had hundreds
  of GRNBoost2 adjacencies) but formed **no** motif-supported regulon in
  any run &mdash; a co-expression/motif failure, not an expression-floor
  one.
* **Expression floor (structural):** **Myc** and **Tp53** are below the
  &ge; 1%-of-cells filter, so they were never candidate regulators and
  draw zero edges in every run. SCENIC *structurally cannot* recover them.
  They are **two of the four `interaction_metabolic` verdict TFs**: the
  data-driven layer cannot corroborate the half of that axis that lacks
  microglial co-expression footing &mdash; precisely the anti-anchoring
  guardrail's "non-recovery = the prior footprint lacks co-expression
  support", reported as-is.

The &ge; 8/10 bar's strictness is itself audited below: the
recurrence-sensitivity curve shows how many regulons survive at each
threshold. The decay is smooth with no cliff at 8, so the locked bar is
strict but not pathological &mdash; a looser bar would admit more
near-misses but the curve gives no principled reason to move it.

```{r scenic-recurrence-fig, fig.width = 7, fig.height = 3.8, fig.cap = "Number of consensus regulons (TFs retaining >=5 high-confidence targets) surviving at each edge-recurrence threshold from 1/10 to 10/10. The dashed line marks the locked >=8/10 bar (51 regulons). The smooth decay shows the bar is strict but not pathological."}
rec_dist <- readr::read_tsv(
  file.path(params$cache_dir, "scenic", "scenic_recurrence_dist.tsv"),
  show_col_types = FALSE)
ggplot(rec_dist, aes(min_runs, n_regulons_ge_min_targets)) +
  geom_vline(xintercept = sc_par$recurrence_threshold, linetype = "dashed",
             colour = "#b40426") +
  geom_line(colour = "grey50") +
  geom_point(size = 2) +
  geom_text(aes(label = n_regulons_ge_min_targets), vjust = -0.8, size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "edge-recurrence threshold (runs out of 10)",
       y = "consensus regulons",
       title = "Recurrence sensitivity of the consensus regulon count") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```

## SCENIC-versus-CollecTRI head-to-head (controlled network swap)

This subsection swaps only the network. Two facts emerge, and the first
governs the second. **First, the data-driven and prior footprints are
almost disjoint gene sets.** Across the `r n_shared` TFs that exist in
both the SCENIC consensus and the CollecTRI prior, the median target-set
Jaccard is **`r sprintf("%.3f", med_jac)`** and the maximum only
**`r sprintf("%.3f", max_jac)`** (`r max_jac_tf`). The recovered Spi1
regulon shares just **`r spi1_ov$n_shared` of its `r spi1_ov$n_union`
union targets** with CollecTRI's Spi1 (`r spi1_ov$n_scenic_targets`
SCENIC targets vs `r spi1_ov$n_collectri_targets` CollecTRI targets).
Near-disjoint target sets are the strongest single piece of evidence that
the layer is genuinely orthogonal rather than a prior re-run &mdash; and
they make activity discordance the expected, not surprising, outcome.

```{r scenic-overlap-table, results = "asis"}
ov_view <- head(sc_over, 10)
knitr::kable(
  ov_view, digits = 4, row.names = FALSE,
  col.names = c("TF", "SCENIC targets", "CollecTRI targets", "shared",
                "union", "Jaccard"),
  caption = paste0(
    "Target-set overlap for the 10 highest-Jaccard TFs shared between the ",
    "SCENIC consensus and the CollecTRI prior (of ", n_shared,
    " shared TFs; median Jaccard ", sprintf("%.3f", med_jac),
    "). Full table in storage/results/scenic_target_overlap.tsv."))
```

**Second, where the prior made its section-14/18 calls, the data-driven
network either disagrees or is silent.** The table below takes each
comparison TF at the contrast where CollecTRI's call is strongest
(smallest BH-FDR on the shared NEBULA *z*) and places the SCENIC verdict
beside it. Of the section-14 verdict TFs, **only Spi1 is also a SCENIC
regulon**, and its activity **sign-flips and loses significance** under
the swap; the rest are CollecTRI-only (absent from the SCENIC network), so
the swap removes the call entirely.

```{r scenic-h2h-table, results = "asis"}
comp_tfs <- unique(sc_par$comparison_tfs$TF)
h2h_comp <- sc_h2h[sc_h2h$TF %in% comp_tfs, ]
# Strongest CollecTRI call per TF: smallest collectri_padj (NA sorted last).
pick <- h2h_comp |>
  dplyr::arrange(TF, collectri_padj) |>
  dplyr::group_by(TF) |>
  dplyr::slice(1L) |>
  dplyr::ungroup()
# Merge only recovery_class; headtohead already carries its own `axis`
# (same combined label as recovery, e.g. Nfkb1 spans two axes) -- pulling
# axis from both would collide into axis.x/axis.y under merge().
ax  <- sc_rec[, c("TF", "recovery_class")]
pick <- merge(pick, ax, by = "TF", all.x = TRUE)
fmt_call <- function(cons, sig) {
  ifelse(is.na(cons), "absent",
         sprintf("%.2f%s", cons, ifelse(sig, "*", "")))
}
h2h_view <- data.frame(
  TF             = pick$TF,
  axis           = pick$axis,
  recovery       = pick$recovery_class,
  in_SCENIC      = ifelse(pick$in_scenic_regulons, "yes", "no"),
  strongest_contrast = pick$contrast,
  CollecTRI      = fmt_call(pick$collectri_cons, pick$collectri_sig),
  SCENIC         = fmt_call(pick$scenic_cons, pick$scenic_sig),
  stringsAsFactors = FALSE)
h2h_view <- h2h_view[order(h2h_view$axis, h2h_view$TF), ]
knitr::kable(
  h2h_view, row.names = FALSE,
  col.names = c("TF", "axis", "recovery class", "in SCENIC net",
                "strongest contrast", "CollecTRI cons", "SCENIC cons"),
  caption = paste0(
    "Controlled head-to-head for the section-14/18 comparison TFs. ",
    "`strongest contrast` is where CollecTRI's call has the smallest ",
    "FDR; CollecTRI / SCENIC columns give the consensus activity there ",
    "(* = FDR < 0.10; `absent` = the TF is not in that network). Both ",
    "sides are decoupleR on the SAME NEBULA z -- only the network ",
    "differs."))
```

The interaction-metabolic axis is **entirely CollecTRI-only**: Myc, Creb1,
Tp53 and Jun are all absent from the SCENIC network (Myc and Tp53 below
the expression floor, Creb1/Jun sub-threshold), so the strongest CollecTRI
call on that axis &mdash; Myc at the interaction contrast (consensus
&minus;6.60, FDR-significant) &mdash; has no SCENIC counterpart at all. On
the amyloid axis, Nfkb1 and Sp3 are CollecTRI-significant at the NLGF
contrasts but unrecovered by SCENIC, while Spi1's CollecTRI activity
(+2.73, significant at `nlgf_in_maptki`) reverses to a small, non-
significant negative SCENIC score. The recovered NF-&kappa;B regulon Rel
has no bare-symbol CollecTRI source, so it carries no head-to-head row
(CollecTRI = absent).

Crucially, the swap does not silence SCENIC &mdash; it **redirects** it.
On the same NEBULA *z*, **`r n_act_sig` of `r n_reg`** SCENIC regulons
reach FDR < 0.10 in at least one contrast. Those significant regulons are
the *large novel* ones (Zbtb20 with 3,140 targets, Fli1, Hif1a, and the
AP-1 / IRF / STAT immune block), not the small section-14-overlap TFs.
The data-driven network has ample power; it simply points at a different,
co-expression-defined TF set. The concordance scatter makes the
orthogonality quantitative.

```{r scenic-h2h-scatter, fig.width = 7, fig.height = 5.2, fig.cap = "SCENIC vs CollecTRI consensus activity on the SAME NEBULA z, for every TF scored by both networks across all five contrasts (one point per TF x contrast). Section-14/18 comparison TFs are highlighted and labelled. Near-zero correlation = the controlled network swap produces orthogonal activity."}
sct <- sc_h2h[!is.na(sc_h2h$scenic_cons) & !is.na(sc_h2h$collectri_cons), ]
sct$is_comp <- sct$TF %in% comp_tfs
rho_s <- suppressWarnings(cor(sct$scenic_cons, sct$collectri_cons,
                              method = "spearman"))
rho_p <- suppressWarnings(cor(sct$scenic_cons, sct$collectri_cons,
                              method = "pearson"))
ggplot(sct, aes(collectri_cons, scenic_cons)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_vline(xintercept = 0, colour = "grey80") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted",
              colour = "grey60") +
  geom_point(aes(colour = is_comp, size = is_comp), alpha = 0.6) +
  ggrepel::geom_text_repel(
    data = sct[sct$is_comp, ],
    aes(label = sprintf("%s (%s)", TF, contrast)),
    size = 2.7, max.overlaps = 30, colour = "#b40426") +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "#b40426"),
                      guide = "none") +
  scale_size_manual(values = c(`FALSE` = 1.1, `TRUE` = 2.2), guide = "none") +
  labs(x = "CollecTRI consensus activity",
       y = "SCENIC consensus activity",
       title = "Controlled network swap: SCENIC vs CollecTRI on the same z",
       subtitle = sprintf(
         "Spearman rho = %.3f, Pearson r = %.3f, n = %d TF x contrast pairs",
         rho_s, rho_p, nrow(sct))) +
  theme_bw()
```

## Native AUCell activity across contrasts

Readout 2 leaves the NEBULA *z* aside and tests the regulons' **own**
per-cell AUCell activity. The 26,104-cell AUCell matrix is pseudobulked to
the 16 `genotype_batch` ids (the locked replicate unit), logit-transformed
(every pooled mean lies strictly interior, 0.005-0.097, so `qlogis` needs
no clamp), and run through the locked 2&times;2 + batch factorial. The
logit choice is immaterial to the conclusions: against raw means the
*t*-statistics correlate at Spearman
`r sprintf("%.3f", sc_par$aucell_logit_vs_raw$spearman_t)` with
`r sprintf("%.1f%%", 100 * sc_par$aucell_logit_vs_raw$sign_agreement)`
sign agreement.

```{r scenic-aucell-counts, results = "asis"}
auc_counts <- sc_auc |>
  dplyr::group_by(contrast) |>
  dplyr::summarise(n_sig = sum(sig), .groups = "drop")
auc_counts <- auc_counts[match(contrast_order, auc_counts$contrast), ]
knitr::kable(
  auc_counts, row.names = FALSE,
  col.names = c("Contrast", "n regulons FDR < 0.10"),
  caption = paste0(
    "Native AUCell activity: number of the ", n_reg, " consensus regulons ",
    "reaching FDR < 0.10 per contrast (2x2 + batch factorial on the ",
    "logit-transformed pseudobulk AUCell means). ", n_auc_sig,
    " distinct regulons are significant in >=1 contrast; none at the ",
    "interaction."))
```

`r n_auc_sig` of `r n_reg` regulons are significant in at least one
contrast, but **`r n_int_aucell` at the interaction**: no data-driven
regulon shows a genotype&times;amyloid interaction at the pseudobulk
level. The amyloid main-effect contrasts carry the signal (14 regulons at
`nlgf_in_p301s`, 7 at `nlgf_in_maptki`), and again the movers are novel
identity/immune regulons (Tcf7l2, Zfp704, Hif1a, the Fos/Junb AP-1 set,
Irf8), not the section-14 leads. The two recovered comparison regulons are
quiet here: Spi1(+) trends negative at both amyloid contrasts
(*t* &approx; &minus;2.2) but does not survive FDR, and Rel(+) is flat
throughout. The heatmap shows the full activity field.

```{r scenic-aucell-heatmap, fig.width = 8, fig.height = 9.5, fig.cap = "Native AUCell activity (limma logFC on logit pseudobulk) for all 51 consensus regulons (rows, ordered by nlgf_in_p301s effect) across the five contrasts (columns). Fill = logFC (blue down, red up); * = FDR < 0.10 within contrast. Signal concentrates on the amyloid main effects; the interaction column carries no significant regulon."}
ord <- sc_auc |>
  dplyr::filter(contrast == "nlgf_in_p301s") |>
  dplyr::arrange(logFC)
hm <- sc_auc
hm$regulon  <- factor(hm$regulon, levels = ord$regulon)
hm$contrast <- factor(hm$contrast, levels = contrast_order)
hm$star <- ifelse(hm$sig, "*", "")
lim <- max(abs(hm$logFC), na.rm = TRUE)
ggplot(hm, aes(contrast, regulon, fill = logFC)) +
  geom_tile(colour = "grey90") +
  geom_text(aes(label = star), size = 3, vjust = 0.78) +
  scale_fill_gradientn(colours = diverging, limits = c(-lim, lim),
                       name = "logFC") +
  labs(x = NULL, y = NULL,
       title = "Native AUCell regulon activity across contrasts") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_text(size = 6.5))
```

## Per-substate regulon activity (NF-&kappa;B focus)

Readout 3 mirrors section 18: the controlled decoupleR swap is run within
each of the four microglia substates on the per-state NEBULA *z*, at the
interaction contrast. The section-18 result this tests against is the
**all-substate NF-&kappa;B attenuation** &mdash; does the *data-driven*
NF-&kappa;B regulon reproduce it? It does not. The recovered NF-&kappa;B
regulon **Rel** (and the recovered amyloid regulon Spi1, shown for
reference) is **non-significant at the interaction in all four substates**;
the closest approach is Rel in the IFN substate (FDR 0.109).

```{r scenic-substate-focus, results = "asis"}
foc <- sc_sub[sc_sub$is_focus, c("substate", "source", "cons_score",
                                 "padj", "sig")]
foc$substate <- factor(foc$substate, levels = substate_order)
foc <- foc[order(foc$source, foc$substate), ]
foc$sig <- ifelse(foc$sig, "yes", "no")
knitr::kable(
  foc, digits = 3, row.names = FALSE,
  col.names = c("substate", "regulon", "consensus", "FDR", "sig"),
  caption = paste0(
    "The two recovered comparison regulons (Rel = NF-kB family; Spi1 = ",
    "amyloid lead) at the interaction contrast, per substate. Neither is ",
    "significant in any substate: the data-driven NF-kB regulon does not ",
    "reproduce the section-18 all-substate attenuation."))
```

That non-reproduction is, however, **not** an absence of data-driven
interaction signal &mdash; it is specific to the section-14/18 TFs. The
per-substate decoupleR readout finds **many** novel regulons significant
at the interaction within substates (22 in homeostatic, 18 in DAM, 19 in
IFN, 24 in proliferative), spanning the same immune/identity programmes
seen elsewhere (Ets1, Klf2, Hif1a, Zbtb20, Mitf, Pax5, the STAT/IRF set).
This stands in deliberate contrast to the native AUCell factorial, which
found zero interaction regulons: the per-state NEBULA *z* is a single-cell
DE statistic with far more power than 16-id pseudobulk means, so the two
readouts probe different quantities (DE-weighted regulon enrichment vs a
shift in mean regulon activity). Both are reported; neither is privileged.

```{r scenic-substate-heatmap, fig.width = 8, fig.height = 8.5, fig.cap = "Per-substate SCENIC regulon activity at the interaction contrast (decoupleR consensus on the per-state NEBULA z), for every regulon significant in >=1 substate plus the two focus regulons. Fill = consensus score (blue down, red up); * = FDR < 0.10 within substate; the focus regulons Rel and Spi1 are tagged (focus) on the axis. They stay flat (no significant cell) while novel immune regulons carry the per-substate interaction signal."}
keep <- unique(c(sc_sub$source[sc_sub$sig],
                 sc_sub$source[sc_sub$is_focus]))
sh <- sc_sub[sc_sub$source %in% keep, ]
# Tag the two focus regulons in the row label itself (a supported,
# device-safe alternative to per-label face styling).
focus_set <- unique(sc_sub$source[sc_sub$is_focus])
sh$row_lab <- ifelse(sh$source %in% focus_set,
                     paste0(sh$source, " (focus)"), sh$source)
# Order rows by mean consensus across substates.
row_ord <- sh |>
  dplyr::group_by(row_lab) |>
  dplyr::summarise(m = mean(cons_score), .groups = "drop") |>
  dplyr::arrange(m)
sh$row_lab  <- factor(sh$row_lab, levels = row_ord$row_lab)
sh$substate <- factor(sh$substate, levels = substate_order)
sh$star <- ifelse(sh$sig, "*", "")
lim2 <- max(abs(sh$cons_score), na.rm = TRUE)
ggplot(sh, aes(substate, row_lab, fill = cons_score)) +
  geom_tile(colour = "grey90") +
  geom_text(aes(label = star), size = 3, vjust = 0.78) +
  scale_fill_gradientn(colours = diverging, limits = c(-lim2, lim2),
                       name = "consensus") +
  labs(x = NULL, y = NULL,
       title = "Per-substate regulon activity at the interaction contrast") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 7))
```

## Verdict (orthogonal corroboration vs honest discordance)

The synthesis is read per comparison axis, with absence reported as
evidence and no axis privileged. The thresholds are restated once:
decoupleR `ulm` significance + `consensus` direction at FDR < 0.10 on the
NEBULA *z* (readouts 1 and 3) and the locked 2&times;2 factorial on logit
AUCell (readout 2); consensus regulons at &ge; 8/10 edge recurrence and
&ge; 5 targets; all regulons activating.

**Overall.** The data-driven layer is **orthogonal to, and largely
discordant with, the section-14 prior.** Target overlap is near-zero
(median Jaccard `r sprintf("%.3f", med_jac)`), the controlled swap
produces near-zero activity correlation, and of the section-14 verdict
TFs only Spi1 is even present in the SCENIC network &mdash; where it
sign-flips and loses significance. This is not a power failure:
`r n_act_sig` of `r n_reg` regulons are significant on the same *z* and
`r n_auc_sig` move in native AUCell, but the movers are novel
co-expression-defined regulons (Zbtb20, Fli1, Hif1a, the AP-1/IRF/STAT
immune programmes), not the literature leads. The honest reading is that
SCENIC corroborates the amyloid Spi1 call at the *motif/recovery* level
only, gives the NF-&kappa;B axis partial recovery (Rel in; Nfkb1, Relb
near), and does **not** recover the interaction-metabolic axis at the
strict bar &mdash; half of it sits below the microglial expression floor.

```{r scenic-verdict-build}
sc_verdict <- tibble::tibble(
  axis = c("amyloid_activation", "interaction_metabolic",
           "nfkb_family_sec18"),
  comparison_tfs = c("Spi1, Nfkb1, Sp3",
                     "Myc, Creb1, Tp53, Jun",
                     "Rela, Nfkb1, Nfkb2, Relb, Rel"),
  scenic_recovery = c(
    "Spi1 recovered (+10/10, 15 targets); Nfkb1 7/10 & Sp3 7/10 near-miss",
    "none recovered; Myc & Tp53 expression-floor (structural), Creb1 6/10 near-miss, Jun 2/10 weak",
    "Rel recovered (+10/10, 9 targets); Relb 10/10 but <5 targets; Nfkb1 7/10 near-miss; Nfkb2 motif-pruned; Rela 1/10 weak"),
  headtohead = c(
    "Spi1 the only shared regulon: activity sign-flips & loses sig (CollecTRI +2.73* -> SCENIC -0.43 ns); shares 2/381 targets",
    "axis entirely CollecTRI-only; Myc interaction -6.60* has no SCENIC counterpart",
    "Rel has no bare-symbol CollecTRI source; Nfkb1/Rela/Sp3 CollecTRI-sig at amyloid but SCENIC-absent"),
  native_aucell = c(
    "Spi1(+) trends negative at amyloid contrasts but ns; recovered regulon carries no main-effect call",
    "no regulon sig at interaction (0/51); recovered axis TFs absent from network",
    "Rel(+) flat across all contrasts"),
  per_substate = c(
    "Spi1 ns at interaction in all 4 substates",
    "section-14 TFs absent, but novel regulons (Ets1/Klf2/Hif1a/Zbtb20) carry strong per-substate interaction signal",
    "Rel ns at interaction in all 4 substates (IFN FDR 0.109 closest); section-18 attenuation NOT reproduced"),
  verdict = c(
    "motif-level recovery, activity-level discordance",
    "axis unrecovered at the strict bar (expression floor); data-driven interaction signal lives in novel regulons",
    "partial regulon recovery, no activity-level corroboration"))
readr::write_tsv(sc_verdict,
                 file.path(params$results_dir, "scenic_verdict.tsv"))
```

```{r scenic-verdict-table, results = "asis"}
knitr::kable(
  sc_verdict[, c("axis", "comparison_tfs", "scenic_recovery", "headtohead",
                 "verdict")],
  col.names = c("Axis", "Comparison TFs", "SCENIC recovery",
                "Head-to-head (vs CollecTRI)", "Verdict"),
  caption = paste0(
    "Per-axis SCENIC verdict. The synaptic-suppression axis has no ",
    "section-14 TF endpoint to compare and so carries no row (consistent ",
    "with the sections 14/20 non-finding for that axis). Full prose incl. ",
    "native-AUCell and per-substate columns in ",
    "storage/results/scenic_verdict.tsv."))
```

The **synaptic-suppression** axis is absent from the table by
construction: section 14 returned no TF verdict for it (it reads as an
engulfment / post-translational programme, not a regulator-driven one), so
there is no prior TF call for SCENIC to confirm or refute &mdash; the same
honest non-finding the TF (14) and causal (20) layers recorded. SCENIC's
largest activity signals do fall on the shared NLGF contrasts but belong
to the novel immune regulon set, not to any synaptic-suppression claim.

**The boundary conditions, restated as the verdict's caveats.** These are
microglial-state regulons from single-nucleus co-expression, identity-
inherited from section 14; the consensus is all-activating, so any
section-14 repressing call is uncomparable by construction rather than
discordant; and the GRNBoost2 step is stochastic, controlled here by the
ten-run &ge; 8/10 consensus rather than a single seed. Within those
bounds the data-driven layer delivers a clean, falsifiable result: run
without a literature prior, this dataset's own microglia reconstruct a
regulatory programme that overlaps the section-14 prior almost not at all
&mdash; corroborating Spi1 and Rel as *recoverable regulons* while
declining to reproduce their prior-based *activity* calls, and locating
the bulk of the regulatory signal in a co-expression-defined TF set the
literature prior never nominated.
