# Microglial activation trajectory: pseudotime dynamics

This section adds the project's first **dynamics / progression** readout and
closes narrative arc **M** (`storage/notes/trajectory_plan.md`). Every
preceding arc reads a **static** quantity: expression DE (section 3), pathway
and module enrichment, the prior-based and data-driven transcription-factor
layers, the kinase and ligand-receptor layers, the causal topology, the
SCENIC regulons, and &mdash; most recently &mdash; discrete cell
**composition** (the GeoMx deconvolution of section 22). None reads the
**ordering** of microglia along an activation continuum. Here a microglial
activation trajectory (homeostatic &rarr; DAM) is inferred from the snRNAseq
itself, every cell is assigned a pseudotime, and the question is whether
**amyloid (NLGF) advances microglia along the homeostatic&rarr;DAM activation
axis**, and whether the **tau background modulates that advance** &mdash; the
same 2&times;2 interaction the rest of the project tracks,
`(NLGF_P301S&minus;P301S)&minus;(NLGF_MAPTKI&minus;MAPTKI)`, now re-expressed
as a *progression* effect rather than an expression, activity, or composition
one.

**Why this is an independent readout, not a restatement of sections 3 or 22.**
Static substate DE asks *which genes change within a fixed cluster set*;
composition (section 22) asks *how many cells fall in each discrete cluster*.
A pseudotime asks *how far along the activation program each cell sits on a
continuum* &mdash; including within-cluster advancement that neither DE nor
discrete composition can see. The honest risk (the load-bearing guardrail) is
that a per-replicate mean-pseudotime shift could be **entirely a composition
shift** (more DAM cells &rArr; higher mean pseudotime), in which case it
reduces to section 22; the design therefore *decomposes* progression from
reshuffling and reports both channels side by side.

**Method, fixed before any contrast was read.** The trajectory is
`slingshot` (Street 2018) run on the batch-corrected **harmony** embedding
truncated to the first `r {tj <- readRDS(file.path(params$cache_dir, "trajectory.rds")); tj$params$n_dims}`
dimensions, with the four microglial `state`s (homeostatic, DAM, IFN,
proliferative) as the cluster labels and the **root fixed at homeostatic**.
The 15-dimension truncation is the documented cell-cycle mitigation: at the
full 30 dimensions the cycling `proliferative` state is pulled in as a
spurious intermediate. The headline lineage is the **clean homeostatic&rarr;DAM
trajectory** built with the two cycling/peripheral side-states (IFN,
proliferative) omitted from lineage inference; the all-states topology is
reported alongside it. Pseudotime is summarised **per replicate**
(`genotype_batch`, the locked 16-id unit) and pushed through the project's
locked 5-contrast 2&times;2 factorial (`~ 0 + genotype + batch`,
`make_contrast_matrix`, limma-trend) exactly as every other modality. The
ordering is cross-checked against three independent methods (scanpy PAGA/DPT,
CellRank2 velocity-free fate probabilities, nested all-states Slingshot) and
the asserted root is validated against three root-free potency proxies
(RNA Shannon entropy, detected-gene count, CytoTRACE2).

**Thresholds, stated before any result is read.** Significance everywhere is
BH-FDR &lt; 0.10 (the project standard), applied across measures within each
contrast. The DAM-onset threshold for the *fraction-past* progression measure
is the global median pseudotime of DAM-labelled cells
(`r sprintf("%.1f", tj$params$dam_threshold)`), fixed once, data-driven.
Pseudotime-scale measures and bounded [0,1] measures are fit in separate
eBayes calls so the shared variance prior is not distorted across scales.

**Caveats carried throughout.** (a) Pseudotime is a single inferred
**activation ordering**, not chronological or developmental time; it is framed
as "activation progression" throughout. (b) The four substates are
transcriptionally close, so a raw per-replicate mean shift is substantially
composition-driven &mdash; the decomposition and the cross-method concordance,
not the raw mean, carry the interpretation. (c) RNA velocity is excluded (no
spliced layers; snRNA capture unreliable). (d) The root is *validated, not
asserted*: as reported below, microglial "homeostatic" is a mature resting
state, not a stem state, so the potency proxies do **not** rank it most
potent &mdash; the trajectory direction rests on the Slingshot
homeostatic&rarr;DAM monotonicity and canonical marker biology, not on a
potency claim.

**Anti-anchoring.** All four states and both lineage topologies are reported
even-handedly; absence of an IFN or proliferative effect is evidence, not
omission. The cross-method pseudotime concordance is reported in full, a
discordant method included rather than smoothed. A null/additive interaction
would be a valid finding; it is not fished for, and the lineage was not
retuned after inspecting which contrast moves.

```{r trajectory-load}
tj   <- readRDS(file.path(params$cache_dir, "trajectory.rds"))
pc   <- tj$per_cell                     # 26,104 x 14 (per-cell pt + potency + umap)
ct   <- tj$contrasts                    # 45 rows: measure x contrast factorial
bg   <- tj$by_genotype                  # 36 rows: measure x genotype mean/sd
dec  <- tj$decomposition                # progression-vs-composition partition
rv   <- tj$root_validation              # 3 potency proxies x per-state ranks
conc <- tj$concordance                  # cross-method Spearman
par  <- tj$params
prov <- tj$python_provenance

geno_order     <- c("MAPTKI", "P301S", "NLGF_MAPTKI", "NLGF_P301S")
contrast_order <- c("nlgf_in_maptki", "nlgf_in_p301s", "interaction",
                    "tau_alone", "tau_in_nlgf")
diverging      <- c("#3a4cc0", "white", "#b40426")
pc$state       <- factor(pc$state,
                         levels = c("homeostatic", "DAM", "IFN", "proliferative"))
pc$genotype    <- factor(pc$genotype, levels = geno_order)
bg$genotype    <- factor(bg$genotype, levels = geno_order)

# Unique-row lookups (each measure x contrast and measure x genotype is unique),
# so prose scalars are read straight from the cache and never drift.
ct_get <- function(m, cn) ct[ct$measure == m & ct$contrast == cn, ]
bg_get <- function(m, g)  bg$mean[bg$measure == m & as.character(bg$genotype) == g]

# Headline scalars.
pt_nm   <- ct_get("mean_pt", "nlgf_in_maptki")
pt_np   <- ct_get("mean_pt", "nlgf_in_p301s")
pt_int  <- ct_get("mean_pt", "interaction")
pt_tau  <- ct_get("mean_pt", "tau_alone")
comp_int <- ct_get("composition_cf", "interaction")
prog_int <- ct_get("progression_cf", "interaction")
fp_int   <- ct_get("frac_past", "interaction")
wh_int   <- ct_get("within_homeostatic", "interaction")
wd_int   <- ct_get("within_DAM", "interaction")
cr_int   <- ct_get("cellrank_dam_fate", "interaction")

# Root validation: rank of the asserted root (homeostatic) in each proxy.
root_rank <- function(p) rv[[p]]$root_rank
ct2_root  <- rv$cytotrace2

# Cross-method concordance pairs.
cc <- conc$pairwise
cc_get <- function(a, b) {
  r <- cc[(cc$method_a == a & cc$method_b == b) |
          (cc$method_a == b & cc$method_b == a), ]
  r$rho[1]
}
```

## Trajectory topology and root validation

Slingshot on the 15-dimension harmony embedding returns a single clean
**homeostatic&rarr;DAM** lineage when the two cycling/peripheral side-states
are omitted, and a single all-states ordering
homeostatic&rarr;proliferative&rarr;IFN&rarr;DAM when they are retained; in
both, **DAM is the terminal state** and homeostatic the fixed root. The two
runs are near-identical where they overlap (Spearman
`r sprintf("%.2f", cc_get("slingshot_clean", "slingshot_all"))` over the
shared cells), so omitting the side-states sharpens the headline axis without
re-ordering it. An independent scanpy PAGA graph over the same embedding
recovers the same broad connectivity (state order
`r paste(prov$paga_state_order, collapse = ", ")`), corroborating the topology
by a method that shares none of Slingshot's curve machinery.

The UMAP below shows the four states and the per-cell pseudotime on the clean
lineage. Pseudotime increases smoothly from the homeostatic body into the DAM
region; the IFN and proliferative cells (grey in the right panel) are off the
clean lineage by construction and carry no clean-lineage pseudotime.

```{r trajectory-umap-state, fig.width = 6.6, fig.height = 5, fig.cap = "Microglial UMAP coloured by the four activation states. Slingshot's clean lineage runs homeostatic -> DAM (root fixed at homeostatic, DAM terminal); IFN and proliferative are treated as cycling/peripheral side-states and omitted from the headline lineage inference."}
plot_trajectory_umap(
  as.matrix(pc[, c("umap_1", "umap_2")]), pc$state,
  colour_lab = "state",
  title = "Microglial activation states on UMAP") +
  scale_colour_brewer(palette = "Dark2") +
  guides(colour = guide_legend(override.aes = list(size = 2.5, alpha = 1)))
```

```{r trajectory-umap-pseudotime, fig.width = 6.6, fig.height = 5, fig.cap = "Same UMAP coloured by clean-lineage (homeostatic -> DAM) pseudotime. Off-lineage IFN/proliferative cells (NA pseudotime) are grey. Pseudotime rises monotonically from the homeostatic body into the DAM territory."}
plot_trajectory_umap(
  as.matrix(pc[, c("umap_1", "umap_2")]), pc$pt_clean,
  colour_lab = "pseudotime",
  title = "Homeostatic -> DAM pseudotime on UMAP")
```

The root is **validated, not asserted**, and the honest result is that the
asserted homeostatic root is *not* the most-potent state by any root-free
proxy. Across RNA Shannon entropy, detected-gene count and CytoTRACE2 (the
load-bearing arbiter), homeostatic ranks `r root_rank("entropy")`,
`r root_rank("n_genes")` and `r root_rank("cytotrace2")` of four respectively,
and is `most_potent = FALSE` in every case (CytoTRACE2 &Delta; to the
next-highest state = `r sprintf("%.3f", ct2_root$delta_to_next)`). This is the
expected confound: activated microglia induce many response genes, inflating
crude potency, and "homeostatic" microglia are a *mature resting* population,
not a developmental stem state. The trajectory direction therefore rests on
the Slingshot homeostatic&rarr;DAM monotonicity (DAM mean pseudotime
`r sprintf("%.1f", dec$global$mu_bar[dec$global$state == "DAM"])` &gt;
homeostatic `r sprintf("%.1f", dec$global$mu_bar[dec$global$state == "homeostatic"])`)
plus canonical marker biology, not on a potency claim &mdash; the pseudotime
is an **activation ordering**, not developmental time.

```{r trajectory-root-table, results = "asis"}
root_tab <- data.frame(
  state = rv$cytotrace2$per_state$state,
  stringsAsFactors = FALSE)
root_tab$`entropy rank`    <- rv$entropy$per_state$rank[match(root_tab$state, rv$entropy$per_state$state)]
root_tab$`n_genes rank`    <- rv$n_genes$per_state$rank[match(root_tab$state, rv$n_genes$per_state$state)]
root_tab$`CytoTRACE2 mean` <- sprintf("%.3f", rv$cytotrace2$per_state$mean_potency[match(root_tab$state, rv$cytotrace2$per_state$state)])
root_tab$`CytoTRACE2 rank` <- rv$cytotrace2$per_state$rank[match(root_tab$state, rv$cytotrace2$per_state$state)]
root_tab <- root_tab[order(root_tab$`CytoTRACE2 rank`), ]
knitr::kable(
  root_tab, row.names = FALSE,
  caption = paste0(
    "Root-free potency by state across three proxies (rank 1 = most potent / ",
    "least differentiated). The asserted homeostatic root ranks last or ",
    "near-last in every proxy, so the lineage is an ACTIVATION ordering, not ",
    "a developmental-potency one; the direction rests on Slingshot ",
    "monotonicity and marker biology. Reported honestly rather than forced."))
```

## Pseudotime distribution by genotype

The per-cell clean-lineage pseudotime, split by the 2&times;2, is the direct
visual of the effect. In both tau backgrounds the introduction of amyloid
shifts the whole distribution rightward &mdash; mass moves from the
homeostatic body toward the DAM tail &mdash; while the two amyloid-free
genotypes (MAPTKI, P301S) sit together at low pseudotime. The mean
clean-lineage pseudotime rises from
`r sprintf("%.1f", bg_get("mean_pt", "MAPTKI"))` (MAPTKI) and
`r sprintf("%.1f", bg_get("mean_pt", "P301S"))` (P301S) to
`r sprintf("%.1f", bg_get("mean_pt", "NLGF_MAPTKI"))` (NLGF_MAPTKI) and
**`r sprintf("%.1f", bg_get("mean_pt", "NLGF_P301S"))`** (NLGF_P301S, the
highest).

```{r trajectory-density, fig.width = 7, fig.height = 4.4, fig.cap = "Per-cell clean-lineage (homeostatic -> DAM) pseudotime density by genotype. Amyloid (NLGF, warm colours) shifts the distribution rightward in both tau backgrounds; the two amyloid-free genotypes overlap at low pseudotime. The NLGF_P301S curve carries the most DAM-end mass."}
plot_pseudotime_density(
  pc, pt_col = "pt_clean", genotype_col = "genotype",
  title = "Activation pseudotime shifts rightward under amyloid")
```

## Factorial interaction test

The per-replicate pseudotime summary is tested with the locked 2&times;2
factorial. The headline measure is the per-replicate mean clean-lineage
pseudotime; amyloid advances it strongly in **both** tau backgrounds
(`nlgf_in_maptki` logFC `r sprintf("%+.2f", pt_nm$logFC)`,
FDR `r sprintf("%.1e", pt_nm$adj.P.Val)`; `nlgf_in_p301s`
`r sprintf("%+.2f", pt_np$logFC)`, FDR `r sprintf("%.1e", pt_np$adj.P.Val)`),
while **tau alone does not** (`tau_alone`
`r sprintf("%+.2f", pt_tau$logFC)`, FDR
`r sprintf("%.2f", pt_tau$adj.P.Val)`, ns). The tau (P301S) background then
**supra-additively amplifies the amyloid advance**: the interaction is
positive and significant (logFC **`r sprintf("%+.2f", pt_int$logFC)`**, FDR
`r sprintf("%.3f", pt_int$adj.P.Val)`), the gap between the two amyloid
contrasts (`r sprintf("%.2f", pt_np$logFC)` &minus;
`r sprintf("%.2f", pt_nm$logFC)`). This is the project's first
tau&times;amyloid interaction that is both positive **and** significant in a
non-expression layer.

```{r trajectory-forest, fig.width = 7, fig.height = 3.8, fig.cap = "Forest of the five canonical contrasts on per-replicate mean clean-lineage pseudotime (logFC +/- approximate Wald 95% CI). Filled = FDR < 0.10. Amyloid (nlgf_in_maptki, nlgf_in_p301s) advances pseudotime strongly in both backgrounds; tau alone does not; the tau x amyloid interaction is positive and significant."}
plot_interaction_forest(
  ct, measure = "mean_pt", padj_cut = par$padj_cut,
  title = "Mean activation-pseudotime: 2x2 factorial contrasts")
```

The same picture holds across every progression measure and is summarised in
the heatmap below: all nine per-replicate measures against all five contrasts.
The two amyloid main-effect columns are saturated; the interaction column is
positive and significant on the genuine progression measures (mean pseudotime,
fraction-past-DAM-threshold, within-homeostatic advancement, and the
progression channel of the decomposition) and null on the composition channel
&mdash; the dissociation that the next subsection makes explicit.

```{r trajectory-heatmap, fig.width = 7.6, fig.height = 5.4, fig.cap = "Factorial contrasts across all nine per-replicate measures. Fill = logFC (blue down, red up); * = FDR < 0.10. Amyloid main effects saturate; the interaction column is significant on the progression measures and null on composition_cf. `observed` reproduces `mean_pt` exactly (law-of-total-expectation check); `mean_pt_all` is the secondary all-states lineage."}
measure_order <- c("mean_pt", "frac_past", "within_homeostatic", "within_DAM",
                   "progression_cf", "composition_cf", "cellrank_dam_fate",
                   "mean_pt_all", "observed")
hm <- ct
hm$measure  <- factor(hm$measure, levels = rev(measure_order))
hm$contrast <- factor(hm$contrast, levels = contrast_order)
hm$star     <- ifelse(hm$sig, "*", "")
lim <- max(abs(hm$logFC), na.rm = TRUE)
ggplot(hm, aes(contrast, measure, fill = logFC)) +
  geom_tile(colour = "grey90") +
  geom_text(aes(label = star), size = 4, vjust = 0.78) +
  scale_fill_gradientn(colours = diverging, limits = c(-lim, lim),
                       name = "logFC") +
  labs(x = NULL, y = NULL,
       title = "Trajectory contrasts across per-replicate measures") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 25, hjust = 1))
```

## Progression versus composition, and cross-method concordance

This is the load-bearing analysis. The per-replicate mean pseudotime is
partitioned into a **composition** channel (vary the between-state fractions,
hold within-state means at the global value) and a **progression** channel
(vary the within-state means, hold the fractions at the global value). If the
interaction lived in the composition channel it would simply restate the
section-22 composition result (more DAM cells); if it lives in the progression
channel it is genuine within-state advancement that no discrete-composition or
DE analysis can see.

The interaction is carried **almost entirely by progression**
(`progression_cf` logFC `r sprintf("%+.2f", prog_int$logFC)`, FDR
`r sprintf("%.3f", prog_int$adj.P.Val)`, significant) and **not by
composition** (`composition_cf` `r sprintf("%+.2f", comp_int$logFC)`, FDR
`r sprintf("%.2f", comp_int$adj.P.Val)`, null) &mdash; the progression channel
recovers `r sprintf("%.0f%%", 100 * prog_int$logFC / pt_int$logFC)` of the
total interaction. The amyloid **main** effects, by contrast, split across
*both* channels (composition `r sprintf("%+.2f", ct_get("composition_cf", "nlgf_in_maptki")$logFC)` /
`r sprintf("%+.2f", ct_get("composition_cf", "nlgf_in_p301s")$logFC)` and
progression `r sprintf("%+.2f", ct_get("progression_cf", "nlgf_in_maptki")$logFC)` /
`r sprintf("%+.2f", ct_get("progression_cf", "nlgf_in_p301s")$logFC)`, all
significant): amyloid both makes more DAM cells *and* advances them, whereas
the tau background adds only the progression piece. Two corroborating
progression measures agree: the fraction of cells past the DAM-onset threshold
has the strongest interaction
(`r sprintf("%+.2f", fp_int$logFC)`, FDR `r sprintf("%.3f", fp_int$adj.P.Val)`;
absolutes `r sprintf("%.0f%%", 100 * bg_get("frac_past", "MAPTKI"))` /
`r sprintf("%.0f%%", 100 * bg_get("frac_past", "P301S"))` at baseline &rarr;
`r sprintf("%.0f%%", 100 * bg_get("frac_past", "NLGF_MAPTKI"))` /
`r sprintf("%.0f%%", 100 * bg_get("frac_past", "NLGF_P301S"))` under amyloid),
and within-homeostatic advancement is significant
(`r sprintf("%+.2f", wh_int$logFC)`, FDR
`r sprintf("%.3f", wh_int$adj.P.Val)`) &mdash; the amplification pushes even
homeostatic-labelled cells rightward, which is advancement that *cannot* be a
composition artefact.

```{r trajectory-decomposition, fig.width = 7.4, fig.height = 3.8, fig.cap = "Progression-vs-composition decomposition of the amyloid and interaction contrasts (logFC +/- approximate 95% CI; filled = FDR < 0.10). Amyloid main effects load on BOTH channels; the tau x amyloid interaction loads on PROGRESSION only (composition null), so the dynamics interaction is non-redundant with the section-22 composition layer."}
dd <- ct[ct$measure %in% c("composition_cf", "progression_cf") &
         ct$contrast %in% c("nlgf_in_maptki", "nlgf_in_p301s", "interaction"), ]
dd$channel  <- factor(ifelse(dd$measure == "composition_cf",
                             "composition", "progression"),
                      levels = c("composition", "progression"))
dd$contrast <- factor(dd$contrast,
                      levels = rev(c("nlgf_in_maptki", "nlgf_in_p301s",
                                     "interaction")))
dd$lo <- dd$logFC - 1.96 * dd$se
dd$hi <- dd$logFC + 1.96 * dd$se
ggplot(dd, aes(logFC, contrast)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.2) +
  geom_point(aes(fill = sig), shape = 21, size = 3) +
  scale_fill_manual(values = c(`TRUE` = "#B0344D", `FALSE` = "white"),
                    name = sprintf("FDR<%.2f", par$padj_cut)) +
  facet_wrap(~ channel) +
  labs(x = "logFC (pseudotime units)", y = NULL,
       title = "Interaction loads on progression, not composition") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```

Cross-method concordance is reported in full, discordance included. Two of
three independent methods corroborate the Slingshot ordering: the nested
all-states Slingshot at Spearman
`r sprintf("%.2f", cc_get("slingshot_clean", "slingshot_all"))` and CellRank2
velocity-free DAM-fate probability at
`r sprintf("%.2f", cc_get("slingshot_clean", "cellrank_dam_fate"))`. The
third, scanpy diffusion pseudotime (DPT), is the outlier
(`r sprintf("%.2f", cc_get("slingshot_clean", "dpt"))`): DPT has weak dynamic
range on this graded activation continuum and is pulled by the peripheral
IFN/proliferative geometry, so it disagrees with both Slingshot *and* CellRank
(DPT&ndash;CellRank `r sprintf("%.2f", cc_get("dpt", "cellrank_dam_fate"))`).
This is exactly the kind of method-sensitivity the maximal-triangulation
design was meant to surface; it is reported, not smoothed. The CellRank
DAM-fate probability itself carries a positive but non-significant interaction
(`r sprintf("%+.3f", cr_int$logFC)`, FDR
`r sprintf("%.2f", cr_int$adj.P.Val)`), direction-concordant with the
pseudotime headline.

```{r trajectory-concordance, fig.width = 5.6, fig.height = 4.6, fig.cap = "Cross-method per-cell pseudotime concordance (Spearman). Slingshot-clean, the nested all-states Slingshot and CellRank2 agree; scanpy DPT is the known-weak outlier on this graded continuum (negative correlation with all curve-based methods)."}
cm  <- conc$cor_mat
ord <- c("slingshot_clean", "slingshot_all", "cellrank_dam_fate", "dpt")
cm  <- cm[ord, ord]
cdf <- as.data.frame(as.table(cm))
names(cdf) <- c("m1", "m2", "rho")
ggplot(cdf, aes(m1, m2, fill = rho)) +
  geom_tile(colour = "grey92") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 3.2) +
  scale_fill_gradientn(colours = diverging, limits = c(-1, 1),
                       name = "Spearman") +
  labs(x = NULL, y = NULL,
       title = "Cross-method pseudotime concordance") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 25, hjust = 1))
```

## Verdict

The synthesis is read per comparison axis, absence reported as evidence and no
axis privileged. Thresholds restated once: BH-FDR &lt; 0.10 on the locked
2&times;2 factorial over per-replicate pseudotime summaries, Slingshot on the
15-dimension harmony embedding, root validated against three potency proxies,
ordering cross-checked against three independent methods.

**Overall.** The dynamics layer delivers a clean, new-evidence-class result on
both the amyloid axis and &mdash; uniquely among the orthogonal layers &mdash;
the divergence axis. Amyloid strongly advances microglia along the
homeostatic&rarr;DAM activation pseudotime in both tau backgrounds
(`nlgf_in_maptki` `r sprintf("%+.2f", pt_nm$logFC)`, `nlgf_in_p301s`
`r sprintf("%+.2f", pt_np$logFC)`, both FDR &lt; 1e-5), and the tau (P301S)
background **supra-additively amplifies** that advance (interaction
`r sprintf("%+.2f", pt_int$logFC)`, FDR `r sprintf("%.3f", pt_int$adj.P.Val)`).
The decomposition shows this interaction is carried by genuine
**progression** (`r sprintf("%+.2f", prog_int$logFC)`, FDR
`r sprintf("%.3f", prog_int$adj.P.Val)`), **not composition**
(`r sprintf("%+.2f", comp_int$logFC)`, null), so it is non-redundant with the
section-22 tissue-composition layer (whose interaction was null) and reveals a
tau&times;amyloid effect invisible to discrete composition. The null
composition-interaction here is exactly consistent with the section-22
SpatialDecon and the human cross-species interaction layers, confirming this
readout is a new channel rather than a restatement.

```{r trajectory-verdict-build}
tj_verdict <- tibble::tibble(
  axis = c("amyloid_activation", "interaction_metabolic",
           "synaptic_suppression"),
  dynamics_endpoint = c(
    "mean homeostatic->DAM pseudotime + fraction-past-DAM-threshold",
    "interaction on pseudotime, decomposed into progression vs composition",
    "no trajectory endpoint (synaptic/engulfment programme)"),
  result = c(
    sprintf(paste0("amyloid advances pseudotime in both backgrounds (logFC ",
                   "%+.2f nlgf_in_maptki, FDR %.1e; %+.2f nlgf_in_p301s, FDR ",
                   "%.1e); mean pt %.1f/%.1f -> %.1f/%.1f; frac-past ",
                   "%.0f%%/%.0f%% -> %.0f%%/%.0f%%"),
            pt_nm$logFC, pt_nm$adj.P.Val, pt_np$logFC, pt_np$adj.P.Val,
            bg_get("mean_pt", "MAPTKI"), bg_get("mean_pt", "P301S"),
            bg_get("mean_pt", "NLGF_MAPTKI"), bg_get("mean_pt", "NLGF_P301S"),
            100 * bg_get("frac_past", "MAPTKI"),
            100 * bg_get("frac_past", "P301S"),
            100 * bg_get("frac_past", "NLGF_MAPTKI"),
            100 * bg_get("frac_past", "NLGF_P301S")),
    sprintf(paste0("interaction %+.2f (FDR %.3f) carried by progression ",
                   "(%+.2f, FDR %.3f) not composition (%+.2f, FDR %.2f); ",
                   "frac-past interaction %+.2f (FDR %.3f)"),
            pt_int$logFC, pt_int$adj.P.Val, prog_int$logFC,
            prog_int$adj.P.Val, comp_int$logFC, comp_int$adj.P.Val,
            fp_int$logFC, fp_int$adj.P.Val),
    "no dynamics endpoint (synaptic axis is not a microglial activation state)"),
  verdict = c(
    "corroborated (new evidence class: activation dynamics)",
    "corroborated as a progression effect (positive, significant, non-composition)",
    "no dynamics endpoint (honest non-finding)"))
readr::write_tsv(tj_verdict,
                 file.path(params$results_dir, "trajectory_verdict.tsv"))
```

```{r trajectory-verdict-table, results = "asis"}
knitr::kable(
  tj_verdict[, c("axis", "dynamics_endpoint", "verdict")],
  col.names = c("Axis", "Dynamics endpoint", "Verdict"),
  caption = paste0(
    "Per-axis activation-trajectory verdict. The amyloid-activation axis is ",
    "corroborated as a dynamics/progression effect; the tau x amyloid ",
    "interaction is corroborated as a PROGRESSION effect (positive, FDR ",
    sprintf("%.3f", pt_int$adj.P.Val), ", composition-null) -- the project's ",
    "first significant interaction in an orthogonal layer; the synaptic axis ",
    "has no trajectory endpoint. Full column in ",
    "storage/results/trajectory_verdict.tsv."))
```

**The boundary conditions, restated as the verdict's caveats.** Pseudotime is
a single inferred activation ordering, not developmental or chronological time,
and the asserted homeostatic root is *not* the most-potent state by any
root-free proxy &mdash; the direction rests on Slingshot monotonicity and
marker biology, so the axis is read as activation progression, never as time
or potency. One of three cross-check methods (DPT) disagrees, a known weakness
of diffusion pseudotime on a graded continuum; the two curve/fate-based
methods agree. The interaction, though positive and significant, is a single
inferred ordering partly entangled with composition, which is why the
decomposition &mdash; isolating the `r sprintf("%.0f%%", 100 * prog_int$logFC / pt_int$logFC)`
progression share &mdash; is the load-bearing evidence and the result is
weighted as **suggestive** corroboration rather than proof. Within those
bounds the layer is a clean, falsifiable dynamics readout: amyloid advances
microglia along the activation continuum in both tau backgrounds, and the
P301S tau background supra-additively amplifies that advance through genuine
within-state progression, not cluster reshuffling.

## Gene-level differential dynamics at the interaction

```{r trajectory-dynamics-load}
td      <- readRDS(file.path(params$cache_dir, "trajectory_dynamics.rds"))
dyn_int <- td$interaction
dyn_vs  <- td$vs_static
dyn_par <- td$params
arc_m   <- td$arc_m_synergy
am_get  <- function(m) arc_m[arc_m$measure == m, ]   # M-002 aggregate synergy

# Significance counts: interaction / omnibus / association (nested), thresholds
# fixed in the cache (dyn_par$padj_cut), read straight so prose never drifts.
sig_genes   <- dyn_int[!is.na(dyn_int$adj.P.Val) & dyn_int$adj.P.Val < dyn_par$padj_cut, ]
n_dyn_sig   <- nrow(sig_genes)
n_omni_sig  <- sum(td$omnibus$adj.P.Val     < dyn_par$padj_cut, na.rm = TRUE)
n_assoc_sig <- sum(td$association$adj.P.Val < dyn_par$padj_cut, na.rm = TRUE)

# Static-vs-dynamic honesty: of the dynamic hits, how many reach static
# (section-24 matched-power Microglia interaction) significance, and the floor
# static FDR among them.
vs_sig          <- dyn_vs[dyn_vs$gene %in% sig_genes$gene, ]
n_static_sig_nb <- sum(vs_sig$stat_nb_adj.P.Val < dyn_par$padj_cut, na.rm = TRUE)
n_static_sig_pb <- sum(vs_sig$stat_pb_adj.P.Val < dyn_par$padj_cut, na.rm = TRUE)
min_static_nb   <- min(vs_sig$stat_nb_adj.P.Val, na.rm = TRUE)

# Ambient / non-microglial-lineage markers -- a heuristic honesty FLAG (genes
# are NOT removed; the raw genome-wide result is reported and the flag annotates
# the contamination caveat). Canonical RBC / endothelial-vascular / myelin sets.
ambient_markers <- c(
  "Hba-a1", "Hba-a2", "Hbb-bs", "Hbb-bt", "Hbb-b1",     # haemoglobin / RBC
  "Ly6c1", "Ramp2", "Cldn5", "Flt1", "Pecam1", "Vtn",   # endothelial / vascular
  "Igfbp7", "Enpp2", "Ttr",                             # vascular / choroid plexus
  "Mal", "Plp1", "Mbp", "Mobp", "Mog")                  # myelin / oligodendrocyte
sig_genes$ambient <- sig_genes$symbol %in% ambient_markers

# Ribosomal / translation genes -- the programme's one coherent enrichment and a
# second, orthogonal confound to the ambient flag (trajectory dynamics of
# translation machinery track cell-state / RNA content, not microglial
# signalling). FLAGGED for honesty, not removed.
sig_genes$ribosomal <- grepl("^(Rps|Rpl)", sig_genes$symbol)
n_ribo_sig <- sum(sig_genes$ribosomal)
ribo_bg    <- mean(grepl("^(Rps|Rpl)", dyn_int$symbol[!is.na(dyn_int$adj.P.Val)]))
ribo_fold  <- (n_ribo_sig / n_dyn_sig) / ribo_bg
```

The interaction above is an **aggregate** rate effect: a per-replicate
mean-pseudotime shift, mechanism-agnostic, naming no gene. This final subsection
deepens it (it post-dates the verdict above) by asking whether a *gene-level*
tau&times;amyloid interaction lives in the **dynamics of expression along the
pseudotime** &mdash; a question two facts make pointed. First, the project's
*static* gene-level interaction &mdash; the same 2&times;2 contrast applied to
expression within a fixed cluster set &mdash; collapses to essentially **zero
genes at matched power in every cell type, microglia included** (section 24).
Second, an aggregate mean shift could in principle arise with no gene having a
tau-modulated, pseudotime-localised amyloid response: it could be pure
reshuffling of cells along a fixed family of curves. The question that sits
exactly in that gap is whether there exist **genes whose amyloid response is
localised to a pseudotime window and is modulated by the tau background** that a
static pseudobulk contrast cannot see. A positive answer would be the project's
first gene-level positive interaction and would hand the aggregate progression
synergy a candidate mechanism; a null would say the synergy lives only in
aggregate rate, with no gene-specific dynamic signature. **Either outcome is a
clean finding; neither is fished for.**

**Method, thresholds stated first.** The readout is lineage differential
expression with **tradeSeq** negative-binomial GAMs (Van den Berge 2020): a
separate smoother of pseudotime is fit per genotype on the clean
homeostatic&rarr;DAM lineage (`conditions` = the four genotypes,
`nknots` = `r dyn_par$nknots`), genome-wide over the
`r format(dyn_par$gene_set_size, big.mark = ",")` genes expressed in at least 1%
of on-lineage cells (`r format(td$n_fitted, big.mark = ",")` fitted; the
remainder rank-deficient). The interaction is a **difference-of-differences Wald
contrast** on the smoother coefficients &mdash; the identical linear algebra
tradeSeq's `conditionTest` uses for a single pairwise pattern difference, with
the pairwise &plusmn;1 replaced by the locked 2&times;2 weights
`(NLGF_P301S&minus;P301S)&minus;(NLGF_MAPTKI&minus;MAPTKI)` per knot &mdash; and
was fixed before any gene was inspected. Significance is BH-FDR &lt;
`r sprintf("%.2f", dyn_par$padj_cut)`. Per the anti-anchoring guardrail the
targeted interaction is reported alongside the **omnibus** `conditionTest` (do
the four smoothers differ at all) and the per-gene **association** test (is the
gene dynamic at all), never the interaction alone. The same arc-M caveats hold
verbatim: pseudotime is a single inferred **activation ordering**, not
developmental time, and the substates are transcriptionally close.

On-lineage cell counts per genotype &mdash; MAPTKI
`r format(td$on_lineage_counts$n_cells[1], big.mark = ",")`, P301S
`r format(td$on_lineage_counts$n_cells[2], big.mark = ",")`, NLGF_MAPTKI
`r format(td$on_lineage_counts$n_cells[3], big.mark = ",")`, NLGF_P301S
`r format(td$on_lineage_counts$n_cells[4], big.mark = ",")` &mdash; are all
well-powered, the amyloid genotypes carrying roughly twice the cells (DAM
expansion); no condition is thin.

The targeted contrast finds **`r n_dyn_sig` genes with a significant
tau&times;amyloid interaction in their pseudotime dynamics** at FDR &lt;
`r sprintf("%.2f", dyn_par$padj_cut)`, correctly nested inside the omnibus
(`r format(n_omni_sig, big.mark = ",")` genotype-varying genes) and the
association test (`r format(n_assoc_sig, big.mark = ",")` dynamic genes) &mdash;
every interaction gene is, as it must be, both genotype-varying and dynamic. The
full set is tabulated below; the per-genotype smoother curves for the strongest
hits follow, where the four genotype lines diverging in *shape* (not merely in
level) is the interaction made visible.

```{r trajectory-dynamics-table}
# DT::datatable of all FDR<0.10 interaction-dynamics genes (mirrors the section-17
# ledger datatable style). effect_magnitude = L2 norm of the diff-of-differences
# across knots; effect_peak = its signed value at the peak-effect pseudotime.
# The static columns are the section-24 matched-power Microglia interaction FDRs
# (NEBULA / pseudobulk) -- the static-vs-dynamic honesty join.
dyn_view <- sig_genes |>
  dplyr::transmute(
    symbol,
    gene,
    waldStat         = round(waldStat, 2),
    df,
    interaction_FDR  = signif(adj.P.Val, 3),
    effect_magnitude = round(effect_l2, 2),
    effect_peak      = round(effect_peak, 2),
    ambient          = ifelse(ambient, "ambient", "")) |>
  dplyr::left_join(
    dplyr::select(dyn_vs, gene,
                  static_NEBULA_FDR     = stat_nb_adj.P.Val,
                  static_pseudobulk_FDR = stat_pb_adj.P.Val),
    by = "gene") |>
  dplyr::mutate(static_NEBULA_FDR     = signif(static_NEBULA_FDR, 3),
                static_pseudobulk_FDR = signif(static_pseudobulk_FDR, 3)) |>
  dplyr::select(-gene) |>
  dplyr::arrange(dplyr::desc(waldStat))

DT::datatable(
  dyn_view,
  rownames   = FALSE,
  filter     = "top",
  extensions = c("Buttons", "FixedHeader"),
  caption    = paste0(
    "All ", n_dyn_sig, " genes with a significant tau x amyloid interaction in ",
    "their homeostatic->DAM pseudotime dynamics (tradeSeq diff-of-differences ",
    "Wald, BH-FDR < ", sprintf("%.2f", dyn_par$padj_cut), "). `ambient` flags ",
    "canonical RBC / endothelial / myelin transcripts (heuristic; genes are NOT ",
    "removed). The two static columns are the section-24 matched-power Microglia ",
    "interaction FDRs -- all >> 0.10, the static-vs-dynamic dissociation."),
  options = list(
    pageLength  = 12,
    lengthMenu  = list(c(10, 12, 25, 50, -1), c("10", "12", "25", "50", "All")),
    autoWidth   = FALSE, scrollX = TRUE, fixedHeader = TRUE,
    dom         = "Bfrtip",
    order       = list(list(1, "desc")),
    buttons     = list("copy", "csv",
                       list(extend = "colvis", text = "Column visibility")))) |>
  DT::formatStyle("ambient",
                  target = "row",
                  backgroundColor = DT::styleEqual("ambient", "#F7E7E7",
                                                   default = "white"))
```

```{r trajectory-dynamics-smoothers, fig.width = 8.4, fig.height = 8.4, fig.cap = "Per-genotype fitted expression smoothers (NB-GAM, 95% CI ribbon) along the homeostatic -> DAM pseudotime for the 12 strongest interaction-dynamics genes. The four genotype curves differing in SHAPE is the difference-of-differences interaction; cool colours = amyloid-free (MAPTKI, P301S), warm = amyloid (NLGF). Panels marked (ambient) carry canonical RBC/endothelial/myelin markers and are shown for honesty, not as microglial mechanism."}
n_panel <- 12L
top_g   <- head(sig_genes[order(-sig_genes$waldStat), ], n_panel)
top_g   <- top_g[top_g$gene %in% names(td$smoothers), ]   # guard (top-20 cached)
panel_lab <- ifelse(top_g$ambient, paste0(top_g$symbol, " (ambient)"), top_g$symbol)
sm_df <- do.call(rbind, lapply(seq_len(nrow(top_g)), function(i) {
  d <- td$smoothers[[top_g$gene[i]]]
  d$symbol <- panel_lab[i]
  d
}))
sm_df$symbol <- factor(sm_df$symbol, levels = panel_lab)
plot_condition_smoothers(
  sm_df, panel_col = "symbol", ncol = 3L,
  title = "Genotype expression smoothers: top interaction-dynamics genes")
```

**The load-bearing comparison is against the static contrast.** Of the
`r n_dyn_sig` genes with a significant *dynamic* interaction,
**`r n_static_sig_nb` reach static significance** under the section-24
matched-power Microglia interaction (NEBULA), and `r n_static_sig_pb` under
matched pseudobulk; the *smallest* static-NEBULA FDR anywhere in the
dynamic-hit set is `r sprintf("%.2f", min_static_nb)`. The two readouts are
therefore essentially disjoint: differential dynamics recovers an entire gene
programme that the static matched contrast &mdash; approximately zero genes,
section 24 &mdash; cannot see at all. The scatter below makes the dissociation
explicit: the dynamic hits (red) sit high on the dynamic axis yet hug zero on
the static axis.

```{r trajectory-dynamics-vs-static, fig.width = 6.6, fig.height = 5, fig.cap = "Static vs dynamic interaction significance, per gene (-log10 BH-FDR). x = section-24 matched-power Microglia interaction (static NEBULA); y = tradeSeq pseudotime-dynamics interaction. Dashed lines = FDR 0.10. The dynamic hits (red) are all static-null (left of the vertical line): the dynamics layer is a non-redundant readout, not a re-detection of the static signal."}
sc <- dyn_vs
sc$dyn_sig <- !is.na(sc$dyn_adj.P.Val) & sc$dyn_adj.P.Val < dyn_par$padj_cut
sc$x <- -log10(pmax(sc$stat_nb_adj.P.Val, 1e-300))
sc$y <- -log10(pmax(sc$dyn_adj.P.Val,     1e-300))
sc <- sc[is.finite(sc$x) & is.finite(sc$y), ]
thr <- -log10(dyn_par$padj_cut)
ggplot(sc, aes(x, y)) +
  geom_point(aes(colour = dyn_sig), size = 0.5, alpha = 0.4) +
  geom_hline(yintercept = thr, linetype = 2, colour = "grey40") +
  geom_vline(xintercept = thr, linetype = 2, colour = "grey40") +
  scale_colour_manual(values = c(`TRUE` = "#B0344D", `FALSE` = "grey78"),
                      name = sprintf("dynamic FDR<%.2f", dyn_par$padj_cut),
                      labels = c(`TRUE` = "interaction-dynamic", `FALSE` = "other")) +
  labs(x = "static interaction  -log10 FDR  (section 24 matched NEBULA)",
       y = "dynamic interaction  -log10 FDR  (tradeSeq diff-of-differences)",
       title = "Dynamics recovers genes the static contrast cannot") +
  theme_classic()
```

**Read against the aggregate result, this is outcome (a):** the gene-level
dynamics recovers a programme static matched-power DE cannot, handing the
aggregate progression synergy (mean pseudotime interaction
`r sprintf("%+.2f", am_get("mean_pt")$logFC)`, FDR
`r sprintf("%.3f", am_get("mean_pt")$adj.P.Val)`; progression channel
`r sprintf("%+.2f", am_get("progression_cf")$logFC)`, FDR
`r sprintf("%.3f", am_get("progression_cf")$adj.P.Val)`; composition channel
null, `r sprintf("%+.2f", am_get("composition_cf")$logFC)`, FDR
`r sprintf("%.2f", am_get("composition_cf")$adj.P.Val)`) a candidate gene-level
signature &mdash; the supra-additive advance is accompanied by a definite set of
genes whose activation-trajectory *shape* is tau-modulated under amyloid, not
merely a faster traversal of fixed curves.

**The load-bearing caveat, carried into the section-17 ledger gate.** The top of
the ranking is a *mixture*. Genuine amyloid- and microglia-relevant genes
(Aplp1, an APP-family member; Ddit4/REDD1, a stress&ndash;mTOR effector; Ctla2a;
Ptgds; C1ql3) sit alongside classic ambient / non-microglial-lineage transcripts
&mdash; haemoglobins (Hbb-bt, Hba-a2), endothelial-vascular markers (Ly6c1,
Ramp2, Enpp2) and myelin (Mal) &mdash; flagged in the table and figure. A
genotype-varying *ambient* gradient along pseudotime would generate exactly this
dynamic-interaction signature with no microglial biology involved, so a
substantial fraction of the programme is plausibly ambient and **no clean
microglial mechanism is asserted here**. A second, orthogonal confound compounds
this: the programme's one coherent enrichment is **ribosomal / translation
genes** (`r n_ribo_sig` of `r n_dyn_sig`,
`r sprintf("%.0f", ribo_fold)`&times; over the fitted background), whose
trajectory dynamics track cell-state and RNA content rather than a microglial
signalling mechanism &mdash; and which are precisely why a nominal Myc-target
overlap here would be spurious rather than evidence of the Gsk3b / Myc effectors
hypothesis 3B names (both of which are *absent* from the 110). Together the
ambient and ribosomal flags cover roughly a quarter of the programme. The honest
reading is that differential dynamics is a genuinely non-redundant layer that
recovers the project's first gene-level positive interaction and gives the
progression synergy a candidate signature, tempered by real ambient- and
ribosomal-contamination caveats &mdash; all of which the integrated claims ledger
(section 17) must weigh before any scored feed.
