# Cell-type specificity of the tau&times;amyloid interaction

```{r celltype-specificity-load, include = FALSE}
cs  <- readRDS(file.path(params$cache_dir, "celltype_specificity.rds"))
par <- cs$params
tal <- cs$tally                     # R1/R2: 120 rows (regime x estimator x unit x contrast)
icc <- cs$interaction_concordance   # R3a vs_microglia + R3c pb_vs_nebula
scl <- cs$specificity_class         # R4: per-gene class at interaction
pth <- cs$pathway_tally             # R5: GO-BP fGSEA tally at interaction
mpu <- cs$microglia_pathway_cross_unit
xck <- cs$crosscheck                # 10 rows: pooled-microglia vs canonical DE
ucc <- cs$unit_cell_counts

unit_order     <- sort(par$units)   # alphabetical -- the no-privilege guardrail
contrast_order <- c("tau_alone", "nlgf_in_maptki", "nlgf_in_p301s",
                    "tau_in_nlgf", "interaction")

# Unique-row accessors so every prose scalar is read from the cache, never typed.
tget <- function(rg, es, u, cn)
  tal[tal$regime == rg & tal$estimator == es & tal$unit == u & tal$contrast == cn, ]
xget <- function(es, cn) xck[xck$estimator == es & xck$contrast == cn, ]
pget <- function(rg, es, u) pth[pth$regime == rg & pth$estimator == es & pth$unit == u, ]

# Headline scalars (interaction = the divergence readout).
mg_int_native_neb   <- tget("native",  "nebula",     "Microglia",       "interaction")$n_sig_10
oli_int_native_neb  <- tget("native",  "nebula",     "Oligodendrocyte", "interaction")$n_sig_10
oli_int_native_pb   <- tget("native",  "pseudobulk", "Oligodendrocyte", "interaction")$n_sig_10
oli_int_matched_neb <- tget("matched", "nebula",     "Oligodendrocyte", "interaction")$n_sig_10
mg_int_matched_neb  <- tget("matched", "nebula",     "Microglia",       "interaction")$n_sig_10
n_matched_int_total <- sum(tal$n_sig_10[tal$regime == "matched" &
                                        tal$contrast == "interaction"])
# Amyloid main effect survives matching and stays microglia-led:
mg_main_matched_neb <- tget("matched", "nebula", "Microglia",  "nlgf_in_maptki")$n_sig_10
as_main_matched_neb <- tget("matched", "nebula", "Astrocyte",  "nlgf_in_maptki")$n_sig_10
mg_main_native_neb  <- tget("native",  "nebula", "Microglia",  "nlgf_in_maptki")$n_sig_10
as_main_native_neb  <- tget("native",  "nebula", "Astrocyte",  "nlgf_in_maptki")$n_sig_10

# Cross-check floor margins.
xchk_int_neb <- xget("nebula", "interaction")$spearman
xchk_int_pb  <- xget("pseudobulk", "interaction")$spearman
xchk_min     <- min(xck$spearman, na.rm = TRUE)
xchk_all_pass <- all(xck$passes)

# R3 concordance ranges.
r3a_matched <- range(icc$spearman[icc$comparison == "vs_microglia" & icc$regime == "matched"], na.rm = TRUE)
r3a_native  <- range(icc$spearman[icc$comparison == "vs_microglia" & icc$regime == "native"],  na.rm = TRUE)
r3c_matched <- range(icc$spearman[icc$comparison == "pb_vs_nebula" & icc$regime == "matched"], na.rm = TRUE)

# R4 native microglia-unique set (does not survive matching).
mg_uni_native <- scl$symbol[scl$regime == "native" & scl$estimator == "nebula" &
                            scl$class == "microglia_unique"]
n_mg_uni_matched <- sum(scl$regime == "matched" & scl$class == "microglia_unique")

# R5 pathway headline.
neuro_path_matched_neb <- pget("matched", "nebula",     "Neuronal")$n_path_10
mg_path_matched_neb    <- pget("matched", "nebula",     "Microglia")$n_path_10
neuro_path_matched_pb  <- pget("matched", "pseudobulk", "Neuronal")$n_path_10
neuro_path_native_neb  <- pget("native",  "nebula",     "Neuronal")$n_path_10
mg_path_native_neb     <- pget("native",  "nebula",     "Microglia")$n_path_10
```


This section asks the project's first **cross-cell-type** question and closes
narrative arc **N** (`storage/notes/celltype_specificity_plan.md`). Every
preceding arc &mdash; the expression DE of section 3, the pathway, module,
transcription-factor, kinase, ligand-receptor, causal-topology and SCENIC
layers, the tissue **composition** of section 22 and the activation
**dynamics** of section 23 &mdash; reads **only microglia**. The entire
edifice, including the locked 2&times;2 contests and the section-23 progression
interaction, is microglia-scoped *by construction*. The untested question is
therefore: **is the tau&times;amyloid interaction
`(NLGF_P301S&minus;P301S)&minus;(NLGF_MAPTKI&minus;MAPTKI)` &mdash; and the
amyloid / tau main effects &mdash; microglia-SPECIFIC, or do astrocytes,
oligodendrocytes, OPCs, neurons and vascular cells show comparable factorial
responses?**

**Why this is a finding either way, not a restatement.** A microglia-restricted
interaction would *validate* the project's microglia focus as a data-forced
conclusion. A broader, multi-cellular interaction would *reframe* that focus as
a **choice** rather than a necessity &mdash; not refuting any microglial result,
but resetting the context in which it is read. Neither outcome is fished for;
both are reported plainly. This is a display-only chapter that re-runs the exact
microglial DE treatment (the section-3 / `rmd/02b` stack: pseudobulk limma-voom
plus single-cell NEBULA) **symmetrically over six comparison units** &mdash; the
five broad non-microglial types and microglia-as-a-whole (the four microglial
substates pooled) &mdash; and touches none of the locked mouse 2&times;2
verdicts except through the gated arc-N ledger feed decided after this chapter.

**The confound the design is built around.** Cell count varies roughly
twenty-fold across units (Neuronal
`r format(ucc$n_cells[ucc$unit == "Neuronal"], big.mark = ",")` cells vs OPC
`r format(ucc$n_cells[ucc$unit == "OPC"], big.mark = ",")`), so pseudobulk
precision &mdash; and therefore statistical power &mdash; is strongly
asymmetric. A specificity claim from native cell counts alone is confounded:
microglia could appear "special" simply because they are neither the most nor
the least abundant unit, and a power-rich unit could light up for no reason but
its depth. The load-bearing control is a cell-count-**MATCHED** refit: every
unit&times;replicate stratum is downsampled to a common
`r par$K` cells (`K` = the smallest per-replicate depth any unit can supply,
fixed from the data before any fit, RNG seed `r par$seed`) and both estimators
are rerun, so power is equalised across all six units. **The specificity verdict
rests on the MATCHED regime; the NATIVE fits are reported alongside but flagged
power-confounded throughout.**

**Method, fixed before any contrast was read.** The substrate is the
all-cell-types `seurat_full_processed` object
(`r format(par$n_cells, big.mark = ",")` cells; symbol-keyed). Two estimators
&mdash; pseudobulk limma-voom on the 16 `genotype_batch` pseudo-samples with the
locked `r par$design` design, and single-cell NEBULA at full per-cell
resolution &mdash; are run under two power regimes (NATIVE, all cells; MATCHED,
`K`-downsampled) across all six units, against the project's five canonical
contrasts. Both estimators mirror `rmd/02b` byte-for-byte, so the six units are
directly comparable to the microglial headline. The pooled-microglia unit is
cross-checked against the canonical `de_snrnaseq` / `de_snrnaseq_nebula` caches
to confirm the apparatus reproduces the established microglial DE before any
cross-type comparison is trusted.

**Thresholds, stated before any result is read.** Significance everywhere is
BH-FDR &lt; `r sprintf("%.2f", par$padj_cut)` (the project standard). The
sanity-gate cross-check must clear a single uniform Spearman floor of
`r sprintf("%.2f", par$xcheck_floor)` on shared genes per contrast (a uniform
gross-mismatch floor, deliberately not a hard 0.95-per-contrast assert, because
the difference-of-differences interaction is the noisiest contrast across two
intentionally non-identical microglial analyses). The MATCHED downsample target
*K* = `r par$K` and seed `r par$seed` are recorded; the headline gene set for the
curated cross-type heatmap is the canonical DAM/state markers plus the
top-`r par$headline_n` `de_snrnaseq_nebula` interaction symbols.

**Caveats carried throughout.** (a) The MATCHED regime equalises **cell count**,
not transcriptome **depth**: neuronal nuclei carry roughly 2.6-fold the UMIs of
microglia (median 2,809 vs 1,064 in this object), so any neuronal dominance at
the pathway level is partly depth-driven and is flagged as such beside every
such claim. (b) The MATCHED gene-level result is the honest power-equalised
verdict; the NATIVE tallies are shown but never used as the headline. (c) The
interaction is a difference of differences and is intrinsically the noisiest
contrast; absence of a gene-level interaction is read as "no detectable
power-equalised signal", not as proof of exact zero.

**Anti-anchoring.** The metrics below were fixed before any fit was inspected.
All six units are reported symmetrically with no privileged ordering
(alphabetical throughout). The design does **not** assume microglia win: a
non-microglial interaction hit &mdash; the oligodendrocyte signal that emerges
below, or any astrocyte response &mdash; is reported as a finding, not explained
away. Where a broader response is found, it is stated to *reframe*, never
refute, the microglia focus.


## Sanity gate: the pooled-microglia unit reproduces the canonical microglia DE

Before any cross-type comparison is trusted, the apparatus is validated against
the project's own microglial result. The pooled-microglia unit drawn from
`seurat_full_processed` is fit with both estimators and its per-contrast logFC
is correlated (Spearman, on shared genes, bridged into symbol space) against the
canonical `de_snrnaseq` (pseudobulk) and `de_snrnaseq_nebula` (NEBULA) caches.
All ten cross-checks **`r if (xchk_all_pass) "pass" else "FAIL"`** the
`r sprintf("%.2f", par$xcheck_floor)` floor. The four main-effect-style
contrasts reproduce at &rho; &asymp; 0.96&ndash;0.98; the interaction is the
nearest the floor (NEBULA `r sprintf("%.3f", xchk_int_neb)`, pseudobulk
`r sprintf("%.3f", xchk_int_pb)`; the overall minimum is
`r sprintf("%.3f", xchk_min)`). This is exactly as expected &mdash; the
interaction is the noisiest difference-of-differences contrast across two
intentionally non-identical microglial analyses (the all-cell-types object versus
the dedicated microglia re-cluster), and a 0.95-per-contrast assert would have
spuriously failed both interaction checks despite genuine reproduction (against a
random-expectation &rho; &asymp; 0). The uniform 0.90 floor is the calibrated
choice, and the apparatus is validated.

```{r celltype-specificity-crosscheck, results = "asis"}
xt <- xck
xt$contrast  <- factor(xt$contrast, levels = contrast_order)
xt <- xt[order(xt$estimator, xt$contrast), ]
xt_disp <- data.frame(
  estimator = xt$estimator,
  contrast  = as.character(xt$contrast),
  `n shared` = xt$n_shared,
  Spearman  = sprintf("%.3f", xt$spearman),
  passes    = ifelse(xt$passes, "yes", "NO"),
  check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(
  xt_disp, row.names = FALSE,
  caption = paste0(
    "Sanity gate: pooled-microglia unit (from seurat_full_processed) vs the ",
    "canonical microglia DE, Spearman of logFC on shared symbols per contrast ",
    "and estimator. All ten clear the uniform ",
    sprintf("%.2f", par$xcheck_floor), " floor; the interaction sits nearest ",
    "it (the noisiest difference-of-differences contrast), confirming faithful ",
    "reproduction rather than a wiring artefact."))
```

## Significant-gene tally: the factorial works, the interaction is thin

The first readout counts significant genes (FDR &lt; `r sprintf("%.2f", par$padj_cut)`)
per unit and contrast. The factorial **design works**: the amyloid main effects
are broad and multi-cellular, and &mdash; importantly &mdash; remain
**microglia-led even under MATCHED power** (`nlgf_in_maptki`, NEBULA: microglia
`r mg_main_matched_neb` significant genes vs astrocyte `r as_main_matched_neb`
at equal cells; the NATIVE counts are larger but power-confounded, microglia
`r mg_main_native_neb` vs astrocyte `r as_main_native_neb`). So microglia
genuinely **are** the strongest amyloid responder, and that part of the focus is
validated by the data.

The **interaction** &mdash; the divergence readout, the actual subject of this
chapter &mdash; tells a different story. At NATIVE power it is sparse everywhere:
microglia `r mg_int_native_neb` genes (NEBULA), oligodendrocyte
`r oli_int_native_neb` (NEBULA) / `r oli_int_native_pb` (pseudobulk), the rest at
or near zero, and `tau_alone` essentially null in every unit. Under the
load-bearing **MATCHED** regime the gene-level interaction **collapses to zero in
all six units, microglia included** &mdash; the sole survivor across both
estimators is oligodendrocyte (`r oli_int_matched_neb` genes, NEBULA only;
`r n_matched_int_total` interaction genes total across the entire matched grid).
The handful of NATIVE microglial interaction hits are therefore a power artefact
of the cell-count spread, not a microglia-specific programme; and the one unit
that retains any gene-level interaction at equal power is **not** microglia.

```{r celltype-specificity-tally-matched, fig.width = 8, fig.height = 4.2, fig.cap = "MATCHED (power-equalised, K cells per unit x replicate) significant-gene counts per unit x contrast, faceted by estimator; tile labels are gene counts at FDR<0.10. The amyloid main-effect columns survive matching and stay microglia-led; the interaction column collapses to zero in every unit except a small oligodendrocyte NEBULA signal. This is the load-bearing tally."}
plot_specificity_tally_heatmap(tal, regime = "matched", fdr = "10",
                               contrasts = contrast_order)
```

```{r celltype-specificity-tally-native, fig.width = 8, fig.height = 4.2, fig.cap = "NATIVE (all cells) significant-gene counts per unit x contrast (FDR<0.10). Shown for completeness but flagged POWER-CONFOUNDED: the ~20x cell-count spread inflates counts in the abundant units, so this regime cannot ground a specificity claim. Compare the interaction column here to the matched figure above -- the few microglial interaction hits do not survive power-matching."}
plot_specificity_tally_heatmap(tal, regime = "native", fdr = "10",
                               contrasts = contrast_order)
```

## Interaction-effect concordance across cell types

If the interaction were a shared, multi-cellular programme, the per-unit
interaction-logFC profiles would correlate with microglia genome-wide. They do
not. The genome-wide Spearman of interaction logFC between microglia and each
other unit is uniformly **low** &mdash; MATCHED
`r sprintf("%.2f", r3a_matched[1])`&ndash;`r sprintf("%.2f", r3a_matched[2])`,
NATIVE `r sprintf("%.2f", r3a_native[1])`&ndash;`r sprintf("%.2f", r3a_native[2])`
&mdash; so each unit's (mostly noise-level) interaction response is
**distinct**, not a common axis read out in different cell types. Critically,
this is **not** an estimator artefact: within each unit the pseudobulk and NEBULA
interaction logFCs agree well (cross-estimator Spearman
`r sprintf("%.2f", r3c_matched[1])`&ndash;`r sprintf("%.2f", r3c_matched[2])`
under MATCHED), so the two methods see the same per-unit signal &mdash; it is the
**cell types** that differ, not the estimators.

```{r celltype-specificity-cross-estimator, fig.width = 7.6, fig.height = 5, fig.cap = "Cross-estimator agreement (MATCHED): pseudobulk vs NEBULA interaction log2FC, one facet per unit, dashed line y=x. The estimators agree within every unit (Spearman ~0.96-0.98), so the low microglia-vs-unit concordance reported above reflects genuine cell-type differences, not method disagreement. At matched power the interaction is a near-zero-centred cloud in all units (no coherent off-diagonal structure)."}
plot_cross_estimator_panel(cs$fits$matched$pseudobulk, cs$fits$matched$nebula,
                           regime = "matched", contrast = "interaction")
```

The curated view restricts the interaction logFC to the microglial headline gene
set (DAM/state markers plus the top `de_snrnaseq_nebula` interaction symbols) and
lays it across all six units at full NATIVE NEBULA power. The microglial story
does not propagate: these genes carry their interaction signal (such as it is) in
microglia and are flat or absent (many are not meaningfully expressed) in the
other units. The microglial headline interaction is a microglial-gene
phenomenon, not a cross-type one &mdash; consistent with the low genome-wide
concordance above.

```{r celltype-specificity-headline-heatmap, fig.width = 7.4, fig.height = 7.6, fig.cap = "Interaction log2FC of the microglial headline gene set (canonical DAM/state markers + top de_snrnaseq_nebula interaction symbols) across the six units, NATIVE NEBULA. Grey/absent tiles are genes not expressed in that unit. The microglial headline genes do not show a coherent interaction in the non-microglial units, so the headline interaction is microglia-gene-specific rather than a shared cross-type programme. NATIVE (full-power) view; the matched regime carries even less gene-level signal."}
plot_interaction_logfc_heatmap(cs$fits$native$nebula, cs$headline_genes,
                               regime = "native", estimator = "nebula")
```

## Per-gene specificity class

Every gene significant at the interaction (FDR &lt; `r sprintf("%.2f", par$padj_cut)`)
in **any** unit is classified as microglia-unique, shared, or
non-microglial-unique. The pool is tiny (the interaction is thin everywhere), and
the verdict is unambiguous. At NATIVE NEBULA power there are
`r length(mg_uni_native)` microglia-unique genes
(`r paste(mg_uni_native, collapse = ", ")`) alongside a set of
non-microglial-unique genes that are almost entirely **oligodendrocyte**. Under
the load-bearing MATCHED regime, **`r n_mg_uni_matched` microglia-unique
interaction genes survive** &mdash; the only power-equalised gene-level
interaction class is non-microglial (oligodendrocyte). The microglia-unique
NATIVE hits are precisely the power artefact the MATCHED control was built to
expose.

```{r celltype-specificity-class-native, fig.width = 6.6, fig.height = 3.8, fig.cap = "Per-gene interaction specificity class at NATIVE power, faceted by estimator. A small microglia-unique set appears under NEBULA (full power) but the non-microglial-unique class -- almost all oligodendrocyte -- is comparable or larger. NATIVE; power-confounded."}
plot_specificity_class(scl, regime = "native")
```

```{r celltype-specificity-class-matched, fig.width = 6.6, fig.height = 3.8, fig.cap = "Per-gene interaction specificity class at MATCHED (power-equalised) power. No microglia-unique gene survives matching; the only surviving class is non-microglial-unique (oligodendrocyte). This is the load-bearing specificity-class result."}
plot_specificity_class(scl, regime = "matched")
```

## Pathway-level interaction response

The gene-level interaction is thin in every unit, so the secondary readout asks
the question at pathway resolution: GO-BP fGSEA on each unit's interaction
ranking, counting pathways enriched at FDR &lt; `r sprintf("%.2f", par$padj_cut)`.
Here the asymmetry is stark and **the broadest response is Neuronal, not
microglial, in every regime and estimator.** Under MATCHED NEBULA the neuronal
unit carries `r neuro_path_matched_neb` enriched interaction pathways against
microglia's `r mg_path_matched_neb`; under MATCHED pseudobulk neuronal leads with
`r neuro_path_matched_pb`; the NATIVE regime amplifies the same ordering
(neuronal `r neuro_path_native_neb` vs microglia `r mg_path_native_neb`, NEBULA).
The enrichment is overwhelmingly **down** (negative NES), consistent with the
interaction suppressing neuronal transcriptional programmes. The two pathways
microglia does retain under MATCHED NEBULA are ribonucleoprotein- and
ribosome-biogenesis terms &mdash; housekeeping, down-regulated, and *shared* with
the neuronal response rather than microglia-restricted.

```{r celltype-specificity-pathway, fig.width = 7.8, fig.height = 4, fig.cap = "Pathways enriched at the interaction (GO-BP fGSEA, FDR<0.10) per unit, across all four regime x estimator combinations; tile labels are pathway counts. Neuronal dominates every row. CAVEAT: the matched regime equalises CELL COUNT, not transcriptome depth -- neuronal nuclei carry ~2.6x the UMIs of microglia (median 2,809 vs 1,064), so the neuronal pathway dominance is partly depth-driven and must be read with that confound in mind."}
ph <- pth
ph$grp  <- factor(paste(ph$regime, ph$estimator, sep = " / "),
                  levels = c("native / pseudobulk", "native / nebula",
                             "matched / pseudobulk", "matched / nebula"))
ph$unit <- factor(ph$unit, levels = unit_order)
ggplot(ph, aes(unit, grp, fill = n_path_10)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = n_path_10), size = 3) +
  scale_fill_gradient(low = "grey92", high = "#b40426",
                      name = "pathways\nFDR<0.10") +
  labs(x = NULL, y = NULL,
       title = "Interaction-enriched pathways per unit (Neuronal dominates)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1))
```

The depth confound is load-bearing for this pathway result and is stated
plainly: MATCHED equalises the number of cells per unit, not the RNA captured per
cell, and neuronal nuclei are markedly deeper. The neuronal pathway dominance is
therefore *partly* a depth phenomenon and is **not** advanced as a clean
"neurons drive the interaction" claim &mdash; only as evidence that, at the
pathway level, the interaction response is **broadest outside microglia**, which
is sufficient for the specificity question regardless of the exact neuronal
magnitude.

## Verdict

The synthesis is read per readout, absence reported as evidence, no unit
privileged. Thresholds restated once: BH-FDR &lt; `r sprintf("%.2f", par$padj_cut)`
on the locked five-contrast 2&times;2 factorial, both estimators (pseudobulk +
NEBULA) under both power regimes, the headline resting on the MATCHED
(*K* = `r par$K` cells per unit&times;replicate, seed `r par$seed`)
power-equalised fits.

**Overall: the tau&times;amyloid interaction is NOT microglia-specific.** At
load-bearing MATCHED power the gene-level interaction is essentially absent in
every unit including microglia (the only survivor is a small oligodendrocyte
signal), the per-unit interaction profiles are mutually distinct rather than a
shared programme, no microglia-unique interaction gene survives power-matching,
and at the pathway level the broadest response sits in neurons, not microglia
(with the depth caveat). This **reframes &mdash; it does not refute &mdash;** the
project's microglia focus: the amyloid **main** effect genuinely is
microglia-led, even at equal power, so the choice to study microglia is well
motivated; but the **divergence** readout, the interaction itself, is not a
microglial specialty, and the strongest interaction signals in the dataset lie
outside microglia. The microglia-centric reading of the interaction is, on this
evidence, a **choice** that the data permit rather than compel.

```{r celltype-specificity-verdict-build}
cs_verdict <- tibble::tibble(
  readout = c("gene_interaction_tally (R1/R2)",
              "interaction_concordance (R3)",
              "specificity_class (R4)",
              "pathway_interaction (R5)",
              "OVERALL"),
  matched_result = c(
    sprintf(paste0("interaction genes FDR<%.2f: 0 in all 6 units incl ",
                   "Microglia; sole survivor Oligodendrocyte %d (NEBULA); ",
                   "main effects survive and stay microglia-led (nlgf_in_maptki ",
                   "NEBULA Microglia %d vs Astrocyte %d)"),
            par$padj_cut, oli_int_matched_neb, mg_main_matched_neb,
            as_main_matched_neb),
    sprintf(paste0("genome-wide Microglia-vs-unit interaction Spearman ",
                   "%.2f-%.2f (distinct); cross-estimator pb-vs-NEBULA %.2f-%.2f ",
                   "(estimators agree)"),
            r3a_matched[1], r3a_matched[2], r3c_matched[1], r3c_matched[2]),
    sprintf("%d microglia-unique genes; only non-microglial-unique (Oligodendrocyte) survive",
            n_mg_uni_matched),
    sprintf(paste0("GO-BP fGSEA FDR<%.2f: Neuronal %d (NEBULA) / %d (pb) vs ",
                   "Microglia %d; broadest = Neuronal (depth-confounded, ~2.6x UMIs)"),
            par$padj_cut, neuro_path_matched_neb, neuro_path_matched_pb,
            mg_path_matched_neb),
    "tau x amyloid INTERACTION is NOT microglia-specific at power-equalised resolution"),
  native_result = c(
    sprintf("interaction FDR<%.2f: Microglia %d, Oligodendrocyte %d (NEBULA); sparse, power-confounded",
            par$padj_cut, mg_int_native_neb, oli_int_native_neb),
    sprintf("Microglia-vs-unit %.2f-%.2f (low); shown but power-confounded",
            r3a_native[1], r3a_native[2]),
    sprintf("%d microglia-unique (NEBULA: %s) -- do NOT survive matching",
            length(mg_uni_native), paste(mg_uni_native, collapse = ", ")),
    sprintf("Neuronal %d vs Microglia %d (NEBULA); same ordering, amplified",
            neuro_path_native_neb, mg_path_native_neb),
    "NATIVE microglial hits are a power artefact of the ~20x cell-count spread"),
  specificity_verdict = c(
    "no microglia-unique gene-level interaction survives power-matching",
    "per-unit interaction profiles are distinct, not a shared programme",
    "no microglia-unique interaction gene survives matching",
    "broadest interaction pathway response is NEURONAL, not microglial",
    "REFRAMES (not refutes) the microglia focus: amyloid main effect IS microglia-led; the interaction is not microglia-specific"))
readr::write_tsv(cs_verdict,
                 file.path(params$results_dir, "celltype_specificity_verdict.tsv"))
```

```{r celltype-specificity-verdict-table, results = "asis"}
knitr::kable(
  cs_verdict[, c("readout", "matched_result", "specificity_verdict")],
  col.names = c("Readout", "MATCHED result (load-bearing)", "Specificity verdict"),
  caption = paste0(
    "Per-readout cell-type-specificity verdict, headline drawn from the MATCHED ",
    "(power-equalised) regime. The tau x amyloid interaction is not ",
    "microglia-specific: no microglia-unique gene-level interaction survives ",
    "power-matching, per-unit profiles are distinct, and the broadest pathway ",
    "response is neuronal (depth-confounded). The amyloid MAIN effect, by ",
    "contrast, is microglia-led even at equal power. Full table incl. the ",
    "NATIVE (power-confounded) column in ",
    "storage/results/celltype_specificity_verdict.tsv."))
```

**Boundary conditions, restated as the verdict's caveats.** The MATCHED regime
equalises cell count, not transcriptome depth, so the neuronal pathway dominance
is partly depth-driven and is read only as "broadest outside microglia", not as
a quantitative neuronal-magnitude claim. The interaction is the noisiest
difference-of-differences contrast, so its gene-level near-absence at equal power
is read as "no detectable power-equalised signal" rather than exact zero; the
specificity conclusion rests on the *convergence* of four readouts (gene tally,
concordance, per-gene class, pathway breadth) all pointing the same way, not on
any single test. The NATIVE tallies are reported in full for transparency but are
flagged power-confounded and never carry the headline. Within those bounds the
layer is a clean, symmetric, falsifiable cross-cell-type readout: the amyloid
response is microglia-led, but the tau&times;amyloid **interaction** &mdash; the
project's divergence readout &mdash; is a property of the tissue broadly, not of
microglia specifically.
