# Cross-modality: spatial, bulk, and divergence checks {#sec-crossmodality}

P4 asks what the non-snRNAseq modalities add to the rebuilt microglia story. The
spatial and bulk layers are not treated as independent winners: GeoMx AOIs are repeated
tissue observations, and the 24-month proteome and phosphoproteome are bulk hippocampus,
not microglia-sorted. The chapter therefore reads them as concordance, caveat and
divergence checks around the P1-P3 spine: amyloid-to-DAM activation, a DAM-composition
tau-amyloid interaction, Myc-linked RNA support, and unsupported NF-kB/Gsk3b mechanism
claims.

```{r}
#| label: crossmodality-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)
targets::tar_source()
targets::tar_load(c(crossmodality_report, crossmodality_figures)) # compact targets only, no heavy modality objects
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

cmr <- crossmodality_report
cmf <- crossmodality_figures
alpha <- cmr$provenance$alpha
focus <- cmr$provenance$focus_contrasts
contrast_labels <- c(
  tau_alone = "tau alone",
  nlgf_in_maptki = "amyloid on MAPTKI",
  nlgf_in_p301s = "amyloid on P301S",
  tau_in_nlgf = "tau in NLGF",
  interaction = "interaction"
)
layer_labels <- c(
  proteome = "proteome",
  phospho_raw = "raw phospho",
  phospho_corrected = "corrected phospho"
)
axis_labels <- c(
  DAM = "DAM",
  homeostatic = "homeostatic",
  synaptic = "synaptic",
  clearance = "clearance",
  complement = "complement",
  antigen_presentation = "antigen presentation",
  MHC_APC = "MHC/APC",
  IFN = "IFN",
  NFkB = "NF-kB",
  Myc = "Myc",
  Proliferative = "proliferative",
  other = "other"
)
fmt_p <- function(p) {
  out <- rep("-", length(p))
  ok <- is.finite(p)
  out[ok] <- ifelse(p[ok] < 0.001,
                    formatC(p[ok], format = "e", digits = 2),
                    formatC(p[ok], format = "f", digits = 3))
  out
}
fmt_num <- function(x, digits = 2) ifelse(is.finite(x), formatC(x, format = "f", digits = digits), "-")
fmt_int <- function(x) format(as.integer(round(x)), big.mark = ",", trim = TRUE)
short_label <- function(x, n = 46L) {
  x <- gsub("^GO[A-Z]*_", "", x)
  x <- gsub("_", " ", x)
  ifelse(nchar(x) > n, paste0(substr(x, 1L, n - 3L), "..."), x)
}
wrap_label <- function(x, width = 22L) {
  vapply(as.character(x), function(s) {
    z <- strwrap(s, width = width)
    if (length(z)) paste(z, collapse = "\n") else s
  }, character(1), USE.NAMES = FALSE)
}
plot_contrast <- function(x, levels = focus, reverse = FALSE) {
  lv <- if (reverse) rev(levels) else levels
  factor(x, levels = lv, labels = unname(contrast_labels[lv]))
}
fmt_axis <- function(x) {
  vapply(strsplit(as.character(x), ";", fixed = TRUE), function(parts) {
    lab <- axis_labels[parts]
    lab[is.na(lab)] <- parts[is.na(lab)]
    paste(unname(lab), collapse = " / ")
  }, character(1), USE.NAMES = FALSE)
}
fmt_pair <- function(x) gsub("_", "-", x, fixed = TRUE)
one_row <- function(d, idx, label) {
  out <- d[idx, , drop = FALSE]
  stopifnot(nrow(out) >= 1L)
  out[1L, , drop = FALSE]
}

geom_counts <- cmr$geomx$counts
bulk_counts <- cmr$bulk$significant_counts
bulk_features <- cmr$bulk$feature_counts
run_index <- cmr$bulk$run_index
decon <- cmr$clearance$spatial_decon
pairs <- cmr$clearance$pair_support
contrast_summary <- cmr$divergence$contrast_summary
path_axis <- cmr$pathway$axis_summary

geo_volc_bins <- cmf$geomx_volcano$bins
geo_volc_bins$contrast <- plot_contrast(geo_volc_bins$contrast)
geo_volc_bins$log_n <- log10(geo_volc_bins$n + 1)
geo_volc_lab <- cmf$geomx_volcano$labels
geo_volc_lab$contrast <- plot_contrast(geo_volc_lab$contrast)

geo_sens <- cmf$geomx_sensitivity
geo_sens <- geo_sens[geo_sens$contrast %in% focus, , drop = FALSE]
geo_sens$contrast <- plot_contrast(geo_sens$contrast)
geo_sens$fit_label <- c(unblocked = "unblocked AOI",
                        collapsed_bio_unit = "bio-unit collapsed")[geo_sens$fit]
geo_sens$fit_label[is.na(geo_sens$fit_label)] <- geo_sens$fit[is.na(geo_sens$fit_label)]
geo_sens$fit_label <- factor(geo_sens$fit_label,
                             levels = c("unblocked AOI", "bio-unit collapsed"))
geo_sens_long <- rbind(
  data.frame(geo_sens[c("fit_label", "contrast")],
             metric = "lost from primary", n = -geo_sens$n_lost,
             stringsAsFactors = FALSE),
  data.frame(geo_sens[c("fit_label", "contrast")],
             metric = "gained in sensitivity", n = geo_sens$n_gained,
             stringsAsFactors = FALSE)
)
geo_sens_long$metric <- factor(geo_sens_long$metric,
                               levels = c("lost from primary", "gained in sensitivity"))

bulk_run_plot <- cmf$bulk_run_index
bulk_run_plot$contrast <- plot_contrast(bulk_run_plot$contrast, levels = names(contrast_labels))
bulk_run_plot$layer_label <- factor(layer_labels[bulk_run_plot$layer],
                                    levels = unname(layer_labels))
bulk_run_plot$cell_label <- ifelse(bulk_run_plot$n_primary_sig > 0,
                                   paste0(bulk_run_plot$n_lost_or_flipped, "/",
                                          bulk_run_plot$n_primary_sig),
                                   "0/0")

pc_bins <- cmf$phospho_raw_corrected$bins
pc_bins$contrast <- plot_contrast(pc_bins$contrast)
pc_bins$log_n <- log10(pc_bins$n + 1)
pc_lab <- cmf$phospho_raw_corrected$labels
pc_lab$best_fdr <- pmin(pc_lab$raw_fdr, pc_lab$corrected_fdr)
pc_lab <- do.call(rbind, lapply(split(pc_lab, pc_lab$contrast, drop = TRUE), function(d) {
  d <- d[order(d$best_fdr, -abs(d$raw_logFC) - abs(d$corrected_logFC),
               d$site_id, method = "radix", na.last = TRUE), , drop = FALSE]
  head(d, 6L)
}))
pc_lab$contrast <- plot_contrast(pc_lab$contrast)
pc_lab$site_label <- wrap_label(gsub("_", " ", pc_lab$site_id, fixed = TRUE), 14L)

anchor_plot <- cmf$anchor_effect_heatmap
anchor_plot <- anchor_plot[anchor_plot$contrast %in% names(contrast_labels), , drop = FALSE]
anchor_plot$key <- paste(anchor_plot$layer, anchor_plot$contrast, anchor_plot$anchor_symbols,
                         sep = "\r")
anchor_plot <- anchor_plot[order(anchor_plot$key, anchor_plot$adj.P.Val, -abs(anchor_plot$t),
                                 anchor_plot$feature, method = "radix", na.last = TRUE),
                           , drop = FALSE]
anchor_plot <- anchor_plot[!duplicated(anchor_plot$key), , drop = FALSE]
anchor_plot$contrast <- plot_contrast(anchor_plot$contrast, levels = names(contrast_labels))
anchor_plot$layer_label <- factor(layer_labels[anchor_plot$layer],
                                  levels = unname(layer_labels))
anchor_plot$anchor_label <- factor(anchor_plot$anchor_symbols,
                                   levels = rev(sort(unique(anchor_plot$anchor_symbols),
                                                     method = "radix")))
anchor_plot$supported <- anchor_plot$adj.P.Val < alpha

clear_plot <- cmf$clearance_pair_grid
clear_plot <- clear_plot[clear_plot$contrast %in% focus, , drop = FALSE]
clear_plot$contrast <- plot_contrast(clear_plot$contrast)
clear_plot$pair_label <- factor(fmt_pair(clear_plot$pair),
                                levels = rev(unique(fmt_pair(clear_plot$pair))))
clear_plot$status_label <- factor(ifelse(clear_plot$status == "earned", "earned", "not earned"),
                                  levels = c("not earned", "earned"))
clear_plot$support_label <- ifelse(clear_plot$n_coherent_supported_modalities > 0,
                                   clear_plot$n_coherent_supported_modalities, "")

sym_all <- cmf$symbol_modality_matrix
sym_all <- sym_all[sym_all$contrast %in% focus, , drop = FALSE]
sym_rank <- do.call(rbind, lapply(split(sym_all, sym_all$symbol, drop = TRUE), function(d) {
  data.frame(symbol = d$symbol[1L],
             max_sig = max(d$n_modalities_sig, na.rm = TRUE),
             max_present = max(d$n_modalities_present, na.rm = TRUE),
             min_fdr = min(d$min_fdr, na.rm = TRUE),
             max_rank = max(d$rank_score, na.rm = TRUE),
             stringsAsFactors = FALSE)
}))
sym_rank <- sym_rank[order(-sym_rank$max_sig, sym_rank$min_fdr, -sym_rank$max_present,
                           sym_rank$symbol, method = "radix", na.last = TRUE), , drop = FALSE]
sym_keep <- head(sym_rank$symbol, 18L)
sym_plot <- sym_all[sym_all$symbol %in% sym_keep, , drop = FALSE]
sym_plot$contrast <- plot_contrast(sym_plot$contrast)
sym_plot$symbol <- factor(sym_plot$symbol, levels = rev(sym_keep))
sym_plot$dir_label <- ifelse(sym_plot$n_modalities_sig == 0, "0",
                             ifelse(sym_plot$direction_call == "mixed", "mix",
                                    ifelse(sym_plot$direction_call == "positive", "+", "-")))

path_fig <- cmf$pathway_axis_heatmap
path_fig <- path_fig[path_fig$contrast %in% focus, , drop = FALSE]
path_fig$contrast <- plot_contrast(path_fig$contrast)
path_fig$axis_label <- fmt_axis(path_fig$axis)
path_fig$set_label <- short_label(path_fig$set, 34L)
path_fig$set_label <- factor(path_fig$set_label, levels = rev(unique(path_fig$set_label)))
path_fig$direction_label <- ifelse(path_fig$mixed_sign, "mix",
                                   ifelse(path_fig$consistent_direction == "positive", "+",
                                          ifelse(path_fig$consistent_direction == "negative", "-", "0")))

geo_int <- one_row(geom_counts, geom_counts$contrast == "interaction", "GeoMx interaction counts")
geo_amy_k <- one_row(geom_counts, geom_counts$contrast == "nlgf_in_maptki", "GeoMx amyloid MAPTKI counts")
geo_amy_p <- one_row(geom_counts, geom_counts$contrast == "nlgf_in_p301s", "GeoMx amyloid P301S counts")
bulk_primary_sig <- sum(run_index$n_primary_sig, na.rm = TRUE)
bulk_lost <- sum(run_index$n_lost_or_flipped, na.rm = TRUE)
bulk_loss_pct <- if (bulk_primary_sig > 0) round(100 * bulk_lost / bulk_primary_sig, 1) else 0
worst_run <- run_index[order(-run_index$n_lost_or_flipped, run_index$layer,
                             match(run_index$contrast, names(contrast_labels)),
                             method = "radix", na.last = TRUE), , drop = FALSE][1L, ]
decon_reasons <- if (length(decon$reasons)) paste(decon$reasons, collapse = "; ") else "no blocking reason recorded"
earned_pairs <- pairs[pairs$status == "earned", , drop = FALSE]
earned_sentence <- if (nrow(earned_pairs)) {
  paste0(paste(unique(earned_pairs$pair), collapse = ", "), " in ",
         paste(contrast_labels[earned_pairs$contrast], collapse = ", "))
} else {
  "no pair"
}
fmt_modalities <- function(x) {
  z <- unlist(strsplit(x, ";", fixed = TRUE), use.names = FALSE)
  z <- z[nzchar(z)]
  z <- gsub("_", " ", z, fixed = TRUE)
  paste(z, collapse = ", ")
}
earned_details <- if (nrow(earned_pairs)) {
  vapply(seq_len(nrow(earned_pairs)), function(i) {
    ep <- earned_pairs[i, , drop = FALSE]
    mods <- fmt_modalities(ep$coherent_supported_modalities)
    paste0(fmt_pair(ep$pair), " in ", contrast_labels[ep$contrast],
           if (nzchar(mods)) paste0(" (supported by ", mods, ")") else "")
  }, character(1))
} else {
  character()
}
earned_detail_sentence <- if (length(earned_details)) {
  paste0(if (length(earned_details) == 1L) "The earned row is " else "The earned rows are ",
         paste(earned_details, collapse = "; "), ".")
} else {
  "No measured pair earns a communication claim under the CCC-lite rule."
}
earned_pair_names <- unique(earned_pairs$pair)
clearance_synthesis_phrase <- if (length(earned_pair_names)) {
  paste0(paste(fmt_pair(earned_pair_names), collapse = ", "),
         if (length(earned_pair_names) == 1L) " earns" else " earn",
         " focused measured clearance support")
} else {
  "no clearance pair earns focused measured clearance support"
}
closing_support_phrase <- if (length(earned_pair_names)) {
  "DAM activation, synaptic suppression and the earned clearance-axis support"
} else {
  "DAM activation and synaptic suppression, with clearance kept descriptive"
}
```

## GeoMx Spatial DE

The GeoMx layer contributes a tissue-spatial transcriptomic readout over
`r fmt_int(cmr$geomx$provenance$n_aoi)` AOIs and
`r fmt_int(cmr$geomx$provenance$n_bio_units)` biological units. Its primary model uses
slide fixed effects plus duplicate-correlation blocking by biological unit, so the AOIs are not
read as independent animals. Amyloid produces large spatial DE counts in both backgrounds
(`r fmt_int(geo_amy_k$n_fdr_0_10)` rows at FDR < `r alpha` on MAPTKI,
`r fmt_int(geo_amy_p$n_fdr_0_10)` on P301S), while the tau-amyloid interaction is much
smaller (`r fmt_int(geo_int$n_fdr_0_10)` rows). That fits the broader story: GeoMx strongly
corroborates the amyloid tissue response, but interaction claims still lean on the microglia
composition and trajectory layers.

```{r}
#| label: fig-crossmodality-geomx-counts
#| fig-width: 8.8
#| fig-height: 4.2
#| fig-cap: "GeoMx spatial DE counts under the blocked AOI model. Bars count primary GeoMx rows with FDR < 0.10, split by logFC direction. The amyloid contrasts dominate; the interaction has far fewer spatial transcriptomic hits."
geo_plot <- geom_counts[geom_counts$contrast %in% focus, , drop = FALSE]
geo_long <- rbind(
  data.frame(contrast = geo_plot$contrast, direction = "up", n = geo_plot$n_up_fdr_0_10),
  data.frame(contrast = geo_plot$contrast, direction = "down", n = geo_plot$n_down_fdr_0_10)
)
geo_long$contrast <- factor(geo_long$contrast, levels = focus, labels = contrast_labels[focus])
geo_long$direction <- factor(geo_long$direction, levels = c("up", "down"))

ggplot(geo_long, aes(contrast, n, fill = direction)) +
  geom_col(width = 0.72) +
  scale_fill_manual(values = c(up = "#B0344D", down = "#3F5A6B"), name = "direction") +
  labs(x = NULL, y = "GeoMx rows at FDR < 0.10", title = "Spatial DE is strongest for amyloid") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| label: fig-crossmodality-geomx-volcano
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "GeoMx primary-model volcanoes across the P4 focus contrasts. Points are pre-binned from the blocked AOI model, and labels mark the strongest rows per contrast. The amyloid contrasts have dense high-significance structure; the interaction is visibly smaller."
ggplot(geo_volc_bins) +
  geom_rect(aes(xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max,
                fill = log_n),
            colour = NA) +
  geom_vline(xintercept = 0, colour = "grey65", linewidth = 0.3) +
  ggrepel::geom_text_repel(
    data = geo_volc_lab,
    aes(effect, neg_log10_p, label = label),
    inherit.aes = FALSE,
    size = 2.25,
    seed = 42L,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.12
  ) +
  facet_wrap(~ contrast, ncol = 2) +
  scale_fill_gradient(low = "grey94", high = "#B0344D", name = "log10(n + 1)") +
  labs(x = "GeoMx log2 fold-change", y = "-log10 p-value",
       title = "GeoMx primary spatial DE by contrast") +
  theme(panel.grid = element_blank())
```

```{r}
#| label: crossmodality-geomx-top
gt <- cmr$geomx$top
gt <- gt[gt$contrast %in% focus, , drop = FALSE]
gt <- gt[order(match(gt$contrast, focus), gt$fdr, -abs(gt$statistic), gt$symbol,
               method = "radix", na.last = TRUE), , drop = FALSE]
gt <- head(gt, 12L)
gtab <- data.frame(
  contrast = contrast_labels[gt$contrast],
  symbol = gt$symbol,
  logFC = fmt_num(gt$effect),
  t = fmt_num(gt$statistic),
  FDR = fmt_p(gt$fdr),
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(gtab, row.names = FALSE,
             caption = "Top GeoMx primary rows across the P4 focus contrasts. These are tissue/AOI DE rows, not cell-type-specific microglia effects.")
```

```{r}
#| label: fig-crossmodality-geomx-sensitivity
#| fig-width: 9.2
#| fig-height: 4.8
#| fig-cap: "GeoMx primary blocked-AOI calls against the two sensitivity fits. Positive bars are rows gained by a sensitivity fit, negative bars are primary rows lost. The collapsed biological-unit fit loses many primary rows, so the chapter privileges the duplicate-correlation blocked primary model rather than reading AOIs as independent animals."
ggplot(geo_sens_long, aes(contrast, n, fill = metric)) +
  geom_hline(yintercept = 0, colour = "grey65") +
  geom_col(width = 0.68) +
  facet_wrap(~ fit_label, nrow = 1) +
  scale_fill_manual(values = c(`lost from primary` = "#3F5A6B",
                               `gained in sensitivity` = "#B0344D"),
                    name = NULL) +
  scale_y_continuous(labels = function(x) abs(x)) +
  labs(x = NULL, y = "GeoMx rows changed at FDR < 0.10",
       title = "Spatial DE support depends on the replicate treatment") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```

## Bulk Proteome And Phospho

The bulk layers use the same 16-run, 24-month hippocampus sample set: total protein, raw
phosphosite and parent-protein-corrected phosphosite. The feature universes are compact enough
to report directly (`r paste(paste(layer_labels[bulk_features$layer], fmt_int(bulk_features$n_features), sep = ": "), collapse = "; ")`).
The run-index sensitivity is load-bearing. Across bulk layers,
`r fmt_int(bulk_primary_sig)` primary FDR < `r alpha` rows are checked, and
`r fmt_int(bulk_lost)` (`r bulk_loss_pct`%) lose support or flip under the additive run-index
sensitivity. The most affected cell is `r layer_labels[worst_run$layer]` in
`r contrast_labels[worst_run$contrast]` (`r fmt_int(worst_run$n_lost_or_flipped)` rows), so
bulk signals below are treated as support only when they survive that caveat.

```{r}
#| label: fig-crossmodality-bulk-counts
#| fig-width: 9
#| fig-height: 4.4
#| fig-cap: "Bulk omics feature counts at FDR < 0.10. Proteome, raw phosphosite and parent-protein-corrected phosphosite layers are 24-month bulk hippocampus, not microglia-sorted. Many primary hits are sensitive to genotype-blocked run order."
bp <- bulk_counts[bulk_counts$contrast %in% focus, , drop = FALSE]
bp$contrast <- factor(bp$contrast, levels = focus, labels = contrast_labels[focus])
bp$layer <- factor(bp$layer, levels = names(layer_labels), labels = unname(layer_labels))

ggplot(bp, aes(contrast, n_fdr_0_10, fill = layer)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.68) +
  scale_fill_manual(values = c("proteome" = "#3F5A6B",
                               "raw phospho" = "#B0344D",
                               "corrected phospho" = "#6B7A3F"),
                    name = NULL) +
  labs(x = NULL, y = "features at FDR < 0.10", title = "Bulk significance is largest in amyloid contrasts") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| label: fig-crossmodality-bulk-run-index
#| fig-width: 9.2
#| fig-height: 4.8
#| fig-cap: "Run-index sensitivity across the 24-month bulk hippocampus layers. Tile fill is the fraction of primary FDR < 0.10 rows that lose support or flip sign after additive run-index adjustment; text is lost or flipped over primary hits. Most bulk signal is run-order-sensitive, so it is corroborative rather than a stand-alone microglial mechanism claim."
ggplot(bulk_run_plot, aes(contrast, layer_label, fill = loss_fraction)) +
  geom_tile(colour = "white", linewidth = 0.45) +
  geom_text(aes(label = cell_label), size = 3) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D", limits = c(0, 1),
                      name = "loss fraction") +
  labs(x = NULL, y = NULL, title = "Bulk run-index sensitivity is load-bearing") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r}
#| label: crossmodality-run-index-table
ri <- run_index[run_index$contrast %in% focus, , drop = FALSE]
ri <- ri[order(match(ri$contrast, focus), ri$layer, method = "radix", na.last = TRUE), , drop = FALSE]
ritab <- data.frame(
  layer = layer_labels[ri$layer],
  contrast = contrast_labels[ri$contrast],
  status = ri$status,
  `primary FDR<0.10` = fmt_int(ri$n_primary_sig),
  `lost or flipped` = fmt_int(ri$n_lost_or_flipped),
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(ritab, row.names = FALSE,
             caption = "Run-index sensitivity for bulk layers. Rows losing support or flipping sign are downgraded in the synthesis.")
```

```{r}
#| label: fig-crossmodality-phospho-correction
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Raw versus parent-protein-corrected phosphosite effects in the 24-month bulk hippocampus layer. Points are pre-binned by site-level log2 fold-change, with selected top sites labelled. Parent correction changes effect geometry but does not turn this bulk phospho layer into a microglia-sorted mechanism assay."
ggplot(pc_bins) +
  geom_abline(slope = 1, intercept = 0, colour = "grey65", linewidth = 0.35) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.25) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.25) +
  geom_rect(aes(xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max,
                fill = log_n),
            colour = NA) +
  ggrepel::geom_text_repel(
    data = pc_lab,
    aes(raw_logFC, corrected_logFC, label = site_label),
    inherit.aes = FALSE,
    size = 2.1,
    seed = 42L,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.12
  ) +
  facet_wrap(~ contrast, ncol = 2) +
  scale_fill_gradient(low = "grey94", high = "#3F5A6B", name = "log10(n + 1)") +
  labs(x = "raw phosphosite log2 fold-change",
       y = "parent-corrected phosphosite log2 fold-change",
       title = "Parent correction changes phosphosite effect geometry") +
  theme(panel.grid = element_blank())
```

```{r}
#| label: fig-crossmodality-anchor-heatmap
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Bulk anchor coverage and effects for clearance, synaptic, complement and Gsk3b-related rows. Each tile is the best available row per anchor symbol, layer and contrast, selected by FDR then absolute t-statistic; outlined points mark FDR < 0.10. Blank cells are unmeasured or filtered in that bulk layer."
ggplot(anchor_plot, aes(contrast, anchor_label, fill = logFC)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_point(data = anchor_plot[anchor_plot$supported, , drop = FALSE],
             shape = 21, size = 1.9, stroke = 0.35, fill = "white", colour = "grey20") +
  facet_wrap(~ layer_label, nrow = 1) +
  scale_fill_gradient2(low = "#3F5A6B", mid = "white", high = "#B0344D",
                       midpoint = 0, name = "logFC") +
  labs(x = NULL, y = NULL, title = "Bulk anchor rows are sparse and layer-dependent") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

## Spatial Composition And Clearance Axis

SpatialDecon remains intentionally skipped here: the live preflight status is
`r decon$status`, with action `r decon$action`. The blocking reasons are:
`r decon_reasons`. This means P4 does not claim spatial abundance or cell counting from GeoMx.
It reports only GeoMx DE plus the measured clearance-axis evidence.

The targeted CCC-lite rule earns `r earned_sentence`; `ccc_called` stays
`r cmr$clearance$verdict$ccc_called`, because no CellChat, LIANA or MultiNicheNet model is run.
`r earned_detail_sentence` Pairs that remain measured but unearned are not promoted to
communication claims.

```{r}
#| label: crossmodality-clearance-table
pt <- pairs[pairs$contrast %in% focus, , drop = FALSE]
pt <- pt[order(pt$pair, match(pt$contrast, focus), method = "radix", na.last = TRUE), , drop = FALSE]
ptab <- data.frame(
  pair = pt$pair,
  contrast = contrast_labels[pt$contrast],
  status = pt$status,
  `supported modalities` = ifelse(pt$coherent_supported_modalities == "",
                                  "-", pt$coherent_supported_modalities),
  `microglia strong` = pt$microglia_strong,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(ptab, row.names = FALSE,
             caption = "Clearance-axis CCC-lite verdicts. This is measured pair support only; full cell-cell communication was not run.")
```

```{r}
#| label: fig-crossmodality-clearance-grid
#| fig-width: 8.4
#| fig-height: 3.8
#| fig-cap: "Clearance-pair CCC-lite support across the focus contrasts. Tile fill marks the conservative earned/not-earned verdict; text is the number of coherent supported broad modality classes, and the point marks a strong whole-microglia anchor. Only Apoe-Trem2 in amyloid on P301S earns measured pair support; no full CCC model is called."
ggplot(clear_plot, aes(contrast, pair_label, fill = status_label)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = support_label), size = 3.2) +
  geom_point(data = clear_plot[clear_plot$microglia_strong, , drop = FALSE],
             shape = 21, fill = "white", colour = "grey20", size = 2.1, stroke = 0.35) +
  scale_fill_manual(values = c(`not earned` = "#D8D8D8", earned = "#B0344D"),
                    name = NULL) +
  labs(x = NULL, y = NULL, title = "Measured clearance-pair support is focused") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank(),
        legend.position = "bottom")
```

## Integrated Divergence

The integration target keeps nulls and mixed signs rather than compressing them into a v1-style
claims ledger. Across the four focus contrasts it carries
`r fmt_int(sum(contrast_summary$n_multimodality))` symbol-contrast rows observed in at least two
broad modality classes, `r fmt_int(sum(contrast_summary$n_any_significant))` rows with any
significant modality, and `r fmt_int(sum(contrast_summary$n_mixed_sign))` mixed-sign rows. Mixed
signs are expected in this design because microglia RNA, spatial tissue RNA, total protein and
phosphosite statistics answer related but non-identical questions.

```{r}
#| label: fig-crossmodality-axis-heatmap
#| fig-width: 9.2
#| fig-height: 5.2
#| fig-cap: "Axis-level pathway integration. Each tile is the best selected pathway row for an axis and contrast; fill counts broad modality classes with FDR-level support, while text gives the dominant direction or mixed-sign call. Rank score is descriptive only."
pa <- path_axis
pa$contrast <- factor(pa$contrast, levels = focus, labels = contrast_labels[focus])
pa$axis_label <- axis_labels[pa$axis]
pa$axis_label[is.na(pa$axis_label)] <- pa$axis[is.na(pa$axis_label)]
pa$axis_label <- factor(pa$axis_label, levels = rev(unique(pa$axis_label)))
pa$direction_label <- ifelse(pa$mixed_sign, "mix",
                             ifelse(pa$consistent_direction == "positive", "+",
                                    ifelse(pa$consistent_direction == "negative", "-", "0")))

ggplot(pa, aes(contrast, axis_label, fill = n_modalities_sig)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = direction_label), size = 3) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D", name = "sig modalities") +
  labs(x = NULL, y = NULL, title = "Cross-modality axis support") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| label: fig-crossmodality-symbol-matrix
#| fig-width: 9.4
#| fig-height: 7
#| fig-cap: "Top named-axis symbols across the focus contrasts. Fill counts broad modality classes with FDR-level support, not layer-level evidence groups; text gives the dominant direction, mixed sign, or 0 when no broad modality class clears FDR. Interaction rows are mostly 0, reinforcing that cross-modality interaction support is limited outside microglia composition."
ggplot(sym_plot, aes(contrast, symbol, fill = n_modalities_sig)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_text(aes(label = dir_label), size = 2.8) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D",
                      breaks = 0:max(sym_plot$n_modalities_sig),
                      name = "sig broad\nmodalities") +
  labs(x = NULL, y = NULL, title = "Symbol-level evidence uses broad modality classes") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r}
#| label: crossmodality-axis-symbols
sx <- cmr$divergence$axis_symbols
sx <- sx[sx$contrast %in% focus & (sx$n_modalities_sig > 0 | sx$contrast == "interaction"),
         , drop = FALSE]
sx <- sx[order(match(sx$contrast, focus), -sx$n_modalities_sig,
               -sx$n_modalities_present, sx$min_fdr, sx$symbol,
               method = "radix", na.last = TRUE), , drop = FALSE]
sx <- head(sx, 18L)
sxtab <- data.frame(
  symbol = sx$symbol,
  contrast = contrast_labels[sx$contrast],
  axis = sx$axis,
  `modalities present` = sx$n_modalities_present,
  `modalities significant` = sx$n_modalities_sig,
  direction = sx$direction_call,
  `minimum FDR` = fmt_p(sx$min_fdr),
  `significant modalities` = ifelse(sx$modalities_sig == "", "-", sx$modalities_sig),
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(sxtab, row.names = FALSE,
             caption = "Named-axis symbol highlights. Interaction anchors are carried even when no modality clears FDR, so nulls stay visible.")
```

```{r}
#| label: fig-crossmodality-pathway-heatmap
#| fig-width: 10
#| fig-height: 7.2
#| fig-cap: "Selected pathway-axis rows by contrast. Fill counts broad modality classes with FDR-level support; text reports the sign call, including mixed signs. These rows are descriptive integration summaries, not a formal meta-analysis or a v1-style convergence score."
ggplot(path_fig, aes(contrast, set_label, fill = n_modalities_sig)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_text(aes(label = direction_label), size = 2.7) +
  facet_wrap(~ axis_label, scales = "free_y", ncol = 2) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D",
                      breaks = 0:max(path_fig$n_modalities_sig),
                      name = "sig broad\nmodalities") +
  labs(x = NULL, y = NULL, title = "Pathway-axis support is strongest in amyloid contrasts") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r}
#| label: crossmodality-pathway-table
ph <- cmr$pathway$highlights
ph <- ph[order(match(ph$contrast, focus), -ph$rank_score, -ph$n_modalities_sig,
               ph$axis, ph$collection, ph$set,
               method = "radix", na.last = TRUE), , drop = FALSE]
ph <- head(ph, 12L)
phtab <- data.frame(
  contrast = contrast_labels[ph$contrast],
  axis = ph$axis,
  set = short_label(ph$set),
  direction = ph$consistent_direction,
  `modalities present` = ph$n_modalities_present,
  `modalities significant` = ph$n_modalities_sig,
  `mixed sign` = ph$mixed_sign,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(phtab, row.names = FALSE,
             caption = "Top pathway rows feeding the integrated view. These rows are descriptive summaries of ranked modality statistics, not a formal meta-analysis.")
```

## P4 Synthesis

Cross-modality evidence strengthens the amyloid-response spine. GeoMx and bulk layers both
show their largest effects in amyloid contrasts, pathway rows reproduce antigen-presentation,
phagocytic/clearance and synaptic axes, and `r clearance_synthesis_phrase`. The same evidence
narrows the claims: the tau-amyloid interaction is
not broadly spatial or bulk-omics significant, SpatialDecon abundance is not earned, full CCC is
not called, and genotype-blocked bulk run order downgrades many proteome/phosphosite rows. In the
closing synthesis, this chapter therefore corroborates `r closing_support_phrase`, while keeping
progression, NF-kB attenuation, Gsk3b and spatial abundance unresolved or unsupported in the
rebuilt data.
