# Cross-modality: spatial, bulk, and divergence checks {#sec-crossmodality}

Cross-modality audit: amyloid response corroborated; interaction support remains microglia-composition-centred.

```{r}
#| label: crossmodality-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)
targets::tar_source()
targets::tar_load(c(crossmodality_report, crossmodality_figures)) # compact targets only, no heavy modality objects
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

cmr <- crossmodality_report
cmf <- crossmodality_figures
alpha <- cmr$provenance$alpha
focus <- cmr$provenance$focus_contrasts
contrast_labels <- c(
  tau_alone = "tau alone",
  nlgf_in_maptki = "amyloid on MAPTKI",
  nlgf_in_p301s = "amyloid on P301S",
  tau_in_nlgf = "tau in NLGF",
  interaction = "interaction"
)
layer_labels <- c(
  proteome = "proteome",
  phospho_raw = "raw phospho",
  phospho_corrected = "corrected phospho"
)
axis_labels <- c(
  DAM = "DAM",
  homeostatic = "homeostatic",
  synaptic = "synaptic",
  clearance = "clearance",
  complement = "complement",
  antigen_presentation = "antigen presentation",
  MHC_APC = "MHC/APC",
  IFN = "IFN",
  NFkB = "NF-kB",
  Myc = "Myc",
  Proliferative = "proliferative",
  other = "other"
)
fmt_p <- function(p) {
  out <- rep("-", length(p))
  ok <- is.finite(p)
  out[ok] <- ifelse(p[ok] < 0.001,
                    formatC(p[ok], format = "e", digits = 2),
                    formatC(p[ok], format = "f", digits = 3))
  out
}
fmt_num <- function(x, digits = 2) ifelse(is.finite(x), formatC(x, format = "f", digits = digits), "-")
fmt_int <- function(x) format(as.integer(round(x)), big.mark = ",", trim = TRUE)
short_label <- function(x, n = 46L) {
  x <- gsub("^GO[A-Z]*_", "", x)
  x <- gsub("_", " ", x)
  ifelse(nchar(x) > n, paste0(substr(x, 1L, n - 3L), "..."), x)
}
wrap_label <- function(x, width = 22L) {
  vapply(as.character(x), function(s) {
    z <- strwrap(s, width = width)
    if (length(z)) paste(z, collapse = "\n") else s
  }, character(1), USE.NAMES = FALSE)
}
plot_contrast <- function(x, levels = focus, reverse = FALSE) {
  lv <- if (reverse) rev(levels) else levels
  factor(x, levels = lv, labels = unname(contrast_labels[lv]))
}
fmt_axis <- function(x) {
  vapply(strsplit(as.character(x), ";", fixed = TRUE), function(parts) {
    lab <- axis_labels[parts]
    lab[is.na(lab)] <- parts[is.na(lab)]
    paste(unname(lab), collapse = " / ")
  }, character(1), USE.NAMES = FALSE)
}
fmt_pair <- function(x) gsub("_", "-", x, fixed = TRUE)
one_row <- function(d, idx, label) {
  out <- d[idx, , drop = FALSE]
  stopifnot(nrow(out) >= 1L)
  out[1L, , drop = FALSE]
}

geom_counts <- cmr$geomx$counts
bulk_counts <- cmr$bulk$significant_counts
bulk_features <- cmr$bulk$feature_counts
run_index <- cmr$bulk$run_index
decon <- cmr$clearance$spatial_decon
pairs <- cmr$clearance$pair_support
contrast_summary <- cmr$divergence$contrast_summary
path_axis <- cmr$pathway$axis_summary

geo_volc_bins <- cmf$geomx_volcano$bins
geo_volc_bins$contrast <- plot_contrast(geo_volc_bins$contrast)
geo_volc_bins$log_n <- log10(geo_volc_bins$n + 1)
geo_volc_lab <- cmf$geomx_volcano$labels
geo_volc_lab$contrast <- plot_contrast(geo_volc_lab$contrast)

geo_sens <- cmf$geomx_sensitivity
geo_sens <- geo_sens[geo_sens$contrast %in% focus, , drop = FALSE]
geo_sens$contrast <- plot_contrast(geo_sens$contrast)
geo_sens$fit_label <- c(unblocked = "unblocked AOI",
                        collapsed_bio_unit = "bio-unit collapsed")[geo_sens$fit]
geo_sens$fit_label[is.na(geo_sens$fit_label)] <- geo_sens$fit[is.na(geo_sens$fit_label)]
geo_sens$fit_label <- factor(geo_sens$fit_label,
                             levels = c("unblocked AOI", "bio-unit collapsed"))
geo_sens_long <- rbind(
  data.frame(geo_sens[c("fit_label", "contrast")],
             metric = "lost from primary", n = -geo_sens$n_lost,
             stringsAsFactors = FALSE),
  data.frame(geo_sens[c("fit_label", "contrast")],
             metric = "gained in sensitivity", n = geo_sens$n_gained,
             stringsAsFactors = FALSE)
)
geo_sens_long$metric <- factor(geo_sens_long$metric,
                               levels = c("lost from primary", "gained in sensitivity"))

bulk_run_plot <- cmf$bulk_run_index
bulk_run_plot$contrast <- plot_contrast(bulk_run_plot$contrast, levels = names(contrast_labels))
bulk_run_plot$layer_label <- factor(layer_labels[bulk_run_plot$layer],
                                    levels = unname(layer_labels))
bulk_run_plot$cell_label <- ifelse(bulk_run_plot$n_primary_sig > 0,
                                   paste0(bulk_run_plot$n_lost_or_flipped, "/",
                                          bulk_run_plot$n_primary_sig),
                                   "0/0")

pc_bins <- cmf$phospho_raw_corrected$bins
pc_bins$contrast <- plot_contrast(pc_bins$contrast)
pc_bins$log_n <- log10(pc_bins$n + 1)
pc_lab <- cmf$phospho_raw_corrected$labels
pc_lab$best_fdr <- pmin(pc_lab$raw_fdr, pc_lab$corrected_fdr)
pc_lab <- do.call(rbind, lapply(split(pc_lab, pc_lab$contrast, drop = TRUE), function(d) {
  d <- d[order(d$best_fdr, -abs(d$raw_logFC) - abs(d$corrected_logFC),
               d$site_id, method = "radix", na.last = TRUE), , drop = FALSE]
  head(d, 6L)
}))
pc_lab$contrast <- plot_contrast(pc_lab$contrast)
pc_lab$site_label <- wrap_label(gsub("_", " ", pc_lab$site_id, fixed = TRUE), 14L)

anchor_plot <- cmf$anchor_effect_heatmap
anchor_plot <- anchor_plot[anchor_plot$contrast %in% names(contrast_labels), , drop = FALSE]
anchor_plot$key <- paste(anchor_plot$layer, anchor_plot$contrast, anchor_plot$anchor_symbols,
                         sep = "\r")
anchor_plot <- anchor_plot[order(anchor_plot$key, anchor_plot$adj.P.Val, -abs(anchor_plot$t),
                                 anchor_plot$feature, method = "radix", na.last = TRUE),
                           , drop = FALSE]
anchor_plot <- anchor_plot[!duplicated(anchor_plot$key), , drop = FALSE]
anchor_plot$contrast <- plot_contrast(anchor_plot$contrast, levels = names(contrast_labels))
anchor_plot$layer_label <- factor(layer_labels[anchor_plot$layer],
                                  levels = unname(layer_labels))
anchor_plot$anchor_label <- factor(anchor_plot$anchor_symbols,
                                   levels = rev(sort(unique(anchor_plot$anchor_symbols),
                                                     method = "radix")))
anchor_plot$supported <- anchor_plot$adj.P.Val < alpha

clear_plot <- cmf$clearance_pair_grid
clear_plot <- clear_plot[clear_plot$contrast %in% focus, , drop = FALSE]
clear_plot$contrast <- plot_contrast(clear_plot$contrast)
clear_plot$pair_label <- factor(fmt_pair(clear_plot$pair),
                                levels = rev(unique(fmt_pair(clear_plot$pair))))
clear_plot$status_label <- factor(ifelse(clear_plot$status == "earned", "earned", "not earned"),
                                  levels = c("not earned", "earned"))
clear_plot$support_label <- ifelse(clear_plot$n_coherent_supported_modalities > 0,
                                   clear_plot$n_coherent_supported_modalities, "")

sym_all <- cmf$symbol_modality_matrix
sym_all <- sym_all[sym_all$contrast %in% focus, , drop = FALSE]
sym_rank <- do.call(rbind, lapply(split(sym_all, sym_all$symbol, drop = TRUE), function(d) {
  data.frame(symbol = d$symbol[1L],
             max_sig = max(d$n_modalities_sig, na.rm = TRUE),
             max_present = max(d$n_modalities_present, na.rm = TRUE),
             min_fdr = min(d$min_fdr, na.rm = TRUE),
             max_rank = max(d$rank_score, na.rm = TRUE),
             stringsAsFactors = FALSE)
}))
sym_rank <- sym_rank[order(-sym_rank$max_sig, sym_rank$min_fdr, -sym_rank$max_present,
                           sym_rank$symbol, method = "radix", na.last = TRUE), , drop = FALSE]
sym_keep <- head(sym_rank$symbol, 18L)
sym_plot <- sym_all[sym_all$symbol %in% sym_keep, , drop = FALSE]
sym_plot$contrast <- plot_contrast(sym_plot$contrast)
sym_plot$symbol <- factor(sym_plot$symbol, levels = rev(sym_keep))
sym_plot$dir_label <- ifelse(sym_plot$n_modalities_sig == 0, "0",
                             ifelse(sym_plot$direction_call == "mixed", "mix",
                                    ifelse(sym_plot$direction_call == "positive", "+", "-")))

path_fig <- cmf$pathway_axis_heatmap
path_fig <- path_fig[path_fig$contrast %in% focus, , drop = FALSE]
path_fig$contrast <- plot_contrast(path_fig$contrast)
path_fig$axis_label <- fmt_axis(path_fig$axis)
path_fig$set_label <- short_label(path_fig$set, 34L)
path_fig$set_label <- factor(path_fig$set_label, levels = rev(unique(path_fig$set_label)))
path_fig$direction_label <- ifelse(path_fig$mixed_sign, "mix",
                                   ifelse(path_fig$consistent_direction == "positive", "+",
                                          ifelse(path_fig$consistent_direction == "negative", "-", "0")))

geo_int <- one_row(geom_counts, geom_counts$contrast == "interaction", "GeoMx interaction counts")
geo_amy_k <- one_row(geom_counts, geom_counts$contrast == "nlgf_in_maptki", "GeoMx amyloid MAPTKI counts")
geo_amy_p <- one_row(geom_counts, geom_counts$contrast == "nlgf_in_p301s", "GeoMx amyloid P301S counts")
bulk_primary_sig <- sum(run_index$n_primary_sig, na.rm = TRUE)
bulk_lost <- sum(run_index$n_lost_or_flipped, na.rm = TRUE)
bulk_loss_pct <- if (bulk_primary_sig > 0) round(100 * bulk_lost / bulk_primary_sig, 1) else 0
worst_run <- run_index[order(-run_index$n_lost_or_flipped, run_index$layer,
                             match(run_index$contrast, names(contrast_labels)),
                             method = "radix", na.last = TRUE), , drop = FALSE][1L, ]
decon_reasons <- if (length(decon$reasons)) paste(decon$reasons, collapse = "; ") else "no blocking reason recorded"
decon_unresolved <- if (is.list(decon$decon) && is.data.frame(decon$decon$unresolved_aoi)) {
  decon$decon$unresolved_aoi
} else {
  data.frame()
}
decon_unresolved_n <- if (nrow(decon_unresolved)) length(unique(decon_unresolved$aoi)) else 0L
decon_unresolved_sentence <- if (decon_unresolved_n > 0L) {
  paste0(decon_unresolved_n, " AOIs have near-zero total beta in the attempted fit")
} else {
  decon_reasons
}
decon_ref <- if (is.data.frame(decon$reference) && nrow(decon$reference)) decon$reference else data.frame()
decon_ref_sentence <- if (nrow(decon_ref)) {
  paste0(
    paste0(decon_ref$arm, " profile ", decon_ref$status, " (",
           fmt_int(decon_ref$n_genes), " genes x ", decon_ref$n_profiles,
           " profiles, max |cor| ", fmt_num(decon_ref$max_abs_correlation, 2), ")"),
    collapse = "; "
  )
} else {
  "reference-profile summary unavailable in this compact bundle"
}
decon_nuclei_sentence <- if (is.list(decon$decon) && is.list(decon$decon$nuclei) &&
                             isFALSE(decon$decon$nuclei$absolute_rescaling_enabled)) {
  paste0("Nuclei-rescaled absolute cell counts remain disabled",
         if (is.finite(decon$decon$nuclei$n_sentinel %||% NA_real_)) {
           paste0(" (", decon$decon$nuclei$n_sentinel, " sentinel nuclei values)")
         } else {
           ""
         })
} else {
  "Nuclei-rescaled absolute cell counts are not used"
}
decon_audit <- if (is.list(decon$residual_audit) && is.data.frame(decon$residual_audit$summary)) {
  decon$residual_audit$summary
} else {
  data.frame()
}
decon_audit_sentence <- if (nrow(decon_audit) && any(is.finite(decon_audit$median_metric))) {
  paste0("residual QC remains available (median RMS residual ",
         paste(paste0(decon_audit$arm, " ", fmt_num(decon_audit$median_metric, 3)),
               collapse = "; "), ")")
} else {
  "residual QC is unavailable"
}
earned_pairs <- pairs[pairs$status == "earned", , drop = FALSE]
earned_sentence <- if (nrow(earned_pairs)) {
  paste0(paste(unique(earned_pairs$pair), collapse = ", "), " in ",
         paste(contrast_labels[earned_pairs$contrast], collapse = ", "))
} else {
  "no pair"
}
fmt_modalities <- function(x) {
  z <- unlist(strsplit(x, ";", fixed = TRUE), use.names = FALSE)
  z <- z[nzchar(z)]
  z <- gsub("_", " ", z, fixed = TRUE)
  paste(z, collapse = ", ")
}
earned_details <- if (nrow(earned_pairs)) {
  vapply(seq_len(nrow(earned_pairs)), function(i) {
    ep <- earned_pairs[i, , drop = FALSE]
    mods <- fmt_modalities(ep$coherent_supported_modalities)
    paste0(fmt_pair(ep$pair), " in ", contrast_labels[ep$contrast],
           if (nzchar(mods)) paste0(" (supported by ", mods, ")") else "")
  }, character(1))
} else {
  character()
}
earned_detail_sentence <- if (length(earned_details)) {
  paste0(if (length(earned_details) == 1L) "The earned row is " else "The earned rows are ",
         paste(earned_details, collapse = "; "), ".")
} else {
  "No measured pair earns a communication claim under the CCC-lite rule."
}
earned_pair_names <- unique(earned_pairs$pair)
closing_support_phrase <- if (length(earned_pair_names)) {
  "DAM activation, synaptic suppression and the earned clearance-axis support"
} else {
  "DAM activation and synaptic suppression, with clearance kept descriptive"
}
cross_board <- cmf$status_board
cross_board$axis <- factor(cross_board$axis, levels = rev(cross_board$axis))
cross_board$status <- factor(cross_board$status,
                             levels = c("corroborative", "focused support", "blocked",
                                        "open caveat", "descriptive"))
cross_board$label <- ifelse(is.finite(cross_board$n),
                            paste0(cross_board$status, "\n", fmt_int(cross_board$n)),
                            as.character(cross_board$status))
```

```{r}
#| label: fig-crossmodality-status-board
#| fig-width: 8.4
#| fig-height: 3.4
#| fig-cap: "Cross-modality status board: SpatialDecon blocked, CCC-lite focused."
ggplot(cross_board, aes(status, axis, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = label), size = 3, lineheight = 0.9) +
  scale_fill_manual(
    values = c(corroborative = "#B0344D", `focused support` = "#6B7A3F",
               blocked = "#3F5A6B", `open caveat` = "#E08754",
               descriptive = "#D8D8D8"),
    guide = "none"
  ) +
  labs(x = NULL, y = NULL, title = "Cross-modality evidence board") +
  theme(panel.grid = element_blank())
```

## GeoMx Spatial DE

GeoMx uses `r fmt_int(cmr$geomx$provenance$n_aoi)` AOIs with bio-unit blocking: amyloid DE is large, interaction DE smaller.

```{r}
#| label: fig-crossmodality-geomx-counts
#| fig-width: 8.8
#| fig-height: 4.2
#| fig-cap: "GeoMx DE counts are largest in amyloid contrasts."
geo_plot <- geom_counts[geom_counts$contrast %in% focus, , drop = FALSE]
geo_long <- rbind(
  data.frame(contrast = geo_plot$contrast, direction = "up", n = geo_plot$n_up_fdr_0_10),
  data.frame(contrast = geo_plot$contrast, direction = "down", n = geo_plot$n_down_fdr_0_10)
)
geo_long$contrast <- factor(geo_long$contrast, levels = focus, labels = contrast_labels[focus])
geo_long$direction <- factor(geo_long$direction, levels = c("up", "down"))

ggplot(geo_long, aes(contrast, n, fill = direction)) +
  geom_col(width = 0.72) +
  scale_fill_manual(values = c(up = "#B0344D", down = "#3F5A6B"), name = "direction") +
  labs(x = NULL, y = "GeoMx rows at FDR < 0.10", title = "Spatial DE is strongest for amyloid") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| label: fig-crossmodality-geomx-volcano
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "GeoMx primary volcanoes show dense amyloid structure."
ggplot(geo_volc_bins) +
  geom_rect(aes(xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max,
                fill = log_n),
            colour = NA) +
  geom_vline(xintercept = 0, colour = "grey65", linewidth = 0.3) +
  ggrepel::geom_text_repel(
    data = geo_volc_lab,
    aes(effect, neg_log10_p, label = label),
    inherit.aes = FALSE,
    size = 2.25,
    seed = 42L,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.12
  ) +
  facet_wrap(~ contrast, ncol = 2) +
  scale_fill_gradient(low = "grey94", high = "#B0344D", name = "log10(n + 1)") +
  labs(x = "GeoMx log2 fold-change", y = "-log10 p-value",
       title = "GeoMx primary spatial DE by contrast") +
  theme(panel.grid = element_blank())
```

```{r}
#| label: fig-crossmodality-geomx-sensitivity
#| fig-width: 9.2
#| fig-height: 4.8
#| fig-cap: "GeoMx sensitivity marks rows gained or lost versus the primary model."
ggplot(geo_sens_long, aes(contrast, n, fill = metric)) +
  geom_hline(yintercept = 0, colour = "grey65") +
  geom_col(width = 0.68) +
  facet_wrap(~ fit_label, nrow = 1) +
  scale_fill_manual(values = c(`lost from primary` = "#3F5A6B",
                               `gained in sensitivity` = "#B0344D"),
                    name = NULL) +
  scale_y_continuous(labels = function(x) abs(x)) +
  labs(x = NULL, y = "GeoMx rows changed at FDR < 0.10",
       title = "Spatial DE support depends on the replicate treatment") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```

## Bulk Proteome And Phospho

Bulk layers are 24-month hippocampus, not microglia-sorted; `r fmt_int(bulk_lost)` of `r fmt_int(bulk_primary_sig)` primary hits lose or flip under run-index sensitivity.

```{r}
#| label: fig-crossmodality-bulk-counts
#| fig-width: 9
#| fig-height: 4.4
#| fig-cap: "Bulk omics FDR counts are corroborative, not microglia-sorted."
bp <- bulk_counts[bulk_counts$contrast %in% focus, , drop = FALSE]
bp$contrast <- factor(bp$contrast, levels = focus, labels = contrast_labels[focus])
bp$layer <- factor(bp$layer, levels = names(layer_labels), labels = unname(layer_labels))

ggplot(bp, aes(contrast, n_fdr_0_10, fill = layer)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.68) +
  scale_fill_manual(values = c("proteome" = "#3F5A6B",
                               "raw phospho" = "#B0344D",
                               "corrected phospho" = "#6B7A3F"),
                    name = NULL) +
  labs(x = NULL, y = "features at FDR < 0.10", title = "Bulk significance is largest in amyloid contrasts") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| label: fig-crossmodality-bulk-run-index
#| fig-width: 9.2
#| fig-height: 4.8
#| fig-cap: "Run-index sensitivity downgrades many bulk hits."
ggplot(bulk_run_plot, aes(contrast, layer_label, fill = loss_fraction)) +
  geom_tile(colour = "white", linewidth = 0.45) +
  geom_text(aes(label = cell_label), size = 3) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D", limits = c(0, 1),
                      name = "loss fraction") +
  labs(x = NULL, y = NULL, title = "Bulk run-index sensitivity is load-bearing") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r}
#| label: fig-crossmodality-phospho-correction
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Parent correction changes bulk phosphosite effect geometry."
ggplot(pc_bins) +
  geom_abline(slope = 1, intercept = 0, colour = "grey65", linewidth = 0.35) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.25) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.25) +
  geom_rect(aes(xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max,
                fill = log_n),
            colour = NA) +
  ggrepel::geom_text_repel(
    data = pc_lab,
    aes(raw_logFC, corrected_logFC, label = site_label),
    inherit.aes = FALSE,
    size = 2.1,
    seed = 42L,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.12
  ) +
  facet_wrap(~ contrast, ncol = 2) +
  scale_fill_gradient(low = "grey94", high = "#3F5A6B", name = "log10(n + 1)") +
  labs(x = "raw phosphosite log2 fold-change",
       y = "parent-corrected phosphosite log2 fold-change",
       title = "Parent correction changes phosphosite effect geometry") +
  theme(panel.grid = element_blank())
```

```{r}
#| label: fig-crossmodality-anchor-heatmap
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Bulk anchor rows are sparse and layer-dependent."
ggplot(anchor_plot, aes(contrast, anchor_label, fill = logFC)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_point(data = anchor_plot[anchor_plot$supported, , drop = FALSE],
             shape = 21, size = 1.9, stroke = 0.35, fill = "white", colour = "grey20") +
  facet_wrap(~ layer_label, nrow = 1) +
  scale_fill_gradient2(low = "#3F5A6B", mid = "white", high = "#B0344D",
                       midpoint = 0, name = "logFC") +
  labs(x = NULL, y = NULL, title = "Bulk anchor rows are sparse and layer-dependent") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

## Spatial Composition And Clearance Axis

SpatialDecon abundance is attempted but blocked (`r decon_unresolved_sentence`); CCC-lite earns `r earned_sentence`, with no full CCC call.

```{r}
#| label: fig-crossmodality-clearance-grid
#| fig-width: 8.4
#| fig-height: 3.8
#| fig-cap: "CCC-lite earns focused measured support; full CCC is absent."
ggplot(clear_plot, aes(contrast, pair_label, fill = status_label)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = support_label), size = 3.2) +
  geom_point(data = clear_plot[clear_plot$microglia_strong, , drop = FALSE],
             shape = 21, fill = "white", colour = "grey20", size = 2.1, stroke = 0.35) +
  scale_fill_manual(values = c(`not earned` = "#D8D8D8", earned = "#B0344D"),
                    name = NULL) +
  labs(x = NULL, y = NULL, title = "Measured clearance-pair support is focused") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank(),
        legend.position = "bottom")
```

## Integrated Divergence

Integration keeps nulls and mixed signs visible instead of scoring convergence.

```{r}
#| label: fig-crossmodality-axis-heatmap
#| fig-width: 9.2
#| fig-height: 5.2
#| fig-cap: "Axis-level support is descriptive and strongest in amyloid contrasts."
pa <- path_axis
pa$contrast <- factor(pa$contrast, levels = focus, labels = contrast_labels[focus])
pa$axis_label <- axis_labels[pa$axis]
pa$axis_label[is.na(pa$axis_label)] <- pa$axis[is.na(pa$axis_label)]
pa$axis_label <- factor(pa$axis_label, levels = rev(unique(pa$axis_label)))
pa$direction_label <- ifelse(pa$mixed_sign, "mix",
                             ifelse(pa$consistent_direction == "positive", "+",
                                    ifelse(pa$consistent_direction == "negative", "-", "0")))

ggplot(pa, aes(contrast, axis_label, fill = n_modalities_sig)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = direction_label), size = 3) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D", name = "sig modalities") +
  labs(x = NULL, y = NULL, title = "Cross-modality axis support") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r}
#| label: fig-crossmodality-symbol-matrix
#| fig-width: 9.4
#| fig-height: 7
#| fig-cap: "Symbol matrix uses broad modality classes and keeps interaction nulls visible."
ggplot(sym_plot, aes(contrast, symbol, fill = n_modalities_sig)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_text(aes(label = dir_label), size = 2.8) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D",
                      breaks = 0:max(sym_plot$n_modalities_sig),
                      name = "sig broad\nmodalities") +
  labs(x = NULL, y = NULL, title = "Symbol-level evidence uses broad modality classes") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r}
#| label: fig-crossmodality-pathway-heatmap
#| fig-width: 10
#| fig-height: 7.2
#| fig-cap: "Pathway-axis summaries are descriptive, not formal meta-analysis."
ggplot(path_fig, aes(contrast, set_label, fill = n_modalities_sig)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_text(aes(label = direction_label), size = 2.7) +
  facet_wrap(~ axis_label, scales = "free_y", ncol = 2) +
  scale_fill_gradient(low = "#EEF1F2", high = "#B0344D",
                      breaks = 0:max(path_fig$n_modalities_sig),
                      name = "sig broad\nmodalities") +
  labs(x = NULL, y = NULL, title = "Pathway-axis support is strongest in amyloid contrasts") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

## Cross-modality status

Cross-modality corroborates `r closing_support_phrase`; SpatialDecon, full CCC, progression, NF-kB and Gsk3b remain unsupported or blocked.
