# Mechanism: Myc signal, NF-kB check, and bulk kinase support {#sec-mechanism}

P3 reframes the mechanism question around the rebuilt P1/P2 result. Amyloid drives a
microglial homeostatic-to-DAM programme, and mutant tau modulates that amyloid response
mainly by expanding the DAM compartment, not by a supported further advance along the
activation ordering. The analyses below ask which RNA pathway, TF and kinase signals
plausibly sit under that state expansion. They are treated as tests, not as assumed winners:
Myc and Gsk3b can be supported or not by the rebuilt data, and NF-kB attenuation is only
called if its predeclared transcript-level gate clears.

```{r}
#| label: mechanism-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)
targets::tar_source()
targets::tar_load(c(mechanism_report, mechanism_figures)) # compact targets only, no heavy Seurat object
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

mrd <- mechanism_report
mfg <- mechanism_figures
pp  <- mrd$pathway_project
go  <- mrd$pathway_go_top
tf  <- mrd$tf_highlights
nf  <- mrd$nfkb$table
kin <- mrd$kinase$table
comp <- mrd$composition_anchor
traj <- mrd$trajectory_anchor
alpha <- mrd$provenance$alpha

fmt_p <- function(p) {
  out <- rep("-", length(p))
  ok <- is.finite(p)
  out[ok] <- ifelse(p[ok] < 0.001,
                    formatC(p[ok], format = "e", digits = 2),
                    formatC(p[ok], format = "f", digits = 3))
  out
}
fmt_num <- function(x, digits = 2) ifelse(is.finite(x), formatC(x, format = "f", digits = digits), "-")
one_row <- function(d, idx, label) {
  out <- d[idx, , drop = FALSE]
  stopifnot(nrow(out) >= 1L)
  out[1, , drop = FALSE]
}

myc_whole <- one_row(tf, tf$population == "whole_microglia" & tf$contrast == "interaction" &
                       tf$source == "Myc", "whole Myc interaction")
myc_dam <- one_row(tf, tf$population == "DAM" & tf$contrast == "interaction" &
                     tf$source == "Myc", "DAM Myc interaction")
dam_int <- one_row(comp, comp$contrast == "interaction" & comp$substate == "DAM",
                   "DAM composition interaction")
traj_comp <- one_row(traj, traj$measure == "comp_cf", "trajectory composition")
traj_prog <- one_row(traj, traj$measure == "progression_cf", "trajectory progression")
proj_dam_int <- one_row(pp, pp$population == "whole_microglia" & pp$contrast == "interaction" &
                          pp$pathway == "DAM", "project DAM interaction")
gsk_int <- one_row(kin, kin$source == "Gsk3b" & kin$contrast == "interaction",
                   "Gsk3b interaction")
gsk_tau_nlgf <- one_row(kin, kin$source == "Gsk3b" & kin$contrast == "tau_in_nlgf",
                        "Gsk3b tau_in_nlgf")
nf_primary <- nf[nf$primary_test %in% TRUE, , drop = FALSE]
stopifnot(nrow(nf_primary) == 2L)
nf_tf <- one_row(nf_primary, nf_primary$test == "tf_family", "NF-kB TF primary")
nf_gsea <- one_row(nf_primary, nf_primary$test == "target_gsea", "NF-kB target primary")
nf_status <- mrd$nfkb$verdict$status
nf_sentence <- if (identical(nf_status, "supported")) {
  "The predeclared NF-kB attenuation gate is supported."
} else if (identical(nf_status, "discordant")) {
  "The predeclared NF-kB attenuation gate is discordant, so P3 does not support the attenuation claim."
} else if (identical(nf_status, "directional_only")) {
  "The predeclared NF-kB attenuation gate is directionally negative but does not clear its FDR rule."
} else {
  "The predeclared NF-kB attenuation gate is not supported."
}
nf_detail_sentence <- if (isTRUE(mrd$nfkb$verdict$primary_discordant)) {
  paste0("The two primary rows disagree in sign: the NF-kB TF-family score is ",
         fmt_num(nf_tf$score), " (primary-family FDR ", fmt_p(nf_tf$primary_family_fdr),
         "), while the sign-aware NF-kB target-GSEA score is ", fmt_num(nf_gsea$score),
         " (primary-family FDR ", fmt_p(nf_gsea$primary_family_fdr), ").")
} else if (isTRUE(mrd$nfkb$verdict$primary_negative)) {
  paste0("Both primary rows are negative: the NF-kB TF-family score is ",
         fmt_num(nf_tf$score), " (primary-family FDR ", fmt_p(nf_tf$primary_family_fdr),
         ") and the sign-aware NF-kB target-GSEA score is ", fmt_num(nf_gsea$score),
         " (primary-family FDR ", fmt_p(nf_gsea$primary_family_fdr), ").")
} else {
  paste0("The primary rows are not concordantly negative: the NF-kB TF-family score is ",
         fmt_num(nf_tf$score), " (primary-family FDR ", fmt_p(nf_tf$primary_family_fdr),
         ") and the sign-aware NF-kB target-GSEA score is ", fmt_num(nf_gsea$score),
         " (primary-family FDR ", fmt_p(nf_gsea$primary_family_fdr), ").")
}
nf_conclusion_sentence <- if (identical(nf_status, "supported")) {
  "That clears the specific attenuation rule."
} else if (identical(nf_status, "directional_only")) {
  "That leaves the evidence directional but below the predeclared FDR rule."
} else {
  "That makes the rebuilt evidence insufficient for the specific attenuation claim."
}
gsk_sentence <- if (isTRUE(gsk_int$significant) || isTRUE(gsk_tau_nlgf$significant)) {
  "Gsk3b has nominal rebuilt kinase support in at least one focal contrast, subject to the bulk-phospho caveats below."
} else {
  "Gsk3b is covered by the kinase prior and phosphosite data, but it is not significant in either focal rebuilt contrast."
}
sig_tau <- kin[kin$contrast == "tau_in_nlgf" & kin$significant %in% TRUE &
                 kin$source != "Gsk3b", , drop = FALSE]
sig_tau <- sig_tau[order(sig_tau$fdr, -abs(sig_tau$score), sig_tau$source,
                         method = "radix", na.last = TRUE), , drop = FALSE]
run_supported_tau <- sum(sig_tau$run_index_supports %in% TRUE)

contrast_labels <- c(tau_alone = "tau alone",
                     nlgf_in_maptki = "amyloid on MAPTKI",
                     nlgf_in_p301s = "amyloid on P301S",
                     tau_in_nlgf = "tau in NLGF",
                     interaction = "interaction")
project_labels <- c(DAM = "DAM",
                    Homeostatic = "homeostatic",
                    IFN = "IFN",
                    MHC_APC = "MHC/APC",
                    NFkB_Activated_Targets = "NF-kB activated targets",
                    NFkB_CollecTRI_Targets = "NF-kB target union",
                    NFkB_Repressed_Targets = "NF-kB repressed targets")
population_labels <- c(whole_microglia = "whole microglia",
                       DAM = "DAM",
                       Homeostatic = "homeostatic")
fmt_contrast <- function(x, rev = FALSE) {
  lv <- if (rev) rev(names(contrast_labels)) else names(contrast_labels)
  factor(x, levels = lv, labels = unname(contrast_labels[lv]))
}
fmt_population <- function(x) {
  out <- population_labels[x]
  out[is.na(out)] <- x[is.na(out)]
  unname(out)
}
short_pathway <- function(x, max_chars = 42L) {
  out <- gsub("^GO[A-Z]*_", "", x)
  out <- gsub("_", " ", out)
  out <- tolower(out)
  too_long <- nchar(out) > max_chars
  out[too_long] <- paste0(substr(out[too_long], 1L, max_chars - 3L), "...")
  out
}

m_project <- mfg$project_pathway_heatmap
m_project$contrast <- fmt_contrast(m_project$contrast)
m_project$population_label <- factor(fmt_population(m_project$population),
                                     levels = unname(population_labels))
m_project$pathway_label <- project_labels[m_project$pathway]
m_project$pathway_label[is.na(m_project$pathway_label)] <-
  m_project$pathway[is.na(m_project$pathway_label)]
m_project$pathway_label <- factor(m_project$pathway_label,
                                  levels = rev(unname(project_labels)))

go_select <- mfg$go_top_dotplot
go_select <- go_select[go_select$contrast %in% names(contrast_labels), , drop = FALSE]
go_select <- go_select[order(go_select$population, go_select$contrast, go_select$fdr,
                             -abs(go_select$NES), go_select$pathway,
                             method = "radix", na.last = TRUE), , drop = FALSE]
go_select <- do.call(rbind, lapply(split(go_select, list(go_select$population, go_select$contrast),
                                         drop = TRUE), head, n = 2L))
go_select$contrast <- fmt_contrast(go_select$contrast)
go_select$population_label <- factor(fmt_population(go_select$population),
                                     levels = unname(population_labels))
go_select$pathway_label <- paste(short_pathway(go_select$pathway), go_select$collection, sep = " / ")
go_select$pathway_label <- factor(go_select$pathway_label,
                                  levels = rev(unique(go_select$pathway_label)))
go_select$supported <- go_select$fdr < alpha

tf_focus <- mfg$tf_lollipop
tf_focus <- tf_focus[tf_focus$source %in% c("Myc", "Nfkb1", "Nfkb2", "Rela", "Relb") &
                       tf_focus$contrast %in% c("interaction", "tau_in_nlgf"), , drop = FALSE]
tf_focus$contrast <- fmt_contrast(tf_focus$contrast)
tf_focus$population_label <- factor(fmt_population(tf_focus$population),
                                    levels = unname(population_labels))
tf_focus$source <- factor(tf_focus$source, levels = rev(c("Myc", "Nfkb1", "Nfkb2", "Rela", "Relb")))
tf_focus$class <- ifelse(tf_focus$source == "Myc", "Myc", "NF-kB family")
tf_focus$supported <- tf_focus$fdr < alpha

nf_plot <- mfg$nfkb_discordance$table
nf_plot <- nf_plot[nf_plot$population == "whole_microglia" &
                     nf_plot$contrast %in% c("interaction", "tau_in_nlgf"), , drop = FALSE]
nf_plot$contrast <- fmt_contrast(nf_plot$contrast)
nf_plot$test_label <- factor(nf_plot$test,
                             levels = c("tf_family", "target_gsea"),
                             labels = c("TF-family", "target GSEA"))
nf_plot$role <- ifelse(nf_plot$primary_test, "primary", "supportive")
nf_plot$tile_label <- paste0(nf_plot$role, "\n", fmt_num(nf_plot$score),
                             "\nFDR ", fmt_p(nf_plot$primary_family_fdr))

kin_plot <- mfg$kinase_heatmap
kin_plot$contrast <- fmt_contrast(kin_plot$contrast)
kin_plot$primary_sig <- kin_plot$significant %in% TRUE
kin_plot$run_supported <- kin_plot$run_index_supports %in% TRUE
kin_order <- kin_plot[order(kin_plot$source != "Gsk3b", kin_plot$fdr,
                            -abs(kin_plot$score), kin_plot$source,
                            method = "radix", na.last = TRUE), "source"]
kin_plot$source <- factor(kin_plot$source, levels = rev(unique(kin_order)))
```

## Pathway Survey

The project gene sets recapitulate the upstream story in RNA. The whole-microglia DAM set is
positive in the tau-amyloid interaction (NES `r fmt_num(proj_dam_int$NES)`, FDR
`r fmt_p(proj_dam_int$padj)`), aligning with the P1 DAM-composition interaction (DAM propeller
ratio `r fmt_num(dam_int$prop_ratio)`, FDR `r fmt_p(dam_int$fdr_global)`) and the P2
composition channel (effect `r fmt_num(traj_comp$coef)`, FDR `r fmt_p(traj_comp$fdr)`). The
progression channel remains non-significant (effect `r fmt_num(traj_prog$coef)`, FDR
`r fmt_p(traj_prog$fdr)`), so the mechanism target is DAM-cell expansion and amyloid-response
modulation, not proven further activation of cells already on the trajectory.

```{r}
#| label: mechanism-project-sets
#| fig-width: 9.5
#| fig-height: 4.8
#| fig-cap: "Focused project gene sets in whole microglia. Tile colour is fgsea NES over the pseudobulk t-statistic ranks; outlined points mark FDR < 0.10 within the project collection and contrast. The DAM set stays positive in the interaction, while the NF-kB target arms are read through the formal gate below."
proj_plot <- pp[pp$population == "whole_microglia", , drop = FALSE]
proj_plot$contrast <- factor(proj_plot$contrast, levels = names(contrast_labels),
                             labels = unname(contrast_labels))
proj_plot$pathway_label <- project_labels[proj_plot$pathway]
proj_plot$pathway_label[is.na(proj_plot$pathway_label)] <- proj_plot$pathway[is.na(proj_plot$pathway_label)]
proj_plot$pathway_label <- factor(proj_plot$pathway_label, levels = rev(unique(project_labels[project_labels %in% proj_plot$pathway_label])))

ggplot(proj_plot, aes(contrast, pathway_label, fill = NES)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_point(data = proj_plot[proj_plot$padj < alpha, , drop = FALSE],
             shape = 21, size = 2, stroke = 0.3, fill = "white", colour = "grey20") +
  scale_fill_gradient2(low = "#3F5A6B", mid = "white", high = "#B0344D", midpoint = 0,
                       name = "NES") +
  labs(x = NULL, y = NULL, title = "Project gene-set activity")
```

```{r}
#| label: fig-mechanism-project-pathway
#| fig-width: 10
#| fig-height: 6.2
#| fig-cap: "Project gene-set survey across populations. Tile colour is fgsea NES; outlined points mark FDR < 0.10 within population and contrast. The interaction pattern is read against the closed P1/P2 result: DAM-cell expansion, not supported progression beyond composition."
ggplot(m_project, aes(contrast, pathway_label, fill = NES)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_point(data = m_project[m_project$fdr < alpha, , drop = FALSE],
             shape = 21, size = 1.8, stroke = 0.3, fill = "white", colour = "grey20") +
  facet_wrap(~ population_label, nrow = 1) +
  scale_fill_gradient2(low = "#3F5A6B", mid = "white", high = "#B0344D",
                       midpoint = 0, name = "NES") +
  labs(x = NULL, y = NULL, title = "Project gene sets across RNA populations") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        panel.grid = element_blank())
```

```{r}
#| label: mechanism-go-table
go_whole_int <- go[go$population == "whole_microglia" & go$contrast == "interaction", , drop = FALSE]
go_whole_int <- go_whole_int[order(go_whole_int$padj, -abs(go_whole_int$NES), go_whole_int$pathway,
                                   method = "radix", na.last = TRUE), , drop = FALSE]
go_tab <- head(go_whole_int, 8L)
go_tab <- data.frame(
  collection = go_tab$collection,
  pathway = gsub("^GO[A-Z]*_", "", go_tab$pathway),
  NES = fmt_num(go_tab$NES),
  FDR = fmt_p(go_tab$padj),
  size = go_tab$size,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(go_tab, row.names = FALSE,
             caption = "Top GO rows for the whole-microglia tau-amyloid interaction. These are survey rows from the bounded fgsea layer, not a separate mechanism claim.")
```

```{r}
#| label: fig-mechanism-go-dotplot
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Compact GO survey from the bounded fgsea layer. Each population-contrast panel contributes its top two GO rows; colour is NES, size is gene-set size, and filled points mark FDR < 0.10. These rows are exploratory context, not a new mechanism call."
ggplot(go_select, aes(contrast, pathway_label)) +
  geom_point(aes(size = size, colour = NES, alpha = supported)) +
  facet_wrap(~ population_label, nrow = 1, scales = "free_y") +
  scale_colour_gradient2(low = "#3F5A6B", mid = "grey88", high = "#B0344D",
                         midpoint = 0, name = "NES") +
  scale_alpha_manual(values = c(`FALSE` = 0.45, `TRUE` = 1), guide = "none") +
  scale_size_continuous(range = c(1.4, 4.2), name = "set size") +
  labs(x = NULL, y = NULL, title = "Top GO rows by population and contrast") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        panel.grid.major.y = element_blank())
```

## TF Activity

The strongest TF result is Myc repression in the interaction. In whole microglia, Myc has
ULM score `r fmt_num(myc_whole$score)` (FDR `r fmt_p(myc_whole$fdr)`); in the DAM substate it is
also negative (score `r fmt_num(myc_dam$score)`, FDR `r fmt_p(myc_dam$fdr)`). This is the
cleanest rebuilt support for the v1 Myc mechanism thread. It should still be read as a
regulatory-activity estimate from static pseudobulk RNA ranks, not as a perturbation experiment.

```{r}
#| label: mechanism-tf-interaction
#| fig-width: 9.5
#| fig-height: 4.8
#| fig-cap: "Top TF activities for the tau-amyloid interaction. Scores are decoupleR ULM activities over CollecTRI, FDR-adjusted within population and contrast. Myc is highlighted because it is a predeclared v1 mechanism candidate and is the strongest rebuilt TF signal."
tf_int <- tf[tf$contrast == "interaction" &
               tf$population %in% c("whole_microglia", "DAM") &
               tf$selection == "top", , drop = FALSE]
tf_int <- tf_int[order(tf_int$population, tf_int$fdr, -abs(tf_int$score), tf_int$source,
                       method = "radix", na.last = TRUE), , drop = FALSE]
tf_int <- do.call(rbind, lapply(split(tf_int, tf_int$population), head, n = 8L))
tf_int$population <- factor(tf_int$population, levels = c("whole_microglia", "DAM"),
                            labels = c("whole microglia", "DAM"))
tf_int$is_myc <- tf_int$source == "Myc"

ggplot(tf_int, aes(score, reorder(source, score), colour = is_myc)) +
  geom_vline(xintercept = 0, colour = "grey75") +
  geom_point(size = 2.2) +
  facet_wrap(~ population, scales = "free_y") +
  scale_colour_manual(values = c(`FALSE` = "grey35", `TRUE` = "#B0344D"), guide = "none") +
  labs(x = "ULM activity score", y = NULL, title = "Interaction TF activity")
```

```{r}
#| label: fig-mechanism-tf-lollipop
#| fig-width: 10
#| fig-height: 5.8
#| fig-cap: "Focused TF candidates across the interaction and tau-in-NLGF contrasts. Myc is consistently negative and supported, while NF-kB-family members do not form a concordant negative attenuation pattern. Filled points mark FDR < 0.10 within population and contrast."
ggplot(tf_focus, aes(score, source, colour = class)) +
  geom_vline(xintercept = 0, colour = "grey75") +
  geom_segment(aes(x = 0, xend = score, yend = source), linewidth = 0.45) +
  geom_point(aes(shape = supported), size = 2.2) +
  facet_grid(population_label ~ contrast) +
  scale_colour_manual(values = c(Myc = "#B0344D", `NF-kB family` = "#3F5A6B"),
                      name = NULL) +
  scale_shape_manual(values = c(`FALSE` = 1, `TRUE` = 16), name = "FDR < 0.10") +
  labs(x = "ULM activity score", y = NULL,
       title = "Myc and NF-kB-family TF activities") +
  theme(legend.position = "bottom")
```

## NF-kB Attenuation

`r nf_sentence` `r nf_detail_sentence` `r nf_conclusion_sentence` Supportive rows, including
tau-in-NLGF and substates, remain descriptive and cannot override the whole-microglia interaction
gate.

```{r}
#| label: mechanism-nfkb-table
nf_tab <- nf[, c("population", "contrast", "test", "score", "p_value", "primary_family_fdr",
                 "detail", "primary_test", "supportive_only"), drop = FALSE]
nf_tab$score <- fmt_num(nf_tab$score)
nf_tab$p_value <- fmt_p(nf_tab$p_value)
nf_tab$primary_family_fdr <- fmt_p(nf_tab$primary_family_fdr)
names(nf_tab) <- c("population", "contrast", "test", "score", "p", "primary FDR",
                   "detail", "primary", "supportive only")
knitr::kable(nf_tab, row.names = FALSE,
             caption = "NF-kB attenuation gate. Only the two whole-microglia interaction rows are primary; all other rows are supportive-only.")
```

```{r}
#| label: fig-mechanism-nfkb-discordance
#| fig-width: 7.2
#| fig-height: 4.2
#| fig-cap: "NF-kB attenuation verdict from the guarded mechanism target. The two whole-microglia interaction rows are primary, and the plot title prints the current target status. Tau-in-NLGF rows are supportive-only and cannot override the primary gate."
ggplot(nf_plot, aes(contrast, test_label, fill = score)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_tile(data = nf_plot[nf_plot$primary_test, , drop = FALSE],
            fill = NA, colour = "#B0344D", linewidth = 1) +
  geom_text(aes(label = tile_label), size = 2.7, lineheight = 0.9) +
  scale_fill_gradient2(low = "#3F5A6B", mid = "white", high = "#B0344D",
                       midpoint = 0, name = "score") +
  labs(x = NULL, y = NULL,
       title = paste("NF-kB attenuation gate:", nf_status)) +
  theme(panel.grid = element_blank())
```

## Gsk3b And Kinase Activity

`r gsk_sentence` The KSN layer has adequate coverage for the specific candidate:
`r mrd$kinase$coverage$gsk3b$matched_sites` Gsk3b substrate sites pass the min-size rule, from
`r format(mrd$kinase$coverage$n_matched_sites, big.mark = ",")` matched KSN sites and
`r mrd$kinase$coverage$kinases_passing_minsize` kinases passing min-size overall. But the rebuilt
Gsk3b interaction score is `r fmt_num(gsk_int$score)` (FDR `r fmt_p(gsk_int$fdr)`) and the
tau-in-NLGF score is `r fmt_num(gsk_tau_nlgf$score)` (FDR `r fmt_p(gsk_tau_nlgf$fdr)`), so the
v1 Gsk3b interaction headline is not recovered here.

The kinase rows that do pass the primary FDR threshold are concentrated in tau-in-NLGF
(`r nrow(sig_tau)` non-Gsk3b kinases at FDR < `r alpha`). None are supported by the additive
run-index sensitivity (`r run_supported_tau` retain FDR support with concordant sign), so the
bulk phosphosite layer is hypothesis-generating. It is 24-month bulk hippocampus
phosphoproteomics, not microglia-sorted, and its 16 runs are genotype-blocked, so it should not
be used as a stand-alone microglial mechanism claim.

```{r}
#| label: mechanism-kinase-table
kin_tab <- kin[kin$contrast %in% c("interaction", "tau_in_nlgf") &
                 (kin$source == "Gsk3b" | kin$significant %in% TRUE), , drop = FALSE]
kin_tab <- kin_tab[order(match(kin_tab$contrast, c("interaction", "tau_in_nlgf")),
                         kin_tab$source != "Gsk3b", kin_tab$fdr, -abs(kin_tab$score),
                         kin_tab$source, method = "radix", na.last = TRUE), , drop = FALSE]
kin_tab <- data.frame(
  kinase = kin_tab$source,
  contrast = kin_tab$contrast,
  score = fmt_num(kin_tab$score),
  FDR = fmt_p(kin_tab$fdr),
  `run-index score` = fmt_num(kin_tab$run_index_score),
  `run-index FDR` = fmt_p(kin_tab$run_index_fdr),
  `run-index supports` = kin_tab$run_index_supports,
  reason = kin_tab$include_reason,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(kin_tab, row.names = FALSE,
             caption = "Focal kinase rows. Gsk3b is carried whether significant or not; other rows are primary FDR < 0.10. Run-index columns are an additive sensitivity for genotype-blocked acquisition order.")
```

```{r}
#| label: fig-mechanism-kinase-heatmap
#| fig-width: 9.5
#| fig-height: 6.2
#| fig-cap: "Kinase activity rows from the 24-month bulk hippocampus phosphosite layer. Gsk3b is carried in every contrast whether significant or not; other kinases appear because they pass primary FDR < 0.10. White circles mark primary support and crosses mark additive run-index support when present."
ggplot(kin_plot, aes(contrast, source, fill = score)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  geom_point(data = kin_plot[kin_plot$primary_sig, , drop = FALSE],
             shape = 21, size = 2.1, stroke = 0.35, fill = "white", colour = "grey20") +
  geom_point(data = kin_plot[kin_plot$run_supported, , drop = FALSE],
             shape = 4, size = 2.6, stroke = 0.8, colour = "black") +
  scale_fill_gradient2(low = "#3F5A6B", mid = "white", high = "#B0344D",
                       midpoint = 0, name = "ULM score") +
  labs(x = NULL, y = NULL,
       title = "Kinase activity and run-index sensitivity") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        panel.grid = element_blank())
```

## Synthesis

The rebuilt mechanism evidence is asymmetric. The RNA layer strongly supports a Myc-linked,
DAM-centred interaction signal, and the pathway layer ties that signal back to the compositional
state expansion established in P1/P2. The targeted NF-kB attenuation check is discordant rather
than supported, and the 24-month bulk phosphosite kinase layer does not recover a significant
Gsk3b interaction. P3 therefore strengthens the DAM/Myc mechanism hypothesis while downgrading
Gsk3b and NF-kB attenuation to unresolved or unsupported within the rebuilt data. The closing
synthesis treats those negatives as part of the result, while the cross-modality chapter audits
whether spatial, total-proteome and broader phosphoproteome views add corroborating context.
