# Mechanism: Myc signal, NF-kB check, and bulk kinase support {#sec-mechanism}

P3 reframes the mechanism question around the rebuilt P1/P2 result. Amyloid drives a
microglial homeostatic-to-DAM programme, and mutant tau modulates that amyloid response
mainly by expanding the DAM compartment, not by a supported further advance along the
activation ordering. The analyses below ask which RNA pathway, TF and kinase signals
plausibly sit under that state expansion. They are treated as tests, not as assumed winners:
Myc and Gsk3b can be supported or not by the rebuilt data, and NF-kB attenuation is only
called if its predeclared transcript-level gate clears.

```{r}
#| label: mechanism-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)
targets::tar_source()
targets::tar_load(mechanism_report)                 # ONE compact target, no heavy Seurat object
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

mrd <- mechanism_report
pp  <- mrd$pathway_project
go  <- mrd$pathway_go_top
tf  <- mrd$tf_highlights
nf  <- mrd$nfkb$table
kin <- mrd$kinase$table
comp <- mrd$composition_anchor
traj <- mrd$trajectory_anchor
alpha <- mrd$provenance$alpha

fmt_p <- function(p) {
  out <- rep("-", length(p))
  ok <- is.finite(p)
  out[ok] <- ifelse(p[ok] < 0.001,
                    formatC(p[ok], format = "e", digits = 2),
                    formatC(p[ok], format = "f", digits = 3))
  out
}
fmt_num <- function(x, digits = 2) ifelse(is.finite(x), formatC(x, format = "f", digits = digits), "-")
one_row <- function(d, idx, label) {
  out <- d[idx, , drop = FALSE]
  stopifnot(nrow(out) >= 1L)
  out[1, , drop = FALSE]
}

myc_whole <- one_row(tf, tf$population == "whole_microglia" & tf$contrast == "interaction" &
                       tf$source == "Myc", "whole Myc interaction")
myc_dam <- one_row(tf, tf$population == "DAM" & tf$contrast == "interaction" &
                     tf$source == "Myc", "DAM Myc interaction")
dam_int <- one_row(comp, comp$contrast == "interaction" & comp$substate == "DAM",
                   "DAM composition interaction")
traj_comp <- one_row(traj, traj$measure == "comp_cf", "trajectory composition")
traj_prog <- one_row(traj, traj$measure == "progression_cf", "trajectory progression")
proj_dam_int <- one_row(pp, pp$population == "whole_microglia" & pp$contrast == "interaction" &
                          pp$pathway == "DAM", "project DAM interaction")
gsk_int <- one_row(kin, kin$source == "Gsk3b" & kin$contrast == "interaction",
                   "Gsk3b interaction")
gsk_tau_nlgf <- one_row(kin, kin$source == "Gsk3b" & kin$contrast == "tau_in_nlgf",
                        "Gsk3b tau_in_nlgf")
nf_primary <- nf[nf$primary_test %in% TRUE, , drop = FALSE]
stopifnot(nrow(nf_primary) == 2L)
nf_tf <- one_row(nf_primary, nf_primary$test == "tf_family", "NF-kB TF primary")
nf_gsea <- one_row(nf_primary, nf_primary$test == "target_gsea", "NF-kB target primary")
nf_status <- mrd$nfkb$verdict$status
nf_sentence <- if (identical(nf_status, "supported")) {
  "The predeclared NF-kB attenuation gate is supported."
} else if (identical(nf_status, "discordant")) {
  "The predeclared NF-kB attenuation gate is discordant, so P3 does not support the attenuation claim."
} else if (identical(nf_status, "directional_only")) {
  "The predeclared NF-kB attenuation gate is directionally negative but does not clear its FDR rule."
} else {
  "The predeclared NF-kB attenuation gate is not supported."
}
nf_detail_sentence <- if (isTRUE(mrd$nfkb$verdict$primary_discordant)) {
  paste0("The two primary rows disagree in sign: the NF-kB TF-family score is ",
         fmt_num(nf_tf$score), " (primary-family FDR ", fmt_p(nf_tf$primary_family_fdr),
         "), while the sign-aware NF-kB target-GSEA score is ", fmt_num(nf_gsea$score),
         " (primary-family FDR ", fmt_p(nf_gsea$primary_family_fdr), ").")
} else if (isTRUE(mrd$nfkb$verdict$primary_negative)) {
  paste0("Both primary rows are negative: the NF-kB TF-family score is ",
         fmt_num(nf_tf$score), " (primary-family FDR ", fmt_p(nf_tf$primary_family_fdr),
         ") and the sign-aware NF-kB target-GSEA score is ", fmt_num(nf_gsea$score),
         " (primary-family FDR ", fmt_p(nf_gsea$primary_family_fdr), ").")
} else {
  paste0("The primary rows are not concordantly negative: the NF-kB TF-family score is ",
         fmt_num(nf_tf$score), " (primary-family FDR ", fmt_p(nf_tf$primary_family_fdr),
         ") and the sign-aware NF-kB target-GSEA score is ", fmt_num(nf_gsea$score),
         " (primary-family FDR ", fmt_p(nf_gsea$primary_family_fdr), ").")
}
nf_conclusion_sentence <- if (identical(nf_status, "supported")) {
  "That clears the specific attenuation rule."
} else if (identical(nf_status, "directional_only")) {
  "That leaves the evidence directional but below the predeclared FDR rule."
} else {
  "That makes the rebuilt evidence insufficient for the specific attenuation claim."
}
gsk_sentence <- if (isTRUE(gsk_int$significant) || isTRUE(gsk_tau_nlgf$significant)) {
  "Gsk3b has nominal rebuilt kinase support in at least one focal contrast, subject to the bulk-phospho caveats below."
} else {
  "Gsk3b is covered by the kinase prior and phosphosite data, but it is not significant in either focal rebuilt contrast."
}
sig_tau <- kin[kin$contrast == "tau_in_nlgf" & kin$significant %in% TRUE &
                 kin$source != "Gsk3b", , drop = FALSE]
sig_tau <- sig_tau[order(sig_tau$fdr, -abs(sig_tau$score), sig_tau$source,
                         method = "radix", na.last = TRUE), , drop = FALSE]
run_supported_tau <- sum(sig_tau$run_index_supports %in% TRUE)

contrast_labels <- c(tau_alone = "tau alone",
                     nlgf_in_maptki = "amyloid on MAPTKI",
                     nlgf_in_p301s = "amyloid on P301S",
                     tau_in_nlgf = "tau in NLGF",
                     interaction = "interaction")
project_labels <- c(DAM = "DAM",
                    Homeostatic = "homeostatic",
                    IFN = "IFN",
                    MHC_APC = "MHC/APC",
                    NFkB_Activated_Targets = "NF-kB activated targets",
                    NFkB_CollecTRI_Targets = "NF-kB target union",
                    NFkB_Repressed_Targets = "NF-kB repressed targets")
```

## Pathway Survey

The project gene sets recapitulate the upstream story in RNA. The whole-microglia DAM set is
positive in the tau-amyloid interaction (NES `r fmt_num(proj_dam_int$NES)`, FDR
`r fmt_p(proj_dam_int$padj)`), aligning with the P1 DAM-composition interaction (DAM propeller
ratio `r fmt_num(dam_int$prop_ratio)`, FDR `r fmt_p(dam_int$fdr_global)`) and the P2
composition channel (effect `r fmt_num(traj_comp$coef)`, FDR `r fmt_p(traj_comp$fdr)`). The
progression channel remains non-significant (effect `r fmt_num(traj_prog$coef)`, FDR
`r fmt_p(traj_prog$fdr)`), so the mechanism target is DAM-cell expansion and amyloid-response
modulation, not proven further activation of cells already on the trajectory.

```{r}
#| label: mechanism-project-sets
#| fig-width: 9.5
#| fig-height: 4.8
#| fig-cap: "Focused project gene sets in whole microglia. Tile colour is fgsea NES over the pseudobulk t-statistic ranks; outlined points mark FDR < 0.10 within the project collection and contrast. The DAM set stays positive in the interaction, while the NF-kB target arms are read through the formal gate below."
proj_plot <- pp[pp$population == "whole_microglia", , drop = FALSE]
proj_plot$contrast <- factor(proj_plot$contrast, levels = names(contrast_labels),
                             labels = unname(contrast_labels))
proj_plot$pathway_label <- project_labels[proj_plot$pathway]
proj_plot$pathway_label[is.na(proj_plot$pathway_label)] <- proj_plot$pathway[is.na(proj_plot$pathway_label)]
proj_plot$pathway_label <- factor(proj_plot$pathway_label, levels = rev(unique(project_labels[project_labels %in% proj_plot$pathway_label])))

ggplot(proj_plot, aes(contrast, pathway_label, fill = NES)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_point(data = proj_plot[proj_plot$padj < alpha, , drop = FALSE],
             shape = 21, size = 2, stroke = 0.3, fill = "white", colour = "grey20") +
  scale_fill_gradient2(low = "#3F5A6B", mid = "white", high = "#B0344D", midpoint = 0,
                       name = "NES") +
  labs(x = NULL, y = NULL, title = "Project gene-set activity")
```

```{r}
#| label: mechanism-go-table
go_whole_int <- go[go$population == "whole_microglia" & go$contrast == "interaction", , drop = FALSE]
go_whole_int <- go_whole_int[order(go_whole_int$padj, -abs(go_whole_int$NES), go_whole_int$pathway,
                                   method = "radix", na.last = TRUE), , drop = FALSE]
go_tab <- head(go_whole_int, 8L)
go_tab <- data.frame(
  collection = go_tab$collection,
  pathway = gsub("^GO[A-Z]*_", "", go_tab$pathway),
  NES = fmt_num(go_tab$NES),
  FDR = fmt_p(go_tab$padj),
  size = go_tab$size,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(go_tab, row.names = FALSE,
             caption = "Top GO rows for the whole-microglia tau-amyloid interaction. These are survey rows from the bounded fgsea layer, not a separate mechanism claim.")
```

## TF Activity

The strongest TF result is Myc repression in the interaction. In whole microglia, Myc has
ULM score `r fmt_num(myc_whole$score)` (FDR `r fmt_p(myc_whole$fdr)`); in the DAM substate it is
also negative (score `r fmt_num(myc_dam$score)`, FDR `r fmt_p(myc_dam$fdr)`). This is the
cleanest rebuilt support for the v1 Myc mechanism thread. It should still be read as a
regulatory-activity estimate from static pseudobulk RNA ranks, not as a perturbation experiment.

```{r}
#| label: mechanism-tf-interaction
#| fig-width: 9.5
#| fig-height: 4.8
#| fig-cap: "Top TF activities for the tau-amyloid interaction. Scores are decoupleR ULM activities over CollecTRI, FDR-adjusted within population and contrast. Myc is highlighted because it is a predeclared v1 mechanism candidate and is the strongest rebuilt TF signal."
tf_int <- tf[tf$contrast == "interaction" &
               tf$population %in% c("whole_microglia", "DAM") &
               tf$selection == "top", , drop = FALSE]
tf_int <- tf_int[order(tf_int$population, tf_int$fdr, -abs(tf_int$score), tf_int$source,
                       method = "radix", na.last = TRUE), , drop = FALSE]
tf_int <- do.call(rbind, lapply(split(tf_int, tf_int$population), head, n = 8L))
tf_int$population <- factor(tf_int$population, levels = c("whole_microglia", "DAM"),
                            labels = c("whole microglia", "DAM"))
tf_int$is_myc <- tf_int$source == "Myc"

ggplot(tf_int, aes(score, reorder(source, score), colour = is_myc)) +
  geom_vline(xintercept = 0, colour = "grey75") +
  geom_point(size = 2.2) +
  facet_wrap(~ population, scales = "free_y") +
  scale_colour_manual(values = c(`FALSE` = "grey35", `TRUE` = "#B0344D"), guide = "none") +
  labs(x = "ULM activity score", y = NULL, title = "Interaction TF activity")
```

## NF-kB Attenuation

`r nf_sentence` `r nf_detail_sentence` `r nf_conclusion_sentence` Supportive rows, including
tau-in-NLGF and substates, remain descriptive and cannot override the whole-microglia interaction
gate.

```{r}
#| label: mechanism-nfkb-table
nf_tab <- nf[, c("population", "contrast", "test", "score", "p_value", "primary_family_fdr",
                 "detail", "primary_test", "supportive_only"), drop = FALSE]
nf_tab$score <- fmt_num(nf_tab$score)
nf_tab$p_value <- fmt_p(nf_tab$p_value)
nf_tab$primary_family_fdr <- fmt_p(nf_tab$primary_family_fdr)
names(nf_tab) <- c("population", "contrast", "test", "score", "p", "primary FDR",
                   "detail", "primary", "supportive only")
knitr::kable(nf_tab, row.names = FALSE,
             caption = "NF-kB attenuation gate. Only the two whole-microglia interaction rows are primary; all other rows are supportive-only.")
```

## Gsk3b And Kinase Activity

`r gsk_sentence` The KSN layer has adequate coverage for the specific candidate:
`r mrd$kinase$coverage$gsk3b$matched_sites` Gsk3b substrate sites pass the min-size rule, from
`r format(mrd$kinase$coverage$n_matched_sites, big.mark = ",")` matched KSN sites and
`r mrd$kinase$coverage$kinases_passing_minsize` kinases passing min-size overall. But the rebuilt
Gsk3b interaction score is `r fmt_num(gsk_int$score)` (FDR `r fmt_p(gsk_int$fdr)`) and the
tau-in-NLGF score is `r fmt_num(gsk_tau_nlgf$score)` (FDR `r fmt_p(gsk_tau_nlgf$fdr)`), so the
v1 Gsk3b interaction headline is not recovered here.

The kinase rows that do pass the primary FDR threshold are concentrated in tau-in-NLGF
(`r nrow(sig_tau)` non-Gsk3b kinases at FDR < `r alpha`). None are supported by the additive
run-index sensitivity (`r run_supported_tau` retain FDR support with concordant sign), so the
bulk phosphosite layer is hypothesis-generating. It is 24-month bulk hippocampus
phosphoproteomics, not microglia-sorted, and its 16 runs are genotype-blocked, so it should not
be used as a stand-alone microglial mechanism claim.

```{r}
#| label: mechanism-kinase-table
kin_tab <- kin[kin$contrast %in% c("interaction", "tau_in_nlgf") &
                 (kin$source == "Gsk3b" | kin$significant %in% TRUE), , drop = FALSE]
kin_tab <- kin_tab[order(match(kin_tab$contrast, c("interaction", "tau_in_nlgf")),
                         kin_tab$source != "Gsk3b", kin_tab$fdr, -abs(kin_tab$score),
                         kin_tab$source, method = "radix", na.last = TRUE), , drop = FALSE]
kin_tab <- data.frame(
  kinase = kin_tab$source,
  contrast = kin_tab$contrast,
  score = fmt_num(kin_tab$score),
  FDR = fmt_p(kin_tab$fdr),
  `run-index score` = fmt_num(kin_tab$run_index_score),
  `run-index FDR` = fmt_p(kin_tab$run_index_fdr),
  `run-index supports` = kin_tab$run_index_supports,
  reason = kin_tab$include_reason,
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE
)
knitr::kable(kin_tab, row.names = FALSE,
             caption = "Focal kinase rows. Gsk3b is carried whether significant or not; other rows are primary FDR < 0.10. Run-index columns are an additive sensitivity for genotype-blocked acquisition order.")
```

## Synthesis

The rebuilt mechanism evidence is asymmetric. The RNA layer strongly supports a Myc-linked,
DAM-centred interaction signal, and the pathway layer ties that signal back to the compositional
state expansion established in P1/P2. The targeted NF-kB attenuation check is discordant rather
than supported, and the 24-month bulk phosphosite kinase layer does not recover a significant
Gsk3b interaction. P3 therefore strengthens the DAM/Myc mechanism hypothesis while downgrading
Gsk3b and NF-kB attenuation to unresolved or unsupported within the rebuilt data. The closing
synthesis treats those negatives as part of the result, while the cross-modality chapter audits
whether spatial, total-proteome and broader phosphoproteome views add corroborating context.
