# Microglia: amyloid drives a homeostatic-to-DAM programme {#sec-microglia}

This chapter is the single-nucleus core of the study. From the reprocessed, batch-integrated
microglia (SCT normalisation, Harmony over batch only) it places each nucleus into a canonical
substate, tests whether substate composition shifts across the genotype contrasts, and runs
pseudobulk differential expression to characterise the activation programme and to weigh the
tau-by-amyloid interaction. The headline, supported three complementary ways below, is that
amyloid (NLGF) drives microglia out of a homeostatic state into disease-associated microglia
(DAM); mutant tau scales the size of that response without, on its own, redrawing the
programme. One measurement caveat frames everything that follows: single-nucleus capture
under-detects a sizeable share of DAM activation transcripts (Thrupp and colleagues, 2020), so
DAM is read as a scored programme rather than thresholded gene by gene, and every differential
test runs on raw counts.

```{r}
#| label: microglia-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)                                    # promote any chunk warning to a render error (gate-critical)
targets::tar_source()                                # constants, marker sets, plot theme + genotype scales
targets::tar_load(c(microglia_report, composition_results, pb_de_microglia,
                    pb_de_substate, symbol_map,
                    microglia_figures))              # all compact targets -> render stays cheap (no 612MB Seurat)
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

mrd  <- microglia_report
cf   <- mrd$cell_frame
comp <- composition_results
pbm  <- pb_de_microglia
pbs  <- pb_de_substate
mf   <- microglia_figures

# substate colours reuse the genotype palette hues -> cohesive with the report identity.
sub_cols <- c(Homeostatic = "#7FA8C1", DAM = "#B0344D", IFN = "#E08754",
              Proliferative = "#6BAF7F", ambiguous = "grey70", unassigned = "grey85")
present_sub <- intersect(names(sub_cols), as.character(unique(cf$substate)))
cf$substate <- factor(as.character(cf$substate), levels = present_sub)
ctr_levels <- c("tau_alone", "nlgf_in_maptki", "nlgf_in_p301s", "tau_in_nlgf", "interaction")
ctr_labels <- c(
  tau_alone = "tau alone",
  nlgf_in_maptki = "amyloid on MAPTKI",
  nlgf_in_p301s = "amyloid on P301S",
  tau_in_nlgf = "tau in NLGF",
  interaction = "tau x amyloid"
)
score_cols <- c(Homeostatic = "#3F5A6B", DAM = "#B0344D", MHC_APC = "#6B7A3F")
score_labels <- c(Homeostatic = "homeostatic", DAM = "DAM", MHC_APC = "MHC-APC")
fmt_contrast <- function(x, rev = FALSE) {
  lv <- if (rev) rev(ctr_levels) else ctr_levels
  factor(x, levels = lv, labels = unname(ctr_labels[lv]))
}

# genotype x substate counts (primary cluster-level labels) -> the compositional shift.
st       <- mrd$provenance$substate_table
dam_pct  <- round(100 * st[, "DAM"] / rowSums(st))               # % DAM per genotype
homeo_pct <- round(100 * st[, "Homeostatic"] / rowSums(st))      # % Homeostatic per genotype

# ensembl -> symbol lookup for DE gene labels (symbol_map rownames are ensembl).
to_symbol <- function(ens) {
  s <- symbol_map$symbol[match(ens, symbol_map$ensembl)]
  ifelse(is.na(s) | s == "", ens, s)
}
fmt_p <- function(p) formatC(p, format = "e", digits = 2)         # compact scientific p-values for tables

# DAM interaction across the two methods -- computed (not hardcoded) so the prose tracks the
# actual cached build, sccomp being run-to-run variable on its few divergent transitions.
int_prop_fdr <- comp$headline$fdr_global[comp$headline$contrast == "interaction"]
int_sc <- if (!is.null(comp$sccomp))
  comp$sccomp[comp$sccomp$contrast == "interaction" & comp$sccomp$substate == "DAM", ] else NULL
int_sentence <- if (is.null(int_sc))
  "it was not re-tested under sccomp on this build (propeller-only)" else
  sprintf("under sccomp it sits near the significance boundary (FDR %s, 95%% credible interval [%.2f, %.2f])",
          fmt_p(int_sc$c_fdr), int_sc$c_lower, int_sc$c_upper)

# Interaction stageR genes carry REAL fold-changes (every one clears the |log2FC| effect bar; what
# they miss is the stricter STANDALONE per-contrast FDR) -> describe them as sub-threshold-per-
# contrast, never "small effect". Compute the descriptors so the prose tracks the cached fit.
lfc <- pbm$thresholds$lfc; fdr <- pbm$thresholds$fdr
int_conf <- pbm$stageR$stage_padj[, "interaction"] <= pbm$stageR$alpha
int_conf[is.na(int_conf)] <- FALSE
int_tt   <- pbm$top$interaction[match(rownames(pbm$stageR$stage_padj)[int_conf],
                                      pbm$top$interaction$gene), ]
n_int_stageR <- nrow(int_tt)
med_int_lfc  <- median(abs(int_tt$logFC))      # typical confirmed-gene effect size (log2 units)
min_int_adjp <- min(int_tt$adj.P.Val)          # most significant interaction gene, standalone per-contrast
# Within-DAM amyloid DE + which substates were fit -> let the prose track the fit, not a hardcode.
dam_fit <- pbs$per_substate$DAM
dam_amyloid_de <- if (identical(dam_fit$status, "fit"))
  sum(abs(dam_fit$top$nlgf_in_p301s$logFC) > lfc & dam_fit$top$nlgf_in_p301s$adj.P.Val < fdr) else NA_integer_
fit_subs <- names(Filter(function(p) identical(p$status, "fit"), pbs$per_substate))

# Back the "no dropped cluster was DAM-high" caveat with the recorded per-cluster DAM medians
# (S2 prune provenance) rather than asserting it; report the genotype/cluster dropout factually.
# Dropped clusters' DAM medians OVERLAP the kept homeostatic clusters -> the honest comparison is
# against the DAM-ENRICHED ceiling (kept_dam_max): no dropped cluster reaches the DAM compartment.
qcr      <- mrd$prune$qc_rationale
dbg      <- mrd$prune$dropped_by_genotype
drp_by_geno <- rowSums(dbg)                                       # absolute nuclei dropped per genotype
drop_frac   <- drp_by_geno / (drp_by_geno + as.integer(table(cf$genotype)[names(drp_by_geno)]))
drop_geno   <- names(which.max(drop_frac))                       # genotype losing the largest FRACTION (here also the largest count)
drop_geno_pct <- round(100 * max(drop_frac))                     # that fraction, as a percent
drop_clu    <- names(which.max(colSums(dbg)))                    # cluster contributing the most nuclei (absolute)
drop_dam_max <- max(qcr[as.character(mrd$prune$dropped), "DAM_med"])                            # top DAM median, dropped
kept_dam_max <- max(qcr[setdiff(rownames(qcr), as.character(mrd$prune$dropped)), "DAM_med"])    # DAM-enriched ceiling, kept

mf_umap <- mf$umap_by_substate
mf_umap$substate <- factor(as.character(mf_umap$substate), levels = present_sub)
mf_umap$genotype <- factor(as.character(mf_umap$genotype), levels = genotype_levels)
mf_score <- mf$score_triptych
mf_score$score_name <- factor(mf_score$score_name, levels = names(score_labels),
                              labels = unname(score_labels))
mf_unit <- mf$unit_composition
mf_unit$substate <- factor(as.character(mf_unit$substate), levels = present_sub)
mf_unit$genotype <- factor(as.character(mf_unit$genotype), levels = genotype_levels)
mf_unit$batch <- factor(mf_unit$batch, levels = sort(unique(mf_unit$batch), method = "radix"))
mf_dist <- mf$score_distribution
mf_dist <- mf_dist[as.character(mf_dist$substate) %in% present_sub &
                     as.character(mf_dist$score_name) %in% names(score_labels), , drop = FALSE]
mf_dist$substate <- factor(as.character(mf_dist$substate), levels = present_sub)
mf_dist$genotype <- factor(as.character(mf_dist$genotype), levels = genotype_levels)
mf_dist$score_name <- factor(as.character(mf_dist$score_name), levels = names(score_labels),
                             labels = unname(score_labels))
mf_conc <- mf$composition_concordance
mf_conc$contrast <- fmt_contrast(mf_conc$contrast)
mf_conc$substate <- factor(mf_conc$substate, levels = present_sub)
mf_conc$status <- ifelse(
  mf_conc$flag, "method flag",
  ifelse(mf_conc$sig_logit,
         ifelse(mf_conc$dir_logit > 0, "primary up", "primary down"),
         "not significant")
)
mf_conc$status <- factor(mf_conc$status,
                         levels = c("primary up", "primary down", "not significant", "method flag"))
mf_conc$label <- ifelse(mf_conc$flag, "flag",
                        ifelse(mf_conc$sig_logit, "FDR", ""))
mf_wv_bins <- mf$whole_de_volcano$bins
mf_wv_bins$contrast <- fmt_contrast(mf_wv_bins$contrast)
mf_wv_bins$log_n <- log10(mf_wv_bins$n + 1)
mf_wv_lab <- mf$whole_de_volcano$labels
mf_wv_lab$contrast <- fmt_contrast(mf_wv_lab$contrast)
mf_audit_counts <- mf$substate_fit_audit$cell_counts
audit_min <- stats::aggregate(n_cells ~ substate, data = mf_audit_counts, FUN = min)
names(audit_min)[2] <- "min_cells_unit"
audit_total <- stats::aggregate(n_cells ~ substate, data = mf_audit_counts, FUN = sum)
names(audit_total)[2] <- "n_cells"
audit_status <- unique(mf$substate_fit_audit$fit_status[c("substate", "status", "reason")])
mf_audit <- Reduce(function(x, y) merge(x, y, by = "substate", all = TRUE, sort = FALSE),
                   list(audit_min, audit_total, audit_status))
mf_audit$substate <- factor(mf_audit$substate, levels = names(sub_cols))
mf_audit$status <- factor(mf_audit$status, levels = c("fit", "skipped"))
mf_ss_counts <- mf$substate_de_volcano$counts
mf_ss_counts <- mf_ss_counts[mf_ss_counts$substate %in% c("Homeostatic", "DAM"), , drop = FALSE]
mf_ss_counts$contrast <- fmt_contrast(mf_ss_counts$contrast)
mf_ss_counts$substate <- factor(mf_ss_counts$substate, levels = c("Homeostatic", "DAM"))
mf_ss_signed <- rbind(
  data.frame(mf_ss_counts[c("contrast", "substate")], direction = "up",
             n_genes = mf_ss_counts$n_up_fdr_0_10, stringsAsFactors = FALSE),
  data.frame(mf_ss_counts[c("contrast", "substate")], direction = "down",
             n_genes = -mf_ss_counts$n_down_fdr_0_10, stringsAsFactors = FALSE)
)
mf_ss_signed$direction <- factor(mf_ss_signed$direction, levels = c("down", "up"))
```

## Substate landscape

UCell scores each nucleus against the canonical homeostatic, DAM, interferon-response (IFN) and
proliferative signatures; a per-signature calibration then assigns a primary substate at the
cluster level (the per-cell argmax is noisier and kept only as a diagnostic). After dropping
`r mrd$prune$n_dropped` contaminant or doublet nuclei (clusters
`r paste(mrd$prune$dropped, collapse = ", ")`), `r format(mrd$n_cells, big.mark = ",")` nuclei
remain. They resolve into `r format(sum(cf$substate == "Homeostatic"), big.mark = ",")`
homeostatic, `r format(sum(cf$substate == "DAM"), big.mark = ",")` DAM and
`r sum(cf$substate == "IFN")` IFN nuclei; no cluster is proliferative-dominant, so that level
carries no cells (sparse cycling markers average out at the cluster level rather than forming a
coherent population). Each substate signature is scored on its genes found in the assay
(`r paste(sprintf("%s %s", names(mrd$provenance$n_used), mrd$provenance$n_used)[2:5], collapse = ", ")` genes used).

```{r}
#| label: umap
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Microglia substate landscape on the batch-integrated UMAP. Left: primary substate label. Right: the calibrated DAM activation score, which forms a continuous homeostatic-to-DAM gradient rather than a discrete switch."
p_sub <- ggplot(cf, aes(umap_1, umap_2, colour = substate)) +
  geom_point(size = 0.25, alpha = 0.6) +
  scale_colour_manual(values = sub_cols, breaks = present_sub, name = "substate") +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Primary substate")

p_dam <- ggplot(cf, aes(umap_1, umap_2, colour = DAM_UCell_z)) +
  geom_point(size = 0.25, alpha = 0.6) +
  scale_colour_gradient(low = "grey85", high = "#B0344D", name = "DAM z") +
  labs(x = "UMAP 1", y = "UMAP 2", title = "DAM activation score")

(p_sub | p_dam)
```

```{r}
#| label: fig-microglia-umap-substate
#| fig-width: 9.2
#| fig-height: 6.4
#| fig-cap: "Genotype-faceted substate map on the same batch-integrated UMAP. Amyloid-bearing genotypes occupy much more of the DAM-labelled region, with NLGF_P301S showing the largest DAM compartment. Labels are cluster-primary substates, not a per-cell hard threshold."
ggplot(mf_umap, aes(umap_1, umap_2, colour = substate)) +
  geom_point(size = 0.18, alpha = 0.55) +
  facet_wrap(~ genotype, ncol = 2) +
  scale_colour_manual(values = sub_cols, breaks = present_sub, name = "substate") +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Substates by genotype") +
  theme(panel.grid = element_blank())
```

```{r}
#| label: fig-microglia-score-triptych
#| fig-width: 10
#| fig-height: 3.5
#| fig-cap: "Binned UMAP score maps for the three plotted signatures. Homeostatic and DAM scores form opposing regions of the activation manifold, while the MHC-APC score marks the antigen-presentation arm within the broader DAM programme."
ggplot(mf_score, aes(umap_1, umap_2, colour = score_mean)) +
  geom_point(aes(size = pmin(n, 6)), shape = 15, alpha = 0.85) +
  facet_wrap(~ score_name, nrow = 1) +
  scale_colour_gradient2(low = "#3F5A6B", mid = "grey92", high = "#B0344D",
                         midpoint = 0, name = "mean z") +
  scale_size_continuous(range = c(0.4, 1.4), guide = "none") +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Signature gradients across the microglia manifold") +
  theme(panel.grid = element_blank())
```

## Amyloid expands the DAM compartment

The compositional shift is unambiguous. Without amyloid the microglia are overwhelmingly
homeostatic (DAM `r dam_pct["MAPTKI"]`% in MAPTKI, `r dam_pct["P301S"]`% in P301S); with amyloid
the DAM compartment swells to `r dam_pct["NLGF_MAPTKI"]`% on the tau-knockout background and
`r dam_pct["NLGF_P301S"]`% on the mutant-tau background, with the homeostatic fraction falling
in step (`r homeo_pct["MAPTKI"]`% and `r homeo_pct["P301S"]`% down to
`r homeo_pct["NLGF_MAPTKI"]`% and `r homeo_pct["NLGF_P301S"]`%). The same shift is visible per
nucleus as a rightward move in the DAM activation score, and it is steeper on the mutant-tau
background, the first hint of the tau-by-amyloid interaction tested below.

```{r}
#| label: composition-shift
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Amyloid drives the homeostatic-to-DAM shift. Left: substate proportions per genotype (cluster-level labels). Right: per-nucleus DAM activation score by genotype; the amyloid-bearing genotypes carry the heavier upper tail."
prop_df <- as.data.frame(prop.table(st, 1))                       # genotype, substate, Freq(=within-genotype proportion)
prop_df <- prop_df[prop_df$substate %in% present_sub, ]
prop_df$genotype <- factor(prop_df$genotype, levels = genotype_levels)
prop_df$substate <- factor(as.character(prop_df$substate), levels = present_sub)

p_prop <- ggplot(prop_df, aes(genotype, Freq, fill = substate)) +
  geom_col() +
  scale_fill_manual(values = sub_cols, breaks = present_sub, name = "substate") +
  labs(x = NULL, y = "proportion of microglia", title = "Substate composition") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_score <- ggplot(cf, aes(genotype, DAM_UCell_z, fill = genotype)) +
  geom_violin(scale = "width") +
  scale_fill_genotype(guide = "none") +
  labs(x = NULL, y = "DAM activation score (z)", title = "DAM score per nucleus") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

(p_prop | p_score)
```

```{r}
#| label: substate-table
counts_tab <- as.data.frame.matrix(st[, present_sub, drop = FALSE])
counts_tab <- cbind(genotype = rownames(counts_tab), counts_tab,
                    `DAM %` = dam_pct[rownames(counts_tab)])
knitr::kable(counts_tab, row.names = FALSE,
             caption = "Microglia per genotype and substate (primary cluster-level labels).")
```

```{r}
#| label: fig-microglia-unit-composition
#| fig-width: 9.5
#| fig-height: 4.6
#| fig-cap: "Composition by the 16 genotype-by-batch replicate units. Each bar is one replicate unit, preserving the inference unit behind the composition and pseudobulk tests. The DAM expansion is visible in every amyloid-bearing background rather than being driven by one batch."
ggplot(mf_unit, aes(batch, proportion, fill = substate)) +
  geom_col(width = 0.78) +
  facet_grid(. ~ genotype) +
  scale_fill_manual(values = sub_cols, breaks = present_sub, name = "substate") +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(x = "batch", y = "within-unit proportion", title = "Substate composition at replicate-unit resolution") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid.major.x = element_blank())
```

```{r}
#| label: fig-microglia-score-distribution
#| fig-width: 10
#| fig-height: 6.2
#| fig-cap: "Score distributions by genotype within each primary substate. Amyloid shifts the DAM score upward and the homeostatic score downward while retaining overlap, consistent with a scored activation programme rather than a binary marker threshold."
ggplot(mf_dist, aes(x_mid, n, colour = genotype, group = genotype)) +
  geom_line(linewidth = 0.45, alpha = 0.9) +
  facet_grid(score_name ~ substate, scales = "free_y") +
  scale_colour_genotype(name = "genotype") +
  labs(x = "signature score (z)", y = "nuclei per bin",
       title = "Signature-score distributions across genotype and substate") +
  theme(panel.grid.minor = element_blank())
```

## Composition is tested across the genotype contrasts

Composition is tested formally with propeller (the locked, reproducible primary, on the logit
transform) and cross-checked with sccomp, a Bayesian beta-binomial model run off-lock when its
Stan backend is present. Both amyloid contrasts move DAM strongly up and the homeostatic
fraction strongly down, with very small tail probabilities (table below). The tau-by-amyloid
`interaction` on DAM is positive and significant under propeller
(FDR `r fmt_p(int_prop_fdr)`); `r int_sentence`,
so it is flagged rather than asserted. A static compositional interaction of this size raises a
further question - does mutant tau also push microglia further along the same activation
programme, or does it simply enlarge the DAM endpoint? The trajectory chapter (@sec-trajectory)
takes that up directly and finds the latter: the synergy is carried by compartment size, not by
cells advancing further along the axis.

```{r}
#| label: composition-forest
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Substate composition effects for DAM and homeostatic microglia across the five contrasts (sccomp posterior on the logit proportion with 95% credible intervals where the Bayesian arm ran, otherwise the propeller t-statistic). DAM and homeostatic move as mirror images; both amyloid contrasts are far from zero, the interaction is small."
ctr_levels <- c("tau_alone", "nlgf_in_maptki", "nlgf_in_p301s", "tau_in_nlgf", "interaction")
if (!is.null(comp$sccomp)) {
  fdf <- comp$sccomp[comp$sccomp$substate %in% c("Homeostatic", "DAM"), ]
  fdf$contrast <- factor(fdf$contrast, levels = rev(ctr_levels))
  ggplot(fdf, aes(c_effect, contrast, colour = substate)) +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_pointrange(aes(xmin = c_lower, xmax = c_upper),
                    position = position_dodge(width = 0.6)) +
    scale_colour_manual(values = sub_cols[c("Homeostatic", "DAM")], name = "substate") +
    labs(x = "effect on logit proportion (sccomp c_effect)", y = NULL,
         title = "Composition effects across contrasts")
} else {
  fdf <- comp$propeller_logit[comp$propeller_logit$substate %in% c("Homeostatic", "DAM"), ]
  fdf$contrast <- factor(fdf$contrast, levels = rev(ctr_levels))
  ggplot(fdf, aes(t, contrast, fill = substate)) +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_col(position = position_dodge(width = 0.6), width = 0.6) +
    scale_fill_manual(values = sub_cols[c("Homeostatic", "DAM")], name = "substate") +
    labs(x = "propeller t-statistic", y = NULL,
         title = "Composition effects across contrasts (propeller)")
}
```

```{r}
#| label: composition-table
pl  <- comp$propeller_logit
dam_pl <- pl[pl$substate == "DAM", ]
dam_pl <- dam_pl[match(ctr_levels, dam_pl$contrast), ]
ctab <- data.frame(
  contrast        = ctr_levels,
  `propeller t`   = round(dam_pl$t, 2),
  `propeller FDR` = fmt_p(dam_pl$fdr_global),
  check.names = FALSE, stringsAsFactors = FALSE)
if (!is.null(comp$sccomp)) {
  sc <- comp$sccomp[comp$sccomp$substate == "DAM", ]
  sc <- sc[match(ctr_levels, sc$contrast), ]
  ctab$`sccomp effect` <- round(sc$c_effect, 2)
  ctab$`sccomp 95% CI` <- sprintf("[%.2f, %.2f]", sc$c_lower, sc$c_upper)
  ctab$`sccomp FDR`    <- fmt_p(sc$c_fdr)
}
knitr::kable(ctab, row.names = FALSE,
             caption = "DAM composition across the five contrasts: propeller (primary, batch-fixed) and, where its Stan backend is present, sccomp (cross-check, random batch). Direction is read from the t-statistic / effect, not the marginal proportion ratio.")

flagged <- comp$concordance[comp$concordance$flag, c("contrast", "substate")]
```

Cross-method concordance flags `r nrow(flagged)` of `r nrow(comp$concordance)`
contrast-by-substate cells (`r paste(sprintf("%s/%s", flagged$contrast, flagged$substate), collapse = "; ")`)`r if (nrow(flagged) > 0 && all(flagged$substate == "IFN")) ", all on the sparse IFN compartment, where a few hundred nuclei let the two methods split on sign or significance" else ""`.
`r if (!any(flagged$contrast == "interaction" & flagged$substate == "DAM")) "The DAM interaction is concordant here - both methods call it significant - so its borderline element is the sccomp effect size (above), not the cross-method agreement. " else ""`Per the pre-declared rule the propeller-logit call stands; the
disagreements are reported, never averaged away.

```{r}
#| label: fig-microglia-composition-concordance
#| fig-width: 8.8
#| fig-height: 3.8
#| fig-cap: "Composition method concordance grid. Colour shows the locked propeller-logit direction and significance; text marks cells pre-declared as cross-method flags. The DAM interaction is a primary positive composition call, while sparse IFN cells carry the main method flags."
ggplot(mf_conc, aes(contrast, substate, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 3) +
  scale_fill_manual(
    values = c(`primary up` = "#B0344D", `primary down` = "#3F5A6B",
               `not significant` = "#D8D8D8", `method flag` = "#B7A06A"),
    name = "primary call"
  ) +
  labs(x = NULL, y = NULL, title = "Composition calls are concordant where the DAM conclusion rests") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

## The amyloid-to-DAM expression programme

Pseudobulk differential expression (limma-voom with quality weights, aggregated to the 16
genotype-by-batch replicate units) confirms the activation programme at the transcript level.
Of `r format(pbm$kept, big.mark = ",")` expressed genes, `r format(pbm$stageR$n_screened, big.mark = ",")`
pass the stageR family screen. The canonical DAM markers move up coherently under amyloid: of
the `r pbm$dam_concordance$nlgf_in_p301s$n_markers_in_fit` markers in the fit,
`r round(100 * pbm$dam_concordance$nlgf_in_maptki$frac_up)`% rise on the tau-knockout background
(mean log2 fold-change `r round(pbm$dam_concordance$nlgf_in_maptki$mean_logFC, 2)`) and
`r round(100 * pbm$dam_concordance$nlgf_in_p301s$frac_up)`% on the mutant-tau background
(mean `r round(pbm$dam_concordance$nlgf_in_p301s$mean_logFC, 2)`), reproducing the original
study's direction. The volcano below shows the strongest amyloid contrast with the DAM markers
picked out.

```{r}
#| label: volcano
#| fig-width: 8
#| fig-height: 5.5
#| fig-cap: "Pseudobulk differential expression for amyloid on the mutant-tau background (NLGF_P301S vs P301S). Canonical DAM markers (crimson) are concentrated among the up-regulated, significant genes; the most significant are labelled. Dashed lines mark the log2 fold-change threshold."
lfc <- pbm$thresholds$lfc; fdr <- pbm$thresholds$fdr
vol <- pbm$top$nlgf_in_p301s
vol$symbol    <- to_symbol(vol$gene)
vol$neglog10p <- -log10(pmax(vol$P.Value, 1e-300))               # cap guards a zero p -> Inf -> ggplot drop-warning
dam_ens <- unique(symbol_map$ensembl[match(canonical_microglia_markers$DAM, symbol_map$symbol)])
dam_ens <- dam_ens[!is.na(dam_ens)]
vol$is_dam <- vol$gene %in% dam_ens
lab <- vol[vol$is_dam & vol$adj.P.Val < fdr & abs(vol$logFC) > lfc, ]
lab <- lab[order(lab$adj.P.Val), ][seq_len(min(12L, nrow(lab))), , drop = FALSE]

ggplot(mapping = aes(logFC, neglog10p)) +
  geom_point(data = vol[!vol$is_dam, ], colour = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = vol[vol$is_dam, ], colour = "#B0344D", size = 1.6) +
  geom_vline(xintercept = c(-lfc, lfc), linetype = "dashed", colour = "grey70") +
  ggrepel::geom_text_repel(data = lab, aes(label = symbol), colour = "#B0344D",
                           size = 3, max.overlaps = Inf, seed = 42L) +
  labs(x = "log2 fold-change (NLGF_P301S vs P301S)", y = "-log10 p-value",
       title = "Amyloid-driven DAM programme")
```

```{r}
#| label: de-counts
sig_tab <- do.call(rbind, lapply(names(pbm$top), function(cn) {
  tt <- pbm$top[[cn]]
  data.frame(
    contrast            = cn,
    `up (large effect)` = sum(tt$logFC >  lfc & tt$adj.P.Val < fdr),
    `down (large effect)` = sum(tt$logFC < -lfc & tt$adj.P.Val < fdr),
    `stageR-confirmed`  = sum(pbm$stageR$stage_padj[, cn] <= pbm$stageR$alpha, na.rm = TRUE),
    check.names = FALSE, stringsAsFactors = FALSE)
}))
knitr::kable(sig_tab, row.names = FALSE,
             caption = sprintf("Differentially expressed genes per contrast. Large-effect = |log2FC| > %.1f and FDR < %.2f; stageR-confirmed counts the family-screen confirmations at any effect size.", lfc, fdr))
```

```{r}
#| label: fig-microglia-whole-volcano
#| fig-width: 10
#| fig-height: 6.5
#| fig-cap: "Whole-microglia pseudobulk volcanoes across the five genotype contrasts. Points are pre-binned for compact rendering; labelled genes are the strongest rows per contrast. Amyloid contrasts carry thousands of FDR < 0.10 rows, whereas the interaction has real stageR-confirmed genes but no standalone large-effect FDR hits."
ggplot(mf_wv_bins, aes(x_mid, y_mid, fill = log_n)) +
  geom_tile(width = median(diff(sort(unique(mf_wv_bins$x_mid)))),
            height = median(diff(sort(unique(mf_wv_bins$y_mid))))) +
  ggrepel::geom_text_repel(
    data = mf_wv_lab,
    aes(effect, neg_log10_p, label = label),
    inherit.aes = FALSE,
    size = 2.2,
    seed = 42L,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.15
  ) +
  facet_wrap(~ contrast, ncol = 3) +
  scale_fill_gradient(low = "grey94", high = "#B0344D", name = "log10(n + 1)") +
  labs(x = "log2 fold-change", y = "-log10 p-value",
       title = "Whole-microglia DE across all contrasts") +
  theme(panel.grid = element_blank())
```

## The tau-by-amyloid interaction is under-powered, not absent

No gene clears the large-effect bar on the `interaction` contrast (0 up, 0 down at
|log2FC| > `r lfc`, FDR < `r fdr`), yet the stageR family screen still confirms
`r n_int_stageR` interaction genes - every one with a substantial fold-change (all above the `r lfc`
log2 effect bar, median `r round(med_int_lfc, 1)`) that the stricter standalone per-contrast test
cannot individually resolve (its most significant gene sits at FDR `r round(min_int_adjp, 2)`), so
this is not an empty result. The difference-of-differences carries roughly
twice the standard error of a main effect, and the pseudobulk design leaves few replicate degrees
of freedom - eBayes moderation lifts the effective total to `r round(pbm$interaction$df)`: the
minimum log2 fold-change detectable at a nominal 80% per-test power for the median gene is
`r round(pbm$interaction$mde, 2)` (median posterior SE `r round(pbm$interaction$median_se, 3)`). The honest reading is therefore that a per-gene
interaction below about `r round(pbm$interaction$mde, 1)` log2 units falls under this design's
nominal detection floor - and the multiple-testing-corrected discovery floor is higher still - so
absence of large-effect genes is not evidence of no interaction. Crucially, the
compositional interaction on DAM *is* significant, which points to a synergy expressed through
compartment size rather than per-gene fold-change. The trajectory chapter (@sec-trajectory) then
separates the two ways that synergy could act - more DAM cells, or cells advancing further along
the activation axis - and resolves it as the former: the tau-amyloid interaction adds DAM cells,
with no statistically supported signal that it advances them further.

## Per-substate differential expression

Repeating the pseudobulk fit within each substate is only valid where every replicate unit
carries enough cells. Under a floor of `r pbs$min_cells` cells per unit, the substates that clear
it on every unit (`r knitr::combine_words(fit_subs)`) are fitted; IFN (as few as
`r min(pbs$cell_counts["IFN", ])` cells in a unit) and the empty proliferative level are left
descriptive-only. Within the DAM compartment alone, amyloid still moves `r dam_amyloid_de` genes
past the large-effect bar (NLGF_P301S vs P301S), confirming the programme is graded - a shift in
what DAM cells express, not merely how many cells are DAM.

```{r}
#| label: substate-de
sub_rows <- lapply(names(pbs$per_substate), function(s) {
  p <- pbs$per_substate[[s]]
  data.frame(
    substate        = s,
    status          = p$status,
    nuclei          = p$n_cells,
    `min cells/unit` = min(pbs$cell_counts[s, ]),
    `genes kept`    = if (identical(p$status, "fit")) p$kept else NA_integer_,
    `interaction MDE` = if (identical(p$status, "fit")) round(p$interaction$mde, 2) else NA_real_,
    check.names = FALSE, stringsAsFactors = FALSE)
})
knitr::kable(do.call(rbind, sub_rows), row.names = FALSE,
             caption = "Per-substate pseudobulk DE under the min-cell floor. Fewer cells give a larger minimum detectable effect (interaction MDE, log2 units), reported honestly.")
```

```{r}
#| label: fig-microglia-substate-audit
#| fig-width: 7.8
#| fig-height: 4.2
#| fig-cap: "Per-substate DE fit audit. Bars show the minimum cells in any replicate unit, with the dashed line marking the pre-set floor for fitting. Homeostatic and DAM pass; IFN and the empty proliferative level remain descriptive."
ggplot(mf_audit, aes(substate, min_cells_unit, fill = status)) +
  geom_col(width = 0.68) +
  geom_hline(yintercept = pbs$min_cells, linetype = "dashed", colour = "#B0344D") +
  geom_text(aes(label = status), vjust = -0.35, size = 3) +
  scale_fill_manual(values = c(fit = "#6B7A3F", skipped = "#B7A06A"), guide = "none") +
  labs(x = NULL, y = "minimum cells per replicate unit",
       title = "Substate fits are limited by replicate-unit cell floors") +
  theme(panel.grid.major.x = element_blank())
```

```{r}
#| label: fig-microglia-substate-volcano
#| fig-width: 9.2
#| fig-height: 4.4
#| fig-cap: "Within-substate DE summary for the fitted homeostatic and DAM compartments. Bars are signed counts of FDR < 0.10 rows (up positive, down negative) for each contrast. Amyloid effects persist within states; the interaction remains much smaller than the amyloid main responses."
ggplot(mf_ss_signed, aes(contrast, n_genes, fill = direction)) +
  geom_hline(yintercept = 0, colour = "grey65") +
  geom_col(width = 0.72) +
  facet_wrap(~ substate, nrow = 1) +
  scale_fill_manual(values = c(down = "#3F5A6B", up = "#B0344D"), name = "direction") +
  labs(x = NULL, y = "signed FDR < 0.10 gene count",
       title = "Within-substate expression shifts keep the amyloid programme visible") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid.major.x = element_blank())
```

## Caveats and provenance

Three caveats travel with these results. First, the Thrupp under-detection already noted: DAM is
scored, not thresholded, and the differential tests run on raw counts, so the attenuation biases
against the DAM signal rather than inventing it. Second, contaminant pruning removed proportionally
more low-quality nuclei from the `r drop_geno` background (about `r drop_geno_pct`% of it, mostly
cluster `r drop_clu`) than from the others. Because every dropped cluster scores low on DAM - the
highest dropped-cluster median (`r round(drop_dam_max, 2)`) falls well short of the DAM-enriched
clusters that drive the headline (medians up to `r round(kept_dam_max, 2)`, all retained) - the
pruning cannot have manufactured the amyloid-to-DAM signal by discarding DAM-high nuclei. Removing
low-DAM nuclei can only nudge a genotype's DAM fraction upward, so this genotype-skewed dropout is
reported in full below rather than hidden. Third,
`r if (!is.null(comp$sccomp)) "the sccomp cross-check is a supportive, off-lock arm: its sampler showed a small number of divergent transitions with otherwise healthy diagnostics, so it corroborates direction and magnitude but the locked propeller call is the primary inference." else "the sccomp Bayesian cross-check did not run on this build (its Stan backend was absent), so the locked propeller call stands alone as the primary inference."`

```{r}
#| label: dropout
knitr::kable(as.data.frame.matrix(mrd$prune$dropped_by_genotype),
             caption = sprintf("Dropped contaminant/doublet nuclei by genotype and cluster (%d of %d nuclei pruned). Reported to show the genotype-associated QC dropout; no dropped cluster was DAM-high.",
                               mrd$prune$n_dropped, mrd$prune$n_dropped + mrd$n_cells))
```

```{r}
#| label: sccomp-diag
#| echo: true
sd <- comp$provenance$sccomp_diagnostics
if (!is.null(sd)) {
  cat(sprintf("sccomp HMC: %d chains, %d post-warmup draws; %d divergent transitions, %d at max tree depth, minimum E-BFMI %.2f.\n",
              sd$n_chains, sd$n_post_warmup_draws, sd$num_divergent, sd$num_max_treedepth, sd$ebfmi_min))
} else {
  cat("sccomp cross-check not run on this build (propeller-only); Stan backend absent.\n")
}
```
