# Microglia: amyloid drives a homeostatic-to-DAM programme {#sec-microglia}

This chapter is the single-nucleus core of the study. From the reprocessed, batch-integrated
microglia (SCT normalisation, Harmony over batch only) it places each nucleus into a canonical
substate, tests whether substate composition shifts across the genotype contrasts, and runs
pseudobulk differential expression to characterise the activation programme and to weigh the
tau-by-amyloid interaction. The headline, supported three independent ways below, is that
amyloid (NLGF) drives microglia out of a homeostatic state into disease-associated microglia
(DAM); mutant tau scales the size of that response without, on its own, redrawing the
programme. One measurement caveat frames everything that follows: single-nucleus capture
under-detects a sizeable share of DAM activation transcripts (Thrupp and colleagues, 2020), so
DAM is read as a scored programme rather than thresholded gene by gene, and every differential
test runs on raw counts.

```{r}
#| label: microglia-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)                                    # promote any chunk warning to a render error (gate-critical)
targets::tar_source()                                # constants, marker sets, plot theme + genotype scales
targets::tar_load(c(microglia_report, composition_results, pb_de_microglia,
                    pb_de_substate, symbol_map))     # all compact targets -> render stays cheap (no 612MB Seurat)
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

mrd  <- microglia_report
cf   <- mrd$cell_frame
comp <- composition_results
pbm  <- pb_de_microglia
pbs  <- pb_de_substate

# substate colours reuse the genotype palette hues -> cohesive with the report identity.
sub_cols <- c(Homeostatic = "#7FA8C1", DAM = "#B0344D", IFN = "#E08754",
              Proliferative = "#6BAF7F", ambiguous = "grey70", unassigned = "grey85")
present_sub <- intersect(names(sub_cols), as.character(unique(cf$substate)))
cf$substate <- factor(as.character(cf$substate), levels = present_sub)

# genotype x substate counts (primary cluster-level labels) -> the compositional shift.
st       <- mrd$provenance$substate_table
dam_pct  <- round(100 * st[, "DAM"] / rowSums(st))               # % DAM per genotype
homeo_pct <- round(100 * st[, "Homeostatic"] / rowSums(st))      # % Homeostatic per genotype

# ensembl -> symbol lookup for DE gene labels (symbol_map rownames are ensembl).
to_symbol <- function(ens) {
  s <- symbol_map$symbol[match(ens, symbol_map$ensembl)]
  ifelse(is.na(s) | s == "", ens, s)
}
fmt_p <- function(p) formatC(p, format = "e", digits = 2)         # compact scientific p-values for tables

# DAM interaction across the two methods -- computed (not hardcoded) so the prose tracks the
# actual cached build, sccomp being run-to-run variable on its few divergent transitions.
int_prop_fdr <- comp$headline$fdr_global[comp$headline$contrast == "interaction"]
int_sc <- if (!is.null(comp$sccomp))
  comp$sccomp[comp$sccomp$contrast == "interaction" & comp$sccomp$substate == "DAM", ] else NULL
int_sentence <- if (is.null(int_sc))
  "it was not re-tested under sccomp on this build (propeller-only)" else
  sprintf("under sccomp it sits near the significance boundary (FDR %s, 95%% credible interval [%.2f, %.2f])",
          fmt_p(int_sc$c_fdr), int_sc$c_lower, int_sc$c_upper)

# Back the "no dropped cluster was DAM-high" caveat with the recorded per-cluster DAM medians
# (S2 prune provenance) rather than asserting it; report the genotype/cluster dropout factually.
# Dropped clusters' DAM medians OVERLAP the kept homeostatic clusters -> the honest comparison is
# against the DAM-ENRICHED ceiling (kept_dam_max): no dropped cluster reaches the DAM compartment.
qcr      <- mrd$prune$qc_rationale
dbg      <- mrd$prune$dropped_by_genotype
drop_geno <- names(which.max(rowSums(dbg)))                       # genotype losing the most nuclei
drop_clu  <- names(which.max(colSums(dbg)))                       # cluster contributing the most
drop_dam_max <- max(qcr[as.character(mrd$prune$dropped), "DAM_med"])                            # top DAM median, dropped
kept_dam_max <- max(qcr[setdiff(rownames(qcr), as.character(mrd$prune$dropped)), "DAM_med"])    # DAM-enriched ceiling, kept
```

## Substate landscape

UCell scores each nucleus against the canonical homeostatic, DAM, interferon-response (IFN) and
proliferative signatures; a per-signature calibration then assigns a primary substate at the
cluster level (the per-cell argmax is noisier and kept only as a diagnostic). After dropping
`r mrd$prune$n_dropped` contaminant or doublet nuclei (clusters
`r paste(mrd$prune$dropped, collapse = ", ")`), `r format(mrd$n_cells, big.mark = ",")` nuclei
remain. They resolve into `r format(sum(cf$substate == "Homeostatic"), big.mark = ",")`
homeostatic, `r format(sum(cf$substate == "DAM"), big.mark = ",")` DAM and
`r sum(cf$substate == "IFN")` IFN nuclei; no cluster is proliferative-dominant, so that level
carries no cells (sparse cycling markers average out at the cluster level rather than forming a
coherent population). Every substate signature maps in full to the assayed genes
(`r paste(sprintf("%s %s", names(mrd$provenance$n_used), mrd$provenance$n_used)[2:5], collapse = ", ")`).

```{r}
#| label: umap
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Microglia substate landscape on the batch-integrated UMAP. Left: primary substate label. Right: the calibrated DAM activation score, which forms a continuous homeostatic-to-DAM gradient rather than a discrete switch."
p_sub <- ggplot(cf, aes(umap_1, umap_2, colour = substate)) +
  geom_point(size = 0.25, alpha = 0.6) +
  scale_colour_manual(values = sub_cols, breaks = present_sub, name = "substate") +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Primary substate")

p_dam <- ggplot(cf, aes(umap_1, umap_2, colour = DAM_UCell_z)) +
  geom_point(size = 0.25, alpha = 0.6) +
  scale_colour_gradient(low = "grey85", high = "#B0344D", name = "DAM z") +
  labs(x = "UMAP 1", y = "UMAP 2", title = "DAM activation score")

(p_sub | p_dam)
```

## Amyloid expands the DAM compartment

The compositional shift is unambiguous. Without amyloid the microglia are overwhelmingly
homeostatic (DAM `r dam_pct["MAPTKI"]`% in MAPTKI, `r dam_pct["P301S"]`% in P301S); with amyloid
the DAM compartment swells to `r dam_pct["NLGF_MAPTKI"]`% on the tau-knockout background and
`r dam_pct["NLGF_P301S"]`% on the mutant-tau background, with the homeostatic fraction falling
in step (`r homeo_pct["MAPTKI"]`% and `r homeo_pct["P301S"]`% down to
`r homeo_pct["NLGF_MAPTKI"]`% and `r homeo_pct["NLGF_P301S"]`%). The same shift is visible per
nucleus as a rightward move in the DAM activation score, and it is steeper on the mutant-tau
background, the first hint of the tau-by-amyloid interaction tested below.

```{r}
#| label: composition-shift
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Amyloid drives the homeostatic-to-DAM shift. Left: substate proportions per genotype (cluster-level labels). Right: per-nucleus DAM activation score by genotype; the amyloid-bearing genotypes carry the heavier upper tail."
prop_df <- as.data.frame(prop.table(st, 1))                       # genotype, substate, Freq(=within-genotype proportion)
prop_df <- prop_df[prop_df$substate %in% present_sub, ]
prop_df$genotype <- factor(prop_df$genotype, levels = genotype_levels)
prop_df$substate <- factor(as.character(prop_df$substate), levels = present_sub)

p_prop <- ggplot(prop_df, aes(genotype, Freq, fill = substate)) +
  geom_col() +
  scale_fill_manual(values = sub_cols, breaks = present_sub, name = "substate") +
  labs(x = NULL, y = "proportion of microglia", title = "Substate composition") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_score <- ggplot(cf, aes(genotype, DAM_UCell_z, fill = genotype)) +
  geom_violin(scale = "width") +
  scale_fill_genotype(guide = "none") +
  labs(x = NULL, y = "DAM activation score (z)", title = "DAM score per nucleus") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

(p_prop | p_score)
```

```{r}
#| label: substate-table
counts_tab <- as.data.frame.matrix(st[, present_sub, drop = FALSE])
counts_tab <- cbind(genotype = rownames(counts_tab), counts_tab,
                    `DAM %` = dam_pct[rownames(counts_tab)])
knitr::kable(counts_tab, row.names = FALSE,
             caption = "Microglia per genotype and substate (primary cluster-level labels).")
```

## Composition is tested across the genotype contrasts

Composition is tested formally with propeller (the locked, reproducible primary, on the logit
transform) and cross-checked with sccomp, a Bayesian beta-binomial model run off-lock when its
Stan backend is present. Both amyloid contrasts move DAM strongly up and the homeostatic
fraction strongly down, with very small tail probabilities (table below). The tau-by-amyloid
`interaction` on DAM is positive and significant under propeller
(FDR `r fmt_p(int_prop_fdr)`); `r int_sentence`,
so it is flagged rather than asserted. A static compositional interaction of this size is
consistent with mutant tau accelerating the same activation programme rather than opening a new
one - a rate question handed to the trajectory phase.

```{r}
#| label: composition-forest
#| fig-width: 8
#| fig-height: 4.5
#| fig-cap: "Substate composition effects (sccomp posterior effect on the logit proportion, with 95% credible intervals) for DAM and homeostatic microglia across the five contrasts. DAM and homeostatic move as mirror images; both amyloid contrasts are far from zero, the interaction is small."
ctr_levels <- c("tau_alone", "nlgf_in_maptki", "nlgf_in_p301s", "tau_in_nlgf", "interaction")
if (!is.null(comp$sccomp)) {
  fdf <- comp$sccomp[comp$sccomp$substate %in% c("Homeostatic", "DAM"), ]
  fdf$contrast <- factor(fdf$contrast, levels = rev(ctr_levels))
  ggplot(fdf, aes(c_effect, contrast, colour = substate)) +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_pointrange(aes(xmin = c_lower, xmax = c_upper),
                    position = position_dodge(width = 0.6)) +
    scale_colour_manual(values = sub_cols[c("Homeostatic", "DAM")], name = "substate") +
    labs(x = "effect on logit proportion (sccomp c_effect)", y = NULL,
         title = "Composition effects across contrasts")
} else {
  fdf <- comp$propeller_logit[comp$propeller_logit$substate %in% c("Homeostatic", "DAM"), ]
  fdf$contrast <- factor(fdf$contrast, levels = rev(ctr_levels))
  ggplot(fdf, aes(t, contrast, fill = substate)) +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_col(position = position_dodge(width = 0.6), width = 0.6) +
    scale_fill_manual(values = sub_cols[c("Homeostatic", "DAM")], name = "substate") +
    labs(x = "propeller t-statistic", y = NULL,
         title = "Composition effects across contrasts (propeller)")
}
```

```{r}
#| label: composition-table
pl  <- comp$propeller_logit
dam_pl <- pl[pl$substate == "DAM", ]
dam_pl <- dam_pl[match(ctr_levels, dam_pl$contrast), ]
ctab <- data.frame(
  contrast        = ctr_levels,
  `propeller t`   = round(dam_pl$t, 2),
  `propeller FDR` = fmt_p(dam_pl$fdr_global),
  check.names = FALSE, stringsAsFactors = FALSE)
if (!is.null(comp$sccomp)) {
  sc <- comp$sccomp[comp$sccomp$substate == "DAM", ]
  sc <- sc[match(ctr_levels, sc$contrast), ]
  ctab$`sccomp effect` <- round(sc$c_effect, 2)
  ctab$`sccomp 95% CI` <- sprintf("[%.2f, %.2f]", sc$c_lower, sc$c_upper)
  ctab$`sccomp FDR`    <- fmt_p(sc$c_fdr)
}
knitr::kable(ctab, row.names = FALSE,
             caption = "DAM composition across the five contrasts: propeller (primary, batch-fixed) and sccomp (cross-check, random batch). Direction is read from the t-statistic / effect, not the marginal proportion ratio.")

flagged <- comp$concordance[comp$concordance$flag, c("contrast", "substate")]
```

Cross-method concordance flags `r nrow(flagged)` of `r nrow(comp$concordance)`
contrast-by-substate cells (`r paste(sprintf("%s/%s", flagged$contrast, flagged$substate), collapse = "; ")`):
the sparse IFN compartment carries most of them, plus the DAM interaction sitting on the
significance boundary. Per the pre-declared rule the propeller-logit call stands; the
disagreements are reported, never averaged away.

## The amyloid-to-DAM expression programme

Pseudobulk differential expression (limma-voom with quality weights, aggregated to the 16
genotype-by-batch replicate units) confirms the activation programme at the transcript level.
Of `r format(pbm$kept, big.mark = ",")` expressed genes, `r format(pbm$stageR$n_screened, big.mark = ",")`
pass the stageR family screen. The canonical DAM markers move up coherently under amyloid: of
the `r pbm$dam_concordance$nlgf_in_p301s$n_markers_in_fit` markers in the fit,
`r round(100 * pbm$dam_concordance$nlgf_in_maptki$frac_up)`% rise on the tau-knockout background
(mean log2 fold-change `r round(pbm$dam_concordance$nlgf_in_maptki$mean_logFC, 2)`) and
`r round(100 * pbm$dam_concordance$nlgf_in_p301s$frac_up)`% on the mutant-tau background
(mean `r round(pbm$dam_concordance$nlgf_in_p301s$mean_logFC, 2)`), reproducing the original
study's direction. The volcano below shows the strongest amyloid contrast with the DAM markers
picked out.

```{r}
#| label: volcano
#| fig-width: 8
#| fig-height: 5.5
#| fig-cap: "Pseudobulk differential expression for amyloid on the mutant-tau background (NLGF_P301S vs P301S). Canonical DAM markers (crimson) are concentrated among the up-regulated, significant genes; the most significant are labelled. Dashed lines mark the log2 fold-change threshold."
lfc <- pbm$thresholds$lfc; fdr <- pbm$thresholds$fdr
vol <- pbm$top$nlgf_in_p301s
vol$symbol    <- to_symbol(vol$gene)
vol$neglog10p <- -log10(pmax(vol$P.Value, 1e-300))               # cap guards a zero p -> Inf -> ggplot drop-warning
dam_ens <- unique(symbol_map$ensembl[match(canonical_microglia_markers$DAM, symbol_map$symbol)])
dam_ens <- dam_ens[!is.na(dam_ens)]
vol$is_dam <- vol$gene %in% dam_ens
lab <- vol[vol$is_dam & vol$adj.P.Val < fdr & abs(vol$logFC) > lfc, ]
lab <- lab[order(lab$adj.P.Val), ][seq_len(min(12L, nrow(lab))), , drop = FALSE]

ggplot(mapping = aes(logFC, neglog10p)) +
  geom_point(data = vol[!vol$is_dam, ], colour = "grey80", size = 0.5, alpha = 0.5) +
  geom_point(data = vol[vol$is_dam, ], colour = "#B0344D", size = 1.6) +
  geom_vline(xintercept = c(-lfc, lfc), linetype = "dashed", colour = "grey70") +
  ggrepel::geom_text_repel(data = lab, aes(label = symbol), colour = "#B0344D",
                           size = 3, max.overlaps = Inf, seed = 42L) +
  labs(x = "log2 fold-change (NLGF_P301S vs P301S)", y = "-log10 p-value",
       title = "Amyloid-driven DAM programme")
```

```{r}
#| label: de-counts
sig_tab <- do.call(rbind, lapply(names(pbm$top), function(cn) {
  tt <- pbm$top[[cn]]
  data.frame(
    contrast            = cn,
    `up (large effect)` = sum(tt$logFC >  lfc & tt$adj.P.Val < fdr),
    `down (large effect)` = sum(tt$logFC < -lfc & tt$adj.P.Val < fdr),
    `stageR-confirmed`  = sum(pbm$stageR$stage_padj[, cn] <= pbm$stageR$alpha, na.rm = TRUE),
    check.names = FALSE, stringsAsFactors = FALSE)
}))
knitr::kable(sig_tab, row.names = FALSE,
             caption = sprintf("Differentially expressed genes per contrast. Large-effect = |log2FC| > %.1f and FDR < %.2f; stageR-confirmed counts the family-screen confirmations at any effect size.", lfc, fdr))
```

## The tau-by-amyloid interaction is under-powered, not absent

No gene clears the large-effect bar on the `interaction` contrast (0 up, 0 down at
|log2FC| > `r lfc`, FDR < `r fdr`), yet the stageR family screen still confirms
`r sum(pbm$stageR$stage_padj[, "interaction"] <= pbm$stageR$alpha, na.rm = TRUE)` genes with
smaller effects, so this is not an empty result. The difference-of-differences carries roughly
twice the standard error of a main effect, and the pseudobulk design leaves few replicate degrees
of freedom - eBayes moderation lifts the effective total to `r round(pbm$interaction$df)`: the
minimum log2 fold-change detectable at 80% power for the median gene is
`r round(pbm$interaction$mde, 2)` (median posterior SE `r round(pbm$interaction$median_se, 3)`). The honest reading is therefore that a per-gene
interaction below about `r round(pbm$interaction$mde, 1)` log2 units would be invisible to this
design - absence of large-effect genes is not evidence of no interaction. Crucially, the
compositional interaction on DAM *is* significant, which points to a synergy expressed as
compartment size and progression rate rather than as per-gene fold-change. That rate question -
does amyloid advance microglia further along the activation trajectory on a mutant-tau
background - is the subject of the next phase.

## Per-substate differential expression

Repeating the pseudobulk fit within each substate is only valid where every replicate unit
carries enough cells. Under a floor of `r pbs$min_cells` cells per unit, homeostatic and DAM
microglia qualify and are fitted; IFN (as few as `r min(pbs$cell_counts["IFN", ])` cells in a
unit) and the absent proliferative level are left descriptive-only. Within the DAM compartment
amyloid still shifts expression, confirming the programme is graded, not merely a change in how
many cells are DAM.

```{r}
#| label: substate-de
sub_rows <- lapply(names(pbs$per_substate), function(s) {
  p <- pbs$per_substate[[s]]
  data.frame(
    substate        = s,
    status          = p$status,
    nuclei          = p$n_cells,
    `min cells/unit` = min(pbs$cell_counts[s, ]),
    `genes kept`    = if (identical(p$status, "fit")) p$kept else NA_integer_,
    `interaction MDE` = if (identical(p$status, "fit")) round(p$interaction$mde, 2) else NA_real_,
    check.names = FALSE, stringsAsFactors = FALSE)
})
knitr::kable(do.call(rbind, sub_rows), row.names = FALSE,
             caption = "Per-substate pseudobulk DE under the min-cell floor. Fewer cells give a larger minimum detectable effect (interaction MDE, log2 units), reported honestly.")
```

## Caveats and provenance

Three caveats travel with these results. First, the Thrupp under-detection already noted: DAM is
scored, not thresholded, and the differential tests run on raw counts, so the attenuation biases
against the DAM signal rather than inventing it. Second, contaminant pruning removed proportionally
more low-quality nuclei from the `r drop_geno` background (mostly cluster `r drop_clu`), but every
dropped cluster scores low on DAM - the highest dropped-cluster median (`r round(drop_dam_max, 2)`)
falls well short of the DAM-enriched clusters that drive the headline (medians up to
`r round(kept_dam_max, 2)`, all retained) - so the amyloid-to-DAM headline cannot be a pruning
artefact; the asymmetry is reported in full below rather than hidden. Third,
the sccomp cross-check is a supportive, off-lock arm: its sampler showed a small number of
divergent transitions with otherwise healthy diagnostics, so it corroborates direction and
magnitude but the locked propeller call is the primary inference.

```{r}
#| label: dropout
knitr::kable(as.data.frame.matrix(mrd$prune$dropped_by_genotype),
             caption = sprintf("Dropped contaminant/doublet nuclei by genotype and cluster (%d of %d nuclei pruned). Reported to show the genotype-associated QC dropout; no dropped cluster was DAM-high.",
                               mrd$prune$n_dropped, mrd$prune$n_dropped + mrd$n_cells))
```

```{r}
#| label: sccomp-diag
#| echo: true
sd <- comp$provenance$sccomp_diagnostics
if (!is.null(sd)) {
  cat(sprintf("sccomp HMC: %d chains, %d post-warmup draws; %d divergent transitions, %d at max tree depth, minimum E-BFMI %.2f.\n",
              sd$n_chains, sd$n_post_warmup_draws, sd$num_divergent, sd$num_max_treedepth, sd$ebfmi_min))
} else {
  cat("sccomp cross-check not run on this build (propeller-only); Stan backend absent.\n")
}
```
