# The amyloid response across tau backgrounds, method by method {#sec-modality}

```{r}
#| label: modality-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)                                    # promote any chunk warning to a render error (gate-critical)
targets::tar_source()                                # constants, plot theme + modality figure helpers
targets::tar_load(modality_scatter_figures)          # compact logFC pairs + functional group scores
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

msf <- modality_scatter_figures

fmt_n <- function(x) format(as.integer(round(x)), big.mark = ",", trim = TRUE)
fmt_num <- function(x, digits = 2) formatC(as.numeric(x), format = "f", digits = digits)
fmt_signed <- function(x, digits = 2) {
  x <- as.numeric(x)
  paste0(ifelse(is.finite(x) & x > 0, "+", ""), fmt_num(x, digits))
}
panel_stats <- do.call(rbind, lapply(msf$order, function(m) {
  d <- msf$panels[[m]]$data
  top <- modality_scatter_label_rows(d)
  data.frame(
    modality = m,
    n = nrow(d),
    pearson = suppressWarnings(stats::cor(d$x, d$y, method = "pearson")),
    spearman = suppressWarnings(stats::cor(d$x, d$y, method = "spearman")),
    slope = unname(stats::coef(stats::lm(y ~ x, data = d))[["x"]]),
    cutoff = attr(d, "offdiag_cutoff", exact = TRUE),
    n_label = nrow(top),
    stringsAsFactors = FALSE
  )
}))
mod_stats <- function(m) panel_stats[panel_stats$modality == m, , drop = FALSE]
mod_group <- function(m, rank = 1L) {
  x <- msf$groups$summary[as.character(msf$groups$summary$modality) == m, , drop = FALSE]
  x <- x[order(-x$rank_score, x$group_priority, x$group_label, method = "radix"), , drop = FALSE]
  x[rank, , drop = FALSE]
}
```

## Non-snRNAseq modality landscape

GeoMx is the transcriptomic comparator closest to the microglia pseudobulk view. Across `r fmt_n(mod_stats("GeoMx")$n)` spatial genes, the amyloid effects on the MAPTKI and P301S backgrounds are strongly aligned (Pearson `r fmt_num(mod_stats("GeoMx")$pearson)`, Spearman `r fmt_num(mod_stats("GeoMx")$spearman)`, OLS slope `r fmt_num(mod_stats("GeoMx")$slope)`). The off-diagonal tail is modest relative to the full feature set (`r fmt_n(mod_stats("GeoMx")$n_label)` labelled genes at |x-y| >= `r fmt_num(mod_stats("GeoMx")$cutoff)`), led by `r as.character(mod_group("GeoMx", 1)$group_label)` (`r as.character(mod_group("GeoMx", 1)$top_features)`) with a MAPTKI-higher aggregate delta of `r fmt_signed(mod_group("GeoMx", 1)$delta)`. That reads as a broad shared tissue-RNA amyloid response with a smaller tau-dependent edge.

The 24M bulk proteome is less coupled between tau backgrounds. It contributes `r fmt_n(mod_stats("Proteome")$n)` protein groups but has weak background-to-background concordance (Pearson `r fmt_num(mod_stats("Proteome")$pearson)`, Spearman `r fmt_num(mod_stats("Proteome")$spearman)`, OLS slope `r fmt_num(mod_stats("Proteome")$slope)`) and a much wider off-diagonal threshold (`r fmt_num(mod_stats("Proteome")$cutoff)`). Only `r fmt_n(mod_stats("Proteome")$n_label)` labels enter the tail. The top-ranked group is `r as.character(mod_group("Proteome", 1)$group_label)` (`r as.character(mod_group("Proteome", 1)$top_features)`), with a P301S-higher aggregate delta of `r fmt_signed(mod_group("Proteome", 1)$delta)`; `r as.character(mod_group("Proteome", 2)$group_label)` is the next strongest category. This layer is therefore a sparse protein-level divergence readout, not a direct microglial composition measurement.

Phosphoproteomics is plotted at the parent-protein level: `r fmt_n(msf$provenance$phospho_site_features)` finite phosphosite rows collapse to `r fmt_n(msf$provenance$phospho_parent_proteins)` parent-protein means. The parent-level amyloid effects are modestly concordant (Pearson `r fmt_num(mod_stats("Phospho")$pearson)`, Spearman `r fmt_num(mod_stats("Phospho")$spearman)`, OLS slope `r fmt_num(mod_stats("Phospho")$slope)`) with `r fmt_n(mod_stats("Phospho")$n_label)` labelled off-diagonal proteins at |x-y| >= `r fmt_num(mod_stats("Phospho")$cutoff)`. The leading category is `r as.character(mod_group("Phospho", 1)$group_label)` (`r as.character(mod_group("Phospho", 1)$top_features)`), P301S-higher by `r fmt_signed(mod_group("Phospho", 1)$delta)`, while `r as.character(mod_group("Phospho", 2)$group_label)` moves MAPTKI-higher. This is a signalling-state layer over parent-protein summaries; the figure deliberately avoids site-specific kinase claims.

## Per-method amyloid effect: tau-KO versus mutant-tau background

```{r}
#| label: fig-modality-amyloid-effect
#| fig-width: 10.5
#| fig-height: 10.4
#| fig-cap: "Per-method amyloid (NLGF) log2 fold change on the tau-KO background (NLGF_MAPTKI vs MAPTKI, y) against the mutant-tau background (NLGF_P301S vs P301S, x). The dashed line is y=x, the null of a tau-independent amyloid response; the tau-by-amyloid interaction contrast is x minus y, so distance from the diagonal ranks its magnitude. Phosphoproteomics points are parent-protein means of finite phosphosite logFC pairs. Labels mark empirical off-diagonal outliers using a standardized within-method rule: unique display labels whose |x-y| exceeds that method's 99th percentile of |x-y|. The solid line is the OLS trend."
#| fig-alt: "Four scatter plots, one per method, each comparing the amyloid log fold change between the two tau backgrounds against a dashed identity line, with an OLS trend and empirical off-diagonal outliers labelled; phosphoproteomics is collapsed to parent-protein means."
panels <- lapply(msf$order, function(m) {
  modality_interaction_scatter(msf$panels[[m]]$data, title = msf$panels[[m]]$title)
})
wrap_plots(panels, ncol = 2)
```

## Off-diagonal functional categories

```{r}
#| label: fig-modality-functional-scores
#| fig-width: 10.5
#| fig-height: 8.2
#| fig-cap: "Within-method 99th-percentile off-diagonal genes and proteins from Figure 6 assigned to one primary functional category per method. Broad role categories are assembled from mouse GO-BP annotations; threshold-passing items without a role-set hit are retained in explicit fallback categories rather than dropped. Phosphoproteomics categories use the same parent-protein mean points displayed in Figure 6. Each segment connects the aggregate amyloid log2 fold change in NLGF_MAPTKI to the aggregate amyloid log2 fold change in NLGF_P301S; segment colour is P301S minus MAPTKI, and point size is the number of scored genes or proteins in the category."
#| fig-alt: "Faceted dumbbell plots show modality-specific functional categories for within-method 99th-percentile off-diagonal genes or proteins, with paired points for aggregate NLGF_MAPTKI and NLGF_P301S amyloid responses connected by a segment coloured by the P301S minus MAPTKI score."
functional_group_score_plot(msf$groups$summary, title = "Functional categories for within-method Q99 off-diagonal genes and proteins")
```
