# The tau-amyloid synergy adds DAM cells rather than advancing them {#sec-trajectory}

The microglia chapter left one question open. Amyloid drives the homeostatic-to-DAM shift, and
the compositional interaction on DAM is significant: on a mutant-tau background amyloid produces
*more* DAM cells than additivity predicts. But a larger DAM compartment and cells that have
travelled *further* into activation are two different claims, and the pseudobulk interaction was
under-powered to separate them. This chapter orders the same nuclei along an activation
pseudotime and asks the distinct question the composition test cannot: beyond making more DAM
cells, does mutant tau push microglia further along the activation axis? Read three ways below,
it does not detectably - no channel gives statistically supported evidence of further advance, and
the tau-amyloid synergy is expressed almost entirely as compartment size rather than as cells
sitting further along the axis. Two framings travel with every number here: pseudotime is an activation
*ordering* (position and extent of advance), not developmental time or potency, and because the
data are a single cross-sectional snapshot, "advance" is a position statement, never a measured
rate.

```{r}
#| label: trajectory-setup
#| include: false
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
})
options(warn = 2)                                    # promote any chunk warning to a render error (gate-critical)
targets::tar_source()                                # constants, plot theme + genotype scales
targets::tar_load(trajectory_report)                 # ONE compact target (no 612MB Seurat) -> render stays cheap
knitr::opts_chunk$set(message = FALSE)
theme_set(theme_tau())

trd   <- trajectory_report
cf    <- trd$cell_frame
on    <- cf[cf$on_lineage, , drop = FALSE]            # on-lineage cells (finite pseudotime)
inter <- trd$interaction
prov  <- trd$provenance
glmm  <- trd$glmm

fmt_p  <- function(p) formatC(p, format = "e", digits = 2)
irow   <- function(m) inter[match(m, inter$measure), ]            # one interaction row by measure
mp_ctr <- function(cn) { r <- trd$weighted_top[[cn]]; r[r$measure == "mean_pt", ] }

# ALL headline numbers inline-computed from the loaded target -> they track the actual cached build
# (R4.6 re-baseline; the per-channel p-values and loadings are drift-prone), never hardcoded.
int_mpt  <- irow("mean_pt")                          # interaction on mean position
int_comp <- irow("comp_cf")                          # composition channel
int_prog <- irow("progression_cf")                   # progression channel (the distinct question)
int_wh   <- irow("within_homeostatic")               # within-homeostatic mean position (composition-robust)
amy_k    <- mp_ctr("nlgf_in_maptki")                 # amyloid main effect, tau-knockout background
amy_p    <- mp_ctr("nlgf_in_p301s")                  # amyloid main effect, mutant-tau background
load_comp  <- prov$composition_loading
load_prog  <- prov$progression_loading
load_cross <- prov$cross_loading
dims_min   <- min(trd$sensitivity$spearman_vs_primary)           # weakest dims/all-retained robustness
omit_pct   <- round(100 * prov$omitted_frac_overall, 1)

# per-cell glmmTMB sentence, computed so the prose adapts if the supportive arm degrades on a build.
glmm_scale <- if (identical(glmm$method, "glmmTMB_beta")) "logit" else "rank-normal"
glmm_sentence <- if (identical(glmm$method, "failed"))
  sprintf("did not yield a usable fit on this build (%s), so the summary-level decomposition stands alone",
          glmm$fail_reason) else
  sprintf("places the interaction at %+.3f on the %s scale (95%% CI [%.3f, %.3f], p %s)",
          glmm$estimate, glmm_scale, glmm$ci_l, glmm$ci_r, fmt_p(glmm$p_value))
# provenance line for the per-cell arm, branched like the sentence: a failed fit carries NA re_sd /
# singular (extractor allows it on the failed branch), so print fail_reason there, never "NA".
glmm_prov <- if (identical(glmm$method, "failed"))
  sprintf("Per-cell sensitivity: %s on %s nuclei / %d units (%s).",
          glmm$method, format(glmm$n_cells, big.mark = ","), glmm$n_units, glmm$fail_reason) else
  sprintf("Per-cell sensitivity: %s on %s nuclei / %d units (random-effect SD %.3f, singular = %s).",
          glmm$method, format(glmm$n_cells, big.mark = ","), glmm$n_units, glmm$re_sd, glmm$singular)

# per-genotype lineage-conditioning audit (omitted IFN/proliferative fraction).
lpu  <- trd$lineage_per_unit
cond <- aggregate(cbind(n_cells, n_on_lineage) ~ genotype, data = lpu, FUN = sum)
cond$genotype <- factor(cond$genotype, levels = genotype_levels)
cond <- cond[order(cond$genotype), ]
cond$`omitted %` <- round(100 * (1 - cond$n_on_lineage / cond$n_cells), 1)
```

## Amyloid drives a large shift along the activation axis

Slingshot orders the homeostatic and DAM nuclei along a single rooted lineage on the
batch-integrated harmony space (the interferon and proliferative nuclei are off this axis and
omitted, recovered as a conditioning check below). Rooting is validated after the fact, not
assumed: pseudotime rises with the DAM score (Spearman
`r round(prov$dam_pt_rho, 2)`) and falls with the homeostatic score
(`r round(prov$homeo_pt_rho, 2)`), so low pseudotime is the homeostatic end and high pseudotime
the DAM end. Against that ordering the amyloid effect is unmistakable and large: it moves mean
pseudotime up by `r round(amy_k$coef, 1)` units on the tau-knockout background
(NLGF_MAPTKI vs MAPTKI, p `r fmt_p(amy_k$p_value)`) and `r round(amy_p$coef, 1)` units on the
mutant-tau background (NLGF_P301S vs P301S, p `r fmt_p(amy_p$p_value)`). Tau on its own moves it
little (`r round(mp_ctr("tau_alone")$coef, 1)`). These amyloid main effects dwarf everything that
follows; the interaction is a second-order adjustment on top of them.

```{r}
#| label: pseudotime-shift
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Amyloid advances microglia along the activation pseudotime. Left: per-nucleus pseudotime distribution by genotype (on-lineage nuclei), with the pooled DAM-onset landmark dashed; the amyloid-bearing genotypes carry far more mass at the DAM end. Right: the interaction in context - the genotype effects on mean pseudotime across the five canonical contrasts (point and 95% CI); the two amyloid contrasts are large, the difference-of-differences interaction (crimson) is a modest positive adjustment."
p_dist <- ggplot(on, aes(genotype, pt_raw, fill = genotype)) +
  geom_violin(scale = "width", colour = "grey30", linewidth = 0.3) +
  geom_hline(yintercept = prov$dam_onset, linetype = "dashed", colour = "grey50") +
  stat_summary(fun = mean, geom = "point", colour = "white", size = 1.6) +
  scale_fill_genotype(guide = "none") +
  labs(x = NULL, y = "activation pseudotime", title = "Pseudotime by genotype") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ctr_levels <- c("tau_alone", "nlgf_in_maptki", "nlgf_in_p301s", "tau_in_nlgf", "interaction")
ov <- do.call(rbind, lapply(ctr_levels, function(cn) {
  r <- mp_ctr(cn)
  data.frame(contrast = cn, coef = r$coef, ci_l = r$ci_l, ci_r = r$ci_r)
}))
ov$contrast <- factor(ov$contrast, levels = rev(ctr_levels))
ov$is_int   <- ov$contrast == "interaction"
p_ctr <- ggplot(ov, aes(coef, contrast, colour = is_int)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_pointrange(aes(xmin = ci_l, xmax = ci_r)) +
  scale_colour_manual(values = c(`FALSE` = "grey40", `TRUE` = "#B0344D"), guide = "none") +
  labs(x = "interaction effect on mean pseudotime", y = NULL,
       title = "Genotype effects on position")

(p_dist | p_ctr)
```

## The interaction is composition, not progression

The difference-of-differences interaction raises mean pseudotime position by
`r round(int_mpt$coef, 1)` units (95% CI [`r round(int_mpt$ci_l, 1)`, `r round(int_mpt$ci_r, 1)`],
ordinary p `r fmt_p(int_mpt$p_value)`, Freedman-Lane permutation p `r round(int_mpt$perm_p, 2)`) -
microglia on the amyloid-plus-mutant-tau background sit a little further along the axis than an
additive model predicts. The exact three-channel shift-share decomposition (composition,
progression, cross) says where that advance comes from, and the two channels point in opposite
directions. The composition channel - the same per-state positions, with cells redistributed
toward the DAM end - carries the whole effect (interaction `r round(int_comp$coef, 2)`,
p `r fmt_p(int_comp$p_value)`; loading `r round(load_comp, 2)` of the position interaction). The
progression channel - cells of a given state sitting further along, the genuinely
abundance-independent quantity - is small, negative and not significant (`r round(int_prog$coef, 2)`,
p `r fmt_p(int_prog$p_value)`, permutation p `r round(int_prog$perm_p, 2)`;
loading `r round(load_prog, 2)`), and the cross term is minor (loading `r round(load_cross, 2)`).
The three loadings sum to one by construction (reconstruction residual
`r signif(prov$recon_resid_max, 2)`).

```{r}
#| label: decomposition
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "The position interaction is compositional. Left: the tau-amyloid interaction on each additive pseudotime channel (point and 95% CI, all in pseudotime units); the composition channel is positive and clears zero, while the progression and within-state channels do not. Right: the shift-share loadings - composition takes more than the whole position interaction, progression is negative, and they sum to one. Mutant tau enlarges the DAM compartment; it does not measurably advance cells within a state."
chan       <- c("mean_pt", "comp_cf", "progression_cf", "cross", "within_homeostatic", "within_dam")
chan_label <- c("mean position", "composition", "progression", "cross",
                "within-homeostatic", "within-DAM")
fd <- inter[match(chan, inter$measure), ]
fd$label <- factor(chan_label, levels = rev(chan_label))
fd$sig   <- fd$ci_l > 0 | fd$ci_r < 0                            # 95% CI excludes zero
p_forest <- ggplot(fd, aes(coef, label, colour = sig)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_pointrange(aes(xmin = ci_l, xmax = ci_r)) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "#B0344D"), guide = "none") +
  labs(x = "interaction effect (pseudotime units)", y = NULL, title = "Interaction by channel")

ld <- data.frame(
  channel = factor(c("composition", "progression", "cross"),
                   levels = c("composition", "progression", "cross")),
  loading = c(load_comp, load_prog, load_cross))
p_load <- ggplot(ld, aes(channel, loading, fill = channel)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, colour = "grey50") +
  scale_fill_manual(values = c(composition = "#B0344D", progression = "#7FA8C1",
                               cross = "grey70"), guide = "none") +
  labs(x = NULL, y = "loading (fraction of position interaction)",
       title = "Decomposition loadings")

(p_forest | p_load)
```

The pre-registered primary endpoints - the progression channel and the within-homeostatic mean
position, both designed to be insensitive to compartment size - are both non-significant
(BH-adjusted FDR `r round(int_prog$fdr, 2)` and `r round(int_wh$fdr, 2)`). So the honest reading
is that there is no statistically supported signal that mutant tau advances microglia further
along the activation axis; what the interaction does, it does by enlarging the DAM compartment.
That is the trajectory-level counterpart of P1's significant DAM-fraction interaction: the same
synergy, now resolved as compositional. The full inference across every endpoint, including the
bounded fraction-past-onset bridges, is tabulated below.

```{r}
#| label: interaction-table
ord  <- c("progression_cf", "within_homeostatic", "mean_pt", "comp_cf", "cross", "within_dam",
          "median_pt", "q90", "frac_past_logit", "frac_past_asin")
nice <- c(progression_cf = "progression", within_homeostatic = "within-homeostatic",
          mean_pt = "mean position", comp_cf = "composition", cross = "cross",
          within_dam = "within-DAM", median_pt = "median position", q90 = "90th percentile",
          frac_past_logit = "fraction past onset (logit)", frac_past_asin = "fraction past onset (asin)")
it <- inter[match(ord, inter$measure), ]
itab <- data.frame(
  endpoint   = nice[it$measure],
  family     = it$family,
  effect     = round(it$coef, 2),
  `95% CI`   = sprintf("[%.2f, %.2f]", it$ci_l, it$ci_r),
  p          = fmt_p(it$p_value),
  `perm p`   = ifelse(is.na(it$perm_p), "-", sprintf("%.2f", it$perm_p)),
  FDR        = round(it$fdr, 3),
  check.names = FALSE, row.names = NULL, stringsAsFactors = FALSE)
knitr::kable(itab, row.names = FALSE,
  caption = "Tau-amyloid interaction on every trajectory endpoint (ordinary t at 9 residual degrees of freedom; Freedman-Lane permutation p where run). The additive channels (progression, composition, cross, mean/median/percentile position, within-state) are in pseudotime units; the fraction-past-onset rows are on the logit and arcsine bridges. BH FDR is computed within the pre-registered primary family and within the exploratory family separately.")
```

## A per-cell model corroborates the position shift

The summary-level test aggregates each of the `r glmm$n_units` genotype-by-batch units to one
number, which is the right scale for the interaction inference but discards the per-cell distribution. As a
supportive cross-check, a per-cell beta generalised linear mixed model
(`pt01 ~ tau * amyloid + batch + (1 | unit)`, `r format(glmm$n_cells, big.mark = ",")` on-lineage
nuclei over `r glmm$n_units` replicate units) `r glmm_sentence`. This agrees that the interaction
raises position. But it models per-cell position directly, so it is the per-cell analogue of mean
pseudotime and is itself composition-conflated: it supports that position rises, not that
progression does. It is reported as supportive only - the summary-level decomposition carries the
composition-versus-progression inference, and there both concordance and discordance with this arm
would be acceptable.

## Reconciliation, robustness and concordance

This re-baselines a v1 result rather than reproducing it. v1's point estimate had loaded the same
interaction mostly onto progression (about `r prov$v1_progression_loading` of the effect) at a
suggestive significance (FDR about `r prov$v1_progression_fdr`); on R `r prov$r_version` the
decomposition attributes it mainly to composition instead, and neither the progression nor the within-state
channel clears significance. The qualitative headline - a real, modest tau-amyloid synergy carried
by the DAM compartment - is unchanged; its mechanistic reading shifts from "further-progressed
cells" toward "more cells". The ordering itself is stable and marker-concordant: refitting on 10 and 20 harmony
dimensions and on all retained nuclei reproduces the primary pseudotime at Spearman
`r round(dims_min, 3)` or better, and the slingshot ordering agrees with the marker score-axis at
Spearman `r round(prov$concordance_rho, 2)` (a moderate-to-large concordance, expected below one
because the transcriptome geometry and the DAM-minus-homeostatic marker contrast are related, not
identical).

```{r}
#| label: concordance
#| fig-width: 6.5
#| fig-height: 4.5
#| fig-cap: "The slingshot pseudotime and the assumption-light marker score-axis (DAM minus homeostatic UCell) order the on-lineage nuclei concordantly; the binned density runs along the diagonal. This shares the marker system that defines the substates, so it guards against gross ordering failure rather than acting as an independent check."
ggplot(on, aes(pt_raw, score_axis_pt)) +
  geom_bin2d(bins = 60) +
  scale_fill_gradient(low = "grey88", high = "#3F5A6B", name = "nuclei", transform = "log10") +
  labs(x = "activation pseudotime", y = "marker score-axis (DAM - homeostatic)",
       title = sprintf("Pseudotime vs marker score-axis (rho = %.2f)", prov$concordance_rho))
```

## Caveats and provenance

Five caveats frame this chapter. First, pseudotime is an activation ordering, not developmental
time or cellular potency: direction rests on slingshot rooting and the validated DAM-marker
monotonicity above, never on a potency proxy. Second, and for the same reason, every effect here
is a statement about position and extent of advance in a cross-sectional snapshot; the rate or
acceleration reading would need longitudinal sampling or RNA velocity, which this design does not
have. Third, the lineage conditions on the homeostatic-to-DAM cells, omitting the interferon and
proliferative nuclei (`r omit_pct`% of microglia overall); were that omission genotype-skewed it
could bias the interaction, so the per-genotype omitted fraction is reported below and is balanced
across genotypes. Fourth, the within-state and composition channels are anchored on cell-weighted
pooled positions, so the decomposition is conditional on that anchoring. Fifth, the substates are
transcriptionally close, which is why
the marker score-axis concordance is moderate rather than near-perfect and is read as a guard, not
an independent validation.

```{r}
#| label: conditioning
knitr::kable(
  data.frame(
    genotype          = as.character(cond$genotype),
    nuclei            = cond$n_cells,
    `on-lineage`      = cond$n_on_lineage,
    `omitted %`       = cond$`omitted %`,
    check.names = FALSE, stringsAsFactors = FALSE),
  row.names = FALSE,
  caption = "Lineage-conditioning audit: nuclei omitted from the homeostatic-to-DAM lineage (interferon and proliferative substates) by genotype. The omitted fraction is low and balanced across genotypes, so conditioning on the lineage barely skews the interaction.")
```
```{r}
#| label: trajectory-provenance
#| echo: true
cat(sprintf("slingshot on harmony[1:%d], rooted at %s -> %s; pseudotime concordant with marker score-axis (rho %.2f, floor %.2f: %s).\nProgression inference: %d-unit weighted factorial, %d Freedman-Lane permutations, seed %d, limma %s, R %s.\n",
  prov$primary_dims, prov$root_substate, prov$terminal_substate, prov$concordance_rho,
  prov$concordance_floor, ifelse(isTRUE(prov$concordant), "pass", "FLAG"),
  nrow(trd$per_unit), prov$n_perm, prov$seed, prov$limma_version, prov$r_version))
cat(glmm_prov, "\n", sep = "")
```
